### 2016-07-29 STJ2012_analysis.Rmd
Import data in phyloseq

---
title: "STJ2012 analysis"
author: "Ross Cunning"
date: "July 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq)
```

## Import data

```{r data_import}
# Import OTU table
OTU <- otu_table(read.table("data/otu_table.tsv", header=T, check.names=F, row.names=1,
                            skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=1,
                                      col.names=c("OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
phy <- phyloseq(OTU, TAX, SAM)
```

### 2016-07-29 STJ2012_analysis.Rmd
run descriptive stats, plot clade abundances

@@
 knitr::opts_chunk$set(echo = TRUE)
-library(phyloseq)
+options(scipen = 4, digits = 2)
+library(phyloseq); library(vegan)
+
+taxcolors <- matrix(c(rgb(141, 211, 199, maxColorValue=255), rgb(255, 255, 179, maxColorValue=255), 
+                      rgb(190, 186, 218, maxColorValue=255), rgb(251, 128, 114, maxColorValue=255), 
+                      rgb(253, 180, 98, maxColorValue=255)), 
+                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF")))
 ```
@@
 # Import taxonomy
-TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=1,
-                                      col.names=c("OTU.ID", "Taxonomy", "Eval", "Subtype"))))
+TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=2,
+                                      col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
 # Import sample data
 SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
+# Build phyloseq object
 phy <- phyloseq(OTU, TAX, SAM)
+phy
 ```
+
+## Descriptive statistics for dataset  
+* Total count in OTU table: `r sum(otu_table(phy))`  
+* Number of samples: `r nsamples(phy)`  
+* Number of OTUs: `r ntaxa(phy)`
+
+#### Distribution of OTU counts and reads per sample  
+```{r phy_histograms, fig.height=3, fig.width=8}
+taxhist <- hist(log10(taxa_sums(phy)), plot=F)
+samhist <- hist(log10(sample_sums(phy)), plot=F)
+
+par(mfrow=c(1, 2), mar=c(3,3,1,1))
+plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+```
+
+* Range of OTU counts: `r range(taxa_sums(phy))`  
+* Number of singletons:  `r length(taxa_sums(phy)[taxa_sums(phy) <= 1])`  
+* Range of reads per sample: `r range(sample_sums(phy))`  
+* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy))` ± `r sd(sample_sums(phy))`  
+* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy))))` ± `r exp(sd(log(sample_sums(phy))))`
+
+```{r reads_per_sample, fig.height=2, fig.width=8}
+plot_bar(phy, x="Sample", y="Abundance")
+```
+
+## Filter dataset  
+#### Filter OTUs by minimum count
+```{r filter_taxa}
+# Set threshold count
+n <- 10
+# Identify OTUs below threshold count
+taxa <- taxa_sums(phy)[which(taxa_sums(phy) >= n)]  
+# Remove taxa below threshold count
+phy.f <- prune_taxa(names(taxa), phy)  
+```
+
+* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`
+
+#### Filter samples by minimum count
+```{r filter_samples}
+# Set threshold number of reads
+sn <- 100
+# Remove samples with fewer reads than threshold
+phy.f <- prune_samples(sample_sums(phy.f)>=sn, phy.f)
+
+```
+
+* Number of samples remaining in dataset: `r nsamples(phy.f) `
+
+## Transform taxa count data
+```{r transform}
+# Convert to proportion (relative abundance)
+phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
+# Apply transformation function
+transform <- function(x) sqrt(x/sum(x))  # Set transformation function
+phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
+```
+
+## Clade composition in all samples
+```{r cladecomp, fig.height=6, fig.width=6}
+samdat <- data.frame(sample_data(phy.f.p))
+samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
+samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
+samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
+samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
+typerelabund <- as.matrix(otu_table(phy.f.p)[order(data.frame(tax_table(phy.f.p))$Subtype), rownames(samdat)])
+shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
+shorebreaks <- which(shorebreaks==F) - 1
+spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
+# Make Barplot
+par(mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7, mfrow=c(1, 1))
+barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i",
+        col=rainbow(75))#,
+        #col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade])
+rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
+axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
+mtext(side=1, "Relative abundance", cex=0.7, line=1)
+legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7,
+       legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
+legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6,
+       legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
+# Add grouping bars for Shore
+for (i in 1:length(shorebreaks)) {
+  lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
+  rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], lwd=0.25, xpd=T)
+}
+# Add lines to separate species and species names
+for (i in 1:length(spbreaks)) {
+  lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
+  text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
+       labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
+                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
+}
+```
+
+# Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
+#   Calculate distances to group centroids in full-dimensional principal coordinate space
+#   Implemented by vegan betadisper(), based on Anderson (2006).
+# Compartments/species:
+```{r }
+sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
+                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid")
+anova(sambd)
+boxplot(sambd)
+sambdmeans <- aggregate(sambd$distances, by=list(group=sambd$group), FUN=mean)
+sambdses <- aggregate(sambd$distances, by=list(group=sambd$group), 
+                      FUN=function(x) sd(x)/sqrt(length(x)))
+sampt <- permutest(sambd, pairwise = T)
+sampt2 <- TukeyHSD(sambd)
+library(multcompView)
+saml <- multcompLetters(sampt$pairwise$permuted)#, Letters=c("e","a", "f","d","c","b"))
+saml2 <- multcompLetters(sampt2$group[, 4])
+
+# Figure: Distances to centroids
+# Coral species
+par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
+plot(sambdmeans$x[order(sambdmeans$x, decreasing=T)], type="n", ylim=c(0, 0.75),
+     ylab="Distance to centroid", xaxt="n", xlab="")
+arrows(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)],
+       1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)],
+       length=0.05, angle=90, code=3)
+points(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)], cex=1, pch=21, bg="white")
+text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75,
+     labels=levels(SAM$Species)[order(sambdmeans$x, decreasing=T)])
+
+text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)] + 0.03,
+     labels=saml$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
+text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)] - 0.03,
+     labels=saml2$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
+
+```
+
+```{r strigosa}
+dstr <- subset_samples(phy.f.p, Species=="strigosa")
+plot_bar(dstr, x="Sample", y="Abundance", fill="OTU")
+plot_bar(dstr, x="Sample", y="Abundance", fill="Subtype")
+# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
+# Also try 100% OTUs
+```

### 2016-07-29 scripts/STJ2012_analysis.Rmd
move .Rmd to scripts directory

--- 2016-07-29 STJ2012_analysis.Rmd
@@
 
-```{r data_import}
+```{r data_import, cache=T}
 # Import OTU table
@@
 # Set threshold number of reads
-sn <- 100
+sn <- 10
 # Remove samples with fewer reads than threshold
@@
 * Number of samples remaining in dataset: `r nsamples(phy.f) `
+
+#### Distribution of OTU counts and reads per sample post-filtering
+```{r phy.f_histograms, fig.height=3, fig.width=8}
+taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
+samhist <- hist(log10(sample_sums(phy.f)), plot=F)
+
+par(mfrow=c(1, 2), mar=c(3,3,1,1))
+plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+```
+
+* Range of OTU counts: `r range(taxa_sums(phy.f))`  
+* Number of singletons:  `r length(taxa_sums(phy.f)[taxa_sums(phy.f) <= 1])`  
+* Range of reads per sample: `r range(sample_sums(phy.f))`  
+* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy.f))` ± `r sd(sample_sums(phy.f))`  
+* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy.f))))` ± `r exp(sd(log(sample_sums(phy.f))))`
+
+```{r reads_per_sample_filtered, fig.height=2, fig.width=8}
+plot_bar(phy.f, x="Sample", y="Abundance")
+```
 
@@
 
-## Clade composition in all samples
-```{r cladecomp, fig.height=6, fig.width=6}
-samdat <- data.frame(sample_data(phy.f.p))
+## Community composition in all samples
+Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
+Each segment of the bar represents an individual OTU.
+In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
+```{r cladecomp, fig.height=6, fig.width=10}
+# Define function to plot Symbiodinium community composition
+composition <- function(phy, col, legend=T) {
+  samdat <- data.frame(sample_data(phy))
   samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
@@
   samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
-typerelabund <- as.matrix(otu_table(phy.f.p)[order(data.frame(tax_table(phy.f.p))$Subtype), rownames(samdat)])
+  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), rownames(samdat)])
   shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
@@
   # Make Barplot
-par(mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7, mfrow=c(1, 1))
-barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i",
-        col=rainbow(75))#,
-        #col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade])
+  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
   rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
@@
   mtext(side=1, "Relative abundance", cex=0.7, line=1)
-legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7,
-       legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
-legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6,
-       legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
+  # Add legend
+  if (legend==T) {
+    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
+    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
+  }
   # Add grouping bars for Shore
@@
   }
+}
+
+par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+composition(phy.f.p, col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade], legend=T)
+composition(phy.f.p, col=rainbow(ntaxa(phy.f.p)), legend=F)
+
 ```
@@
 # Compartments/species:
-```{r }
+```{r beta, cache=T}
 sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 

### 2016-07-29 analysis/STJ2012_analysis.Rmd
rename scripts/ -> analysis/

--- 2016-07-29 scripts/STJ2012_analysis.Rmd
@@
 ```{r setup, include=FALSE}
+knitr::opts_knit$set(root.dir = normalizePath(".."))
 knitr::opts_chunk$set(echo = TRUE)
+
 options(scipen = 4, digits = 2)

### 2016-07-29 analysis/STJ2012_analysis.Rmd
add TOC, organize headings

@@
 date: "July 28, 2016"
-output: html_document
+output: 
+  html_document: 
+    toc: yes
 ---
@@
 
-## Import data
+# Data pre-processing
+## Import dataset
 
@@
 
-## Descriptive statistics for dataset  
+#### Descriptive statistics  
 * Total count in OTU table: `r sum(otu_table(phy))`  
@@
 #### Distribution of OTU counts and reads per sample  
-```{r phy_histograms, fig.height=3, fig.width=8}
+```{r phy_histograms, fig.height=2, fig.width=8}
 taxhist <- hist(log10(taxa_sums(phy)), plot=F)
@@
 
-#### Distribution of OTU counts and reads per sample post-filtering
-```{r phy.f_histograms, fig.height=3, fig.width=8}
+#### Descriptive statistics of filtered dataset
+* Total count in OTU table: `r sum(otu_table(phy.f))`  
+* Number of samples: `r nsamples(phy.f)`  
+* Number of OTUs: `r ntaxa(phy.f)`
+
+#### Distribution of OTU counts and reads of filtered dataset
+```{r phy.f_histograms, fig.height=2, fig.width=8}
 taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
@@
 
-## Transform taxa count data
+## Transform count data
 ```{r transform}
@@
 
-## Community composition in all samples
+-----
+
+# Data analysis
+## *Symbiodinium* community composition barplots
 Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
@@
 In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
-```{r cladecomp, fig.height=6, fig.width=10}
+```{r cladecomp, fig.height=6, fig.width=8}
 # Define function to plot Symbiodinium community composition
@@
 
+## Beta diversity
+
+```{r beta, cache=T}
 # Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
@@
 # Compartments/species:
-```{r beta, cache=T}
 sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 

### 2016-07-29 analysis/STJ2012_analysis.Rmd
update betadisper analysis

@@
 options(scipen = 4, digits = 2)
-library(phyloseq); library(vegan)
+library(phyloseq); library(vegan); library(multcompView)
 
@@
 # Data analysis
-## *Symbiodinium* community composition barplots
+## Community composition barplots
 Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
@@
   samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
-  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), rownames(samdat)])
+  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
+                                           rownames(samdat)])
   shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
@@
   if (legend==T) {
-    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
-    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
+    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
+           cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
+    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
+           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
+           x.intersp=0.3)
   }
@@
     lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
-    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], lwd=0.25, xpd=T)
+    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
+         lwd=0.25, xpd=T)
   }
@@
 ## Beta diversity
+Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group.
+Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test.
+This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
+```{r betadisper, cache=T, fig.width=4, fig.height=4}
+# Calculate distances to species centroids for each sample in principal coordinate space
+sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
+                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid",
+                    bias.adjust=TRUE)
+# Calculate each species' mean distance to centroid and standard error
+sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
+sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
+                      FUN=function(x) sd(x)/sqrt(length(x)))$x
+# Determine order of decreasing mean dispersion
+sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
+# Update betadisper object with species in decreasing order of mean dispersion
+sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
+                    group=factor(data.frame(sample_data(phy.f.t))$Species, levels=as.character(sambdsumm.ord$group)), type="centroid", bias.adjust=TRUE)
+# Use permutations to perform pairwise comparisons of group mean dispersions
+sampt <- permutest(sambd, pairwise=T)
+saml <- multcompLetters(sampt$pairwise$permuted)
 
-```{r beta, cache=T}
-# Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
-#   Calculate distances to group centroids in full-dimensional principal coordinate space
-#   Implemented by vegan betadisper(), based on Anderson (2006).
-# Compartments/species:
-sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
-                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid")
-anova(sambd)
-boxplot(sambd)
-sambdmeans <- aggregate(sambd$distances, by=list(group=sambd$group), FUN=mean)
-sambdses <- aggregate(sambd$distances, by=list(group=sambd$group), 
-                      FUN=function(x) sd(x)/sqrt(length(x)))
-sampt <- permutest(sambd, pairwise = T)
-sampt2 <- TukeyHSD(sambd)
-library(multcompView)
-saml <- multcompLetters(sampt$pairwise$permuted)#, Letters=c("e","a", "f","d","c","b"))
-saml2 <- multcompLetters(sampt2$group[, 4])
-
-# Figure: Distances to centroids
-# Coral species
+# Figure: Distance to centroids
 par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
-plot(sambdmeans$x[order(sambdmeans$x, decreasing=T)], type="n", ylim=c(0, 0.75),
-     ylab="Distance to centroid", xaxt="n", xlab="")
-arrows(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)],
-       1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)],
-       length=0.05, angle=90, code=3)
-points(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)], cex=1, pch=21, bg="white")
+with(sambdsumm.ord, {
+  plot(mean, type="n", ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
   text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-     labels=levels(SAM$Species)[order(sambdmeans$x, decreasing=T)])
-
-text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)] + 0.03,
-     labels=saml$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
-text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)] - 0.03,
-     labels=saml2$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
-
+       labels=levels(SAM$Species)[order(sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=saml$Letters[as.character(group)], cex=0.5)
+})
 ```

### 2016-07-29 analysis/STJ2012_analysis.Rmd
add PERMANOVA analyses

@@
 
+## Differences between species and shores
+Whether *Symbiodinium* communities differ among species, or differ between north vs. south shores within species, is evaluated here using permutational analysis of variance (PERMANOVA).
+
+```{r permanovas, cache=T}
+# PERMANOVA for differences among species? Post-hoc: which spp are different?
+
+# Compute within- and between-pool statistics for each Species
+shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy.f.t))$Species), 
+                                              c("Species", "n", "overall", "within", 
+                                                "between", "R2", "bd", "p")), ncol=8, nrow=10))
+for (i in 1:nlevels(data.frame(sample_data(phy.f.t))$Species)) {
+  sp <- levels(data.frame(sample_data(phy.f.t))$Species)[i]
+  shorestats$Species[i] <- sp
+  phy.f.t.species <- subset_samples(phy.f.t, Species==sp)  # Set temp phyloseq object
+  shorestats$n[i] <- nsamples(phy.f.t.species)
+  md <- meandist(phyloseq::distance(phy.f.t.species, "bray"), 
+                 sample_data(phy.f.t.species)$Shore)
+  shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
+  shorestats$between[i] <- summary(md)$B  # Mean dissimilarity between Pools
+  shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
+  if (nlevels(as(sample_data(phy.f.t.species), "data.frame")$Shore) > 1) {
+    set.seed(70235)
+    permanova <- adonis(phyloseq::distance(phy.f.t.species, "bray") ~ Shore, 
+                        data=as(sample_data(phy.f.t.species), "data.frame"), permutations=9999)
+    shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
+    shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
+    bd <- betadisper(phyloseq::distance(phy.f.t.species, "bray"),
+                     group=as(sample_data(phy.f.t.species), "data.frame")$Shore)
+    shorestats$bd[i] <- TukeyHSD(bd)$group[4]
+  } else {
+    shorestats$R2[i] <- NA
+    shorestats$p[i] <- NA
+    shorestats$bd[i] <- NA
+  }
+}
+knitr::kable(shorestats)  # dendrogyra, porites furcata, and siderastrea siderea differ by shore
+```
+
 ## Beta diversity

### 2016-07-30 analysis/STJ2012_analysis.Rmd
filter samples before OTUS; pairwise spp permanovas

--- 2016-07-29 analysis/STJ2012_analysis.Rmd
@@
 ## Filter dataset
-#### Filter OTUs by minimum count
-```{r filter_taxa}
-# Set threshold count
-n <- 10
-# Identify OTUs below threshold count
-taxa <- taxa_sums(phy)[which(taxa_sums(phy) >= n)]  
-# Remove taxa below threshold count
-phy.f <- prune_taxa(names(taxa), phy)  
-```
-
-* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`
-
 #### Filter samples by minimum count
@@
 # Remove samples with fewer reads than threshold
-phy.f <- prune_samples(sample_sums(phy.f)>=sn, phy.f)
+phy.f <- prune_samples(sample_sums(phy)>=sn, phy)
 
@@
 * Number of samples remaining in dataset: `r nsamples(phy.f) `
+
+#### Filter OTUs by minimum count
+```{r filter_taxa}
+# Set threshold count
+n <- 3
+# Identify OTUs below threshold count
+taxa <- taxa_sums(phy.f)[which(taxa_sums(phy.f) >= n)]  
+# Remove taxa below threshold count
+phy.f <- prune_taxa(names(taxa), phy.f)  
+```
+
+* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`
 
@@
 
+## Differences between species and shores
+Whether *Symbiodinium* communities differ among species, or differ between north vs. south shores within species, is evaluated here using permutational analysis of variance (PERMANOVA).
+
+```{r permanovas, cache=T}
+# PERMANOVA for differences among species? Post-hoc: which spp are different?
+
+permanova.spp.t <- adonis(phyloseq::distance(phy.f.t, "bray") ~ Species, 
+                          data=as(sample_data(phy.f.t), "data.frame"), permutations=9999)
+
+dmat <- as(phyloseq::distance(phy.f.t, "bray"), "matrix")
+df <- as(sample_data(phy.f.t), "data.frame")
+
+pairs <- data.frame(t(combn(levels(df$Species), 2)), stringsAsFactors=F)
+p.pairs <- setNames(rep(NA, nrow(pairs)), with(pairs, paste(X1, X2, sep='-')))
+
+for (i in 1:nrow(pairs)) {
+  dfs <- subset(df, Species %in% c(pairs[i,1], pairs[i,2]))
+  dmats <- dmat[rownames(dfs), rownames(dfs)]
+  set.seed(152470)
+  permanova <- adonis(dmats ~ Species, data=dfs, permutations=999)
+  p.pairs[i] <- permanova$aov.tab$`Pr(>F)`[1]
+}
+
+multcompLetters(p.pairs)
+
+# Compute within- and between-pool statistics for each Species
+shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy.f.t))$Species), 
+                                              c("Species", "n", "overall", "within", 
+                                                "between", "R2", "bd", "p")), ncol=8, nrow=10))
+for (i in 1:nlevels(data.frame(sample_data(phy.f.t))$Species)) {
+  sp <- levels(data.frame(sample_data(phy.f.t))$Species)[i]
+  shorestats$Species[i] <- sp
+  phy.f.t.species <- subset_samples(phy.f.t, Species==sp)  # Set temp phyloseq object
+  shorestats$n[i] <- nsamples(phy.f.t.species)
+  md <- meandist(phyloseq::distance(phy.f.t.species, "bray"), 
+                 sample_data(phy.f.t.species)$Shore)
+  shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
+  shorestats$between[i] <- summary(md)$B  # Mean dissimilarity between Pools
+  shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
+  if (nlevels(as(sample_data(phy.f.t.species), "data.frame")$Shore) > 1) {
+    set.seed(70235)
+    permanova <- adonis(phyloseq::distance(phy.f.t.species, "bray") ~ Shore, 
+                        data=as(sample_data(phy.f.t.species), "data.frame"), permutations=999)
+    shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
+    shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
+    bd <- betadisper(phyloseq::distance(phy.f.t.species, "bray"),
+                     group=as(sample_data(phy.f.t.species), "data.frame")$Shore)
+    shorestats$bd[i] <- TukeyHSD(bd)$group[4]
+  } else {
+    shorestats$R2[i] <- NA
+    shorestats$p[i] <- NA
+    shorestats$bd[i] <- NA
+  }
+}
+knitr::kable(shorestats)  # dendrogyra, porites furcata, and siderastrea siderea differ by shore
+```
+
 ## Beta diversity
@@
 ```
+
+## Niche breadth of *Symbiodinium*
+```{r nichebreadth}
+# Mean dispersion of clades
+otubd <- betadisper(d=vegdist(otu_table(phy.f.t), method="bray"), 
+                    group=data.frame(tax_table(phy.f.t))$Clade, type="centroid")
+otubd
+
+# Number of samples each OTU appears in
+nsamp <- data.frame(n=rowSums(otu_table(phy.f.t) != 0))
+nsamp$Subtype <- data.frame(tax_table(phy.f.t))[rownames(nsamp), "Subtype"]
+nsamp[order(nsamp$n, decreasing=T), ]
+nsamp
+?order
+
+```
+
+plot_richness(phy.f, x="Species")

### 2016-07-31 analysis/STJ2012_analysis.Rmd
add 100% OTU dataset

@@
   html_document: 
+    self_contained: no
     toc: yes
@@
 
-options(scipen = 4, digits = 2)
+options(scipen = 3, digits = 2)
 library(phyloseq); library(vegan); library(multcompView)
@@
 ```{r data_import, cache=T}
-# Import OTU table
-OTU <- otu_table(read.table("data/otu_table.tsv", header=T, check.names=F, row.names=1,
+# Import OTU tables
+OTU97 <- otu_table(read.table("data/97_otu_table.tsv", header=T, check.names=F, row.names=1,
+                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
+OTU100 <- otu_table(read.table("data/100_otu_table.tsv", header=T, check.names=F, row.names=1,
                                skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
 # Import taxonomy
-TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=2,
+TAX97 <- tax_table(as.matrix(read.table("data/97_tax_assignments.txt", sep="\t", row.names=2,
+                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
+TAX100 <- tax_table(as.matrix(read.table("data/100_tax_assignments.txt", sep="\t", row.names=2,
                                          col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
@@
 SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
-# Build phyloseq object
-phy <- phyloseq(OTU, TAX, SAM)
-phy
+# Build phyloseq objects
+phy97 <- phyloseq(OTU97, TAX97, SAM)
+phy100 <- phyloseq(OTU100, TAX100, SAM)
 ```
 
-#### Descriptive statistics  
-* Total count in OTU table: `r sum(otu_table(phy))`  
-* Number of samples: `r nsamples(phy)`  
-* Number of OTUs: `r ntaxa(phy)`
+#### Descriptive statistics and distribution of OTU counts
+```{r phy_stats, fig.height=4, fig.width=8}
+# Compute summary statistics
+stats <- function(phy) {
+  return(
+    data.frame(rangeotus=paste0(range(taxa_sums(phy)), collapse=" - "),
+               nsingles=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
+               rangerps=paste0(range(sample_sums(phy)), collapse=" - "),
+               amrps=paste(as.integer(mean(sample_sums(phy))), 
+                           as.integer(sd(sample_sums(phy))), sep=" ± "),
+               gmrps=paste(as.integer(exp(mean(log(sample_sums(phy))))),
+                           as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "))
+  )
+}
+stats97 <- stats(phy97)
+stats100 <- stats(phy100)
 
-#### Distribution of OTU counts and reads per sample  
-```{r phy_histograms, fig.height=2, fig.width=8}
-taxhist <- hist(log10(taxa_sums(phy)), plot=F)
-samhist <- hist(log10(sample_sums(phy)), plot=F)
+# Create and plot histograms
+phy <- phy97
+taxhist97 <- hist(log10(taxa_sums(phy97)), plot=F)
+samhist97 <- hist(log10(sample_sums(phy97)), plot=F)
 
-par(mfrow=c(1, 2), mar=c(3,3,1,1))
-plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+taxhist100 <- hist(log10(taxa_sums(phy100)), plot=F)
+samhist100 <- hist(log10(sample_sums(phy100)), plot=F)
+
+par(mfrow=c(2, 2), mar=c(3,3,1,1))
+plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
@@
 
-* Range of OTU counts: `r range(taxa_sums(phy))`  
-* Number of singletons:  `r length(taxa_sums(phy)[taxa_sums(phy) <= 1])`  
-* Range of reads per sample: `r range(sample_sums(phy))`  
-* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy))` ± `r sd(sample_sums(phy))`  
-* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy))))` ± `r exp(sd(log(sample_sums(phy))))`
-
-```{r reads_per_sample, fig.height=2, fig.width=8}
-plot_bar(phy, x="Sample", y="Abundance")
-```
+|                                  |    97% OTUs               |    100% OTUs                |
+| -------------------------------- | ------------------------- | --------------------------- |
+| Total count in OTU table         | `r sum(otu_table(phy97))` |  `r sum(otu_table(phy100))` |
+| Number of OTUs                   | `r ntaxa(phy97)`          |  `r ntaxa(phy100)`          |
+| Range of OTU counts              | `r stats97$rangeotus`     | `r stats100$rangeotus`      |
+| Number of singleton OTUs         | `r stats97$nsingles`      | `r stats100$nsingles`       |
+| Number of samples                | `r nsamples(phy97)`       |  `r nsamples(phy100)`       |
+| Range of reads per sample        | `r stats97$rangerps`      | `r stats100$rangerps`       |
+| Mean (±SD) reads per sample (arithmetic) | `r stats97$amrps` | `r stats100$amrps`          |
+| Mean (±SD) reads per sample (geometric)  | `r stats97$gmrps` | `r stats100$gmrps`          |
 
@@
 # Remove samples with fewer reads than threshold
-phy.f <- prune_samples(sample_sums(phy)>=sn, phy)
+phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
+phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)
 
@@
 
-* Number of samples remaining in dataset: `r nsamples(phy.f) `
+* Number of samples remaining in 97% OTU dataset: `r nsamples(phy97.f) `
+* Number of samples remaining in 100% OTU dataset: `r nsamples(phy100.f) `
 
@@
 # Set threshold count
-n <- 3
+n <- 1
 # Identify OTUs below threshold count
-taxa <- taxa_sums(phy.f)[which(taxa_sums(phy.f) >= n)]  
+taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]  
+taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
 # Remove taxa below threshold count
-phy.f <- prune_taxa(names(taxa), phy.f)  
+phy97.f <- prune_taxa(names(taxa97), phy97.f)  
+phy100.f <- prune_taxa(names(taxa100), phy100.f)  
 ```
 
-* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`
+* Number of OTUs remaining in 97% dataset: `r ntaxa(phy97.f)`
+* Number of OTUs remaining in 97% dataset: `r ntaxa(phy100.f)`
 
-#### Descriptive statistics of filtered dataset
-* Total count in OTU table: `r sum(otu_table(phy.f))`  
-* Number of samples: `r nsamples(phy.f)`  
-* Number of OTUs: `r ntaxa(phy.f)`
+#### Filtered data: descriptive statistics and distribution of OTUs
+```{r phy.f_histograms, fig.height=4, fig.width=8}
+# Compute summary statistics
+stats97.f <- stats(phy97.f)
+stats100.f <- stats(phy100.f)
 
-#### Distribution of OTU counts and reads of filtered dataset
-```{r phy.f_histograms, fig.height=2, fig.width=8}
-taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
-samhist <- hist(log10(sample_sums(phy.f)), plot=F)
+# Create and plot histograms
+taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
+taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
+samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
+samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)
 
-par(mfrow=c(1, 2), mar=c(3,3,1,1))
-plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+par(mfrow=c(2, 2), mar=c(3,3,1,1))
+plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
@@
 
-* Range of OTU counts: `r range(taxa_sums(phy.f))`  
-* Number of singletons:  `r length(taxa_sums(phy.f)[taxa_sums(phy.f) <= 1])`  
-* Range of reads per sample: `r range(sample_sums(phy.f))`  
-* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy.f))` ± `r sd(sample_sums(phy.f))`  
-* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy.f))))` ± `r exp(sd(log(sample_sums(phy.f))))`
-
-```{r reads_per_sample_filtered, fig.height=2, fig.width=8}
-plot_bar(phy.f, x="Sample", y="Abundance")
-```
+|                                  |    97% OTUs                     |    100% OTUs                |
+| -------------------------------- | ------------------------------- | --------------------------- |
+| Total count in OTU table         | `r sum(otu_table(phy97.f))`     | `r sum(otu_table(phy100.f))`|
+| Number of OTUs                   | `r ntaxa(phy97.f)`              | `r ntaxa(phy100.f)`         |
+| Range of OTU counts              | `r stats97.f$rangeotus`         | `r stats100.f$rangeotus`    |
+| Number of singleton OTUs         | `r stats97.f$nsingles`          | `r stats100.f$nsingles`     |
+| Number of samples                | `r nsamples(phy97.f)`           | `r nsamples(phy100.f)`      |
+| Range of reads per sample        | `r stats97.f$rangerps`          | `r stats100.f$rangerps`     |
+| Mean (±SD) reads per sample (arithmetic) | `r stats97.f$amrps`     | `r stats100.f$amrps`        |
+| Mean (±SD) reads per sample (geometric)  | `r stats97.f$gmrps`     | `r stats100.f$gmrps`        |
 
@@
 # Convert to proportion (relative abundance)
-phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
+phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
+phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
 # Apply transformation function
 transform <- function(x) sqrt(x/sum(x))  # Set transformation function
-phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
+phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
+phy100.f.t <- transform_sample_counts(phy100.f, transform)
 ```
@@
 par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
-composition(phy.f.p, col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade], legend=T)
-composition(phy.f.p, col=rainbow(ntaxa(phy.f.p)), legend=F)
+composition(phy97.f.p, col=taxcolors[data.frame(tax_table(phy97.f.p))[order(data.frame(tax_table(phy97.f.p))$Subtype), ]$Clade], legend=T)
+composition(phy97.f.p, col=rainbow(ntaxa(phy97.f.p)), legend=F)
 
+```
+
+```{r, cache=T}
+par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+composition(phy100.f.p, col=taxcolors[data.frame(tax_table(phy100.f.p))[order(data.frame(tax_table(phy100.f.p))$Subtype), ]$Clade], legend=T)
+composition(phy100.f.p, col=rainbow(ntaxa(phy100.f.p)), legend=F)
+phy.f.p <- phy97.f.p
+phy.f.t <- phy97.f.t
 ```
@@
 ```{r strigosa}
-dstr <- subset_samples(phy.f.p, Species=="strigosa")
-plot_bar(dstr, x="Sample", y="Abundance", fill="OTU")
-plot_bar(dstr, x="Sample", y="Abundance", fill="Subtype")
+dstr97 <- subset_samples(phy97.f.p, Species=="strigosa")
+plot_bar(dstr97, x="Sample", y="Abundance", fill="OTU")
+plot_bar(dstr97, x="Sample", y="Abundance", fill="Subtype")
 # D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
-# Also try 100% OTUs
+# Also try 100% OTUs -- must filter more first...
+#dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
+#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
+#dstr100.f.p.f <- prune_taxa(rownames(otu_table(dstr100.f.p))[which(rowSums(otu_table(dstr100.f.p)) > 0.01)],
+#                            dstr100.f.p)
+#  
+#plot_bar(dstr100.f.p.f, x="Sample", y="Abundance", fill="OTU")
+#plot_bar(dstr100, x="Sample", y="Abundance", fill="Subtype")
+#sample_names(dstr100.f)
+#subset_samples(dstr100, Sample=="7")
+#otu_table(dstr100.f)
+#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
+#colSums(otu_table(dstr100.f.p)[rowSums(otu_table(dstr100.f.p)) > 0.1, ])
+
 ```

### 2016-07-31 analysis/STJ2012_analysis.Rmd
Add overview and bioinformatics description

@@
   html_document: 
-    self_contained: no
     toc: yes
@@
 
+# Overview
+  This project investigates *Symbiodinium* communities in scleractinian corals sampled at various sites on the north and south shores of St.
+John, US Virgin Islands in August 2012.
+The ITS2 region of nrDNA isolated from coral samples was amplified and sequenced by paired-end 300bp reads on an Illumina MiSeq platform.
+Sequence data is processed through a custom bioinformatic pipeline and resulting OTU tables are analyzed by various multivariate statistical approaches to ask questions regarding alpha and beta diversity of *Symbiodinium* communities within and across coral species.
+The major aims of this work are to 1) characterize *Symbiodinium* communities in a range of coral species using high-throughput sequencing (many for the first time), and 2) to forward novel bioinformatic and statistical approaches to better understand *Symbiodinium* ecology.
+
+# Bioinformatics
+  A [custom pipeline](sequence_processing.txt) is used here to process ITS2 sequence data.
+Briefly, paired reads were merged using the [illumina-utils](http://www.github.com/meren/illumina-utils) and filtered to keep only those with 3 mismatches or less.
+Chimeras were then removed by [usearch](http://www.drive5.com/usearch).
+From here, sequence data was processed using both of two approaches:
+
+1.
+__97% OTU clustering within samples.__ With this approach, sequences from each individual sample were clustered at 97% similarity and the most abundant sequence from each cluster was selected as the representative sequence and assigned taxonomy by BLAST using a [custom ITS2 reference database](../data/Unaligned_ITS2_Database_23April13.fasta).
+Singletons were excluded.
+Identical OTUs across samples were then merged to create a global OTU table.
+By clustering at 97% _within_ samples, as opposed to across the entire dataset, OTUs from different samples that have _different representative sequences_ (i.e., different most abundant sequence within the 97% cluster) will remain separate even if these sequences are >97% similar to each other.
+This approach is used because _Symbiodinium_ ITS2 sequences are know to be multi-copy and vary intragenomically, and distinct _Symbiodinium_ types may contain identical sets of ITS2 sequences that occur in different relative proportions within the genome; therefore, selecting the most abundant sequence variant _within_ a sample prevents a potentially distinct _Symbiodinium_ type from being collapsed into an OTU with a different representative sequence that happens to be more abundant in other samples within the dataset.
+2.
+__100% OTU clustering.__ All ITS2 sequences were clustered at 100% identity and assigned taxonomy by BLAST to the [custom reference database](../data/Unaligned_ITS2_Database_23April13.fasta).
+Singletons were excluded.
+
+OTU tables and taxonomic assignments from both of these bioinformatic approaches are used in comparative downstream analyses.
+
 # Data pre-processing
 ## Import dataset
-
+The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object to facilitate downstream analyses.
 ```{r data_import, cache=T}
@@
 # Set threshold count
-n <- 1
+n <- 3
 # Identify OTUs below threshold count

### 2016-08-01 analysis/STJ2012_analysis.Rmd
continue integrating 100% OTU data analysis

@@
 # Set threshold count
-n <- 3
+n <- 10
 # Identify OTUs below threshold count
@@
 In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
-```{r cladecomp, fig.height=6, fig.width=8}
+```{r cladecomp97, cache=T, fig.height=6, fig.width=8}
 # Define function to plot Symbiodinium community composition
@@
 par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+
+# Plot composition of 97% OTUs colored by clade
 composition(phy97.f.p, col=taxcolors[data.frame(tax_table(phy97.f.p))[order(data.frame(tax_table(phy97.f.p))$Subtype), ]$Clade], legend=T)
+
+# Plot composition of 97% OTUs colored by OTU
 composition(phy97.f.p, col=rainbow(ntaxa(phy97.f.p)), legend=F)
@@
 
-```{r, cache=T}
+```{r cladecomp100, cache=T, fig.height=6, fig.width=8}
 par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+# Plot composition of 100% OTUs colored by clade
 composition(phy100.f.p, col=taxcolors[data.frame(tax_table(phy100.f.p))[order(data.frame(tax_table(phy100.f.p))$Subtype), ]$Clade], legend=T)
+
+# Plot composition of 100% OTUs colored by OTU
 composition(phy100.f.p, col=rainbow(ntaxa(phy100.f.p)), legend=F)
-phy.f.p <- phy97.f.p
-phy.f.t <- phy97.f.t
 ```
 
-## Differences between species and shores
-Whether *Symbiodinium* communities differ among species, or differ between north vs. south shores within species, is evaluated here using permutational analysis of variance (PERMANOVA).
+## Differences between species
+Whether *Symbiodinium* communities differ among species is evaluated using permutational analysis of variance (PERMANOVA).
 
 ```{r permanovas, cache=T}
-# PERMANOVA for differences among species? Post-hoc: which spp are different?
-
-permanova.spp.t <- adonis(phyloseq::distance(phy.f.t, "bray") ~ Species, 
-                          data=as(sample_data(phy.f.t), "data.frame"), permutations=9999)
-
-dmat <- as(phyloseq::distance(phy.f.t, "bray"), "matrix")
-df <- as(sample_data(phy.f.t), "data.frame")
-
+# PERMANOVA for difference among species
+sppdiffs <- function(phy) {
+  permanova.spp.t <- adonis(phyloseq::distance(phy, "bray") ~ Species, 
+                          data=as(sample_data(phy), "data.frame"), permutations=9999)
+  permanova.spp.t
+  # Pairwise PERMANOVA tests for differences between species
+  dmat <- as(phyloseq::distance(phy, "bray"), "matrix")
+  df <- as(sample_data(phy), "data.frame")
   pairs <- data.frame(t(combn(levels(df$Species), 2)), stringsAsFactors=F)
   p.pairs <- setNames(rep(NA, nrow(pairs)), with(pairs, paste(X1, X2, sep='-')))
-
   for (i in 1:nrow(pairs)) {
@@
   }
+  # Return letters signifying differences 
+  return(multcompLetters(p.pairs))
+}
 
-multcompLetters(p.pairs)
+sppdiffs97 <- sppdiffs(phy97.f.t)
+sppdiffs100 <- sppdiffs(phy100.f.t)
 
-# Compute within- and between-pool statistics for each Species
-shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy.f.t))$Species), 
+sppdiffs <- data.frame("97% OTUs"=sppdiffs97$Letters, "100% OTUs"=sppdiffs100$Letters, check.names=F)
+knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
+```
+
+## Differences between shores
+Whether *Symbiodinium* communities differ between north vs. south shores within species is tested here using PERMANOVA.
+
+```{r shorediffs, cache=T}
+# Compute within- and between-shore statistics for each Species
+shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy97.f.t))$Species), 
                                               c("Species", "n", "overall", "within", 
                                                 "between", "R2", "bd", "p")), ncol=8, nrow=10))
-for (i in 1:nlevels(data.frame(sample_data(phy.f.t))$Species)) {
-  sp <- levels(data.frame(sample_data(phy.f.t))$Species)[i]
+for (i in 1:nlevels(data.frame(sample_data(phy97.f.t))$Species)) {
+  sp <- levels(data.frame(sample_data(phy97.f.t))$Species)[i]
   shorestats$Species[i] <- sp
-  phy.f.t.species <- subset_samples(phy.f.t, Species==sp)  # Set temp phyloseq object
-  shorestats$n[i] <- nsamples(phy.f.t.species)
-  md <- meandist(phyloseq::distance(phy.f.t.species, "bray"), 
-                 sample_data(phy.f.t.species)$Shore)
+  phy97.f.t.species <- subset_samples(phy97.f.t, Species==sp)  # Set temp phyloseq object
+  shorestats$n[i] <- nsamples(phy97.f.t.species)
+  md <- meandist(phyloseq::distance(phy97.f.t.species, "bray"), 
+                 sample_data(phy97.f.t.species)$Shore)
   shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
@@
   shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
-  if (nlevels(as(sample_data(phy.f.t.species), "data.frame")$Shore) > 1) {
+  if (nlevels(as(sample_data(phy97.f.t.species), "data.frame")$Shore) > 1) {
     set.seed(70235)
-    permanova <- adonis(phyloseq::distance(phy.f.t.species, "bray") ~ Shore, 
-                        data=as(sample_data(phy.f.t.species), "data.frame"), permutations=999)
+    permanova <- adonis(phyloseq::distance(phy97.f.t.species, "bray") ~ Shore, 
+                        data=as(sample_data(phy97.f.t.species), "data.frame"), permutations=999)
     shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
     shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
-    bd <- betadisper(phyloseq::distance(phy.f.t.species, "bray"),
-                     group=as(sample_data(phy.f.t.species), "data.frame")$Shore)
+    bd <- betadisper(phyloseq::distance(phy97.f.t.species, "bray"),
+                     group=as(sample_data(phy97.f.t.species), "data.frame")$Shore)
     shorestats$bd[i] <- TukeyHSD(bd)$group[4]
@@
 # Calculate distances to species centroids for each sample in principal coordinate space
-sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
-                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid",
+sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
+                    group=data.frame(sample_data(phy97.f.t))$Species, type="centroid",
                     bias.adjust=TRUE)
@@
 # Update betadisper object with species in decreasing order of mean dispersion
-sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
-                    group=factor(data.frame(sample_data(phy.f.t))$Species, levels=as.character(sambdsumm.ord$group)), type="centroid", bias.adjust=TRUE)
+sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
+                    group=factor(data.frame(sample_data(phy97.f.t))$Species, levels=as.character(sambdsumm.ord$group)), type="centroid", bias.adjust=TRUE)
 # Use permutations to perform pairwise comparisons of group mean dispersions
@@
 # Also try 100% OTUs -- must filter more first...
-#dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
-#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
+dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
+dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
 #dstr100.f.p.f <- prune_taxa(rownames(otu_table(dstr100.f.p))[which(rowSums(otu_table(dstr100.f.p)) > 0.01)],
@@
 
-## Niche breadth of *Symbiodinium*
-```{r nichebreadth}
-# Mean dispersion of clades
-otubd <- betadisper(d=vegdist(otu_table(phy.f.t), method="bray"), 
-                    group=data.frame(tax_table(phy.f.t))$Clade, type="centroid")
-otubd
+```{r otu_barplots_custom}
+# 100%-OTUs barplot 
+par(mar=c(2,3,4,1))
+samdat <- data.frame(sample_data(dstr100.f.p))
+typerelabund <- as.matrix(otu_table(dstr100.f.p)[order(data.frame(tax_table(dstr100.f.p))$Subtype), 
+                                                 rownames(samdat)])
 
-# Number of samples each OTU appears in
-nsamp <- data.frame(n=rowSums(otu_table(phy.f.t) != 0))
-nsamp$Subtype <- data.frame(tax_table(phy.f.t))[rownames(nsamp), "Subtype"]
-nsamp[order(nsamp$n, decreasing=T), ]
-nsamp
-?order
+# Get info for plotting OTU names and blast hits on top of barplot
+blasthits <- as.character(data.frame(tax_table(dstr100.f.p))[order(data.frame(tax_table(dstr100.f.p))$Subtype)
+                                                             , "Subtype"])
+names <- as.character(rownames(data.frame(tax_table(dstr100.f.p)))[order(data.frame(tax_table(dstr100.f.p))$Subtype)])
+divides <- apply(typerelabund, 2, cumsum)
+heights <- diff(divides)
+heights[heights < 0.04] <- NA
+
+# Plot barplot and OTU names and blast hits
+bars <- barplot(typerelabund, col=rainbow(941), border=NA, las=1, xlab="Sample", ylab="", cex.axis=0.75, cex.names=0.75)
+mtext(side=2, text = "Relative Abundance", line=2)
+for (i in 1:length(bars)) {
+  text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
+       divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
+       labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
+       cex=0.6)
+}
+
+mtext(side=3, "B.) 100% OTUs", xpd=T, adj=0, line=2.5, font=2)
 
 ```
-
-plot_richness(phy.f, x="Species")

### 2016-08-01 analysis/STJ2012_analysis.Rmd
test shore differences for both 97 and 100 OTUs

@@
 ```{r shorediffs, cache=T}
+shorediffs <- function(phy) {
   # Compute within- and between-shore statistics for each Species
-shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy97.f.t))$Species), 
+  shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy))$Species), 
                                                 c("Species", "n", "overall", "within", 
                                                   "between", "R2", "bd", "p")), ncol=8, nrow=10))
-for (i in 1:nlevels(data.frame(sample_data(phy97.f.t))$Species)) {
-  sp <- levels(data.frame(sample_data(phy97.f.t))$Species)[i]
+  for (i in 1:nlevels(data.frame(sample_data(phy))$Species)) {
+    sp <<- levels(data.frame(sample_data(phy))$Species)[i]
     shorestats$Species[i] <- sp
-  phy97.f.t.species <- subset_samples(phy97.f.t, Species==sp)  # Set temp phyloseq object
-  shorestats$n[i] <- nsamples(phy97.f.t.species)
-  md <- meandist(phyloseq::distance(phy97.f.t.species, "bray"), 
-                 sample_data(phy97.f.t.species)$Shore)
+    phy.species <- subset_samples(phy, Species==sp)  # Set temp phyloseq object
+    shorestats$n[i] <- nsamples(phy.species)
+    md <- meandist(phyloseq::distance(phy.species, "bray"), 
+                   sample_data(phy.species)$Shore)
     shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
@@
     shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
-  if (nlevels(as(sample_data(phy97.f.t.species), "data.frame")$Shore) > 1) {
+    if (nlevels(as(sample_data(phy.species), "data.frame")$Shore) > 1) {
       set.seed(70235)
-    permanova <- adonis(phyloseq::distance(phy97.f.t.species, "bray") ~ Shore, 
-                        data=as(sample_data(phy97.f.t.species), "data.frame"), permutations=999)
+      permanova <- adonis(phyloseq::distance(phy.species, "bray") ~ Shore, 
+                          data=as(sample_data(phy.species), "data.frame"), permutations=999)
       shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
       shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
-    bd <- betadisper(phyloseq::distance(phy97.f.t.species, "bray"),
-                     group=as(sample_data(phy97.f.t.species), "data.frame")$Shore)
+      bd <- betadisper(phyloseq::distance(phy.species, "bray"),
+                       group=as(sample_data(phy.species), "data.frame")$Shore)
       shorestats$bd[i] <- TukeyHSD(bd)$group[4]
@@
   }
-knitr::kable(shorestats)  # dendrogyra, porites furcata, and siderastrea siderea differ by shore
+  return(shorestats)
+}
+
+shorestats97 <- shorediffs(phy97.f.t)
+shorestats100 <- shorediffs(phy100.f.t)
+
+knitr::kable(shorestats97, row.names=F, caption="97% OTU data: differences within and between shore for each species")
+knitr::kable(shorestats100, row.names=F, caption="100% OTU data: differences within and between shore for each species")
 ```

### 2016-08-01 analysis/STJ2012_analysis.Rmd
add individual spp barplots

@@
 #### Descriptive statistics and distribution of OTU counts
-```{r phy_stats, fig.height=4, fig.width=8}
+```{r phy_stats, cache=T, fig.height=4, fig.width=8}
 # Compute summary statistics
@@
 #### Filter samples by minimum count
-```{r filter_samples}
+```{r filter_samples, cache=T}
 # Set threshold number of reads
@@
 phy100.f <- prune_taxa(names(taxa100), phy100.f)
+
+# Refilter samples in case removal of OTUs caused any samples to drop below threshold
+phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
+phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
 ```
@@
 This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
-```{r betadisper, cache=T, fig.width=4, fig.height=4}
+```{r betadisper, cache=T, fig.width=8, fig.height=4}
+betad <- function(phy) {
   # Calculate distances to species centroids for each sample in principal coordinate space
-sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
-                    group=data.frame(sample_data(phy97.f.t))$Species, type="centroid",
+  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
+                      group=data.frame(sample_data(phy))$Species, type="centroid",
                       bias.adjust=TRUE)
@@
   # Update betadisper object with species in decreasing order of mean dispersion
-sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
-                    group=factor(data.frame(sample_data(phy97.f.t))$Species, levels=as.character(sambdsumm.ord$group)), type="centroid", bias.adjust=TRUE)
+  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
+                      group=factor(data.frame(sample_data(phy))$Species,
+                                   levels=as.character(sambdsumm.ord$group)), 
+                      type="centroid", bias.adjust=TRUE)
   # Use permutations to perform pairwise comparisons of group mean dispersions
@@
   saml <- multcompLetters(sampt$pairwise$permuted)
+  # Return results as a list
+  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
+}
+
+betad97 <- betad(phy97.f.t)
+betad100 <- betad(phy100.f.t)
 
 # Figure: Distance to centroids
-par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
-with(sambdsumm.ord, {
-  plot(mean, type="n", ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+par(mfrow=c(1,2), mar=c(3.5,4,2,1), lwd=1)
+with(betad97$sambdsumm.ord, {
+  plot(mean, type="n", main="97% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
   arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
@@
   text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=saml$Letters[as.character(group)], cex=0.5)
+       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
+})
+with(betad100$sambdsumm.ord, {
+  plot(mean, type="n", main="100% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
 })
@@
 
-```{r strigosa}
-dstr97 <- subset_samples(phy97.f.p, Species=="strigosa")
-plot_bar(dstr97, x="Sample", y="Abundance", fill="OTU")
-plot_bar(dstr97, x="Sample", y="Abundance", fill="Subtype")
-# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
-# Also try 100% OTUs -- must filter more first...
-dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
-dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
-#dstr100.f.p.f <- prune_taxa(rownames(otu_table(dstr100.f.p))[which(rowSums(otu_table(dstr100.f.p)) > 0.01)],
-#                            dstr100.f.p)
-#  
-#plot_bar(dstr100.f.p.f, x="Sample", y="Abundance", fill="OTU")
-#plot_bar(dstr100, x="Sample", y="Abundance", fill="Subtype")
-#sample_names(dstr100.f)
-#subset_samples(dstr100, Sample=="7")
-#otu_table(dstr100.f)
-#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
-#colSums(otu_table(dstr100.f.p)[rowSums(otu_table(dstr100.f.p)) > 0.1, ])
-
-```
-
-```{r otu_barplots_custom}
-# 100%-OTUs barplot 
-par(mar=c(2,3,4,1))
-samdat <- data.frame(sample_data(dstr100.f.p))
-typerelabund <- as.matrix(otu_table(dstr100.f.p)[order(data.frame(tax_table(dstr100.f.p))$Subtype), 
+```{r otu_barplots_custom, fig.width=8, fig.height=8}
+otubarplot <- function(phy, main) {
+  samdat <- data.frame(sample_data(phy))
+  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                                  rownames(samdat)])
-
+  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
+                                                                order(Shore, Site, InputFileName))])]
   # Get info for plotting OTU names and blast hits on top of barplot
-blasthits <- as.character(data.frame(tax_table(dstr100.f.p))[order(data.frame(tax_table(dstr100.f.p))$Subtype)
-                                                             , "Subtype"])
-names <- as.character(rownames(data.frame(tax_table(dstr100.f.p)))[order(data.frame(tax_table(dstr100.f.p))$Subtype)])
+  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
+  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
+  samples <- colnames(typerelabund)
+  shores <- data.frame(sample_data(phy))[samples, "Shore"]
+  sites <- data.frame(sample_data(phy))[samples, "Site"]
   divides <- apply(typerelabund, 2, cumsum)
@@
   heights[heights < 0.04] <- NA
-
   # Plot barplot and OTU names and blast hits
-bars <- barplot(typerelabund, col=rainbow(941), border=NA, las=1, xlab="Sample", ylab="", cex.axis=0.75, cex.names=0.75)
+  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
+                  xlab="", ylab="", cex.axis=0.75, cex.names=0.75, xaxt="n")
+  text(bars, -0.06, labels=samples, xpd=T, cex=0.75)
+  text(bars, -0.10, labels=shores, xpd=T, cex=0.75)
+  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T, cex=0.75)
+  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, cex=0.75, pos=2)
   mtext(side=2, text = "Relative Abundance", line=2)
@@
          labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
-       cex=0.6)
+         cex=0.5)
   }
+  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
+}
+```
 
-mtext(side=3, "B.) 100% OTUs", xpd=T, adj=0, line=2.5, font=2)
+## Individual coral species
+### *Diploria strigosa*
+```{r dstr, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Diploria strigosa
+dstr97.f.p <- subset_samples(phy97.f.p, Species=="strigosa")
+dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
+# Plot custom barplots for Diploria strigosa
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(dstr97.f.p, main="97% OTUs")
+otubarplot(dstr100.f.p, main="100% OTUs")
+```
+
+### *Porites furcata*
+```{r pfur, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Porites furcata
+pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
+pfur100.f.p <- subset_samples(phy100.f.p, Species=="furcata")
+# Plot custom barplots for Porites furcata
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(pfur97.f.p, main="97% OTUs")
+otubarplot(pfur100.f.p, main="100% OTUs")
+```
+
+### *Orbicella annularis*
+```{r oann, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Orbicella annularis
+oann97.f.p <- subset_samples(phy97.f.p, Species=="annularis")
+oann100.f.p <- subset_samples(phy100.f.p, Species=="annularis")
+# Plot custom barplots for Orbicella annularis
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(oann97.f.p, main="97% OTUs")
+otubarplot(oann100.f.p, main="100% OTUs")
+```
+
+### *Millepora alcicornis*
+```{r malc, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Millepora alcicornis
+malc97.f.p <- subset_samples(phy97.f.p, Species=="alcicornis")
+malc100.f.p <- subset_samples(phy100.f.p, Species=="alcicornis")
+# Plot custom barplots for Millepora alcicornis
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(malc97.f.p, main="97% OTUs")
+otubarplot(malc100.f.p, main="100% OTUs")
+```
+
+### *Siderastrea siderea*
+```{r ssid, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Siderastrea siderea
+ssid97.f.p <- subset_samples(phy97.f.p, Species=="siderea")
+ssid100.f.p <- subset_samples(phy100.f.p, Species=="siderea")
+# Plot custom barplots for Siderastrea siderea
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(ssid97.f.p, main="97% OTUs")
+otubarplot(ssid100.f.p, main="100% OTUs")
+```
+
+### *Favia fragum*
+```{r ffra, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Favia fragum
+ffra97.f.p <- subset_samples(phy97.f.p, Species=="fragum")
+ffra100.f.p <- subset_samples(phy100.f.p, Species=="fragum")
+# Plot custom barplots for Favia fragum
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(ffra97.f.p, main="97% OTUs")
+otubarplot(ffra100.f.p, main="100% OTUs")
+```
+
+### *Siderastrea radians*
+```{r srad, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Siderastrea radians
+srad97.f.p <- subset_samples(phy97.f.p, Species=="radians")
+srad100.f.p <- subset_samples(phy100.f.p, Species=="radians")
+# Plot custom barplots for Siderastrea radians
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(srad97.f.p, main="97% OTUs")
+otubarplot(srad100.f.p, main="100% OTUs")
+```
+
+### *Porites astreoides*
+```{r past, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Porites astreoides
+past97.f.p <- subset_samples(phy97.f.p, Species=="astreoides")
+past100.f.p <- subset_samples(phy100.f.p, Species=="astreoides")
+# Plot custom barplots for Porites astreoides
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(past97.f.p, main="97% OTUs")
+otubarplot(past100.f.p, main="100% OTUs")
+```
+
+### *Dendrogyra cylindrus*
+```{r dcyl, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Dendrogyra cylindrus
+dcyl97.f.p <- subset_samples(phy97.f.p, Species=="cylindrus")
+dcyl100.f.p <- subset_samples(phy100.f.p, Species=="cylindrus")
+# Plot custom barplots for Dendrogyra cylindrus
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(dcyl97.f.p, main="97% OTUs")
+otubarplot(dcyl100.f.p, main="100% OTUs")
+```
+
+### *Montastraea cavernosa*
+```{r mcav, cache=T, fig.height=8, fig.width=8}
+# Create subsetted phyloseq objects for Montastraea cavernosa
+mcav97.f.p <- subset_samples(phy97.f.p, Species=="cavernosa")
+mcav100.f.p <- subset_samples(phy100.f.p, Species=="cavernosa")
+# Plot custom barplots for Montastraea cavernosa
+par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
+otubarplot(mcav97.f.p, main="97% OTUs")
+otubarplot(mcav100.f.p, main="100% OTUs")
+```
 

### 2016-08-02 merge_qc_reads.sh
add shell scripts for bioinformatic pipeline

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+# Unzip sequence data archive
+unzip data/*.zip -d data/fastq
+
+# Generate config files for merging reads with illumina-utils
+mkdir -p data/merge
+mv data/fastq_list.txt data/fastq/fastq_list.txt
+iu-gen-configs data/fastq/fastq_list.txt -o data/merge
+
+# Merge paired reads for each sample (in parallel)
+parallel iu-merge-pairs {} --min-overlap-size 150 --enforce-Q30-check --marker-gene-stringent ::: data/merge/*.ini
+
+# Filter sequences - keep only those with 3 mismatches or less (in parallel)
+parallel iu-filter-merged-reads {} --max-mismatches 3 ::: data/merge/*_MERGED
+
+# Add QIIME labels
+add_qiime_labels.py -m data/mapping_file.txt -i data/merge/ -c InputFileName -o data/fasta
+
+# Chimera checking
+identify_chimeric_seqs.py -i data/fasta/combined_seqs.fna --suppress_usearch61_ref -m usearch61 -o data/fasta/usearch61_chimeras
+filter_fasta.py -f data/fasta/combined_seqs.fna -o data/fasta/combined_seqs_chimera_filtered.fasta -s data/fasta/usearch61_chimeras/chimeras.txt -n

### 2016-08-02 otus_100.sh

@@ -1,0 +1,32 @@
+#!/bin/bash
+
+# Cluster entire dataset at 100% similarity
+pick_otus.py -i data/fasta/combined_seqs_chimera_filtered.fasta -s 1.0 -o data/otus_100/
+
+# Remove singletons before picking rep set and assigning taxonomy (to save time)
+awk '$3 ~ /./ {print}' data/otus_100/combined_seqs_chimera_filtered_otus.txt > data/otus_100/nosingles_otus.txt
+
+# Get rep set of 100% OTUs
+pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
+-f data/fasta/combined_seqs_chimera_filtered.fasta \
+-o data/otus_100/rep_set.fasta
+
+# Assign taxonomy
+assign_taxonomy.py -i data/otus_100/rep_set.fasta -m blast \
+-r data/Unaligned_ITS2_Database_31July16.fasta \
+-t data/id_to_taxonomy_31July16.txt \
+-o data/otus_100/blast_taxonomy
+
+# Make list of "no blast hits"
+awk '/No blast hit/' data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/otus_100/blast_taxonomy/no_blast_hits.txt
+
+# Make OTU table excluding no blast hits
+make_otu_table.py -i data/otus_100/nosingles_otus.txt \
+-t data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt \
+-e data/otus_100/blast_taxonomy/no_blast_hits.txt \
+-o data/otus_100/100_otu_table.biom
+rm data/100_otu_table.tsv  # Delete old OTU .tsv if present
+biom convert -i data/otus_100/100_otu_table.biom -o data/100_otu_table.tsv --to-tsv
+
+# Add clade to tax table and copy to data directory
+cut -d$'\t' -f4- data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt | cut -c-1 | paste - data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/100_tax_assignments.txt

### 2016-08-02 otus_97.sh

@@ -1,0 +1,55 @@
+#!/bin/bash
+
+# Split into individual sample files
+rm -r data/fasta/bysample
+split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_chimera_filtered.fasta -o data/fasta/bysample
+
+# OTU clustering at 97% similarity for each sample (in parallel)
+clust97() {
+	s=`echo $1 | grep -Eoh '/[0-9]{1,4}' | cut -c 2-`
+	pick_otus.py -i $1 -s 0.97 --denovo_otu_id_prefix $s'_denovo' -o data/otus_97
+}
+export -f clust97
+parallel clust97 ::: data/fasta/bysample/*.fasta
+
+# Pick representative sequence set for each sample
+repset97() {
+	s=`echo $1 | grep -Eoh '/[0-9]{1,4}' | cut -c 2-`
+	pick_rep_set.py -i $1 -m most_abundant -f 'data/fasta/bysample/'$s'.fasta' \
+	-o 'data/otus_97/'$s'_rep_set.fasta'
+}
+export -f repset97
+parallel repset97 ::: data/otus_97/[0-9]*_otus.txt
+
+# Merge all rep sets together
+cat data/otus_97/[0-9]*_rep_set.fasta > data/otus_97/all_rep_set.fasta
+
+# Cluster all 97% rep sets at 100% identity
+pick_otus.py -i data/otus_97/all_rep_set.fasta -s 1.0 -o data/otus_97
+
+# Get rep set of 100% OTUs and assign taxonomy
+pick_rep_set.py -i data/otus_97/all_rep_set_otus.txt -f data/otus_97/all_rep_set.fasta \
+-o data/otus_97/all_rep_set_rep_set.fasta
+
+# Assign taxonomy
+assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/Unaligned_ITS2_Database_31July16.fasta -t data/id_to_taxonomy_31July16.txt -m blast -o data/otus_97/blast_taxonomy
+
+# Make list of "no blast hits"
+awk '/No blast hit/' data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt > data/otus_97/blast_taxonomy/no_blast_hits.txt
+ 
+# Concatenate 97% OTU maps and merge with 100% OTU map
+cat data/otus_97/*_otus.txt > data/otus_97/all_97_otus.txt
[…]

### 2016-08-02 merge_qc_reads.sh
keep fastq_list.txt in data dir

@@ -5,7 +5,7 @@ unzip data/*.zip -d data/fastq
 
 # Generate config files for merging reads with illumina-utils
 mkdir -p data/merge
-mv data/fastq_list.txt data/fastq/fastq_list.txt
+cp data/fastq_list.txt data/fastq/fastq_list.txt
 iu-gen-configs data/fastq/fastq_list.txt -o data/merge
 
 # Merge paired reads for each sample (in parallel)

### 2016-08-07 get_ITS2_database.sh
generate new ITS2 database from NCBI

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+# Get Subtype and Accession number links from old database and use accession number to download sequences from NCBI
+cat data/Unaligned_ITS2_Database_31July16.fasta | awk '/>/ {print}' | \
+while read n
+do
+	acn=${n##*_}
+	echo $n
+	curl "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta"
+done > data/ITS2_Database.fasta
+	
+# Reformat names of new database to be only accession numbers & subtypes and put sequences on one line
+awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' data/ITS2_Database.fasta > data/ITS2_Database_inline.fasta
+awk '!/^>gi/ {print}' data/ITS2_Database_inline.fasta > data/ITS2_Database_ready.fasta
+
+
+# Trim primers from database sequences using cutadapt
+# Trim forward primers using cutadapt
+#   Allow error rate of 15% (0 indels/mismatches)
+cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/ITS2_Database_ready.fasta -o data/ITS2_Database_trimF.fasta
+# Trim forward primers again (may be multiple internal primer sequences), but do not discard sequences that do not contain primer
+cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/ITS2_Database_trimF.fasta -o data/ITS2_Database_trimF2.fasta
+# Trim reverse primers using cutadapt
+#   Do not remove sequences that do not have reverse primer sequence
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2.fasta -o data/ITS2_Database_trimF2_trimR.fasta
+# Trim reverse primers again for any remaining internally
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db.fasta
+
+# Clean up intermediate files
+rm data/ITS2_Database.fasta
+rm data/ITS2_Database_inline.fasta
+rm data/ITS2_Database_ready.fasta
+rm data/ITS2_Database_trimF.fasta
+rm data/ITS2_Database_trimF2.fasta
+rm data/ITS2_Database_trimF2_trimR.fasta

### 2016-08-07 otus_100.sh
trim primers, use new ITS2db

@@ -1,7 +1,7 @@
 #!/bin/bash
 
 # Cluster entire dataset at 100% similarity
-pick_otus.py -i data/fasta/combined_seqs_chimera_filtered.fasta -s 1.0 -o data/otus_100/
+pick_otus.py -i data/fasta/combined_seqs_trimmed.fasta -s 1.0 -o data/otus_100/
 
 # Remove singletons before picking rep set and assigning taxonomy (to save time)
 awk '$3 ~ /./ {print}' data/otus_100/combined_seqs_chimera_filtered_otus.txt > data/otus_100/nosingles_otus.txt
@@ -13,7 +13,7 @@ pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
 
 # Assign taxonomy
 assign_taxonomy.py -i data/otus_100/rep_set.fasta -m blast \
--r data/Unaligned_ITS2_Database_31July16.fasta \
+-r data/ITS2db.fasta \
 -t data/id_to_taxonomy_31July16.txt \
 -o data/otus_100/blast_taxonomy
 

### 2016-08-07 otus_97.sh

@@ -2,7 +2,7 @@
 
 # Split into individual sample files
 rm -r data/fasta/bysample
-split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_chimera_filtered.fasta -o data/fasta/bysample
+split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_trimmed.fasta -o data/fasta/bysample
 
 # OTU clustering at 97% similarity for each sample (in parallel)
 clust97() {
@@ -32,7 +32,7 @@ pick_rep_set.py -i data/otus_97/all_rep_set_otus.txt -f data/otus_97/all_rep_set
 -o data/otus_97/all_rep_set_rep_set.fasta
 
 # Assign taxonomy
-assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/Unaligned_ITS2_Database_31July16.fasta -t data/id_to_taxonomy_31July16.txt -m blast -o data/otus_97/blast_taxonomy
+assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/ITS2db.fasta -t data/id_to_taxonomy_31July16.txt -m blast -o data/otus_97/blast_taxonomy
 
 # Make list of "no blast hits"
 awk '/No blast hit/' data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt > data/otus_97/blast_taxonomy/no_blast_hits.txt

### 2016-08-07 merge_qc_reads.sh
trim primer sequences

@@ -20,3 +20,19 @@ add_qiime_labels.py -m data/mapping_file.txt -i data/merge/ -c InputFileName -o 
 # Chimera checking
 identify_chimeric_seqs.py -i data/fasta/combined_seqs.fna --suppress_usearch61_ref -m usearch61 -o data/fasta/usearch61_chimeras
 filter_fasta.py -f data/fasta/combined_seqs.fna -o data/fasta/combined_seqs_chimera_filtered.fasta -s data/fasta/usearch61_chimeras/chimeras.txt -n
+
+# Trim primer sequences using cutadapt
+# Trim forward primers (allow error rate of 15% (3 indels/mismatches))
+cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/fasta/combined_seqs_chimera_filtered.fasta -o data/fasta/trimF.fasta
+# Trim forward primers again (may be multiple internal primer sequences), but do not discard sequences that do not contain primer
+cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/fasta/trimF.fasta -o data/fasta/trimF2.fasta
+# Trim reverse primers using cutadapt
+#   Do not remove sequences that do not have reverse primer sequence
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/fasta/trimF2.fasta -o data/fasta/trimF2_trimR.fasta
+# Trim reverse primers again for any remaining internally
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/fasta/trimF2_trimR.fasta -o data/fasta/combined_seqs_trimmed.fasta
+
+# Remove intermediate files
+rm data/fasta/trimF.fasta
+rm data/fasta/trimF2.fasta
+rm data/fasta/trimF2_trimR.fasta

### 2016-08-07 get_ITS2_database.sh
add new seqs to database

@@ -1,7 +1,10 @@
 #!/bin/bash
 
+awk '/>/ {print}' data/Unaligned_ITS2_Database_31July16.fasta > data/seqIDs.txt
+cat data/addseqs.txt >> data/seqIDs.txt
+
 # Get Subtype and Accession number links from old database and use accession number to download sequences from NCBI
-cat data/Unaligned_ITS2_Database_31July16.fasta | awk '/>/ {print}' | \
+cat data/seqIDs.txt | \
 while read n
 do
 	acn=${n##*_}
@@ -27,6 +30,7 @@ cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2.fasta -o dat
 cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db.fasta
 
 # Clean up intermediate files
+rm data/seqIDs.txt
 rm data/ITS2_Database.fasta
 rm data/ITS2_Database_inline.fasta
 rm data/ITS2_Database_ready.fasta

### 2016-08-07 get_ITS2_database.sh
remove .1 from accn#s, create id_to_taxonomy

@@ -1,6 +1,7 @@
 #!/bin/bash
 
-awk '/>/ {print}' data/Unaligned_ITS2_Database_31July16.fasta > data/seqIDs.txt
+awk '/>/ {print}' data/Unaligned_ITS2_Database_31July16.fasta > data/seqIDs1.txt
+sed 's/\.1$//' data/seqIDs1.txt > data/seqIDs.txt  # remove trailing '.1' on some accession numbers
 cat data/addseqs.txt >> data/seqIDs.txt
 
 # Get Subtype and Accession number links from old database and use accession number to download sequences from NCBI
@@ -29,7 +30,11 @@ cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2.fasta -o dat
 # Trim reverse primers again for any remaining internally
 cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db.fasta
 
+# Generate id_to_taxonomy file
+sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3 Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
+
 # Clean up intermediate files
+rm data/seqIDs1.txt
 rm data/seqIDs.txt
 rm data/ITS2_Database.fasta
 rm data/ITS2_Database_inline.fasta

### 2016-08-07 get_ITS2_database.sh
use tab delimiter in id_to_taxonomy

@@ -31,7 +31,8 @@ cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2.fasta -o dat
 cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db.fasta
 
 # Generate id_to_taxonomy file
-sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3 Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
+TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
+sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3'"${TAB}"'Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
 
 # Clean up intermediate files
 rm data/seqIDs1.txt

### 2016-08-07 otus_100.sh
use new id_to_tax

@@ -14,7 +14,7 @@ pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
 # Assign taxonomy
 assign_taxonomy.py -i data/otus_100/rep_set.fasta -m blast \
 -r data/ITS2db.fasta \
--t data/id_to_taxonomy_31July16.txt \
+-t data/id_to_taxonomy.txt \
 -o data/otus_100/blast_taxonomy
 
 # Make list of "no blast hits"

### 2016-08-07 otus_97.sh

@@ -32,7 +32,7 @@ pick_rep_set.py -i data/otus_97/all_rep_set_otus.txt -f data/otus_97/all_rep_set
 -o data/otus_97/all_rep_set_rep_set.fasta
 
 # Assign taxonomy
-assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/ITS2db.fasta -t data/id_to_taxonomy_31July16.txt -m blast -o data/otus_97/blast_taxonomy
+assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/ITS2db.fasta -t data/id_to_taxonomy.txt -m blast -o data/otus_97/blast_taxonomy
 
 # Make list of "no blast hits"
 awk '/No blast hit/' data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt > data/otus_97/blast_taxonomy/no_blast_hits.txt

### 2016-08-08 get_ITS2_database.sh
download seqs from NCBI in parallel

@@ -1,17 +1,23 @@
 #!/bin/bash
 
+# Get Subtype and Accession number links from old database and new ones in addseqs
+
 awk '/>/ {print}' data/Unaligned_ITS2_Database_31July16.fasta > data/seqIDs1.txt
 sed 's/\.1$//' data/seqIDs1.txt > data/seqIDs.txt  # remove trailing '.1' on some accession numbers
 cat data/addseqs.txt >> data/seqIDs.txt
 
-# Get Subtype and Accession number links from old database and use accession number to download sequences from NCBI
-cat data/seqIDs.txt | \
-while read n
-do
+# use accession number to download sequences from NCBI in parallel
+getseq() {
+	n=$1
 	acn=${n##*_}
-	echo $n
-	curl "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta"
-done > data/ITS2_Database.fasta
+	echo $1 > data/dbseqs/${acn}.fasta
+	curl "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta" >> data/dbseqs/${acn}.fasta
+}
+export -f getseq
+rm -r data/dbseqs; mkdir data/dbseqs
+parallel -a data/seqIDs.txt getseq 
+cat data/dbseqs/*.fasta > data/ITS2_Database.fasta
+
 	
 # Reformat names of new database to be only accession numbers & subtypes and put sequences on one line
 awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' data/ITS2_Database.fasta > data/ITS2_Database_inline.fasta
@@ -37,6 +43,7 @@ sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3'"
 # Clean up intermediate files
 rm data/seqIDs1.txt
 rm data/seqIDs.txt
+rm -r data/dbseqs
 rm data/ITS2_Database.fasta
 rm data/ITS2_Database_inline.fasta
 rm data/ITS2_Database_ready.fasta

### 2016-08-08 otus_97.sh
move tax assignment to new script

@@ -32,10 +32,10 @@ pick_rep_set.py -i data/otus_97/all_rep_set_otus.txt -f data/otus_97/all_rep_set
 -o data/otus_97/all_rep_set_rep_set.fasta
 
 # Assign taxonomy
-assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/ITS2db.fasta -t data/id_to_taxonomy.txt -m blast -o data/otus_97/blast_taxonomy
+#assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/ITS2db.fasta -t data/id_to_taxonomy.txt -m blast -o data/otus_97/blast_taxonomy
 
 # Make list of "no blast hits"
-awk '/No blast hit/' data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt > data/otus_97/blast_taxonomy/no_blast_hits.txt
+#awk '/No blast hit/' data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt > data/otus_97/blast_taxonomy/no_blast_hits.txt
  
 # Concatenate 97% OTU maps and merge with 100% OTU map
 cat data/otus_97/*_otus.txt > data/otus_97/all_97_otus.txt
@@ -45,11 +45,11 @@ merge_otu_maps.py -i data/otus_97/all_97_otus.txt,data/otus_97/all_rep_set_otus.
 awk '$3 ~ /./ {print}' data/otus_97/merged_otu_map.txt > data/otus_97/nosingles_otus.txt
 
 # Make OTU table excluding no blast hits
-make_otu_table.py -i data/otus_97/nosingles_otus.txt \
--t data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt \
--e data/otus_97/blast_taxonomy/no_blast_hits.txt -o data/otus_97/97_otu_table.biom
-rm data/97_otu_table.tsv  # delete old OTU .tsv if present
-biom convert -i data/otus_97/97_otu_table.biom -o data/97_otu_table.tsv --to-tsv
+#make_otu_table.py -i data/otus_97/nosingles_otus.txt \
+#-t data/otus_97/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt \
+#-e data/otus_97/blast_taxonomy/no_blast_hits.txt -o data/otus_97/97_otu_table.biom
[…]

### 2016-08-08 tax_results.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+DIR=dirname $1
+
+
+# Assign taxonomy
+assign_taxonomy.py -i $1 -r data/ITS2db.fasta -t data/id_to_taxonomy.txt -m blast -o $DIR/blast_taxonomy
+
+# Make list of "no blast hits"
+awk '/No blast hit/' $DIR/blast_taxonomy/*_tax_assignments.txt > $DIR/blast_taxonomy/no_blast_hits.txt
+
+# Make OTU table excluding no blast hits
+make_otu_table.py -i $DIR/nosingles_otus.txt \
+-t $DIR/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt \
+-e $DIR/blast_taxonomy/no_blast_hits.txt -o $DIR/$2.biom
+rm $DIR/*.tsv  # delete old OTU .tsv if present
+biom convert -i $DIR/$2.biom -o data/$2.tsv --to-tsv
+
+# Add clade to tax table and copy to data directory
+cut -d$'\t' -f4- $DIR/blast_taxonomy/*_tax_assignments.txt | cut -c-1 | paste - $DIR/blast_taxonomy/*_tax_assignments.txt > data/$2_tax_assignments.txt

### 2016-08-08 tax_results.sh
fix filepath error

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-DIR=dirname $1
+DIR=$(dirname $1)
 
 
 # Assign taxonomy

### 2016-08-08 otus_100.sh
move tax assignment to tax_results.sh

@@ -12,21 +12,21 @@ pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
 -o data/otus_100/rep_set.fasta
 
 # Assign taxonomy
-assign_taxonomy.py -i data/otus_100/rep_set.fasta -m blast \
--r data/ITS2db.fasta \
--t data/id_to_taxonomy.txt \
--o data/otus_100/blast_taxonomy
+#assign_taxonomy.py -i data/otus_100/rep_set.fasta -m blast \
+#-r data/ITS2db.fasta \
+#-t data/id_to_taxonomy.txt \
+#-o data/otus_100/blast_taxonomy
 
-# Make list of "no blast hits"
-awk '/No blast hit/' data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/otus_100/blast_taxonomy/no_blast_hits.txt
+## Make list of "no blast hits"
+#awk '/No blast hit/' data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/otus_100/blast_taxonomy/no_blast_hits.txt
 
-# Make OTU table excluding no blast hits
-make_otu_table.py -i data/otus_100/nosingles_otus.txt \
--t data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt \
--e data/otus_100/blast_taxonomy/no_blast_hits.txt \
--o data/otus_100/100_otu_table.biom
-rm data/100_otu_table.tsv  # Delete old OTU .tsv if present
-biom convert -i data/otus_100/100_otu_table.biom -o data/100_otu_table.tsv --to-tsv
+## Make OTU table excluding no blast hits
+#make_otu_table.py -i data/otus_100/nosingles_otus.txt \
+#-t data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt \
+#-e data/otus_100/blast_taxonomy/no_blast_hits.txt \
+#-o data/otus_100/100_otu_table.biom
+#rm data/100_otu_table.tsv  # Delete old OTU .tsv if present
+#biom convert -i data/otus_100/100_otu_table.biom -o data/100_otu_table.tsv --to-tsv
 
-# Add clade to tax table and copy to data directory
-cut -d$'\t' -f4- data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt | cut -c-1 | paste - data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/100_tax_assignments.txt
[…]

### 2016-08-08 tax_results.sh

@@ -1,6 +1,7 @@
 #!/bin/bash
 
 DIR=$(dirname $1)
+OUT=$2
 
 
 # Assign taxonomy
@@ -11,10 +12,10 @@ awk '/No blast hit/' $DIR/blast_taxonomy/*_tax_assignments.txt > $DIR/blast_taxo
 
 # Make OTU table excluding no blast hits
 make_otu_table.py -i $DIR/nosingles_otus.txt \
--t $DIR/blast_taxonomy/all_rep_set_rep_set_tax_assignments.txt \
--e $DIR/blast_taxonomy/no_blast_hits.txt -o $DIR/$2.biom
+-t $DIR/blast_taxonomy/*_tax_assignments.txt \
+-e $DIR/blast_taxonomy/no_blast_hits.txt -o $DIR/$OUT.biom
 rm $DIR/*.tsv  # delete old OTU .tsv if present
-biom convert -i $DIR/$2.biom -o data/$2.tsv --to-tsv
+biom convert -i $DIR/$OUT.biom -o data/$OUT.tsv --to-tsv
 
 # Add clade to tax table and copy to data directory
-cut -d$'\t' -f4- $DIR/blast_taxonomy/*_tax_assignments.txt | cut -c-1 | paste - $DIR/blast_taxonomy/*_tax_assignments.txt > data/$2_tax_assignments.txt
+cut -d$'\t' -f4- $DIR/blast_taxonomy/*_tax_assignments.txt | cut -c-1 | paste - $DIR/blast_taxonomy/*_tax_assignments.txt > data/$OUT_tax_assignments.txt

### 2016-08-08 bioinformatics_run.sh
script to run all bioinformatics at once

@@ -1,0 +1,18 @@
+
+# Prepare ITS2 reference database
+./get_ITS2_database.sh
+
+# Merge, QC, and trim Illumina reads
+./merge_qc_reads.sh
+
+# Cluster 97% OTUs within samples
+./otus_97.sh
+
+# Assign taxonomy and generate OTU table for 97% OTUs
+./tax_results.sh data/otus_97/all_rep_set_rep_set.fasta 97_otus
+
+# Cluster 100% OTUs
+./otus_100.sh
+
+# Assign taxonomy and generate OTU table for 97% OTUs
+./tax_results.sh data/otus_100/rep_set.fasta 100_otus

### 2016-08-08 tax_results.sh
fix filepath error

@@ -14,8 +14,8 @@ awk '/No blast hit/' $DIR/blast_taxonomy/*_tax_assignments.txt > $DIR/blast_taxo
 make_otu_table.py -i $DIR/nosingles_otus.txt \
 -t $DIR/blast_taxonomy/*_tax_assignments.txt \
 -e $DIR/blast_taxonomy/no_blast_hits.txt -o $DIR/$OUT.biom
-rm $DIR/*.tsv  # delete old OTU .tsv if present
+rm -f  $DIR/*.tsv  # delete old OTU .tsv if present
 biom convert -i $DIR/$OUT.biom -o data/$OUT.tsv --to-tsv
 
 # Add clade to tax table and copy to data directory
-cut -d$'\t' -f4- $DIR/blast_taxonomy/*_tax_assignments.txt | cut -c-1 | paste - $DIR/blast_taxonomy/*_tax_assignments.txt > data/$OUT_tax_assignments.txt
+cut -d$'\t' -f4- $DIR/blast_taxonomy/*_tax_assignments.txt | cut -c-1 | paste - $DIR/blast_taxonomy/*_tax_assignments.txt > 'data/'$OUT'_tax_assignments.txt'

### 2016-08-08 bioinformatics_run.sh
run 97% global clustering

@@ -5,14 +5,32 @@
 # Merge, QC, and trim Illumina reads
 ./merge_qc_reads.sh
 
+
+
+# -------------------------------
 # Cluster 97% OTUs within samples
 ./otus_97.sh
 
-# Assign taxonomy and generate OTU table for 97% OTUs
-./tax_results.sh data/otus_97/all_rep_set_rep_set.fasta 97_otus
+# Assign taxonomy and generate OTU table for 97% within-sample OTUs
+./tax_results.sh data/otus_97/97_otus_rep_set.fasta 97_otus
+# -------------------------------
 
+
+
+# -------------------------------
+# Cluster 97% OTUs within samples
+./otus_97_bysample.sh
+
+# Assign taxonomy and generate OTU table for 97% within-sample OTUs
+./tax_results.sh data/otus_97_bysample/all_rep_set_rep_set.fasta 97_otus
+# -------------------------------
+
+
+
+# -------------------------------
 # Cluster 100% OTUs
 ./otus_100.sh
 
 # Assign taxonomy and generate OTU table for 97% OTUs
-./tax_results.sh data/otus_100/rep_set.fasta 100_otus
+./tax_results.sh data/otus_100/100_otus_rep_set.fasta 100_otus
+# -------------------------------

### 2016-08-08 otus_97.sh

@@ -1,55 +1,12 @@
 #!/bin/bash
 
-# Split into individual sample files
-rm -r data/fasta/bysample
-split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_trimmed.fasta -o data/fasta/bysample
+# Cluster entire dataset at 97% similarity
+pick_otus.py -i data/fasta/combined_seqs_trimmed.fasta -s 0.97 -o data/otus_97/
 
-# OTU clustering at 97% similarity for each sample (in parallel)
-clust97() {
-	s=`echo $1 | grep -Eoh '/[0-9]{1,4}' | cut -c 2-`
-	pick_otus.py -i $1 -s 0.97 --denovo_otu_id_prefix $s'_denovo' --optimal_uclust -o data/otus_97
-}
-export -f clust97
-parallel clust97 ::: data/fasta/bysample/*.fasta
+# Remove singletons before picking rep set and assigning taxonomy (to save time)
+awk '$3 ~ /./ {print}' data/otus_97/combined_seqs_chimera_filtered_otus.txt > data/otus_97/nosingles_otus.txt
 
-# Pick representative sequence set for each sample
-repset97() {
-	s=`echo $1 | grep -Eoh '/[0-9]{1,4}' | cut -c 2-`
-	pick_rep_set.py -i $1 -m most_abundant -f 'data/fasta/bysample/'$s'.fasta' \
-	-o 'data/otus_97/'$s'_rep_set.fasta'
-}
-export -f repset97
-parallel repset97 ::: data/otus_97/[0-9]*_otus.txt
-
-# Merge all rep sets together
-cat data/otus_97/[0-9]*_rep_set.fasta > data/otus_97/all_rep_set.fasta
-
-# Cluster all 97% rep sets at 100% identity
-pick_otus.py -i data/otus_97/all_rep_set.fasta -s 1.0 --optimal_uclust -o data/otus_97
-
-# Get rep set of 100% OTUs and assign taxonomy
-pick_rep_set.py -i data/otus_97/all_rep_set_otus.txt -f data/otus_97/all_rep_set.fasta \
--o data/otus_97/all_rep_set_rep_set.fasta
-
-# Assign taxonomy
-#assign_taxonomy.py -i data/otus_97/all_rep_set_rep_set.fasta -r data/ITS2db.fasta -t data/id_to_taxonomy.txt -m blast -o data/otus_97/blast_taxonomy
-
[…]

### 2016-08-08 otus_97_bysample.sh

@@ -1,0 +1,39 @@
+#!/bin/bash
+
+# Split into individual sample files
+rm -rf data/fasta/bysample
+split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_trimmed.fasta -o data/fasta/bysample
+
+# OTU clustering at 97% similarity for each sample (in parallel)
+clust97() {
+	s=`echo $1 | grep -Eoh '/[0-9]{1,4}' | cut -c 2-`
+	pick_otus.py -i $1 -s 0.97 --denovo_otu_id_prefix $s'_denovo' --optimal_uclust -o data/otus_97_bysample
+}
+export -f clust97
+parallel clust97 ::: data/fasta/bysample/*.fasta
+
+# Pick representative sequence set for each sample
+repset97() {
+	s=`echo $1 | grep -Eoh '/[0-9]{1,4}' | cut -c 2-`
+	pick_rep_set.py -i $1 -m most_abundant -f 'data/fasta/bysample/'$s'.fasta' \
+	-o 'data/otus_97_bysample/'$s'_rep_set.fasta'
+}
+export -f repset97
+parallel repset97 ::: data/otus_97_bysample/[0-9]*_otus.txt
+
+# Merge all rep sets together
+cat data/otus_97_bysample/[0-9]*_rep_set.fasta > data/otus_97_bysample/all_rep_set.fasta
+
+# Cluster all 97% rep sets at 100% identity
+pick_otus.py -i data/otus_97_bysample/all_rep_set.fasta -s 1.0 --optimal_uclust -o data/otus_97_bysample
+
+# Get rep set of 100% OTUs and assign taxonomy
+pick_rep_set.py -i data/otus_97_bysample/all_rep_set_otus.txt -f data/otus_97_bysample/all_rep_set.fasta \
+-o data/otus_97_bysample/all_rep_set_rep_set.fasta
+
+# Concatenate 97% OTU maps and merge with 100% OTU map
+cat data/otus_97_bysample/*_otus.txt > data/otus_97_bysample/all_97_otus.txt
+merge_otu_maps.py -i data/otus_97_bysample/all_97_otus.txt,data/otus_97_bysample/all_rep_set_otus.txt -o data/otus_97_bysample/merged_otu_map.txt
+
+# Remove singletons
+awk '$3 ~ /./ {print}' data/otus_97_bysample/merged_otu_map.txt > data/otus_97_bysample/nosingles_otus.txt

### 2016-08-08 otus_100.sh
remove old code

@@ -9,24 +9,4 @@ awk '$3 ~ /./ {print}' data/otus_100/combined_seqs_chimera_filtered_otus.txt > d
 # Get rep set of 100% OTUs
 pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
 -f data/fasta/combined_seqs_chimera_filtered.fasta \
--o data/otus_100/rep_set.fasta
-
-# Assign taxonomy
-#assign_taxonomy.py -i data/otus_100/rep_set.fasta -m blast \
-#-r data/ITS2db.fasta \
-#-t data/id_to_taxonomy.txt \
-#-o data/otus_100/blast_taxonomy
-
-## Make list of "no blast hits"
-#awk '/No blast hit/' data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/otus_100/blast_taxonomy/no_blast_hits.txt
-
-## Make OTU table excluding no blast hits
-#make_otu_table.py -i data/otus_100/nosingles_otus.txt \
-#-t data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt \
-#-e data/otus_100/blast_taxonomy/no_blast_hits.txt \
-#-o data/otus_100/100_otu_table.biom
-#rm data/100_otu_table.tsv  # Delete old OTU .tsv if present
-#biom convert -i data/otus_100/100_otu_table.biom -o data/100_otu_table.tsv --to-tsv
-
-## Add clade to tax table and copy to data directory
-#cut -d$'\t' -f4- data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt | cut -c-1 | paste - data/otus_100/blast_taxonomy/rep_set_tax_assignments.txt > data/100_tax_assignments.txt
+-o data/otus_100/100_otus_rep_set.fasta

### 2016-08-09 bioinformatics_run.sh
fix filepath errors

@@ -8,10 +8,10 @@
 
 
 # -------------------------------
-# Cluster 97% OTUs within samples
+# Cluster 97% OTUs
 ./otus_97.sh
 
-# Assign taxonomy and generate OTU table for 97% within-sample OTUs
+# Assign taxonomy and generate OTU table for 97% OTUs
 ./tax_results.sh data/otus_97/97_otus_rep_set.fasta 97_otus
 # -------------------------------
 
@@ -22,7 +22,7 @@
 ./otus_97_bysample.sh
 
 # Assign taxonomy and generate OTU table for 97% within-sample OTUs
-./tax_results.sh data/otus_97_bysample/all_rep_set_rep_set.fasta 97_otus
+./tax_results.sh data/otus_97_bysample/all_rep_set_rep_set.fasta 97_otus_bysample
 # -------------------------------
 
 
@@ -31,6 +31,6 @@
 # Cluster 100% OTUs
 ./otus_100.sh
 
-# Assign taxonomy and generate OTU table for 97% OTUs
+# Assign taxonomy and generate OTU table for 100% OTUs
 ./tax_results.sh data/otus_100/100_otus_rep_set.fasta 100_otus
 # -------------------------------

### 2016-08-09 otus_100.sh

@@ -4,9 +4,9 @@
 pick_otus.py -i data/fasta/combined_seqs_trimmed.fasta -s 1.0 -o data/otus_100/
 
 # Remove singletons before picking rep set and assigning taxonomy (to save time)
-awk '$3 ~ /./ {print}' data/otus_100/combined_seqs_chimera_filtered_otus.txt > data/otus_100/nosingles_otus.txt
+awk '$3 ~ /./ {print}' data/otus_100/combined_seqs_trimmed_otus.txt > data/otus_100/nosingles_otus.txt
 
 # Get rep set of 100% OTUs
 pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
--f data/fasta/combined_seqs_chimera_filtered.fasta \
+-f data/fasta/combined_seqs_trimmed.fasta \
 -o data/otus_100/100_otus_rep_set.fasta

### 2016-08-09 otus_97.sh

@@ -4,9 +4,9 @@
 pick_otus.py -i data/fasta/combined_seqs_trimmed.fasta -s 0.97 -o data/otus_97/
 
 # Remove singletons before picking rep set and assigning taxonomy (to save time)
-awk '$3 ~ /./ {print}' data/otus_97/combined_seqs_chimera_filtered_otus.txt > data/otus_97/nosingles_otus.txt
+awk '$3 ~ /./ {print}' data/otus_97/combined_seqs_trimmed_otus.txt > data/otus_97/nosingles_otus.txt
 
 # Get rep set of 97% OTUs
 pick_rep_set.py -i data/otus_97/nosingles_otus.txt \
--f data/fasta/combined_seqs_chimera_filtered.fasta \
+-f data/fasta/combined_seqs_trimmed.fasta \
 -o data/otus_97/97_otus_rep_set.fasta

### 2016-08-09 get_ITS2_database.sh
use independent list of accn_nos to build ITS2db

@@ -1,12 +1,6 @@
 #!/bin/bash
 
-# Get Subtype and Accession number links from old database and new ones in addseqs
-
-awk '/>/ {print}' data/Unaligned_ITS2_Database_31July16.fasta > data/seqIDs1.txt
-sed 's/\.1$//' data/seqIDs1.txt > data/seqIDs.txt  # remove trailing '.1' on some accession numbers
-cat data/addseqs.txt >> data/seqIDs.txt
-
-# use accession number to download sequences from NCBI in parallel
+# use accession numbers to download sequences from NCBI in parallel
 getseq() {
 	n=$1
 	acn=${n##*_}
@@ -14,8 +8,8 @@ getseq() {
 	curl "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta" >> data/dbseqs/${acn}.fasta
 }
 export -f getseq
-rm -r data/dbseqs; mkdir data/dbseqs
-parallel -a data/seqIDs.txt getseq 
+rm -rf data/dbseqs; mkdir data/dbseqs
+parallel -a data/accn_nos.txt getseq 
 cat data/dbseqs/*.fasta > data/ITS2_Database.fasta
 
 	
@@ -41,8 +35,6 @@ TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
 sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3'"${TAB}"'Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
 
 # Clean up intermediate files
-rm data/seqIDs1.txt
-rm data/seqIDs.txt
 rm -r data/dbseqs
 rm data/ITS2_Database.fasta
 rm data/ITS2_Database_inline.fasta

### 2016-08-09 otus_97.sh
--optimal uclust

@@ -1,7 +1,7 @@
 #!/bin/bash
 
 # Cluster entire dataset at 97% similarity
-pick_otus.py -i data/fasta/combined_seqs_trimmed.fasta -s 0.97 -o data/otus_97/
+pick_otus.py -i data/fasta/combined_seqs_trimmed.fasta -s 0.97 --optimal_uclust -o data/otus_97/
 
 # Remove singletons before picking rep set and assigning taxonomy (to save time)
 awk '$3 ~ /./ {print}' data/otus_97/combined_seqs_trimmed_otus.txt > data/otus_97/nosingles_otus.txt

### 2016-08-10 get_ITS2_database.sh
remove Ns from database sequences

@@ -28,7 +28,11 @@ cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/ITS2_Database_trimF.fasta -o data/
 #   Do not remove sequences that do not have reverse primer sequence
 cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2.fasta -o data/ITS2_Database_trimF2_trimR.fasta
 # Trim reverse primers again for any remaining internally
-cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db.fasta
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db_trimmed.fasta
+
+# Remove Ns if present in sequences -- noting that several were deposited in GenBank using Ns as gaps...
+awk '!/>/ { gsub("N","") }; { print $0 }' data/ITS2db_trimmed.fasta > data/ITS2db.fasta
+
 
 # Generate id_to_taxonomy file
 TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
@@ -42,3 +46,4 @@ rm data/ITS2_Database_ready.fasta
 rm data/ITS2_Database_trimF.fasta
 rm data/ITS2_Database_trimF2.fasta
 rm data/ITS2_Database_trimF2_trimR.fasta
+rm data/ITS2db_trimmed.fasta

### 2016-08-10 analysis/STJ2012_analysis.Rmd
both 97% OTU and 97% within-sample OTU analyses

--- 2016-08-01 analysis/STJ2012_analysis.Rmd
@@
 options(scipen = 3, digits = 2)
-library(phyloseq); library(vegan); library(multcompView)
 
-taxcolors <- matrix(c(rgb(141, 211, 199, maxColorValue=255), rgb(255, 255, 179, maxColorValue=255), 
-                      rgb(190, 186, 218, maxColorValue=255), rgb(251, 128, 114, maxColorValue=255), 
-                      rgb(253, 180, 98, maxColorValue=255)), 
-                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF")))
+library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr)
+
+taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
+                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
 ```
@@
 # Import OTU tables
-OTU97 <- otu_table(read.table("data/97_otu_table.tsv", header=T, check.names=F, row.names=1,
+OTU97 <- otu_table(read.table("data/97_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU100 <- otu_table(read.table("data/100_otu_table.tsv", header=T, check.names=F, row.names=1,
+OTU97bs <- otu_table(read.table("data/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
+                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
+OTU100 <- otu_table(read.table("data/100_otus.tsv", header=T, check.names=F, row.names=1,
                                skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
 # Import taxonomy
-TAX97 <- tax_table(as.matrix(read.table("data/97_tax_assignments.txt", sep="\t", row.names=2,
+TAX97 <- tax_table(as.matrix(read.table("data/97_otus_tax_assignments.txt", sep="\t", row.names=2,
                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX100 <- tax_table(as.matrix(read.table("data/100_tax_assignments.txt", sep="\t", row.names=2,
+TAX97bs <- tax_table(as.matrix(read.table("data/97_otus_bysample_tax_assignments.txt", sep="\t", row.names=2,
+                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
+TAX100 <- tax_table(as.matrix(read.table("data/100_otus_tax_assignments.txt", sep="\t", row.names=2,
                                          col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
@@
 phy97 <- phyloseq(OTU97, TAX97, SAM)
+phy97bs <- phyloseq(OTU97bs, TAX97bs, SAM)
 phy100 <- phyloseq(OTU100, TAX100, SAM)
@@
   return(
-    data.frame(rangeotus=paste0(range(taxa_sums(phy)), collapse=" - "),
-               nsingles=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
-               rangerps=paste0(range(sample_sums(phy)), collapse=" - "),
-               amrps=paste(as.integer(mean(sample_sums(phy))), 
+    data.frame(
+      'Total count in OTU table'=sum(otu_table(phy)),
+      'Number of OTUs'=ntaxa(phy),
+      'Range of OTU counts'=paste0(range(taxa_sums(phy)), collapse=" - "),
+      'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
+      'Number of samples'=nsamples(phy),
+      'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
+      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
                                                      as.integer(sd(sample_sums(phy))), sep=" ± "),
-               gmrps=paste(as.integer(exp(mean(log(sample_sums(phy))))),
-                           as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "))
+      'Geometric mean (±SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
+                                                    as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "),
+      check.names=F)
   )
 }
-stats97 <- stats(phy97)
-stats100 <- stats(phy100)
+stats97 <- data.frame(`97% OTUs`=t(stats(phy97)), check.names=F)
+stats97bs <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs)), check.names=F)
+stats100 <- data.frame(`100% OTUs`=t(stats(phy100)), check.names=F)
 
 # Create and plot histograms
-phy <- phy97
 taxhist97 <- hist(log10(taxa_sums(phy97)), plot=F)
 samhist97 <- hist(log10(sample_sums(phy97)), plot=F)
+
+taxhist97bs <- hist(log10(taxa_sums(phy97bs)), plot=F)
+samhist97bs <- hist(log10(sample_sums(phy97bs)), plot=F)
 
@@
 
-par(mfrow=c(2, 2), mar=c(3,3,1,1))
-plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+par(mfrow=c(2, 3), mar=c(3,3,1,1))
+plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+plot(taxhist97bs, col="black", main="97% within-sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+plot(samhist97bs, col="black", main="97% within-sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+
+# Create stats table
+knitr::kable(cbind(stats97, stats97bs, stats100))
 ```
-
-|                                  |    97% OTUs               |    100% OTUs                |
-| -------------------------------- | ------------------------- | --------------------------- |
-| Total count in OTU table         | `r sum(otu_table(phy97))` |  `r sum(otu_table(phy100))` |
-| Number of OTUs                   | `r ntaxa(phy97)`          |  `r ntaxa(phy100)`          |
-| Range of OTU counts              | `r stats97$rangeotus`     | `r stats100$rangeotus`      |
-| Number of singleton OTUs         | `r stats97$nsingles`      | `r stats100$nsingles`       |
-| Number of samples                | `r nsamples(phy97)`       |  `r nsamples(phy100)`       |
-| Range of reads per sample        | `r stats97$rangerps`      | `r stats100$rangerps`       |
-| Mean (±SD) reads per sample (arithmetic) | `r stats97$amrps` | `r stats100$amrps`          |
-| Mean (±SD) reads per sample (geometric)  | `r stats97$gmrps` | `r stats100$gmrps`          |
 
@@
 #### Filter samples by minimum count
-```{r filter_samples}
+```{r filter_samples, cache=T}
 # Set threshold number of reads
-sn <- 10
+sn <- 0
 # Remove samples with fewer reads than threshold
 phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
+phy97bs.f <- prune_samples(sample_sums(phy97bs)>=sn, phy97bs)
 phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)
@@
 ```
-
-* Number of samples remaining in 97% OTU dataset: `r nsamples(phy97.f) `
-* Number of samples remaining in 100% OTU dataset: `r nsamples(phy100.f) `
 
@@
 taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
+taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]  
 taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
@@
 phy97.f <- prune_taxa(names(taxa97), phy97.f)
+phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
 phy100.f <- prune_taxa(names(taxa100), phy100.f)
+
+# Refilter samples in case removal of OTUs caused any samples to drop below threshold
+sn <- 200
+phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
+phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
+phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
 ```
-
-* Number of OTUs remaining in 97% dataset: `r ntaxa(phy97.f)`
-* Number of OTUs remaining in 97% dataset: `r ntaxa(phy100.f)`
 
@@
 # Compute summary statistics
-stats97.f <- stats(phy97.f)
-stats100.f <- stats(phy100.f)
+stats97.f <- data.frame(`97% OTUs`=t(stats(phy97.f)), check.names=F)
+stats97bs.f <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs.f)), check.names=F)
+stats100.f <- data.frame(`100% OTUs`=t(stats(phy100.f)), check.names=F)
 
@@
 taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
+taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
 taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
 samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
+samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
 samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)
 
-par(mfrow=c(2, 2), mar=c(3,3,1,1))
+par(mfrow=c(2, 3), mar=c(3,3,1,1))
 plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(taxhist97bs, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
@@
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97bs, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
 plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+
+# Create stats table
+knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
 ```
-
-|                                  |    97% OTUs                     |    100% OTUs                |
-| -------------------------------- | ------------------------------- | --------------------------- |
-| Total count in OTU table         | `r sum(otu_table(phy97.f))`     | `r sum(otu_table(phy100.f))`|
-| Number of OTUs                   | `r ntaxa(phy97.f)`              | `r ntaxa(phy100.f)`         |
-| Range of OTU counts              | `r stats97.f$rangeotus`         | `r stats100.f$rangeotus`    |
-| Number of singleton OTUs         | `r stats97.f$nsingles`          | `r stats100.f$nsingles`     |
-| Number of samples                | `r nsamples(phy97.f)`           | `r nsamples(phy100.f)`      |
-| Range of reads per sample        | `r stats97.f$rangerps`          | `r stats100.f$rangerps`     |
-| Mean (±SD) reads per sample (arithmetic) | `r stats97.f$amrps`     | `r stats100.f$amrps`        |
-| Mean (±SD) reads per sample (geometric)  | `r stats97.f$gmrps`     | `r stats100.f$gmrps`        |
 
@@
 phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
+phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
 phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
@@
 phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
+phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
 phy100.f.t <- transform_sample_counts(phy100.f, transform)
@@
 
+# Plot composition of 97% within-sample OTUs colored by clade
+composition(phy97bs.f.p, col=taxcolors[data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade], legend=T)
+
+# Plot composition of 97% within-sample OTUs colored by OTU
+composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)
+
 ```
@@
 sppdiffs97 <- sppdiffs(phy97.f.t)
+sppdiffs97bs <- sppdiffs(phy97bs.f.t)
 sppdiffs100 <- sppdiffs(phy100.f.t)
 
-sppdiffs <- data.frame("97% OTUs"=sppdiffs97$Letters, "100% OTUs"=sppdiffs100$Letters, check.names=F)
+sppdiffs <- data.frame("97% OTUs"=sppdiffs97$Letters, "97% within-sample OTUs"=sppdiffs97bs$Letters,
+                       "100% OTUs"=sppdiffs100$Letters, check.names=F)
 knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
@@
     shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
-  if (nlevels(as(sample_data(phy97.f.t.species), "data.frame")$Shore) > 1) {
-    set.seed(70235)
-    permanova <- adonis(phyloseq::distance(phy97.f.t.species, "bray") ~ Shore, 
-                        data=as(sample_data(phy97.f.t.species), "data.frame"), permutations=999)
+    if (nlevels(as(sample_data(phy.species), "data.frame")$Shore) > 1) {
+      #set.seed(70235)
+      set.seed(43789)
+      permanova <- adonis(phyloseq::distance(phy.species, "bray") ~ Shore, 
+                          data=as(sample_data(phy.species), "data.frame"), permutations=999)
       shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
       shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
-    bd <- betadisper(phyloseq::distance(phy97.f.t.species, "bray"),
-                     group=as(sample_data(phy97.f.t.species), "data.frame")$Shore)
+      bd <- betadisper(phyloseq::distance(phy.species, "bray"),
+                       group=as(sample_data(phy.species), "data.frame")$Shore)
       shorestats$bd[i] <- TukeyHSD(bd)$group[4]
@@
   }
-knitr::kable(shorestats)  # dendrogyra, porites furcata, and siderastrea siderea differ by shore
+  return(shorestats)
+}
+
+shorestats97 <- shorediffs(phy97.f.t)
+shorestats97bs <- shorediffs(phy97bs.f.t)
+shorediffs(phy97bs.f.p)
+shorestats100 <- shorediffs(phy100.f.t)
+
+knitr::kable(shorestats97, row.names=F, 
+             caption="97% OTU data: differences within and between shore for each species")
+knitr::kable(shorestats97bs, row.names=F, 
+             caption="97% within-sample OTU data: differences within and between shore for each species")
+knitr::kable(shorestats100, row.names=F, 
+             caption="100% OTU data: differences within and between shore for each species")
 ```
@@
 This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
-```{r betadisper, cache=T, fig.width=4, fig.height=4}
+```{r betadisper, cache=T, fig.width=10, fig.height=4}
+betad <- function(phy) {
   # Calculate distances to species centroids for each sample in principal coordinate space
-sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
-                    group=data.frame(sample_data(phy97.f.t))$Species, type="centroid",
+  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
+                      group=data.frame(sample_data(phy))$Species, type="centroid",
                       bias.adjust=TRUE)
@@
   saml <- multcompLetters(sampt$pairwise$permuted)
+  # Return results as a list
+  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
+}
+
+betad97 <- betad(phy97.f.t)
+betad97bs <- betad(phy97bs.f.t)
+betad100 <- betad(phy100.f.t)
 
 # Figure: Distance to centroids
-par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
-with(sambdsumm.ord, {
-  plot(mean, type="n", ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+par(mfrow=c(1,3), mar=c(3.5,4,2,1), lwd=1)
+with(betad97$sambdsumm.ord, {
+  plot(mean, type="n", main="97% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
   arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
@@
   text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=saml$Letters[as.character(group)], cex=0.5)
+       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
+})
+with(betad97bs$sambdsumm.ord, {
+  plot(mean, type="n", main="97% within-sample OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
+})
+with(betad100$sambdsumm.ord, {
+  plot(mean, type="n", main="100% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
 })
@@
   heights[heights < 0.04] <- NA
-
   # Plot barplot and OTU names and blast hits
-bars <- barplot(typerelabund, col=rainbow(941), border=NA, las=1, xlab="Sample", ylab="", cex.axis=0.75, cex.names=0.75)
+  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
+                  xlab="", ylab="", xaxt="n")
+  text(bars, -0.06, labels=samples, xpd=T)
+  text(bars, -0.10, labels=shores, xpd=T)
+  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T)
+  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, pos=2)
   mtext(side=2, text = "Relative Abundance", line=2)
@@
          labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
-       cex=0.6)
+         cex=0.75)
   }
+  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
+}
+```
 
-mtext(side=3, "B.) 100% OTUs", xpd=T, adj=0, line=2.5, font=2)
+## Individual coral species
+### *Diploria strigosa*
+```{r dstr, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Diploria strigosa
+dstr97.f.p <- subset_samples(phy97.f.p, Species=="strigosa")
+dstr97bs.f.p <- subset_samples(phy97bs.f.p, Species=="strigosa")
+dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
+# Plot custom barplots for Diploria strigosa
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(dstr97.f.p, main="97% OTUs")
+otubarplot(dstr97bs.f.p, main="97% within-sample OTUs")
+otubarplot(dstr100.f.p, main="100% OTUs")
+```
+
+### *Porites furcata*
+```{r pfur, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Porites furcata
+pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
+pfur97bs.f.p <- subset_samples(phy97bs.f.p, Species=="furcata")
+pfur100.f.p <- subset_samples(phy100.f.p, Species=="furcata")
+# Plot custom barplots for Porites furcata
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(pfur97.f.p, main="97% OTUs")
+otubarplot(pfur97bs.f.p, main="97% within-sample OTUs")
+otubarplot(pfur100.f.p, main="100% OTUs")
+```
+
+### *Orbicella annularis*
+```{r oann, cache=F, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Orbicella annularis
+oann97.f.p <- subset_samples(phy97.f.p, Species=="annularis")
+oann97bs.f.p <- subset_samples(phy97bs.f.p, Species=="annularis")
+oann100.f.p <- subset_samples(phy100.f.p, Species=="annularis")
+# Plot custom barplots for Orbicella annularis
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(oann97.f.p, main="97% OTUs")
+otubarplot(oann97bs.f.p, main="97% within-sample OTUs")
+otubarplot(oann100.f.p, main="100% OTUs")
+```
+
+### *Millepora alcicornis*
+```{r malc, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Millepora alcicornis
+malc97.f.p <- subset_samples(phy97.f.p, Species=="alcicornis")
+malc97bs.f.p <- subset_samples(phy97bs.f.p, Species=="alcicornis")
+malc100.f.p <- subset_samples(phy100.f.p, Species=="alcicornis")
+# Plot custom barplots for Millepora alcicornis
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(malc97.f.p, main="97% OTUs")
+otubarplot(malc97bs.f.p, main="97% within-sample OTUs")
+otubarplot(malc100.f.p, main="100% OTUs")
+```
+
+### *Siderastrea siderea*
+```{r ssid, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Siderastrea siderea
+ssid97.f.p <- subset_samples(phy97.f.p, Species=="siderea")
+ssid97bs.f.p <- subset_samples(phy97bs.f.p, Species=="siderea")
+ssid100.f.p <- subset_samples(phy100.f.p, Species=="siderea")
+# Plot custom barplots for Siderastrea siderea
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(ssid97.f.p, main="97% OTUs")
+otubarplot(ssid97bs.f.p, main="97% within-sample OTUs")
+otubarplot(ssid100.f.p, main="100% OTUs")
+```
+
+### *Favia fragum*
+```{r ffra, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Favia fragum
+ffra97.f.p <- subset_samples(phy97.f.p, Species=="fragum")
+ffra97bs.f.p <- subset_samples(phy97bs.f.p, Species=="fragum")
+ffra100.f.p <- subset_samples(phy100.f.p, Species=="fragum")
+# Plot custom barplots for Favia fragum
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(ffra97.f.p, main="97% OTUs")
+otubarplot(ffra97bs.f.p, main="97% within-sample OTUs")
+otubarplot(ffra100.f.p, main="100% OTUs")
+```
+
+### *Siderastrea radians*
+```{r srad, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Siderastrea radians
+srad97.f.p <- subset_samples(phy97.f.p, Species=="radians")
+srad97bs.f.p <- subset_samples(phy97bs.f.p, Species=="radians")
+srad100.f.p <- subset_samples(phy100.f.p, Species=="radians")
+# Plot custom barplots for Siderastrea radians
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(srad97.f.p, main="97% OTUs")
+otubarplot(srad97bs.f.p, main="97% within-sample OTUs")
+otubarplot(srad100.f.p, main="100% OTUs")
+```
+
+### *Porites astreoides*
+```{r past, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Porites astreoides
+past97.f.p <- subset_samples(phy97.f.p, Species=="astreoides")
+past97bs.f.p <- subset_samples(phy97bs.f.p, Species=="astreoides")
+past100.f.p <- subset_samples(phy100.f.p, Species=="astreoides")
+# Plot custom barplots for Porites astreoides
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(past97.f.p, main="97% OTUs")
+otubarplot(past97bs.f.p, main="97% within-sample OTUs")
+otubarplot(past100.f.p, main="100% OTUs")
+```
+
+### *Dendrogyra cylindrus*
+```{r dcyl, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Dendrogyra cylindrus
+dcyl97.f.p <- subset_samples(phy97.f.p, Species=="cylindrus")
+dcyl97bs.f.p <- subset_samples(phy97bs.f.p, Species=="cylindrus")
+dcyl100.f.p <- subset_samples(phy100.f.p, Species=="cylindrus")
+# Plot custom barplots for Dendrogyra cylindrus
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(dcyl97.f.p, main="97% OTUs")
+otubarplot(dcyl97bs.f.p, main="97% within-sample OTUs")
+otubarplot(dcyl100.f.p, main="100% OTUs")
+```
+
+### *Montastraea cavernosa*
+```{r mcav, cache=T, fig.height=12, fig.width=8}
+# Create subsetted phyloseq objects for Montastraea cavernosa
+mcav97.f.p <- subset_samples(phy97.f.p, Species=="cavernosa")
+mcav97bs.f.p <- subset_samples(phy97bs.f.p, Species=="cavernosa")
+mcav100.f.p <- subset_samples(phy100.f.p, Species=="cavernosa")
+# Plot custom barplots for Montastraea cavernosa
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+otubarplot(mcav97.f.p, main="97% OTUs")
+otubarplot(mcav97bs.f.p, main="97% within-sample OTUs")
+otubarplot(mcav100.f.p, main="100% OTUs")
+```
 
 ```
+
+# Network analysis from tutorial
+```{r networks}
+make.net <- function(physeq, n) {
+  # Remove taxa that are not present in subset
+  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
+  # Convert OTU table to "edges" table (weight = relative abundance)
+  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
+  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
+  edges <- edges[edges$weight>n, ]
+  # Create "nodes" table from sample and tax data
+  sdf <- data.frame(sample_data(physeq))
+  sdf$id <- rownames(sdf)
+  tdf <- data.frame(tax_table(physeq))
+  tdf$id <- rownames(tdf)
+  nodes <- merge(sdf, tdf, all=T)
+  # Create network
+  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+  return(net)
+}
+dstr.net <- make.net(dstr97.f.p, 0)
+
+plot(dstr.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(dstr.net)$weight)^0.2,
+     vertex.label=substr(V(dstr.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(dstr.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+pfur.net <- make.net(pfur97.f.p, 0)
+plot(pfur.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(pfur.net)$weight)^0.2,
+     vertex.label=substr(V(pfur.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(pfur.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+oann97.f.p <- prune_taxa(rowSums(otu_table(oann97.f.p))!=0, oann97.f.p)
+oann.net <- make.net(oann97.f.p, 0)
+plot(oann.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(oann.net)$weight)^0.2,
+     vertex.label=substr(V(oann.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(oann.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+malc.net <- make.net(malc97.f.p, 0)
+plot(malc.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(malc.net)$weight)^0.2,
+     vertex.label=substr(V(malc.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(malc.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+ssid.net <- make.net(ssid97.f.p, 0)
+plot(ssid.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(ssid.net)$weight)^0.2,
+     vertex.label=substr(V(ssid.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(ssid.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+ffra.net <- make.net(ffra97.f.p, 0)
+plot(ffra.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(ffra.net)$weight)^0.2,
+     vertex.label=substr(V(ffra.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(ffra.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+srad.net <- make.net(srad97.f.p, 0)
+plot(srad.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(srad.net)$weight)^0.2,
+     vertex.label=substr(V(srad.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(srad.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+past.net <- make.net(past97.f.p, 0)
+plot(past.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(past.net)$weight)^0.2,
+     vertex.label=substr(V(past.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(past.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+dcyl.net <- make.net(dcyl97.f.p, 0)
+plot(dcyl.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(dcyl.net)$weight)^0.2,
+     vertex.label=substr(V(dcyl.net)$Shore, 1, 4),
+     vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+mcav.net <- make.net(mcav97.f.p, 0)
+plot(mcav.net,
+     edge.curved=0.1, 
+     edge.width=10*(E(mcav.net)$weight)^0.2,
+     vertex.label=substr(V(mcav.net)$Subtype, 1, 4),
+     vertex.color=taxcolors[factor(V(mcav.net)$Clade, levels=c("A","B","C","D","F","G"))])
+
+# Compute node degrees (#links) and use that to set node size:
+deg <- degree(mcav.net, mode="all")
+V(mcav.net)$size <- 20
+
+plot(mcav.net, layout=layout_with_fr(mcav.net))
+
+# Get network with one node for each species.
+# Edge weight is the mean relative abundance of OTU.
+# Convert OTU table to "edges" table (weight = relative abundance)
+  otudf <- data.frame(t(otu_table(phy97.f.p)), check.names=F)
+  # Average relative abundance of OTU within species
+  #sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), FUN=mean)
+  # Percent of samples of species that contain OTU at more than 1% relative abundance
+  sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), 
+                     FUN=function(x) length(x[x>0.01])/length(x))
+
+  rownames(sp.ag) <- sp.ag$Species
+  sp.ag <- sp.ag[, -1]
+  edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
+  edges <- droplevels(edges[edges$weight>0, ])
+  # Create "nodes" table from sample and tax data
+  sdf <- data.frame(id=levels(data.frame(sample_data(phy97.f.p))$Species))
+  tdf <- data.frame(tax_table(phy97.f.p))[rownames(data.frame(tax_table(phy97.f.p))) %in% edges$target, ]
+  tdf$id <- rownames(tdf)
+  nodes <- merge(sdf, tdf, all=T)
+  nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
+  # Create network
+  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+
+  # Best one for now ---------
+  V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
+  V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))) #5*sqrt(degree(net))
+  V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
+                         unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
+  V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
+                                                  "astreoides"="P.ast.", "cavernosa"="M.cav.",
+                                                  "cylindrus"="D.cyl.", "fragum"="F.fra.", 
+                                                  "furcata"="P.fur.", "radians"="S.rad.", 
+                                                  "siderea"="S.sid.", "strigosa"="D.str."))
+  V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
+                         taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
+  E(net)$color <- "gray60"
+  E(net)$width <- 15 * E(net)$weight
+  set.seed(12347)
+  l <- layout.fruchterman.reingold(net)
+  l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
+  plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
+       vertex.label.color="black")
+  legend(x=-1.05, y=-0.92, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
+       col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
+  # --------------------------
+
+```

### 2016-08-10 get_ITS2_database.sh
create and use trimmed ITS2 database

@@ -38,6 +38,47 @@ awk '!/>/ { gsub("N","") }; { print $0 }' data/ITS2db_trimmed.fasta > data/ITS2d
 TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
 sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3'"${TAB}"'Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
 
+
+awk '/>A/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeA_align.fasta
+o-smart-trim --min-percent 80 -E -o data/cladeA_align_trimE.fasta data/cladeA_align.fasta
+o-smart-trim --min-percent 80 -S -o data/cladeA_align_trimES.fasta data/cladeA_align_trimE.fasta
+
+awk '/>B/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeB_align.fasta
+o-smart-trim --min-percent 80 -E -o data/cladeB_align_trimE.fasta data/cladeB_align.fasta
+o-smart-trim --min-percent 80 -S -o data/cladeB_align_trimES.fasta data/cladeB_align_trimE.fasta
+
+awk '/>C/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeC_align.fasta
+o-smart-trim --min-percent 80 -E -o data/cladeC_align_trimE.fasta data/cladeC_align.fasta
+o-smart-trim --min-percent 80 -S -o data/cladeC_align_trimES.fasta data/cladeC_align_trimE.fasta
+
+awk '/>D/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeD_align.fasta
+o-smart-trim --min-percent 80 -E -o data/cladeD_align_trimE.fasta data/cladeD_align.fasta
+o-smart-trim --min-percent 80 -S -o data/cladeD_align_trimES.fasta data/cladeD_align_trimE.fasta
+
+awk '/>E/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeE_align.fasta
+o-smart-trim --min-percent 80 -E -o data/cladeE_align_trimE.fasta data/cladeE_align.fasta
[…]

### 2016-08-10 tax_results.sh
use ITS2db_trimmed to assign taxonomy

@@ -5,7 +5,7 @@ OUT=$2
 
 
 # Assign taxonomy
-assign_taxonomy.py -i $1 -r data/ITS2db.fasta -t data/id_to_taxonomy.txt -m blast -o $DIR/blast_taxonomy
+assign_taxonomy.py -i $1 -r data/ITS2db_trimmed.fasta -t data/id_to_taxonomy.txt -m blast -o $DIR/blast_taxonomy
 
 # Make list of "no blast hits"
 awk '/No blast hit/' $DIR/blast_taxonomy/*_tax_assignments.txt > $DIR/blast_taxonomy/no_blast_hits.txt

### 2016-08-11 get_ITS2_database.sh
trim db seqs to 90% character presence

@@ -40,40 +40,40 @@ sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3'"
 
 
 awk '/>A/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeA_align.fasta
-o-smart-trim --min-percent 80 -E -o data/cladeA_align_trimE.fasta data/cladeA_align.fasta
-o-smart-trim --min-percent 80 -S -o data/cladeA_align_trimES.fasta data/cladeA_align_trimE.fasta
+o-smart-trim --min-percent 90 -E -o data/cladeA_align_trimE.fasta data/cladeA_align.fasta
+o-smart-trim --min-percent 90 -S -o data/cladeA_align_trimES.fasta data/cladeA_align_trimE.fasta
 
 awk '/>B/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeB_align.fasta
-o-smart-trim --min-percent 80 -E -o data/cladeB_align_trimE.fasta data/cladeB_align.fasta
-o-smart-trim --min-percent 80 -S -o data/cladeB_align_trimES.fasta data/cladeB_align_trimE.fasta
+o-smart-trim --min-percent 90 -E -o data/cladeB_align_trimE.fasta data/cladeB_align.fasta
+o-smart-trim --min-percent 90 -S -o data/cladeB_align_trimES.fasta data/cladeB_align_trimE.fasta
 
 awk '/>C/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeC_align.fasta
-o-smart-trim --min-percent 80 -E -o data/cladeC_align_trimE.fasta data/cladeC_align.fasta
-o-smart-trim --min-percent 80 -S -o data/cladeC_align_trimES.fasta data/cladeC_align_trimE.fasta
+o-smart-trim --min-percent 90 -E -o data/cladeC_align_trimE.fasta data/cladeC_align.fasta
+o-smart-trim --min-percent 90 -S -o data/cladeC_align_trimES.fasta data/cladeC_align_trimE.fasta
 
 awk '/>D/ {print; getline; print}' data/ITS2db.fasta | muscle -out data/cladeD_align.fasta
-o-smart-trim --min-percent 80 -E -o data/cladeD_align_trimE.fasta data/cladeD_align.fasta
[…]

### 2016-08-11 tax_results.sh
add BLAST e-value cutoff of 1e-30

@@ -5,7 +5,7 @@ OUT=$2
 
 
 # Assign taxonomy
-assign_taxonomy.py -i $1 -r data/ITS2db_trimmed.fasta -t data/id_to_taxonomy.txt -m blast -o $DIR/blast_taxonomy
+assign_taxonomy.py -i $1 -r data/ITS2db_trimmed.fasta -t data/id_to_taxonomy.txt -m blast -e 1e-30 -o $DIR/blast_taxonomy
 
 # Make list of "no blast hits"
 awk '/No blast hit/' $DIR/blast_taxonomy/*_tax_assignments.txt > $DIR/blast_taxonomy/no_blast_hits.txt

### 2016-08-11 analysis/STJ2012_analysis.Rmd
make networks based on bysample OTUs

@@
     legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
-           cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
+           cex=0.7, legend=c("A", "B", "C", "D", "F", "G"), fill=taxcolors, x.intersp=0.5)
     legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
@@
 # Plot composition of 97% within-sample OTUs colored by clade
-composition(phy97bs.f.p, col=taxcolors[data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade], legend=T)
+composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
 
@@
 
-```
-
 # Network analysis from tutorial
@@
 }
-dstr.net <- make.net(dstr97.f.p, 0)
+dstr.net <- make.net(dstr97bs.f.p, 0)
 
+set.seed(5453890)
 plot(dstr.net,
@@
 
-pfur.net <- make.net(pfur97.f.p, 0)
+pfur.net <- make.net(pfur97bs.f.p, 0)
 plot(pfur.net,
@@
 
-oann97.f.p <- prune_taxa(rowSums(otu_table(oann97.f.p))!=0, oann97.f.p)
-oann.net <- make.net(oann97.f.p, 0)
+oann.net <- make.net(oann97bs.f.p, 0)
 plot(oann.net,
@@
 
-malc.net <- make.net(malc97.f.p, 0)
+malc.net <- make.net(malc97bs.f.p, 0)
 plot(malc.net,
@@
 
-ssid.net <- make.net(ssid97.f.p, 0)
+set.seed(9147)
+ssid.net <- make.net(ssid97bs.f.p, 0)
 plot(ssid.net,
@@
 
-ffra.net <- make.net(ffra97.f.p, 0)
+ffra.net <- make.net(ffra97bs.f.p, 0)
 plot(ffra.net,
@@
 
-srad.net <- make.net(srad97.f.p, 0)
+srad.net <- make.net(srad97bs.f.p, 0)
 plot(srad.net,
@@
 
-past.net <- make.net(past97.f.p, 0)
+past.net <- make.net(past97bs.f.p, 0)
 plot(past.net,
@@
 
-dcyl.net <- make.net(dcyl97.f.p, 0)
+dcyl.net <- make.net(dcyl97bs.f.p, 0)
 plot(dcyl.net,
@@
      edge.width=10*(E(dcyl.net)$weight)^0.2,
-     vertex.label=substr(V(dcyl.net)$Shore, 1, 4),
+     vertex.label=substr(V(dcyl.net)$Subtype, 1, 4),
      vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])
 
-mcav.net <- make.net(mcav97.f.p, 0)
+mcav.net <- make.net(mcav97bs.f.p, 0)
 plot(mcav.net,
@@
 # Convert OTU table to "edges" table (weight = relative abundance)
-  otudf <- data.frame(t(otu_table(phy97.f.p)), check.names=F)
+  otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
   # Average relative abundance of OTU within species
@@
   # Percent of samples of species that contain OTU at more than 1% relative abundance
-  sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), 
+  sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                      FUN=function(x) length(x[x>0.01])/length(x))
@@
   # Create "nodes" table from sample and tax data
-  sdf <- data.frame(id=levels(data.frame(sample_data(phy97.f.p))$Species))
-  tdf <- data.frame(tax_table(phy97.f.p))[rownames(data.frame(tax_table(phy97.f.p))) %in% edges$target, ]
+  sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
+  tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
   tdf$id <- rownames(tdf)
@@
   set.seed(12347)
+  set.seed(12364)
+  set.seed(12374)
   l <- layout.fruchterman.reingold(net)
   l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
+  par(mar=c(0,0,0,0))
   plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
        vertex.label.color="black")
-  legend(x=-1.05, y=-0.92, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
-       col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
-  # --------------------------
-
+  #legend(x=-1.05, y=1.1, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
+  #     col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
 ```

### 2016-08-13 get_ITS2_database.sh
rename intermediate fasta to avoid same name as product

@@ -28,10 +28,10 @@ cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/ITS2_Database_trimF.fasta -o data/
 #   Do not remove sequences that do not have reverse primer sequence
 cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2.fasta -o data/ITS2_Database_trimF2_trimR.fasta
 # Trim reverse primers again for any remaining internally
-cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db_trimmed.fasta
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2_Database_trimF2_trimR.fasta -o data/ITS2db_trimFR.fasta
 
 # Remove Ns if present in sequences -- noting that several were deposited in GenBank using Ns as gaps...
-awk '!/>/ { gsub("N","") }; { print $0 }' data/ITS2db_trimmed.fasta > data/ITS2db.fasta
+awk '!/>/ { gsub("N","") }; { print $0 }' data/ITS2db_trimFR.fasta > data/ITS2db.fasta
 
 
 # Generate id_to_taxonomy file
@@ -87,4 +87,5 @@ rm data/ITS2_Database_ready.fasta
 rm data/ITS2_Database_trimF.fasta
 rm data/ITS2_Database_trimF2.fasta
 rm data/ITS2_Database_trimF2_trimR.fasta
+rm data/ITS2db_trimFR.fasta
 rm data/clade*

### 2016-08-14 analysis/STJ2012_analysis.Rmd
move R functions to separate scripts

@@
 # Compute summary statistics
-stats <- function(phy) {
-  return(
-    data.frame(
-      'Total count in OTU table'=sum(otu_table(phy)),
-      'Number of OTUs'=ntaxa(phy),
-      'Range of OTU counts'=paste0(range(taxa_sums(phy)), collapse=" - "),
-      'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
-      'Number of samples'=nsamples(phy),
-      'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
-      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
-                                                     as.integer(sd(sample_sums(phy))), sep=" ± "),
-      'Geometric mean (±SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
-                                                    as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "),
-      check.names=F)
-  )
-}
-stats97 <- data.frame(`97% OTUs`=t(stats(phy97)), check.names=F)
-stats97bs <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs)), check.names=F)
-stats100 <- data.frame(`100% OTUs`=t(stats(phy100)), check.names=F)
+source("R/phystats.R")
+
+stats97 <- data.frame(`97% OTUs`=t(phystats(phy97)), check.names=F)
+stats97bs <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs)), check.names=F)
+stats100 <- data.frame(`100% OTUs`=t(phystats(phy100)), check.names=F)
 
@@
 # Compute summary statistics
-stats97.f <- data.frame(`97% OTUs`=t(stats(phy97.f)), check.names=F)
-stats97bs.f <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs.f)), check.names=F)
-stats100.f <- data.frame(`100% OTUs`=t(stats(phy100.f)), check.names=F)
+stats97.f <- data.frame(`97% OTUs`=t(phystats(phy97.f)), check.names=F)
+stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
+stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)
 
@@
 # Define function to plot Symbiodinium community composition
-composition <- function(phy, col, legend=T) {
-  samdat <- data.frame(sample_data(phy))
-  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
-  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
-  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
-  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
-  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
-                                           rownames(samdat)])
-  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
-  shorebreaks <- which(shorebreaks==F) - 1
-  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
-  # Make Barplot
-  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
-  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
-  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
-  mtext(side=1, "Relative abundance", cex=0.7, line=1)
-  # Add legend
-  if (legend==T) {
-    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
-           cex=0.7, legend=c("A", "B", "C", "D", "F", "G"), fill=taxcolors, x.intersp=0.5)
-    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
-           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
-           x.intersp=0.3)
-  }
-  # Add grouping bars for Shore
-  for (i in 1:length(shorebreaks)) {
-    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
-    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
-         lwd=0.25, xpd=T)
-  }
-  # Add lines to separate species and species names
-  for (i in 1:length(spbreaks)) {
-    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
-    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
-        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
-                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
-  }
-}
+source("R/composition.R")
 
@@
 ```{r shorediffs, cache=T}
-shorediffs <- function(phy) {
-  # Compute within- and between-shore statistics for each Species
-  shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy))$Species), 
-                                                c("Species", "n", "overall", "within", 
-                                                  "between", "R2", "bd", "p")), ncol=8, nrow=10))
-  for (i in 1:nlevels(data.frame(sample_data(phy))$Species)) {
-    sp <<- levels(data.frame(sample_data(phy))$Species)[i]
-    shorestats$Species[i] <- sp
-    phy.species <- subset_samples(phy, Species==sp)  # Set temp phyloseq object
-    shorestats$n[i] <- nsamples(phy.species)
-    md <- meandist(phyloseq::distance(phy.species, "bray"), 
-                   sample_data(phy.species)$Shore)
-    shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
-    shorestats$between[i] <- summary(md)$B  # Mean dissimilarity between Pools
-    shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
-    if (nlevels(as(sample_data(phy.species), "data.frame")$Shore) > 1) {
-      #set.seed(70235)
+source("R/perms.R")
 set.seed(43789)
-      permanova <- adonis(phyloseq::distance(phy.species, "bray") ~ Shore, 
-                          data=as(sample_data(phy.species), "data.frame"), permutations=999)
-      shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
-      shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
-      bd <- betadisper(phyloseq::distance(phy.species, "bray"),
-                       group=as(sample_data(phy.species), "data.frame")$Shore)
-      shorestats$bd[i] <- TukeyHSD(bd)$group[4]
-    } else {
-      shorestats$R2[i] <- NA
-      shorestats$p[i] <- NA
-      shorestats$bd[i] <- NA
-    }
-  }
-  return(shorestats)
-}
-
-shorestats97 <- shorediffs(phy97.f.t)
-shorestats97bs <- shorediffs(phy97bs.f.t)
-shorediffs(phy97bs.f.p)
-shorestats100 <- shorediffs(phy100.f.t)
+shorestats97 <- perms(phy97.f.t, group="Species", trt="Shore")
+shorestats97bs <- perms(phy97bs.f.t, group="Species", trt="Shore")
+shorestats100 <- perms(phy100.f.t, group="Species", trt="Shore")
 
@@
 ```{r betadisper, cache=T, fig.width=10, fig.height=4}
-betad <- function(phy) {
-  # Calculate distances to species centroids for each sample in principal coordinate space
-  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
-                      group=data.frame(sample_data(phy))$Species, type="centroid",
-                      bias.adjust=TRUE)
-  # Calculate each species' mean distance to centroid and standard error
-  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
-  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
-                            FUN=function(x) sd(x)/sqrt(length(x)))$x
-  # Determine order of decreasing mean dispersion
-  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
-  # Update betadisper object with species in decreasing order of mean dispersion
-  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
-                      group=factor(data.frame(sample_data(phy))$Species,
-                                   levels=as.character(sambdsumm.ord$group)), 
-                      type="centroid", bias.adjust=TRUE)
-  # Use permutations to perform pairwise comparisons of group mean dispersions
-  sampt <- permutest(sambd, pairwise=T)
-  saml <- multcompLetters(sampt$pairwise$permuted)
-  # Return results as a list
-  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
-}
+source("R/betad.R")
 
-betad97 <- betad(phy97.f.t)
-betad97bs <- betad(phy97bs.f.t)
-betad100 <- betad(phy100.f.t)
+betad97 <- betad(phy97.f.t, group="Species")
+betad97bs <- betad(phy97bs.f.t, group="Species")
+betad100 <- betad(phy100.f.t, group="Species")
 
@@
 })
-```
-
-```{r otu_barplots_custom, fig.width=8, fig.height=8}
-otubarplot <- function(phy, main) {
-  samdat <- data.frame(sample_data(phy))
-  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
-                                                 rownames(samdat)])
-  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
-                                                                order(Shore, Site, InputFileName))])]
-  # Get info for plotting OTU names and blast hits on top of barplot
-  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
-  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
-  samples <- colnames(typerelabund)
-  shores <- data.frame(sample_data(phy))[samples, "Shore"]
-  sites <- data.frame(sample_data(phy))[samples, "Site"]
-  divides <- apply(typerelabund, 2, cumsum)
-  heights <- diff(divides)
-  heights[heights < 0.04] <- NA
-  # Plot barplot and OTU names and blast hits
-  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
-                  xlab="", ylab="", xaxt="n")
-  text(bars, -0.06, labels=samples, xpd=T)
-  text(bars, -0.10, labels=shores, xpd=T)
-  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T)
-  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, pos=2)
-  mtext(side=2, text = "Relative Abundance", line=2)
-  for (i in 1:length(bars)) {
-    text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
-         divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
-         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
-         cex=0.75)
-  }
-  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
-}
 ```
@@
 ```{r networks}
-make.net <- function(physeq, n) {
-  # Remove taxa that are not present in subset
-  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
-  # Convert OTU table to "edges" table (weight = relative abundance)
-  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
-  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
-  edges <- edges[edges$weight>n, ]
-  # Create "nodes" table from sample and tax data
-  sdf <- data.frame(sample_data(physeq))
-  sdf$id <- rownames(sdf)
-  tdf <- data.frame(tax_table(physeq))
-  tdf$id <- rownames(tdf)
-  nodes <- merge(sdf, tdf, all=T)
-  # Create network
-  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
-  return(net)
-}
-dstr.net <- make.net(dstr97bs.f.p, 0)
+
+dstr.net <- makenet(dstr97bs.f.p, 0)
 
@@
 
-pfur.net <- make.net(pfur97bs.f.p, 0)
+pfur.net <- makenet(pfur97bs.f.p, 0)
 plot(pfur.net,
@@
 
-oann.net <- make.net(oann97bs.f.p, 0)
+oann.net <- makenet(oann97bs.f.p, 0)
 plot(oann.net,
@@
 
-malc.net <- make.net(malc97bs.f.p, 0)
+malc.net <- makenet(malc97bs.f.p, 0)
 plot(malc.net,
@@
 set.seed(9147)
-ssid.net <- make.net(ssid97bs.f.p, 0)
+ssid.net <- makenet(ssid97bs.f.p, 0)
 plot(ssid.net,
@@
 
-ffra.net <- make.net(ffra97bs.f.p, 0)
+ffra.net <- makenet(ffra97bs.f.p, 0)
 plot(ffra.net,
@@
 
-srad.net <- make.net(srad97bs.f.p, 0)
+srad.net <- makenet(srad97bs.f.p, 0)
 plot(srad.net,
@@
 
-past.net <- make.net(past97bs.f.p, 0)
+past.net <- makenet(past97bs.f.p, 0)
 plot(past.net,
@@
 
-dcyl.net <- make.net(dcyl97bs.f.p, 0)
+dcyl.net <- makenet(dcyl97bs.f.p, 0)
 plot(dcyl.net,
@@
 
-mcav.net <- make.net(mcav97bs.f.p, 0)
+mcav.net <- makenet(mcav97bs.f.p, 0)
 plot(mcav.net,

### 2016-08-14 R/betad.R

@@ -1,0 +1,22 @@
+betad <- function(phy, group) {
+  # Calculate distances to species centroids for each sample in principal coordinate space
+  samdat <- data.frame(sample_data(phy))
+  braydist <- vegdist(t(otu_table(phy)), method="bray")
+  sambd <- betadisper(d=braydist, group=samdat[, group], 
+                      type="centroid", bias.adjust=TRUE)
+  # Calculate each species' mean distance to centroid and standard error
+  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
+  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
+                            FUN=function(x) sd(x)/sqrt(length(x)))$x
+  # Determine order of decreasing mean dispersion
+  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
+  # Update betadisper object with species in decreasing order of mean dispersion
+  sambd <- betadisper(d=braydist, 
+                      group=factor(samdat[, group], levels=as.character(sambdsumm.ord$group)), 
+                      type="centroid", bias.adjust=TRUE)
+  # Use permutations to perform pairwise comparisons of group mean dispersions
+  sampt <- permutest(sambd, pairwise=T)
+  saml <- multcompLetters(sampt$pairwise$permuted)
+  # Return results as a list
+  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
+}

### 2016-08-14 R/composition.R

@@ -1,0 +1,38 @@
+composition <- function(phy, col, legend=T) {
+  samdat <- data.frame(sample_data(phy))
+  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
+  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
+  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
+  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
+  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
+                                           rownames(samdat)])
+  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
+  shorebreaks <- which(shorebreaks==F) - 1
+  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
+  # Make Barplot
+  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
+  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
+  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
+  mtext(side=1, "Relative abundance", cex=0.7, line=1)
+  # Add legend
+  if (legend==T) {
+    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
+           cex=0.7, legend=c("A", "B", "C", "D", "F", "G"), fill=taxcolors, x.intersp=0.5)
+    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
+           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
+           x.intersp=0.3)
+  }
+  # Add grouping bars for Shore
+  for (i in 1:length(shorebreaks)) {
+    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
+    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
[…]

### 2016-08-14 R/makenet.R

@@ -1,0 +1,17 @@
+makenet <- function(physeq, n) {
+  # Remove taxa that are not present in subset
+  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
+  # Convert OTU table to "edges" table (weight = relative abundance)
+  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
+  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
+  edges <- edges[edges$weight>n, ]
+  # Create "nodes" table from sample and tax data
+  sdf <- data.frame(sample_data(physeq))
+  sdf$id <- rownames(sdf)
+  tdf <- data.frame(tax_table(physeq))
+  tdf$id <- rownames(tdf)
+  nodes <- merge(sdf, tdf, all=T)
+  # Create network
+  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+  return(net)
+}

### 2016-08-14 R/otubarplot.R

@@ -1,0 +1,31 @@
+otubarplot <- function(phy, main) {
+  samdat <- data.frame(sample_data(phy))
+  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
+                                           rownames(samdat)])
+  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
+                                                                order(Shore, Site, InputFileName))])]
+  # Get info for plotting OTU names and blast hits on top of barplot
+  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
+  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
+  samples <- colnames(typerelabund)
+  shores <- data.frame(sample_data(phy))[samples, "Shore"]
+  sites <- data.frame(sample_data(phy))[samples, "Site"]
+  divides <- apply(typerelabund, 2, cumsum)
+  heights <- diff(divides)
+  heights[heights < 0.04] <- NA
+  # Plot barplot and OTU names and blast hits
+  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
+                  xlab="", ylab="", xaxt="n")
+  text(bars, -0.06, labels=samples, xpd=T)
+  text(bars, -0.10, labels=shores, xpd=T)
+  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T)
+  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, pos=2)
+  mtext(side=2, text = "Relative Abundance", line=2)
+  for (i in 1:length(bars)) {
+    text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
+         divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
+         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
+         cex=0.75)
+  }
+  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
+}

### 2016-08-14 R/perms.R

@@ -1,0 +1,40 @@
+# Function to test differences in community composition within and between groups by PERMANOVA
+
+perms <- function(phy, group, trt) {
+  groups <- data.frame(sample_data(phy))[, group]
+  # Create data frame to populate with results
+  resdf <- data.frame(
+    matrix(ncol=8, nrow=10,
+           dimnames=list(levels(data.frame(sample_data(phy))[, group]), 
+                         c(group, "n", "overall", "within", "between", "R2", "bd", "p")))
+    )
+  # Compute within- and between-treatment statistics for each group
+  for (i in 1:nlevels(groups)) {
+    group.name <<- levels(groups)[i]
+    resdf[i, group] <- group.name
+    # make subsetted phyloseq object for individual group
+    phy.group <- phy 
+    samdat <- data.frame(sample_data(phy))
+    sample_data(phy.group) <- sample_data(samdat[samdat[[group]]==group.name, ])
+    samdat <- as(sample_data(phy.group), "data.frame") # sample data for subsetted phyloseq object
+    # calculate distances
+    braydist <- phyloseq::distance(phy.group, "bray")
+    resdf$n[i] <- nsamples(phy.group)
+    md <- meandist(braydist, samdat[, trt])
+    resdf$within[i] <- summary(md)$W  # Weighted mean dissimilarity within treatments
+    resdf$between[i] <- summary(md)$B  # Mean dissimilarity between treatments
+    resdf$overall[i] <- summary(md)$D  # Overall dissimilarity
+    if (nlevels(as(sample_data(phy.group), "data.frame")[, trt]) > 1) {
+      permanova <- adonis(braydist ~ get(trt), data=samdat, permutations=999)
+      resdf$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
+      resdf$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
+      bd <- betadisper(braydist, group=samdat[, trt])
+      resdf$bd[i] <- TukeyHSD(bd)$group[4]
+    } else {
+      resdf$R2[i] <- NA
+      resdf$p[i] <- NA
+      resdf$bd[i] <- NA
+    }
+  }
+  return(resdf)
+}

### 2016-08-14 R/phystats.R

@@ -1,0 +1,18 @@
+# Function to calculate summary statistics from phyloseq object
+
+phystats <- function(phy) {
+  return(
+    data.frame(
+      'Total count in OTU table'=sum(otu_table(phy)),
+      'Number of OTUs'=ntaxa(phy),
+      'Range of OTU counts'=paste0(range(taxa_sums(phy)), collapse=" - "),
+      'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
+      'Number of samples'=nsamples(phy),
+      'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
+      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
+                                                     as.integer(sd(sample_sums(phy))), sep=" ± "),
+      'Geometric mean (±SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
+                                                    as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "),
+      check.names=F)
+  )
+}

### 2016-08-14 bioinformatics_run.sh
replace with Makefile

@@ -1,36 +1,0 @@
-
-# Prepare ITS2 reference database
-./get_ITS2_database.sh
-
-# Merge, QC, and trim Illumina reads
-./merge_qc_reads.sh
-
-
-
-# -------------------------------
-# Cluster 97% OTUs
-./otus_97.sh
-
-# Assign taxonomy and generate OTU table for 97% OTUs
-./tax_results.sh data/otus_97/97_otus_rep_set.fasta 97_otus
-# -------------------------------
-
-
-
-# -------------------------------
-# Cluster 97% OTUs within samples
-./otus_97_bysample.sh
-
-# Assign taxonomy and generate OTU table for 97% within-sample OTUs
-./tax_results.sh data/otus_97_bysample/all_rep_set_rep_set.fasta 97_otus_bysample
-# -------------------------------
-
-
-
-# -------------------------------
-# Cluster 100% OTUs
-./otus_100.sh
-
-# Assign taxonomy and generate OTU table for 100% OTUs
-./tax_results.sh data/otus_100/100_otus_rep_set.fasta 100_otus
-# -------------------------------

### 2016-08-15 R/STJ_map.R
move map to R directory

@@ -1,7 +1,7 @@
 # St. John Map
 
 library(rgdal)
-stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line")  # units are in meters
+stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
 stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long
 
 sitecoords <- data.frame(
@@ -12,10 +12,11 @@ sitecoords <- data.frame(
 
 viers <- c(-64.722697, 18.322277)
 
-pdf(file="figures/map.pdf", height=3, width=4)
+#pdf(file="figures/map.pdf", height=3, width=4)
 par(mar=c(0,0,0,0))
 plot(stjmap, lwd=1, col="gray40",
      xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
+box()
 text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
 axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
 axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
@@ -27,26 +28,7 @@ with(sitecoords, {
 })
 points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
 text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
-dev.off()
+#dev.off()
 
 
 
-
-text(x=seq(-157.88, -157.73, 0.04), y=21.403, labels=seq(-157.88, -157.73, 0.04), cex=0.5)
-axis(side=3, at=seq(-157.88, -157.80, 0.04), line=0, tck=0.01, labels=FALSE)
-text(x=seq(-157.88, -157.80, 0.04), y=21.527, labels=seq(-157.88, -157.80, 0.04), cex=0.5)
-axis(side=2, at=seq(21.41, 21.6, 0.04), tck=0.01)
-text(x=-157.892, y=seq(21.41, 21.5, 0.04), labels=seq(21.41, 21.5, 0.04), cex=0.5)
-axis(side=4, at=seq(21.41, 21.6, 0.04), tck=0.01)
[…]

### 2016-08-15 R/makenet.R
network functions

@@ -1,4 +1,7 @@
 makenet <- function(physeq, n) {
+  require(phyloseq)
+  require(reshape2)
+  require(igraph)
   # Remove taxa that are not present in subset
   physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
   # Convert OTU table to "edges" table (weight = relative abundance)

### 2016-08-15 R/plotnet.R

@@ -1,0 +1,9 @@
+plotnet <- function(net) {
+  plot(net,
+       edge.curved=0.1, 
+       edge.width=15*(E(net)$weight)^0.35,
+       vertex.label=unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)),
+       vertex.color=ifelse(is.na(V(net)$Clade), "white",
+                           taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))]),
+       vertex.size=ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))))
+}

### 2016-08-15 analysis/STJ2012_analysis.Rmd
reorganize Rmd

@@
 ---
-title: "STJ2012 analysis"
+title: "*Symbiodinium* communities in reef corals of St. John, U.S.V.I. (2012)"
+subtitle: "Data analysis report"
 author: "Ross Cunning"
-date: "July 28, 2016"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
@@
     toc: yes
+    toc_depth: 2
+    toc_float:
+      collapsed: true
+    code_folding: hide
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+options(scipen = 3, digits = 2)
+```
 
-options(scipen = 3, digits = 2)
-
+```{r setup_2, include=FALSE}
+# Load all R scripts in "R/" directory
+sapply(list.files(path="R", pattern="*.R", full.names=TRUE), source, .GlobalEnv)
+# Load package libraries
 library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr)
-
+# Set colors for plotting clades
 taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
@@
 # Overview
-  This project investigates *Symbiodinium* communities in scleractinian corals sampled at various sites on the north and south shores of St.
+  This project investigates *Symbiodinium* communities in scleractinian and milleporine corals sampled at various sites on the north and south shores of St.
 John, US Virgin Islands in August 2012.
-The ITS2 region of nrDNA isolated from coral samples was amplified and sequenced by paired-end 300bp reads on an Illumina MiSeq platform.
-Sequence data is processed through a custom bioinformatic pipeline and resulting OTU tables are analyzed by various multivariate statistical approaches to ask questions regarding alpha and beta diversity of *Symbiodinium* communities within and across coral species.
-The major aims of this work are to 1) characterize *Symbiodinium* communities in a range of coral species using high-throughput sequencing (many for the first time), and 2) to forward novel bioinformatic and statistical approaches to better understand *Symbiodinium* ecology.
+The ITS2 region of nrDNA isolated from coral samples was amplified and sequenced by paired-end 300 bp reads on an Illumina MiSeq platform.
+Sequence data is processed through a custom bioinformatic pipeline and resulting OTU tables are statistically analyzed to ask questions regarding diversity and community structure of *Symbiodinium* within and across coral species and between shores.
+  The major aims of this work are to:
+
+1. characterize *Symbiodinium* communities in a range of coral species using high-throughput sequencing (many for the first time)
+2. forward novel bioinformatic (within-sample clustering) and statistical (beta diversity/distance to centroid, network analysis) approaches for the use of ITS2 data in *Symbiodinium* ecology.
+
+```{r STJmap, echo=F, message=F, cache=T, fig.height=3, fig.width=4, fig.align="center", fig.cap="Sampling locations in St. John"}
+source("R/STJ_map.R")
+```
 
 # Bioinformatics
-  A [custom pipeline](sequence_processing.txt) is used here to process ITS2 sequence data.
-Briefly, paired reads were merged using the [illumina-utils](http://www.github.com/meren/illumina-utils) and filtered to keep only those with 3 mismatches or less.
-Chimeras were then removed by [usearch](http://www.drive5.com/usearch).
-From here, sequence data was processed using both of two approaches:
+  A custom pipeline is used here to process ITS2 sequence data.
+Briefly, paired reads were merged using [illumina-utils](http://www.github.com/meren/illumina-utils) and filtered to keep only those with 3 mismatches or less.
+Chimeras were then removed by [usearch](http://www.drive5.com/usearch), and singletons were removed.
+From here, sequence data was processed using each of three approaches:
 
 1.
-__97% OTU clustering within samples.__ With this approach, sequences from each individual sample were clustered at 97% similarity and the most abundant sequence from each cluster was selected as the representative sequence and assigned taxonomy by BLAST using a [custom ITS2 reference database](../data/Unaligned_ITS2_Database_23April13.fasta).
-Singletons were excluded.
-Identical OTUs across samples were then merged to create a global OTU table.
+__100% OTU clustering.__ All ITS2 sequences from all samples were clustered at 100% identity using *uclust* and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
+2.
+__97% OTU clustering.__ All ITS2 sequences from all samples were clustered at 97% identity using *uclust* and the most abundant sequence within each OTU was selected as the representative sequence and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
+3.
+__97% OTU clustering within samples.__ Sequences from each individual sample were clustered separately at 97% similarity and the most abundant sequence from each cluster was selected as the representative sequence and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
+OTUs with identical representative sequences were then merged across samples to create a global OTU table.
 By clustering at 97% _within_ samples, as opposed to across the entire dataset, OTUs from different samples that have _different representative sequences_ (i.e., different most abundant sequence within the 97% cluster) will remain separate even if these sequences are >97% similar to each other.
-This approach is used because _Symbiodinium_ ITS2 sequences are know to be multi-copy and vary intragenomically, and distinct _Symbiodinium_ types may contain identical sets of ITS2 sequences that occur in different relative proportions within the genome; therefore, selecting the most abundant sequence variant _within_ a sample prevents a potentially distinct _Symbiodinium_ type from being collapsed into an OTU with a different representative sequence that happens to be more abundant in other samples within the dataset.
-2.
-__100% OTU clustering.__ All ITS2 sequences were clustered at 100% identity and assigned taxonomy by BLAST to the [custom reference database](../data/Unaligned_ITS2_Database_23April13.fasta).
-Singletons were excluded.
+This approach is used because _Symbiodinium_ ITS2 sequences are know to be multi-copy and intragenomically variable, and distinct _Symbiodinium_ types may contain some of the same ITS2 sequence variants in different relative proportions.
+Therefore, selecting the most abundant sequence variant _within_ a sample prevents a potentially distinct _Symbiodinium_ type from being collapsed into an OTU with a different representative sequence that happens to be more abundant in other samples within the dataset.
+For more discussion on this approach, see [this entry in my lab notebook](http://jrcunning.github.io/labnotebook//2016/02/12/ITS2_clustering.html).
 
-OTU tables and taxonomic assignments from both of these bioinformatic approaches are used in comparative downstream analyses.
+OTU tables and taxonomic assignments from each of these bioinformatic approaches are used in comparative downstream analyses.
 
 # Data pre-processing
-## Import dataset
-The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object to facilitate downstream analyses.
+### Import dataset
+The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object (class 'phyloseq') to facilitate downstream analyses.
 ```{r data_import, cache=T}
@@
 
-#### Descriptive statistics and distribution of OTU counts
+### Filter dataset
+```{r filter_dataset, cache=T}
+# Filter OTUs by minimum count
+# Set threshold count
+n <- 10
+# Identify OTUs below threshold count
+taxa97 <- taxa_sums(phy97)[which(taxa_sums(phy97) >= n)]
+taxa97bs <- taxa_sums(phy97bs)[which(taxa_sums(phy97bs) >= n)]  
+taxa100 <- taxa_sums(phy100)[which(taxa_sums(phy100) >= n)]
+# Remove taxa below threshold count
+phy97.f <- prune_taxa(names(taxa97), phy97)
+phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs)
+phy100.f <- prune_taxa(names(taxa100), phy100)
+
+# Filter samples by minimum count
+# Set threshold number of reads
+sn <- 200
+# Remove samples with fewer reads than threshold
+phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
+phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
+phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
+```
+* Minimum count to retain OTU: `r n`
+* Minimum count to retain sample: `r sn`
+
+### Descriptive statistics {.tabset}
+
+#### Filtered dataset
+```{r phy.f_histograms, cache=T, fig.height=4, fig.width=8}
+# Compute summary statistics
+stats97.f <- data.frame(`97% OTUs`=t(phystats(phy97.f)), check.names=F)
+stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
+stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)
+
+# Create and plot histograms
+taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
+taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
+taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
+samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
+samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
+samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)
+
+par(mfrow=c(2, 3), mar=c(3,3,1,1))
+plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(taxhist97bs, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97bs, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+
+# Create stats table
+knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
+```
+
+#### Raw dataset
 ```{r phy_stats, cache=T, fig.height=4, fig.width=8}
 # Compute summary statistics
-source("R/phystats.R")
-
 stats97 <- data.frame(`97% OTUs`=t(phystats(phy97)), check.names=F)
@@
 
-## Filter dataset
-#### Filter samples by minimum count
-```{r filter_samples, cache=T}
-# Set threshold number of reads
-sn <- 0
-# Remove samples with fewer reads than threshold
-phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
-phy97bs.f <- prune_samples(sample_sums(phy97bs)>=sn, phy97bs)
-phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)
-
-```
-
-#### Filter OTUs by minimum count
-```{r filter_taxa}
-# Set threshold count
-n <- 10
-# Identify OTUs below threshold count
-taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
-taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]  
-taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
-# Remove taxa below threshold count
-phy97.f <- prune_taxa(names(taxa97), phy97.f)
-phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
-phy100.f <- prune_taxa(names(taxa100), phy100.f)
-
-# Refilter samples in case removal of OTUs caused any samples to drop below threshold
-sn <- 200
-phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
-phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
-phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
-```
-
-#### Filtered data: descriptive statistics and distribution of OTUs
-```{r phy.f_histograms, fig.height=4, fig.width=8}
-# Compute summary statistics
-stats97.f <- data.frame(`97% OTUs`=t(phystats(phy97.f)), check.names=F)
-stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
-stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)
-
-# Create and plot histograms
-taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
-taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
-taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
-samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
-samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
-samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)
-
-par(mfrow=c(2, 3), mar=c(3,3,1,1))
-plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(taxhist97bs, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-plot(samhist97bs, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-
-# Create stats table
-knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
-```
-
-## Transform count data
+### Transform count data
+Count data are transformed to both relative abundance (proportions) and square-root proportions for downstream statistical analyses.
 ```{r transform}
@@
 
-# Data analysis
-## Community composition barplots
-Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
-Each segment of the bar represents an individual OTU.
-In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
-```{r cladecomp97, cache=T, fig.height=6, fig.width=8}
-# Define function to plot Symbiodinium community composition
-source("R/composition.R")
+# *Symbiodinium* in each coral
+For each coral species, barplots are presented showing the relative abundance of OTUs obtained by 100%, 97%, and 97%-within-sample clustering.
+OTUs comprising more than 4% of a sample are labeled with the unique OTU number and the *Symbiodinium* subtype and NCBI GenBank accession number of the closest BLAST hit for that OTU in the reference database.
+OTU numbers and barplot colors are NOT comparable across the three clustering methods.
 
-par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
-
-# Plot composition of 97% OTUs colored by clade
-composition(phy97.f.p, col=taxcolors[data.frame(tax_table(phy97.f.p))[order(data.frame(tax_table(phy97.f.p))$Subtype), ]$Clade], legend=T)
-
-# Plot composition of 97% OTUs colored by OTU
-composition(phy97.f.p, col=rainbow(ntaxa(phy97.f.p)), legend=F)
-
-# Plot composition of 97% within-sample OTUs colored by clade
-composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
-
-# Plot composition of 97% within-sample OTUs colored by OTU
-composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)
-
-```
-
-```{r cladecomp100, cache=T, fig.height=6, fig.width=8}
-par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
-# Plot composition of 100% OTUs colored by clade
-composition(phy100.f.p, col=taxcolors[data.frame(tax_table(phy100.f.p))[order(data.frame(tax_table(phy100.f.p))$Subtype), ]$Clade], legend=T)
-
-# Plot composition of 100% OTUs colored by OTU
-composition(phy100.f.p, col=rainbow(ntaxa(phy100.f.p)), legend=F)
-```
-
-## Differences between species
-Whether *Symbiodinium* communities differ among species is evaluated using permutational analysis of variance (PERMANOVA).
-
-```{r permanovas, cache=T}
-# PERMANOVA for difference among species
-sppdiffs <- function(phy) {
-  permanova.spp.t <- adonis(phyloseq::distance(phy, "bray") ~ Species, 
-                          data=as(sample_data(phy), "data.frame"), permutations=9999)
-  permanova.spp.t
-  # Pairwise PERMANOVA tests for differences between species
-  dmat <- as(phyloseq::distance(phy, "bray"), "matrix")
-  df <- as(sample_data(phy), "data.frame")
-  pairs <- data.frame(t(combn(levels(df$Species), 2)), stringsAsFactors=F)
-  p.pairs <- setNames(rep(NA, nrow(pairs)), with(pairs, paste(X1, X2, sep='-')))
-  for (i in 1:nrow(pairs)) {
-    dfs <- subset(df, Species %in% c(pairs[i,1], pairs[i,2]))
-    dmats <- dmat[rownames(dfs), rownames(dfs)]
-    set.seed(152470)
-    permanova <- adonis(dmats ~ Species, data=dfs, permutations=999)
-    p.pairs[i] <- permanova$aov.tab$`Pr(>F)`[1]
-  }
-  # Return letters signifying differences 
-  return(multcompLetters(p.pairs))
-}
-
-sppdiffs97 <- sppdiffs(phy97.f.t)
-sppdiffs97bs <- sppdiffs(phy97bs.f.t)
-sppdiffs100 <- sppdiffs(phy100.f.t)
-
-sppdiffs <- data.frame("97% OTUs"=sppdiffs97$Letters, "97% within-sample OTUs"=sppdiffs97bs$Letters,
-                       "100% OTUs"=sppdiffs100$Letters, check.names=F)
-knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
-```
-
-## Differences between shores
-Whether *Symbiodinium* communities differ between north vs. south shores within species is tested here using PERMANOVA.
-
-```{r shorediffs, cache=T}
-source("R/perms.R")
-set.seed(43789)
-shorestats97 <- perms(phy97.f.t, group="Species", trt="Shore")
-shorestats97bs <- perms(phy97bs.f.t, group="Species", trt="Shore")
-shorestats100 <- perms(phy100.f.t, group="Species", trt="Shore")
-
-knitr::kable(shorestats97, row.names=F, 
-             caption="97% OTU data: differences within and between shore for each species")
-knitr::kable(shorestats97bs, row.names=F, 
-             caption="97% within-sample OTU data: differences within and between shore for each species")
-knitr::kable(shorestats100, row.names=F, 
-             caption="100% OTU data: differences within and between shore for each species")
-```
-
-## Beta diversity
-Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group.
-Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test.
-This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
-```{r betadisper, cache=T, fig.width=10, fig.height=4}
-source("R/betad.R")
-
-betad97 <- betad(phy97.f.t, group="Species")
-betad97bs <- betad(phy97bs.f.t, group="Species")
-betad100 <- betad(phy100.f.t, group="Species")
-
-# Figure: Distance to centroids
-par(mfrow=c(1,3), mar=c(3.5,4,2,1), lwd=1)
-with(betad97$sambdsumm.ord, {
-  plot(mean, type="n", main="97% OTUs",
-       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
-  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
-  points(1:10, mean, cex=1, pch=21, bg="white")
-  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
-})
-with(betad97bs$sambdsumm.ord, {
-  plot(mean, type="n", main="97% within-sample OTUs",
-       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
-  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
-  points(1:10, mean, cex=1, pch=21, bg="white")
-  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
-})
-with(betad100$sambdsumm.ord, {
-  plot(mean, type="n", main="100% OTUs",
-       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
-  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
-  points(1:10, mean, cex=1, pch=21, bg="white")
-  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
-})
-```
-
-## Individual coral species
-### *Diploria strigosa*
+## *Diploria strigosa*
 ```{r dstr, cache=T, fig.height=12, fig.width=8}
@@
 
-### *Porites furcata*
+```{r dstr.net, cache=T, fig.height=6, fig.width=6}
+dstr.net <- makenet(dstr97bs.f.p, 0)
+set.seed(54538)
+plotnet(dstr.net)
+```
+
+#### Network analysis for *D. strigosa*
+#### Number of OTUs: `r table(is.na(V(dstr.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+-----
+
+## *Porites furcata*
 ```{r pfur, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r pfur.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+pfur.net <- makenet(pfur97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(pfur.net)
+```
 
-### *Orbicella annularis*
+#### Network analysis for *P. furcata*
+#### Number of OTUs: `r table(is.na(V(pfur.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Orbicella annularis*
 ```{r oann, cache=F, fig.height=12, fig.width=8}
@@
 ```
+```{r oann.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+oann.net <- makenet(oann97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(oann.net)
+```
 
-### *Millepora alcicornis*
+#### Network analysis for *O. annularis*
+#### Number of OTUs: `r table(is.na(V(oann.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Millepora alcicornis*
 ```{r malc, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r malc.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+malc.net <- makenet(malc97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(malc.net)
+```
 
-### *Siderastrea siderea*
+#### Network analysis for *M. alcicornis*
+#### Number of OTUs: `r table(is.na(V(malc.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Siderastrea siderea*
 ```{r ssid, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r ssid.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+ssid.net <- makenet(ssid97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(834)
+plotnet(ssid.net)
+```
 
-### *Favia fragum*
+#### Network analysis for *S. siderea*
+#### Number of OTUs: `r table(is.na(V(ssid.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Favia fragum*
 ```{r ffra, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r ffra.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+ffra.net <- makenet(ffra97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(ffra.net)
+```
 
-### *Siderastrea radians*
+#### Network analysis for *F. fragum*
+#### Number of OTUs: `r table(is.na(V(ffra.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Siderastrea radians*
 ```{r srad, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r srad.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+srad.net <- makenet(srad97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(srad.net)
+```
 
-### *Porites astreoides*
+#### Network analysis for *S. radians*
+#### Number of OTUs: `r table(is.na(V(srad.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Porites astreoides*
 ```{r past, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r past.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+past.net <- makenet(past97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(past.net)
+```
 
-### *Dendrogyra cylindrus*
+#### Network analysis for *P. astreoides*
+#### Number of OTUs: `r table(is.na(V(past.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Dendrogyra cylindrus*
 ```{r dcyl, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r dcyl.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+dcyl.net <- makenet(dcyl97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(dcyl.net)
+```
 
-### *Montastraea cavernosa*
+#### Network analysis for *D. cylindrus*
+#### Number of OTUs: `r table(is.na(V(dcyl.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+## *Montastraea cavernosa*
 ```{r mcav, cache=T, fig.height=12, fig.width=8}
@@
 ```
+```{r mcav.net, cache=T, fig.height=6, fig.width=6}
+# Network analysis
+mcav.net <- makenet(mcav97bs.f.p, 0)
+par(mar=c(0,0,0,0))
+set.seed(54538)
+plotnet(mcav.net)
+```
 
-# Network analysis from tutorial
+#### Network analysis for *M. cavernosa*
+#### Number of OTUs: `r table(is.na(V(mcav.net)$Clade))[[1]]`
+White circles represent coral samples, colored circles represent *Symbiodinium* OTUs (97% within-sample). Thickness of lines corresponds to the relative abundance of an OTU within a sample.
+
+# All species barplot
+Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
+Each segment of the bar represents an individual OTU and is colored by clade.
+This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
+```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
+par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+# Plot composition of 97% within-sample OTUs colored by clade
+composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
+# Plot composition of 97% within-sample OTUs colored by OTU
+#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)
+
+```
+
+# All species network analysis
+In this network, each coral species is collapsed into a single node (white squares), and connected to every *Symbiodinium* OTU (colored circles) that comprises at least 1% relative abundance within at least one individual of that species.
+Thus, very low abundance OTUs are not represented.
+The thickness of connections between coral species and OTUs is scaled by the relative proportion of samples of that species in which that OTU was present at ≥ 1% relative abundance.
+The size of *Symbiodinium* OTU nodes in the network is scaled by the number of species in which they occur.
 ```{r networks}
-
-dstr.net <- makenet(dstr97bs.f.p, 0)
-
-set.seed(5453890)
-plot(dstr.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(dstr.net)$weight)^0.2,
-     vertex.label=substr(V(dstr.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(dstr.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-pfur.net <- makenet(pfur97bs.f.p, 0)
-plot(pfur.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(pfur.net)$weight)^0.2,
-     vertex.label=substr(V(pfur.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(pfur.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-oann.net <- makenet(oann97bs.f.p, 0)
-plot(oann.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(oann.net)$weight)^0.2,
-     vertex.label=substr(V(oann.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(oann.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-malc.net <- makenet(malc97bs.f.p, 0)
-plot(malc.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(malc.net)$weight)^0.2,
-     vertex.label=substr(V(malc.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(malc.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-set.seed(9147)
-ssid.net <- makenet(ssid97bs.f.p, 0)
-plot(ssid.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(ssid.net)$weight)^0.2,
-     vertex.label=substr(V(ssid.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(ssid.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-ffra.net <- makenet(ffra97bs.f.p, 0)
-plot(ffra.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(ffra.net)$weight)^0.2,
-     vertex.label=substr(V(ffra.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(ffra.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-srad.net <- makenet(srad97bs.f.p, 0)
-plot(srad.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(srad.net)$weight)^0.2,
-     vertex.label=substr(V(srad.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(srad.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-past.net <- makenet(past97bs.f.p, 0)
-plot(past.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(past.net)$weight)^0.2,
-     vertex.label=substr(V(past.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(past.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-dcyl.net <- makenet(dcyl97bs.f.p, 0)
-plot(dcyl.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(dcyl.net)$weight)^0.2,
-     vertex.label=substr(V(dcyl.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-mcav.net <- makenet(mcav97bs.f.p, 0)
-plot(mcav.net,
-     edge.curved=0.1, 
-     edge.width=10*(E(mcav.net)$weight)^0.2,
-     vertex.label=substr(V(mcav.net)$Subtype, 1, 4),
-     vertex.color=taxcolors[factor(V(mcav.net)$Clade, levels=c("A","B","C","D","F","G"))])
-
-# Compute node degrees (#links) and use that to set node size:
-deg <- degree(mcav.net, mode="all")
-V(mcav.net)$size <- 20
-
-plot(mcav.net, layout=layout_with_fr(mcav.net))
-
 # Get network with one node for each species.
@@
 ```
+
+# Differences between species
+Whether *Symbiodinium* communities differ among species is evaluated using permutational analysis of variance (PERMANOVA).
+
+```{r permanovas, cache=T}
+# PERMANOVA for difference among species
+sppdiffs <- function(phy) {
+  permanova.spp.t <- adonis(phyloseq::distance(phy, "bray") ~ Species, 
+                          data=as(sample_data(phy), "data.frame"), permutations=9999)
+  permanova.spp.t
+  # Pairwise PERMANOVA tests for differences between species
+  dmat <- as(phyloseq::distance(phy, "bray"), "matrix")
+  df <- as(sample_data(phy), "data.frame")
+  pairs <- data.frame(t(combn(levels(df$Species), 2)), stringsAsFactors=F)
+  p.pairs <- setNames(rep(NA, nrow(pairs)), with(pairs, paste(X1, X2, sep='-')))
+  for (i in 1:nrow(pairs)) {
+    dfs <- subset(df, Species %in% c(pairs[i,1], pairs[i,2]))
+    dmats <- dmat[rownames(dfs), rownames(dfs)]
+    set.seed(152470)
+    permanova <- adonis(dmats ~ Species, data=dfs, permutations=999)
+    p.pairs[i] <- permanova$aov.tab$`Pr(>F)`[1]
+  }
+  # Return letters signifying differences 
+  return(multcompLetters(p.pairs))
+}
+
+sppdiffs97 <- sppdiffs(phy97.f.t)
+sppdiffs97bs <- sppdiffs(phy97bs.f.t)
+sppdiffs100 <- sppdiffs(phy100.f.t)
+
+sppdiffs <- data.frame("97% within-sample OTUs"=sppdiffs97bs$Letters,
+                       "100% OTUs"=sppdiffs100$Letters, 
+                       "97% OTUs"=sppdiffs97$Letters, check.names=F)
+knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
+```
+
+# Differences between shores {.tabset}
+Whether *Symbiodinium* communities differ between north vs. south shores within species is tested here using PERMANOVA.
+
+## 97% within-sample OTUs
+```{r shorediffs97bs, cache=T}
+set.seed(43789)
+shorestats97bs <- perms(phy97bs.f.t, group="Species", trt="Shore")
+knitr::kable(shorestats97bs, row.names=F, 
+             caption="97% within-sample OTU data: differences within and between shore for each species")
+```
+
+## 100% OTUs
+```{r shorediffs100, cache=T}
+set.seed(43789)
+shorestats100 <- perms(phy100.f.t, group="Species", trt="Shore")
+knitr::kable(shorestats100, row.names=F, 
+             caption="100% OTU data: differences within and between shore for each species")
+```
+
+## 97% OTUs
+```{r shorediffs97, cache=T}
+set.seed(43789)
+shorestats97 <- perms(phy97.f.t, group="Species", trt="Shore")
+knitr::kable(shorestats97, row.names=F, 
+             caption="97% OTU data: differences within and between shore for each species")
+```
+
+# Beta diversity {.tabset}
+Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species.
+Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test.
+This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
+In this context, beta diveristy is analogous to 'flexibility' in symbiosis -- how different can corals of the same species be in terms of their symbiont communities?
+
+## 97% within-sample OTUs
+```{r betad97bs, cache=T, fig.width=6, fig.height=6}
+betad97bs <- betad(phy97bs.f.t, group="Species")
+with(betad97bs$sambdsumm.ord, {
+  plot(mean, type="n", main="97% within-sample OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
+})
+```
+
+## 100% OTUs
+```{r betad100, cache=T, fig.width=6, fig.height=6}
+betad100 <- betad(phy100.f.t, group="Species")
+with(betad100$sambdsumm.ord, {
+  plot(mean, type="n", main="100% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
+})
+```
+
+## 97% OTUs
+```{r betad97, cache=T, fig.width=6, fig.height=6}
+betad97 <- betad(phy97.f.t, group="Species")
+# Figure: Distance to centroids
+with(betad97$sambdsumm.ord, {
+  plot(mean, type="n", main="97% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
+})
+```

### 2016-08-15 analysis/STJ2012_analysis.Rmd
add results text

@@
 ## *Diploria strigosa*  
+Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, denovo235.
+In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- denovo73, denovo231, and denovo10.
+Look at the 100% OTU dataset to see the composition of unique sequence variants in these samples.
+Indeed, these samples are dominated by **different sequence variants**.
+However, in the 97% OTU approach, they have all been collapsed into the same OTU, while in the 97% within-sample clustering approach, samples with different dominant sequence variants are maintained as separate OTUs, even though these different sequence variants are more than 97% similar to each other.
+I suggest that the  within-sample clustering approach is a better way of treating *Symbiodinium* ITS2 sequence data.
 ```{r dstr, cache=T, fig.height=12, fig.width=8}
@@
 Thus, very low abundance OTUs are not represented.
-The thickness of connections between coral species and OTUs is scaled by the relative proportion of samples of that species in which that OTU was present at ≥ 1% relative abundance.
-The size of *Symbiodinium* OTU nodes in the network is scaled by the number of species in which they occur.
+The thickness of connections between coral species and OTUs is scaled by the relative proportion of samples of that species in which that OTU was present at ≥1% relative abundance (i.e., the frequency of presence at ≥1%).
+The size of *Symbiodinium* OTU nodes in the network is scaled by the number of species in which they occur, and colored according to clade.
 ```{r networks}
@@
 
+##### Results:
+- At the 97% within-sample OTU scale, *Symbiodinium* communities are **not significantly different** in the corals *M.
+alcicornis*, *O.
+annularis*, and *F.
+fragum*, as these corals are all dominantly associated with the same *Symbiodinium* OTU (BLAST identity=B1).
+These corals are **also not different** from *D.
+cylindrus* or *D.
+strigosa*, which also dominantly associated with the same B1 OTU, although the latter two species are different from each other.
+Meanwhile, *M.
+cavernosa* and *S.
+siderea* also share similar symbiont communities dominated by *Symbiodinium* C3.
+Finally, *S.
+radians*, *P.
+furcata*, and *P.
+astreoides* each have distinct, **significantly different** *Symbiodinium* communities.
+- At the 100% OTU scale, *Symbiodinium* communities were **significantly differentiated** in all species except for *O.
+annularis* and *D.
+strigosa* (sharing B1-dominated communities), and *S.
+siderea* and *M.
+cavernosa* (sharing C3-dominated communities).
+
 # Differences between shores {.tabset}
@@
 set.seed(43789)
+set.seed(43789)
 shorestats97bs <- perms(phy97bs.f.t, group="Species", trt="Shore")
-knitr::kable(shorestats97bs, row.names=F, 
+knitr::kable(shorestats97bs, digits=3, row.names=F, 
              caption="97% within-sample OTU data: differences within and between shore for each species")
@@
 shorestats100 <- perms(phy100.f.t, group="Species", trt="Shore")
-knitr::kable(shorestats100, row.names=F, 
+knitr::kable(shorestats100, digits=3, row.names=F, 
              caption="100% OTU data: differences within and between shore for each species")
@@
 shorestats97 <- perms(phy97.f.t, group="Species", trt="Shore")
-knitr::kable(shorestats97, row.names=F, 
+knitr::kable(shorestats97, digits=3, row.names=F, 
              caption="97% OTU data: differences within and between shore for each species")
@@
 
+##### Results:
+- __*P. furcata* is the only species whose symbiont communities are significantly different between the north and south shores__, at both the 97% within-sample OTU and 100% OTU scales.
+
 # Beta diversity {.tabset}
-Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species.
+- Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species.
 Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test.
 This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
-In this context, beta diveristy is analogous to 'flexibility' in symbiosis -- how different can corals of the same species be in terms of their symbiont communities?
+- In this context, beta diveristy is analogous to 'flexibility' in symbiosis -- how different can corals of the same species be in terms of their symbiont communities? 
+- The species with **highest beta diversity** are *D.
+strigosa*, *P.
+furcata*, and *M.
+alcicornis*, followed by *O.
+annularis* and *S.
+siderea*.
+Species with the **lowest beta diversity** are *F.
+fragum*, *D.
+cylindrus*, *S.
+radians*, *P.
+astreoides*, and *M.
+cavernosa*.
 

### 2016-08-15 analysis/STJ2012_analysis.Rmd
add main findings

@@
 ```
+
+# Main findings
+1.
+Symbiont communities are generally consistent with previous work on these species in the Caribbean, but high-throughput sequencing provides a much **more complete picture** of these communities’ composition.
+2. South shore *P. furcata* are more likely to have clade A instead of C, and *S. siderea* more likely to have clade D instead of C.
+3. **Within-sample OTU clustering** is more appropriate than clustering across samples.
+4. We can apply **new types of analysis** with this much data, including network analysis and statistical tests on beta diversity.

### 2016-08-16 analysis/STJ2012_analysis.Rmd
add background and dominant symbiont networks, reorganize Rmd

@@
 
-# Overview
+# Project
   This project investigates *Symbiodinium* communities in scleractinian and milleporine corals sampled at various sites on the north and south shores of St.
@@
 phy100.f.t <- transform_sample_counts(phy100.f, transform)
+```
+
+# Clade overview
+Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
+Each segment of the bar represents an individual OTU and is colored by clade.
+This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
+```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
+par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+# Plot composition of 97% within-sample OTUs colored by clade
+composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
+# Plot composition of 97% within-sample OTUs colored by OTU
+#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)
+
 ```
@@
 
-# All species barplot
-Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore.
-Each segment of the bar represents an individual OTU and is colored by clade.
-This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
-```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
-par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
-# Plot composition of 97% within-sample OTUs colored by clade
-composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
-# Plot composition of 97% within-sample OTUs colored by OTU
-#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)
-
-```
-
-# All species network analysis
-In this network, each coral species is collapsed into a single node (white squares), and connected to every *Symbiodinium* OTU (colored circles) that comprises at least 1% relative abundance within at least one individual of that species.
+# Network analyses
+## "Resident" symbionts
+#### (>1% relative abundance)
+In this network, each coral species is collapsed into a single node (white squares), and connected to every *Symbiodinium* OTU (colored circles) that comprises **at least 1% relative abundance** within at least one individual of that species.
 Thus, very low abundance OTUs are not represented.
@@
 # Get network with one node for each species.
-# Edge weight is the mean relative abundance of OTU.
-# Convert OTU table to "edges" table (weight = relative abundance)
 otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
-# Average relative abundance of OTU within species
-#sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), FUN=mean)
 # Percent of samples of species that contain OTU at more than 1% relative abundance
@@
                    FUN=function(x) length(x[x>0.01])/length(x))
-
 rownames(sp.ag) <- sp.ag$Species
@@
 #     col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
+```
+
+## Dominant symbionts
+#### (>50% relative abundance)
+```{r dom_networks}
+# Get network with one node for each species.
+otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
+# Percent of samples of species that contain OTU at more than 1% relative abundance
+sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
+                   FUN=function(x) length(x[x>0.5])/length(x))
+rownames(sp.ag) <- sp.ag$Species
+sp.ag <- sp.ag[, -1]
+edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
+edges <- droplevels(edges[edges$weight>0, ])
+# Create "nodes" table from sample and tax data
+sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
+tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
+tdf$id <- rownames(tdf)
+nodes <- merge(sdf, tdf, all=T)
+nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
+# Create network
+net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+
+# Best one for now ---------
+V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
+V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 10*(degree(net))^0.75) #5*sqrt(degree(net))
+V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
+                       unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
+V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
+                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
+                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
+                                                "furcata"="P.fur.", "radians"="S.rad.", 
+                                                "siderea"="S.sid.", "strigosa"="D.str."))
+V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
+                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
+E(net)$color <- "gray60"
+E(net)$width <- 15 * E(net)$weight
+set.seed(12347)
+set.seed(12364)
+set.seed(12374)
+l <- layout.fruchterman.reingold(net)
+l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
+par(mar=c(0,0,0,0))
+plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
+     vertex.label.color="black")
+#legend(x=-1.05, y=1.1, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
+#     col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
+```
+
+## Background symbionts
+#### (<1% relative abundance)
+```{r bg_net}
+# Get network with one node for each species.
+otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
+# Percent of samples of species that contain OTU at less than 1% relative abundance (but more than 0%)
+# Only record if bkgd symbiont was detected in more than one individual
+sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
+                   FUN=function(x) ifelse(length(x[x<0.01 & x>0])>1, # if more than 1 individual has bkgd symbiont
+                                          length(x[x<0.01 & x>0])/length(x), 0))
+sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
+                   FUN=function(x) length(x[x<0.01 & x>0])/length(x))
+sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
+                   FUN=function(x) any(x>0 & x<0.01))
+rownames(sp.ag) <- sp.ag$Species
+sp.ag <- sp.ag[, -1]
+edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
+edges <- droplevels(edges[edges$weight>0, ])
+# Create "nodes" table from sample and tax data
+sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
+tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
+tdf$id <- rownames(tdf)
+nodes <- merge(sdf, tdf, all=T)
+nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
+# Create network
+net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+
+# Best one for now ---------
+V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
+V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 5*degree(net)) #5*sqrt(degree(net))
+V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
+                       unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
+V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
+                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
+                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
+                                                "furcata"="P.fur.", "radians"="S.rad.", 
+                                                "siderea"="S.sid.", "strigosa"="D.str."))
+V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
+                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
+E(net)$color <- "gray60"
+E(net)$width <- 1
+set.seed(12347)
+set.seed(12364)
+set.seed(12374)
+l <- layout.fruchterman.reingold(net)
+l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
+par(mar=c(0,0,0,0))
+#plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
+#     vertex.label.color="black")
+#plot(net, layout=layout_nicely, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
+#     vertex.label.color="black")
+#plot(net, layout=layout_as_tree(net, circular=T))
+#plot(net, layout=layout_with_dh)
+
+net2 <- delete.vertices(net, degree(net)<=3)
+V(net2)$size <- ifelse(is.na(V(net2)$Clade), 15, 4*degree(net2))
+set.seed(42381)
+l2 <- layout.fruchterman.reingold(net2)
+l2 <- norm_coords(l2, ymin=-1, ymax=1, xmin=-1, xmax=1)
+plot(net2, rescale=F, layout=l2*1, edge.curved=0.1, vertex.label.cex=V(net2)$size/20,
+     vertex.label.color="black")
+#plot(net2, layout=layout_as_tree(net2, circular=T), edge.curved=0.1)
+#plot(net2, layout=layout_as_star(net2, center = 18))
+#plot(net2, layout=layout_in_circle(net2))
+#plot(net2, layout=layout_with_kk)
+#plot(net2, layout=layout_as_tree(net2, root=16, circular=T))
+#plot(net2, layout=layout_with_dh)
+#?layout_as_tree
+#V(net2)
+
+#cor<-induced.subgraph(net2,V(net2)$type==1)
+#sym<-induced.subgraph(net2,V(net2)$type==2)
+#polar.layout <- function(radii, angles) {
+#    cbind(radii*cos(angles), -radii*sin(angles))        
+#}
+#cor <- which(V(net2)$type == 1)
+#sym <- which(V(net2)$type == 2)
+#radii <- ifelse(V(net2)$type == 1, 2, 1)
+#angles <- rep.int(0, vcount(net2))
+#angles[sym] <- (1:length(sym)-1) * 2 * pi / length(sym)
+#angles[cor] <- (1:length(cor)-1) * 2 * pi / length(cor)
+#layout <- polar.layout(radii, angles)
+#plot(net2, layout=layout)
+```
+
+# Beta diversity {.tabset}
+- Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species.
+Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test.
+This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
+- In this context, beta diveristy is analogous to 'flexibility' in symbiosis -- how different can corals of the same species be in terms of their symbiont communities? 
+- The species with **highest beta diversity** are *D.
+strigosa*, *P.
+furcata*, and *M.
+alcicornis*, followed by *O.
+annularis* and *S.
+siderea*.
+Species with the **lowest beta diversity** are *F.
+fragum*, *D.
+cylindrus*, *S.
+radians*, *P.
+astreoides*, and *M.
+cavernosa*.
+
+## 97% within-sample OTUs
+```{r betad97bs, cache=T, fig.width=6, fig.height=6}
+betad97bs <- betad(phy97bs.f.t, group="Species")
+with(betad97bs$sambdsumm.ord, {
+  plot(mean, type="n", main="97% within-sample OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
+})
+```
+
+## 100% OTUs
+```{r betad100, cache=T, fig.width=6, fig.height=6}
+betad100 <- betad(phy100.f.t, group="Species")
+with(betad100$sambdsumm.ord, {
+  plot(mean, type="n", main="100% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
+})
+```
+
+## 97% OTUs
+```{r betad97, cache=T, fig.width=6, fig.height=6}
+betad97 <- betad(phy97.f.t, group="Species")
+# Figure: Distance to centroids
+with(betad97$sambdsumm.ord, {
+  plot(mean, type="n", main="97% OTUs",
+       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
+  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
+  points(1:10, mean, cex=1, pch=21, bg="white")
+  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
+       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
+  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
+})
 ```
@@
 
-# Beta diversity {.tabset}
-- Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species.
-Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test.
-This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
-- In this context, beta diveristy is analogous to 'flexibility' in symbiosis -- how different can corals of the same species be in terms of their symbiont communities? 
-- The species with **highest beta diversity** are *D.
-strigosa*, *P.
-furcata*, and *M.
-alcicornis*, followed by *O.
-annularis* and *S.
-siderea*.
-Species with the **lowest beta diversity** are *F.
-fragum*, *D.
-cylindrus*, *S.
-radians*, *P.
-astreoides*, and *M.
-cavernosa*.
-
-## 97% within-sample OTUs
-```{r betad97bs, cache=T, fig.width=6, fig.height=6}
-betad97bs <- betad(phy97bs.f.t, group="Species")
-with(betad97bs$sambdsumm.ord, {
-  plot(mean, type="n", main="97% within-sample OTUs",
-       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
-  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
-  points(1:10, mean, cex=1, pch=21, bg="white")
-  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
-})
-```
-
-## 100% OTUs
-```{r betad100, cache=T, fig.width=6, fig.height=6}
-betad100 <- betad(phy100.f.t, group="Species")
-with(betad100$sambdsumm.ord, {
-  plot(mean, type="n", main="100% OTUs",
-       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
-  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
-  points(1:10, mean, cex=1, pch=21, bg="white")
-  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
-})
-```
-
-## 97% OTUs
-```{r betad97, cache=T, fig.width=6, fig.height=6}
-betad97 <- betad(phy97.f.t, group="Species")
-# Figure: Distance to centroids
-with(betad97$sambdsumm.ord, {
-  plot(mean, type="n", main="97% OTUs",
-       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
-  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
-  points(1:10, mean, cex=1, pch=21, bg="white")
-  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
-  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
-})
-```
-
 # Main findings

### 2016-08-17 Shell/otus_97_byspecies.sh
add 97_byspecies clustering

@@ -1,0 +1,51 @@
+#!/bin/bash
+
+rm -rf data/otus_97_byspecies
+# Split into individual sample files
+rm -rf data/fasta/bysample
+split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_trimmed.fasta -o data/fasta/bysample
+
+rm -rf data/fasta/byspecies; mkdir data/fasta/byspecies
+# Get SampleIDs for each species
+cd data/fasta/bysample
+awk -F"\t" '{print $1".fasta" > $6"_IDs.txt"}' ../../mapping_file.txt
+rm -f Species_IDs.txt
+for spp in `ls *_IDs.txt`; do
+	cat $(cat $spp) > ../byspecies/${spp%%_*}.fasta
+done
+cd ../../..
+
+# OTU clustering at 97% similarity for each species (in parallel)
+clust97() {
+	s=`basename $1`
+	pick_otus.py -i $1 -s 0.97 --denovo_otu_id_prefix $s'_denovo' --optimal_uclust -o data/otus_97_byspecies
+}
+export -f clust97
+parallel -j5 clust97 ::: data/fasta/byspecies/*.fasta
+
+# Pick representative sequence set for each sample
+repset97() {
+	filename=`basename $1`
+	s=${filename%_*}
+	pick_rep_set.py -i $1 -m most_abundant -f 'data/fasta/byspecies/'$s'.fasta' \
+	-o 'data/otus_97_byspecies/'$s'_rep_set.fasta'
+}
+export -f repset97
+parallel repset97 ::: data/otus_97_byspecies/*_otus.txt
+
+cat data/otus_97_byspecies/*otus.txt > data/otus_97_byspecies/all_97_otus.txt
+# Merge all rep sets together
+cat data/otus_97_byspecies/*_rep_set.fasta > data/otus_97_byspecies/all_rep_set.fasta
+
+# Cluster all 97% rep sets at 100% identity
+pick_otus.py -i data/otus_97_byspecies/all_rep_set.fasta -s 1.0 --optimal_uclust -o data/otus_97_byspecies
+
+# Get rep set of 100% OTUs and assign taxonomy
+pick_rep_set.py -i data/otus_97_byspecies/all_rep_set_otus.txt -f data/otus_97_byspecies/all_rep_set.fasta \
+-o data/otus_97_byspecies/all_rep_set_rep_set.fasta
+
[…]

### 2016-08-17 analysis/STJ2012_analysis.Rmd
add 97% within-species OTU analysis

@@
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
+OTU97bsp <- otu_table(read.table("data/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
+                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
 OTU100 <- otu_table(read.table("data/100_otus.tsv", header=T, check.names=F, row.names=1,
@@
 TAX97bs <- tax_table(as.matrix(read.table("data/97_otus_bysample_tax_assignments.txt", sep="\t", row.names=2,
+                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
+TAX97bsp <- tax_table(as.matrix(read.table("data/97_otus_byspecies_tax_assignments.txt", sep="\t", row.names=2,
                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
@@
 phy97bs <- phyloseq(OTU97bs, TAX97bs, SAM)
+phy97bsp <- phyloseq(OTU97bsp, TAX97bsp, SAM)
 phy100 <- phyloseq(OTU100, TAX100, SAM)
@@
 taxa97bs <- taxa_sums(phy97bs)[which(taxa_sums(phy97bs) >= n)]
+taxa97bsp <- taxa_sums(phy97bsp)[which(taxa_sums(phy97bsp) >= n)] 
 taxa100 <- taxa_sums(phy100)[which(taxa_sums(phy100) >= n)]
@@
 phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs)
+phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp)
 phy100.f <- prune_taxa(names(taxa100), phy100)
@@
 phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
+phy97bsp.f <- prune_samples(sample_sums(phy97bsp.f)>=sn, phy97bsp.f)
 phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
@@
 stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
+stats97bsp.f <- data.frame(`97% within-species OTUs`=t(phystats(phy97bsp.f)), check.names=F)
 stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)
@@
 taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
+taxhist97bsp <- hist(log10(taxa_sums(phy97bsp.f)), plot=F)
 taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
@@
 samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
+samhist97bsp <- hist(log10(sample_sums(phy97bsp.f)), plot=F)
 samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)
 
-par(mfrow=c(2, 3), mar=c(3,3,1,1))
+par(mfrow=c(2, 4), mar=c(3,3,1,1))
 plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(taxhist97bs, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+plot(taxhist97bs, col="black", main="97% sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(taxhist97bsp, col="black", main="97% species OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
@@
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-plot(samhist97bs, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+plot(samhist97bs, col="black", main="97% sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97bsp, col="black", main="97% species OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
@@
 # Create stats table
-knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
+knitr::kable(cbind(stats97.f, stats97bs.f, stats97bsp.f, stats100.f))
 ```
@@
 stats97bs <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs)), check.names=F)
+stats97bsp <- data.frame(`97% within-species OTUs`=t(phystats(phy97bsp)), check.names=F)
 stats100 <- data.frame(`100% OTUs`=t(phystats(phy100)), check.names=F)
@@
 
+taxhist97bsp <- hist(log10(taxa_sums(phy97bsp)), plot=F)
+samhist97bsp <- hist(log10(sample_sums(phy97bsp)), plot=F)
+
 taxhist100 <- hist(log10(taxa_sums(phy100)), plot=F)
@@
 
-par(mfrow=c(2, 3), mar=c(3,3,1,1))
+par(mfrow=c(2, 4), mar=c(3,3,1,1))
 plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
 plot(taxhist97bs, col="black", main="97% within-sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+plot(taxhist97bsp, col="black", main="97% within-species OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
@@
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+plot(samhist97bs, col="black", main="97% within-species OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
+     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
 plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
 # Create stats table
-knitr::kable(cbind(stats97, stats97bs, stats100))
+knitr::kable(cbind(stats97, stats97bs, stats97bsp, stats100))
 ```
@@
 phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
+phy97bsp.f.p <- transform_sample_counts(phy97bsp.f, function(x) x/sum(x))
 phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
@@
 phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
+phy97bsp.f.t <- transform_sample_counts(phy97bsp.f, transform) 
 phy100.f.t <- transform_sample_counts(phy100.f, transform)
@@
 I suggest that the  within-sample clustering approach is a better way of treating *Symbiodinium* ITS2 sequence data.
-```{r dstr, cache=T, fig.height=12, fig.width=8}
+```{r dstr, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Diploria strigosa
@@
 dstr97bs.f.p <- subset_samples(phy97bs.f.p, Species=="strigosa")
+dstr97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="strigosa")
 dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
 # Plot custom barplots for Diploria strigosa
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(dstr97.f.p, main="97% OTUs")
+otubarplot(dstr97bsp.f.p, main="97% within-species OTUs")
 otubarplot(dstr97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Porites furcata*
-```{r pfur, cache=T, fig.height=12, fig.width=8}
+```{r pfur, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Porites furcata
 pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
+pfur97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="furcata")
 pfur97bs.f.p <- subset_samples(phy97bs.f.p, Species=="furcata")
@@
 # Plot custom barplots for Porites furcata
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(pfur97.f.p, main="97% OTUs")
+otubarplot(pfur97bsp.f.p, main="97% within-species OTUs")
 otubarplot(pfur97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Orbicella annularis*
-```{r oann, cache=F, fig.height=12, fig.width=8}
+```{r oann, cache=F, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Orbicella annularis
 oann97.f.p <- subset_samples(phy97.f.p, Species=="annularis")
+oann97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="annularis")
 oann97bs.f.p <- subset_samples(phy97bs.f.p, Species=="annularis")
@@
 # Plot custom barplots for Orbicella annularis
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(oann97.f.p, main="97% OTUs")
+otubarplot(oann97bsp.f.p, main="97% within-species OTUs")
 otubarplot(oann97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Millepora alcicornis*
-```{r malc, cache=T, fig.height=12, fig.width=8}
+```{r malc, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Millepora alcicornis
 malc97.f.p <- subset_samples(phy97.f.p, Species=="alcicornis")
+malc97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="alcicornis")
 malc97bs.f.p <- subset_samples(phy97bs.f.p, Species=="alcicornis")
@@
 # Plot custom barplots for Millepora alcicornis
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(malc97.f.p, main="97% OTUs")
+otubarplot(malc97bsp.f.p, main="97% within-sample OTUs")
 otubarplot(malc97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Siderastrea siderea*
-```{r ssid, cache=T, fig.height=12, fig.width=8}
+```{r ssid, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Siderastrea siderea
 ssid97.f.p <- subset_samples(phy97.f.p, Species=="siderea")
+ssid97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="siderea")
 ssid97bs.f.p <- subset_samples(phy97bs.f.p, Species=="siderea")
@@
 # Plot custom barplots for Siderastrea siderea
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(ssid97.f.p, main="97% OTUs")
+otubarplot(ssid97bsp.f.p, main="97% within-species OTUs")
 otubarplot(ssid97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Favia fragum*
-```{r ffra, cache=T, fig.height=12, fig.width=8}
+```{r ffra, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Favia fragum
 ffra97.f.p <- subset_samples(phy97.f.p, Species=="fragum")
+ffra97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="fragum")
 ffra97bs.f.p <- subset_samples(phy97bs.f.p, Species=="fragum")
@@
 # Plot custom barplots for Favia fragum
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(ffra97.f.p, main="97% OTUs")
+otubarplot(ffra97bsp.f.p, main="97% within-species OTUs")
 otubarplot(ffra97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Siderastrea radians*
-```{r srad, cache=T, fig.height=12, fig.width=8}
+```{r srad, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Siderastrea radians
 srad97.f.p <- subset_samples(phy97.f.p, Species=="radians")
+srad97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="radians")
 srad97bs.f.p <- subset_samples(phy97bs.f.p, Species=="radians")
@@
 # Plot custom barplots for Siderastrea radians
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(srad97.f.p, main="97% OTUs")
+otubarplot(srad97bsp.f.p, main="97% within-species OTUs")
 otubarplot(srad97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Porites astreoides*
-```{r past, cache=T, fig.height=12, fig.width=8}
+```{r past, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Porites astreoides
 past97.f.p <- subset_samples(phy97.f.p, Species=="astreoides")
+past97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="astreoides")
 past97bs.f.p <- subset_samples(phy97bs.f.p, Species=="astreoides")
@@
 # Plot custom barplots for Porites astreoides
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(past97.f.p, main="97% OTUs")
+otubarplot(past97bsp.f.p, main="97% within-species OTUs")
 otubarplot(past97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Dendrogyra cylindrus*
-```{r dcyl, cache=T, fig.height=12, fig.width=8}
+```{r dcyl, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Dendrogyra cylindrus
 dcyl97.f.p <- subset_samples(phy97.f.p, Species=="cylindrus")
+dcyl97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="cylindrus")
 dcyl97bs.f.p <- subset_samples(phy97bs.f.p, Species=="cylindrus")
@@
 # Plot custom barplots for Dendrogyra cylindrus
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(dcyl97.f.p, main="97% OTUs")
+otubarplot(dcyl97bsp.f.p, main="97% within-species OTUs")
 otubarplot(dcyl97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Montastraea cavernosa*
-```{r mcav, cache=T, fig.height=12, fig.width=8}
+```{r mcav, cache=T, fig.height=16, fig.width=8}
 # Create subsetted phyloseq objects for Montastraea cavernosa
 mcav97.f.p <- subset_samples(phy97.f.p, Species=="cavernosa")
+mcav97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="cavernosa")
 mcav97bs.f.p <- subset_samples(phy97bs.f.p, Species=="cavernosa")
@@
 # Plot custom barplots for Montastraea cavernosa
-par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(mcav97.f.p, main="97% OTUs")
+otubarplot(mcav97bsp.f.p, main="97% within-species OTUs")
 otubarplot(mcav97bs.f.p, main="97% within-sample OTUs")

### 2016-08-24 Shell/merge_qc_reads.sh
remove seqs <250bp

@@ -22,15 +22,14 @@ identify_chimeric_seqs.py -i data/fasta/combined_seqs.fna --suppress_usearch61_r
 filter_fasta.py -f data/fasta/combined_seqs.fna -o data/fasta/combined_seqs_chimera_filtered.fasta -s data/fasta/usearch61_chimeras/chimeras.txt -n
 
 # Trim primer sequences using cutadapt
-# Trim forward primers (allow error rate of 15% (3 indels/mismatches))
-cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/fasta/combined_seqs_chimera_filtered.fasta -o data/fasta/trimF.fasta
+# Trim forward primers (allow error rate of 15% (3 indels/mismatches)) and keep only trimmed sequences
+cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/fasta/combined_seqs_chimera_filtered.fasta --trimmed-only -o data/fasta/trimF.fasta
 # Trim forward primers again (may be multiple internal primer sequences), but do not discard sequences that do not contain primer
 cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/fasta/trimF.fasta -o data/fasta/trimF2.fasta
-# Trim reverse primers using cutadapt
-#   Do not remove sequences that do not have reverse primer sequence
-cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/fasta/trimF2.fasta -o data/fasta/trimF2_trimR.fasta
-# Trim reverse primers again for any remaining internally
-cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/fasta/trimF2_trimR.fasta -o data/fasta/combined_seqs_trimmed.fasta
+# Trim reverse primers using cutadapt and keep only trimmed sequences
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/fasta/trimF2.fasta --trimmed-only -o data/fasta/trimF2_trimR.fasta
+# Trim reverse primers again for any remaining internally and remove sequences shorter than 250 bp
+cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/fasta/trimF2_trimR.fasta -m 250 -o data/fasta/combined_seqs_trimmed.fasta
 
 # Remove intermediate files
 rm data/fasta/trimF.fasta

### 2016-08-24 analysis/STJ2012_analysis.Rmd

@@
 phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
+
+# Filter OTUs by minimum count again in case any dropped below threshold after filtering samples
+# Identify OTUs below threshold count
+taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
+taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
+taxa97bsp <- taxa_sums(phy97bsp.f)[which(taxa_sums(phy97bsp.f) >= n)] 
+taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
+# Remove taxa below threshold count
+phy97.f <- prune_taxa(names(taxa97), phy97.f)
+phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
+phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp.f)
+phy100.f <- prune_taxa(names(taxa100), phy100.f)
 ```
@@
 ```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
+cladeAbund <- aggregate(data.frame(RelAbund=rowSums(otu_table(phy97bs.f.p))),
+                        by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
+sum(cladeAbund$RelAbund)
+cladeAbund$Prop <- prop.table(cladeAbund$RelAbund)
+pie(cladeAbund$Prop, col=taxcolors)
+dotchart(cladeAbund$Prop)
+bars <- barplot(cladeAbund$Prop*100, col=taxcolors, space=0,
+                names.arg=cladeAbund$Clade, xlab=expression(paste(italic('Symbiodinium'), " Clade")),
+                ylab="Relative abundance (%)")
+text(bars, cladeAbund$Prop*100+2, labels=round(cladeAbund$Prop*100, 1), xpd=T)
+
+cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
+cladeAbund
+
 par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
@@
 The size of *Symbiodinium* OTU nodes in the network is scaled by the number of species in which they occur, and colored according to clade.
-```{r networks}
+```{r networks, fig.width=5, fig.height=5, fig.align="center"}
 # Get network with one node for each species.
@@
 #### (>50% relative abundance)
-```{r dom_networks}
+```{r dom_networks, fig.width=5, fig.height=5, fig.align="center"}
 # Get network with one node for each species.
 otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
-# Percent of samples of species that contain OTU at more than 1% relative abundance
+# Percent of samples of species that contain OTU at more than 50% relative abundance
 sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
@@
 #### (<1% relative abundance)
-```{r bg_net}
+```{r bk.net, fig.width=5, fig.height=5, fig.align="center"}
+# filter out OTUs that only occurred in one sample
+phy97bs.f.p2 <- filter_taxa(phy97bs.f.p, function(x) sum(x>0) > 1, prune=TRUE)
 # Get network with one node for each species.
-otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
+otudf <- data.frame(t(otu_table(phy97bs.f.p2)), check.names=F)
 # Percent of samples of species that contain OTU at less than 1% relative abundance (but more than 0%)
-# Only record if bkgd symbiont was detected in more than one individual
-sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
-                   FUN=function(x) ifelse(length(x[x<0.01 & x>0])>1, # if more than 1 individual has bkgd symbiont
-                                          length(x[x<0.01 & x>0])/length(x), 0))
-sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
-                   FUN=function(x) length(x[x<0.01 & x>0])/length(x))
-sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
+
+sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p2))$Species), 
                    FUN=function(x) any(x>0 & x<0.01))
@@
 # Create "nodes" table from sample and tax data
-sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
-tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
+sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p2))$Species))
+tdf <- data.frame(tax_table(phy97bs.f.p2))[rownames(data.frame(tax_table(phy97bs.f.p2))) %in% edges$target, ]
 tdf$id <- rownames(tdf)
@@
 
-# Best one for now ---------
+# Edit vertex attributes
 V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
@@
 E(net)$color <- "gray60"
-E(net)$width <- 1
-set.seed(12347)
-set.seed(12364)
-set.seed(12374)
-l <- layout.fruchterman.reingold(net)
-l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
+E(net)$width <- E(net)$weight * 2
+
+# Merge D nodes (assuming they are all D1a/trenchii)
+# Get original node map (sequence starting at one to the total number of nodes)
+nodemap <- seq(1:length(V(net)))
+# Make a copy of the nodemap that will be modified
+nodemapnew <- nodemap
+# Find all nodes beginning with D, and give them the same node ID number
+nodemapnew[grep("^D", V(net)$Subtype)] <- min(grep("^D", V(net)$Subtype))
+# Replace the now missing values in the node map so node IDs remain sequential
+while (max(setdiff(nodemap, nodemapnew)) < max(nodemap)) {
+  nodemapnew[which(nodemapnew==max(setdiff(nodemap, nodemapnew))+1)] <- min(setdiff(nodemap, nodemapnew))
+}
+# Contract the network using the new nodemap (nodes with same node ID are merged)
+net2 <- contract(net, nodemapnew, vertex.attr.comb=list(size="sum", "last"))
+# Simplify network so that multiple edges are combined
+net2 <- simplify(net2, remove.multiple=T, edge.attr.comb=list(weight="mean", width="mean", "ignore"))
+
+# remove symbionts that are only connected to one or two coral species
+net3 <- delete.vertices(net2, degree(net2)<=2) # & V(net2)$type==2)
+V(net3)$size <- ifelse(is.na(V(net3)$Clade), 15, 1.5*degree(net3)^1.5)
+set.seed(42384)
+l3 <- layout.fruchterman.reingold(net3, weights=E(net3)$weight)
+l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
 par(mar=c(0,0,0,0))
-#plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
-#     vertex.label.color="black")
-#plot(net, layout=layout_nicely, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
-#     vertex.label.color="black")
-#plot(net, layout=layout_as_tree(net, circular=T))
-#plot(net, layout=layout_with_dh)
+plot(net3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(net3)$size/20,
+     vertex.label.color="black")
+points(0.9,0.9, pch=0, cex=4)
+text(0.9,0.9, "O.ann.", cex=0.5)
 
-net2 <- delete.vertices(net, degree(net)<=3)
-V(net2)$size <- ifelse(is.na(V(net2)$Clade), 15, 4*degree(net2))
-set.seed(42381)
-l2 <- layout.fruchterman.reingold(net2)
-l2 <- norm_coords(l2, ymin=-1, ymax=1, xmin=-1, xmax=1)
-plot(net2, rescale=F, layout=l2*1, edge.curved=0.1, vertex.label.cex=V(net2)$size/20,
-     vertex.label.color="black")
-#plot(net2, layout=layout_as_tree(net2, circular=T), edge.curved=0.1)
-#plot(net2, layout=layout_as_star(net2, center = 18))
-#plot(net2, layout=layout_in_circle(net2))
-#plot(net2, layout=layout_with_kk)
-#plot(net2, layout=layout_as_tree(net2, root=16, circular=T))
-#plot(net2, layout=layout_with_dh)
-#?layout_as_tree
-#V(net2)
-
-#cor<-induced.subgraph(net2,V(net2)$type==1)
-#sym<-induced.subgraph(net2,V(net2)$type==2)
-#polar.layout <- function(radii, angles) {
-#    cbind(radii*cos(angles), -radii*sin(angles))        
-#}
-#cor <- which(V(net2)$type == 1)
-#sym <- which(V(net2)$type == 2)
-#radii <- ifelse(V(net2)$type == 1, 2, 1)
-#angles <- rep.int(0, vcount(net2))
-#angles[sym] <- (1:length(sym)-1) * 2 * pi / length(sym)
-#angles[cor] <- (1:length(cor)-1) * 2 * pi / length(cor)
-#layout <- polar.layout(radii, angles)
-#plot(net2, layout=layout)
 ```

### 2016-08-24 Shell/get_ITS2_database.sh
get Green2014 seqs within get_ITS2_database.sh

@@ -17,6 +17,15 @@ cat data/dbseqs/*.fasta > data/ITS2_Database.fasta
 awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' data/ITS2_Database.fasta > data/ITS2_Database_inline.fasta
 awk '!/^>gi/ {print}' data/ITS2_Database_inline.fasta > data/ITS2_Database_ready.fasta
 
+# Download sequences from supplementary material of Green et al. 2014 and add to database
+curl https://dfzljdn9uc3pi.cloudfront.net/2014/386/1/DataS1_PerlScripts_alignment_Bioinformatics.zip > data/green.zip
+unzip -p data/green.zip DataS1* > data/greenseqs.aln
+awk '/Haplotype/ {a[$1] = a[$1]"\n"$2}END{for(i in a){print ">B1."i""a[i]}}' data/greenseqs.aln | sed 's/-//g' | \
+awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' | \
+sed 's/Haplotype\(.*$\)/\1_Green2014/' > data/greenseqs.fasta
+rm data/green.zip
+rm data/greenseqs.aln
+cat data/greenseqs.fasta >> data/ITS2_Database_ready.fasta
 
 # Trim primers from database sequences using cutadapt
 # Trim forward primers using cutadapt
@@ -36,7 +45,7 @@ awk '!/>/ { gsub("N","") }; { print $0 }' data/ITS2db_trimFR.fasta > data/ITS2db
 
 # Generate id_to_taxonomy file
 TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
-sed -e 's/>\([A-Z]\)\(.*\)\(_[A-Z][A-Z][0-9][0-9][0-9][0-9][0-9][0-9]\)/\1\2\3'"${TAB}"'Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
[…]

### 2016-08-24 Shell/get_ITS2_database.sh
use permanent PeerJ link to Green2014 data

@@ -18,7 +18,7 @@ awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { print
 awk '!/^>gi/ {print}' data/ITS2_Database_inline.fasta > data/ITS2_Database_ready.fasta
 
 # Download sequences from supplementary material of Green et al. 2014 and add to database
-curl https://dfzljdn9uc3pi.cloudfront.net/2014/386/1/DataS1_PerlScripts_alignment_Bioinformatics.zip > data/green.zip
+curl https://peerj.com/articles/386/DataS1_PerlScripts_alignment_Bioinformatics.zip > data/green.zip
 unzip -p data/green.zip DataS1* > data/greenseqs.aln
 awk '/Haplotype/ {a[$1] = a[$1]"\n"$2}END{for(i in a){print ">B1."i""a[i]}}' data/greenseqs.aln | sed 's/-//g' | \
 awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' | \

### 2016-08-26 Shell/get_ITS2_database.sh
identify duplicates in trimmed database

@@ -99,3 +99,20 @@ rm data/ITS2_Database_trimF2_trimR.fasta
 rm data/ITS2db_trimFR.fasta
 rm data/clade*
 rm data/greenseqs.fasta
+
+
+### Identify identical reference sequences (post-trimming)
+# Get database sequences that are not uniqueq
+usearch -derep_fulllength data/ITS2db_trimmed.fasta -output data/ITS2db_trimmed_notunique.fasta -sizeout -minuniquesize 2
+# Get fasta file of all the non-unique sequences
+usearch -usearch_global data/ITS2db_trimmed.fasta -db data/ITS2db_trimmed_notunique.fasta -strand plus -id 1.0 \
+-matched data/ITS2db_trimmed_notuniques.fasta
+# Re-cluster non-unique sequences at 100% identity to get list of which ones are identical
+pick_otus.py -i data/ITS2db_trimmed_notuniques.fasta -s 1.0 -o data
+# list of identical sequence groups is in output file: data/ITS2db_trimmed_notuniques_otus.txt
+
+# Clean up
+rm data/ITS2db_trimmed_notunique.fasta
+rm data/ITS2db_trimmed_notuniques_clusters.uc
+rm data/ITS2db_trimmed_notuniques.fasta
+rm data/ITS2db_trimmed_notuniques_otus.log

### 2016-08-26 Shell/tax_local_global.sh
script both local and global alignment tax assignment

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+dir=$(dirname $1)
+
+### Get blast results (local alignment) for each sequence
+formatdb -p F -i data/ITS2db_trimmed.fasta
+blastall -p blastn -i $1 -d data/ITS2db_trimmed.fasta -b 1 -m 8 > $dir/blast_results.txt
+# Sort by query sequence and then E-value in descending order, then get only the first hist for each query sequence
+sort -k1,1 -k11,11nr $dir/blast_results.txt | sort -u -k1,1 > $dir/blast_results_filtered.txt
+# cleanup 
+rm -f data/ITS2db_trimmed.fasta.*
+
+### Get top needle hit (global alignment) for each sequence
+# split fasta file into separate file for each sequence to feed to get_needle
+rm -rf $dir/needle; mkdir $dir/needle
+split -l 2 $1 $dir/needle/
+# rename files according to OTU name
+for i in $dir/needle/*
+do
+d="$(head -1 "$i" | cut -d " " -f1 | cut -c 2-).fasta";
+mv "$i" $dir/needle/$d
+done
+# build function to get top needle hit from a sequence in above format
+get_needle() {
+	needle -gapopen 10.0 -gapextend 0.5 -endopen 0.0 -endextend 0.0 -asequence $1 -bsequence data/ITS2db_trimmed.fasta -outfile $1.out -aformat score 
+}
+export -f get_needle
+# feed each line of tab-separated sequences to the get_needle function
+parallel get_needle :::  $dir/needle/*.fasta
+# get top hit for each sequence and write to file
+parallel "sort -nt'(' -k2 {} | tail -n 1" ::: $dir/needle/*.out | sort -k1,1 > $dir/needle_results.txt
+
+
+## merge together blast results and needle results to compare
+join $dir/blast_results_filtered.txt $dir/needle_results.txt > $dir/tax_results.txt

### 2016-08-26 Shell/otus_97_bysample.sh
include all OTUs in OTU table, regardless of taxonomy/BLAST

@@ -37,3 +37,8 @@ merge_otu_maps.py -i data/otus_97_bysample/all_97_otus.txt,data/otus_97_bysample
 
 # Remove singletons
 awk '$3 ~ /./ {print}' data/otus_97_bysample/merged_otu_map.txt > data/otus_97_bysample/nosingles_otus.txt
+
+# Make OTU table
+make_otu_table.py -i data/otus_97_bysample/nosingles_otus.txt -o data/otus_97_bysample/97_otus_bysample.biom
+rm -f  data/otus_97_bysample/97_otus_bysample.tsv  # delete old OTU .tsv if present
+biom convert -i data/otus_97_bysample/97_otus_bysample.biom -o data/otus_97_bysample/97_otus_bysample.tsv --to-tsv

### 2016-08-26 Shell/tax_local_global.sh
local-global tax for 97 & 100 OTUs, cleanup after needle

@@ -13,7 +13,7 @@ rm -f data/ITS2db_trimmed.fasta.*
 ### Get top needle hit (global alignment) for each sequence
 # split fasta file into separate file for each sequence to feed to get_needle
 rm -rf $dir/needle; mkdir $dir/needle
-split -l 2 $1 $dir/needle/
+split -a 10 -l 2 $1 $dir/needle/
 # rename files according to OTU name
 for i in $dir/needle/*
 do
@@ -33,3 +33,6 @@ parallel "sort -nt'(' -k2 {} | tail -n 1" ::: $dir/needle/*.out | sort -k1,1 > $
 
 ## merge together blast results and needle results to compare
 join $dir/blast_results_filtered.txt $dir/needle_results.txt > $dir/tax_results.txt
+
+# Clean up
+rm -rf $dir/needle

### 2016-08-26 Shell/otus_100.sh
make otu tables including all OTUs

@@ -10,3 +10,12 @@ awk '$3 ~ /./ {print}' data/otus_100/combined_seqs_trimmed_otus.txt > data/otus_
 pick_rep_set.py -i data/otus_100/nosingles_otus.txt \
 -f data/fasta/combined_seqs_trimmed.fasta \
 -o data/otus_100/100_otus_rep_set.fasta
+
+# Make OTU table
+make_otu_table.py -i data/otus_100/nosingles_otus.txt -o data/otus_100/100_otus.biom
+rm -f  data/otus_100/100_otus.tsv  # delete old OTU .tsv if present
+biom convert -i data/otus_100/100_otus.biom -o data/otus_100/100_otus.tsv --to-tsv
+
+
+
+

### 2016-08-26 Shell/otus_97.sh

@@ -10,3 +10,8 @@ awk '$3 ~ /./ {print}' data/otus_97/combined_seqs_trimmed_otus.txt > data/otus_9
 pick_rep_set.py -i data/otus_97/nosingles_otus.txt \
 -f data/fasta/combined_seqs_trimmed.fasta \
 -o data/otus_97/97_otus_rep_set.fasta
+
+# Make OTU table
+make_otu_table.py -i data/otus_97/nosingles_otus.txt -o data/otus_97/97_otus.biom
+rm -f  data/otus_97/97_otus.tsv  # delete old OTU .tsv if present
+biom convert -i data/otus_97/97_otus.biom -o data/otus_97/97_otus.tsv --to-tsv

### 2016-08-26 Shell/tax_local_global.sh
do not rename split output (too many files for 100otus)

@@ -15,18 +15,18 @@ rm -f data/ITS2db_trimmed.fasta.*
 rm -rf $dir/needle; mkdir $dir/needle
 split -a 10 -l 2 $1 $dir/needle/
 # rename files according to OTU name
-for i in $dir/needle/*
-do
-d="$(head -1 "$i" | cut -d " " -f1 | cut -c 2-).fasta";
-mv "$i" $dir/needle/$d
-done
+#for i in $dir/needle/*
+#do
+#d="$(head -1 "$i" | cut -d " " -f1 | cut -c 2-).fasta";
+#mv "$i" $dir/needle/$d
+#done
 # build function to get top needle hit from a sequence in above format
 get_needle() {
 	needle -gapopen 10.0 -gapextend 0.5 -endopen 0.0 -endextend 0.0 -asequence $1 -bsequence data/ITS2db_trimmed.fasta -outfile $1.out -aformat score 
 }
 export -f get_needle
 # feed each line of tab-separated sequences to the get_needle function
-parallel get_needle :::  $dir/needle/*.fasta
+parallel get_needle :::  $dir/needle/*
 # get top hit for each sequence and write to file
 parallel "sort -nt'(' -k2 {} | tail -n 1" ::: $dir/needle/*.out | sort -k1,1 > $dir/needle_results.txt
 

### 2016-08-27 Shell/prep_ITS2db.sh
separate scripts for downloading and trimming database sequences

@@ -1,21 +1,5 @@
 #!/bin/bash
-
-# use accession numbers to download sequences from NCBI in parallel
-getseq() {
-	n=$1
-	acn=${n##*_}
-	echo $1 > data/dbseqs/${acn}.fasta
-	curl "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta" >> data/dbseqs/${acn}.fasta
-}
-export -f getseq
-rm -rf data/dbseqs; mkdir data/dbseqs
-parallel -a data/accn_nos.txt getseq 
-cat data/dbseqs/*.fasta > data/ITS2_Database.fasta
-
 	
-# Reformat names of new database to be only accession numbers & subtypes and put sequences on one line
-awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' data/ITS2_Database.fasta > data/ITS2_Database_inline.fasta
-awk '!/^>gi/ {print}' data/ITS2_Database_inline.fasta > data/ITS2_Database_ready.fasta
 
 # Download sequences from supplementary material of Green et al. 2014 and add to database
 curl https://peerj.com/articles/386/DataS1_PerlScripts_alignment_Bioinformatics.zip > data/green.zip
@@ -25,27 +9,29 @@ awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { print
 sed 's/Haplotype\(.*$\)/\1_Green2014/' > data/greenseqs.fasta
 rm data/green.zip
 rm data/greenseqs.aln
-cat data/greenseqs.fasta >> data/ITS2_Database_ready.fasta
+# If Green sequences are not already added to ITS2db_raw, add them.
+! grep -q "Green" data/ITS2db_raw.fasta && cat data/greenseqs.fasta >> data/ITS2db_raw.fasta
 
 # Trim primers from database sequences using cutadapt
 # Trim forward primers using cutadapt
 #   Allow error rate of 15% (0 indels/mismatches)
-cutadapt -g GTGAATTGCAGAACTCCGTG -e 0.15 data/ITS2_Database_ready.fasta -o data/ITS2_Database_trimF.fasta
[…]

### 2016-08-27 analysis/STJ2012_analysis.Rmd
begin pairwiseAlignment analysis

@@
 knitr::opts_chunk$set(echo = TRUE)
-options(scipen = 3, digits = 2)
+options(scipen = 3, digits = 9)
 ```
@@
 ```{r data_import, cache=T}
+
+```
+
+```{r filter_taxa}
+# Import taxonomic assignment data from local and global alignments
+taxcolnames <- c("OTU", "BLAST.hit", "BLAST.pct.id", "BLAST.align.length", "BLAST.mismatches", "BLAST.gapopen", 
+                 "BLAST.q.start", "BLAST.q.end", "BLAST.s.start", "BLAST.s.end", "BLAST.e.val", "BLAST.bit.score", 
+                 "needle.hit", "needle.length", "needle.score")
+
+TAX97 <- read.table("data/otus_97/tax_results.txt", sep=" ", col.names=taxcolnames)
+TAX97$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX97$needle.score))
+
+TAX97bs <- read.table("data/otus_97_bysample/tax_results.txt", sep=" ", col.names=taxcolnames)
+TAX97bs$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX97bs$needle.score))
+
+TAX97bsp <- read.table("data/otus_97_byspecies/tax_results.txt", sep=" ", col.names=taxcolnames)
+TAX97bsp$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX97bsp$needle.score))
+
+TAX100 <- read.table("data/otus_100/tax_results.txt", sep=" ", col.names=taxcolnames)
+TAX100$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX100$needle.score))
+
+# Merge identical taxonomic assignments (because some reference sequences are not unique...)
+ident <- readLines("data/ITS2db_trimmed_notuniques_otus.txt")
+ident <- strsplit(ident, split="\t")
+names(ident) <- c("C3_multiple", "B19_multiple", "H1_multiple", "C33_multiple", "H5_multiple", "A2_multiple",
+                  "B21/B38_multiple", "C35/C26_multiple", "A1/A13_multiple", "F5.2_multiple", "D1_multiple", 
+                  "A10_multiple", "B2_multiple", "A1.2_multiple", "C1c/C45_multiple", "B1_multiple",
+                  "C91_multiple", "C82_multiple", "D1a_multiple", "A4a_multiple", "C1_multiple",
+                  "F5.2d_multiple", "C21/C32_multiple", "A3/A6_multiple", "C26/C35a_multiple")
+
+merge.identical.hits <- function(x) {
+  sapply(as.character(x), FUN=function(y) {
+    ifelse(y %in% unlist(ident), 
+           gsub("[0-9]*$", "", names(unlist(ident))[match(y, unlist(ident))]), 
+           y)
+    })
+}
+
+TAX97$BLAST.matched <- merge.identical.hits(TAX97$BLAST.hit)
+TAX97$needle.matched <- merge.identical.hits(TAX97$needle.hit)
+
+# OTUs for which BLAST and needle gave different assignments
+nomatch <- subset(TAX97, BLAST.matched!=needle.matched)
+nomatch
+
+# Get percent identity of global alignment to needle hits
+# import reference sequences
+ref <- readDNAStringSet("data/ITS2db_trimmed.fasta")
+# import otu sequences
+otus97 <- readDNAStringSet("data/otus_97/97_otus_rep_set.fasta")
+names(otus97) <- gsub(" .*$", "", names(otus97))
+# align each otu sequence to its best needle hit, using same parameters as command line needle
+# "global-local" forces whole reference/pattern sequence to be used, but subject/query can be a concescutive subsequence -- this is because queries may include slightly longer region of gene than references because references were trimmed by o-smart-trim.
+this means that query sequence can have more bases than the reference at beginning or end of the sequence (=end gaps in alignment) without penalties.
+EDNAFULL <- as.matrix(read.table("data/EDNAFULL", skip=8, header=T, row.names=1))
+n.psa <- pairwiseAlignment(subject=otus97[TAX97$OTU], pattern=ref[TAX97$needle.hit], type="global-local",
+                           substitutionMatrix=EDNAFULL,
+                           gapOpening=10.0, gapExtension=0.5)
+
+# calculate percent similarity to reference with indels counting as single difference regardless of length
+TAX97 <- within(TAX97, {
+  needle.hit.length <- width(pattern(n.psa))
+  needle.mismatches <- nmismatch(n.psa)
+  needle.insertions <- unlist(lapply(as.list(insertion(n.psa)), length))
+  needle.deletions <- unlist(lapply(as.list(deletion(n.psa)), length))
+  needle.similarity <- (needle.hit.length-needle.mismatches-needle.insertions-needle.deletions)/needle.hit.length
+})
+
+g.psa <- pairwiseAlignment(subject=otus97[TAX97$OTU], pattern=ref[TAX97$needle.hit], type="global",
+                           substitutionMatrix=EDNAFULL,
+                           gapOpening=10.0, gapExtension=0.5)
+
+# calculate percent similarity to reference with indels counting as single difference regardless of length
+TAX97 <- within(TAX97, {
+  gneedle.hit.length <- width(pattern(g.psa))
+  gneedle.mismatches <- nmismatch(g.psa)
+  gneedle.insertions <- unlist(lapply(as.list(insertion(g.psa)), length))
+  gneedle.deletions <- unlist(lapply(as.list(deletion(g.psa)), length))
+  gneedle.similarity <- (gneedle.hit.length-gneedle.mismatches-gneedle.insertions-gneedle.deletions)/gneedle.hit.length
+})
+
+aligned(n.psa[])
+n.psa
+aligned(n.psa[1])
+TAX97$needle.hit.length <- width(pattern(n.psa))
+TAX97$needle.mismatches <- nmismatch(n.psa)
+TAX97$needle.insertions <- unlist(lapply(as.list(insertion(n.psa)), length))
+TAX97$needle.deletions <- unlist(lapply(as.list(deletion(n.psa)), length))
+TAX97$needle.similarity <- with(TAX97, (needle.hit.length - needle.mismatches - needle.insertions - needle.deletions) / needle.hit.length)
+
+with(TAX97[115, ], {
+  aln <<- pairwiseAlignment(subject=otus97[OTU], pattern=ref[needle.hit], type="global-local",
+                            substitutionMatrix=EDNAFULL,
+                            gapOpening=10.0, gapExtension=0.5)
+})
+aln
+
+aln <- pairwiseAlignment(subject="ACTTCTCTATCTATCTTGGTACG", pattern="AATTCTCGGTACG", type="global")
+aln2 <- pairwiseAlignment(subject="ACTTCTACG", pattern="AATTCTCGGTACG", type="global")
+aln
+aln2
+width(aligned(aln))
+nmatch(aln)/width(aligned(aln))
+
+# calculate pct id of all positions relative to reference/(pattern)
+nmatch(aln)/width(pattern(aln))
+# calculate pct id counting indels as single difference regardless of length = GAST definition
+(width(pattern(aln)) - nmismatch(aln) - length(unlist(insertion(aln))) - length(unlist(deletion(aln))))/width(pattern(aln))
+
+stringDist(aln)
+
+nmatch(aln)
+nmismatch(aln)
+width(nindel(aln))
+length(unlist(insertion(aln))); length(unlist(deletion(aln)))
+length(unlist(insertion(aln)))
+length(unlist(deletion(aln)))
+insertion(aln)
+deletion(aln)
+unlist(nindel(aln))
+nindel(aln)$insertion
+str(nindel(aln))
+b <- n.psa[11]
+as.list(indel(subject(b)))
+as.list(indel(subject(aln)))
+length(as.data.frame(deletion(b))[[1]])
+indel(b)
+length(unlist(insertion(b)))
+
+class(EDNAFULL)
+n.psa[47]
+summary(n.psa[47])
+nmatch(n.psa[47])
+nindel(n.psa[47])
+nmismatch(n.psa[47])
+nchar(pattern(n.psa[2]))
+246/255
+nmismatch(n.psa[2])
+
+```
+
+```{r build_phyloseq}
 # Import OTU tables
-OTU97 <- otu_table(read.table("data/97_otus.tsv", header=T, check.names=F, row.names=1,
+OTU97 <- otu_table(read.table("data/otus_97/97_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU97bs <- otu_table(read.table("data/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
+OTU97bs <- otu_table(read.table("data/otus_97_bysample/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU97bsp <- otu_table(read.table("data/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
-                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU100 <- otu_table(read.table("data/100_otus.tsv", header=T, check.names=F, row.names=1,
+#OTU97bsp <- otu_table(read.table("data/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
+#                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
+OTU100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
                                skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
 # Import taxonomy
-TAX97 <- tax_table(as.matrix(read.table("data/97_otus_tax_assignments.txt", sep="\t", row.names=2,
-                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX97bs <- tax_table(as.matrix(read.table("data/97_otus_bysample_tax_assignments.txt", sep="\t", row.names=2,
-                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX97bsp <- tax_table(as.matrix(read.table("data/97_otus_byspecies_tax_assignments.txt", sep="\t", row.names=2,
-                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX100 <- tax_table(as.matrix(read.table("data/100_otus_tax_assignments.txt", sep="\t", row.names=2,
-                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
+
 # Import sample data
@@
 ```{r filter_dataset, cache=T}
+
 # Filter OTUs by minimum count
@@
 otudf <- data.frame(t(otu_table(phy97bs.f.p2)), check.names=F)
-# Percent of samples of species that contain OTU at less than 1% relative abundance (but more than 0%)
-
+# Whether any samples of species contain OTU at less than 1% relative abundance (but more than 0%)
 sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p2))$Species), 
@@
 edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
-edges <- droplevels(edges[edges$weight>0, ])
+edges <- droplevels(edges[edges$weight>0, ])  # only keep if was observed in more than N colony of a spp
 # Create "nodes" table from sample and tax data
@@
 text(0.9,0.9, "O.ann.", cex=0.5)
+
+set.seed(42399)
+l3 <- layout_nicely(net3)
+l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
+par(mar=c(0,0,0,0))
+V(net3)$size <- ifelse(is.na(V(net3)$Clade), 15, 10*degree(net3)^0.2)
+plot(net3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(net3)$size/20,
+     vertex.label.color="black")
 

### 2016-08-28 Shell/prep_ITS2db.sh
dereplicate trimmed ITS2db

@@ -93,6 +93,11 @@ usearch -usearch_global data/ITS2db_trimmed.fasta -db data/ITS2db_trimmed_notuni
 pick_otus.py -i data/ITS2db_trimmed_notuniques.fasta -s 1.0 -o data
 # list of identical sequence groups is in output file: data/ITS2db_trimmed_notuniques_otus.txt
 
+# Dereplicate database
+usearch -derep_fulllength data/ITS2db_trimmed.fasta -output data/ITS2db_trimmed_derep.fasta -sizeout
+
+
+
 # Clean up
 rm data/ITS2db_trimmed_notunique.fasta
 rm data/ITS2db_trimmed_notuniques_clusters.uc

### 2016-08-28 R/run_nw.R
script to run needleman-wunsch

@@ -1,0 +1,65 @@
+# get command line arguments
+args = commandArgs(trailingOnly=TRUE)
+
+# test if there is at least one argument: if not, return an error
+if (length(args) < 2) {
+  stop("must specify query sequences and reference sequences", call.=FALSE)
+}
+
+# Import query sequences
+library(Biostrings)
+otus <- readDNAStringSet(args[1])
+names(otus) <- gsub(" .*$", "", names(otus))
+
+# Import reference sequences
+ref <- readDNAStringSet(args[2])
+
+# import substitution matrix
+EDNAFULL <- as.matrix(read.table("data/EDNAFULL", skip=8, header=T, row.names=1))
+
+# Build function to find which reference sequences produces best alignment score
+get.best <- function(x) {
+  scores <- pairwiseAlignment(subject=x, pattern=ref, type="global-local", scoreOnly=TRUE,
+                           substitutionMatrix=EDNAFULL,
+                           gapOpening=10.0, gapExtension=0.5)
+  best <- paste(names(ref)[which(scores==max(scores))], collapse=";")
+  return(best)
+}
+
+# Set up for parallel computation
+library(parallel)
+cl <- makeCluster(detectCores() - 1)  # Initiate cluster
+clusterEvalQ(cl, library(Biostrings))  # Make Biostrings library available to cluster
+clusterExport(cl, c("EDNAFULL", "ref", "otus", "get.best"))  # Export objects to cluster
+
+# Run function in parallel to get best hit for each sequence
+best <- parSapply(cl, otus, FUN=get.best)
+stopCluster(cl)  # Stop cluster
+
+# Collect results in a data frame
+results <- data.frame(otu=names(otus), hits=best, stringsAsFactors = FALSE)
+# If multiple hit, select first one
+results$hit <- unlist(lapply(strsplit(results$hits, split=";;"), "[", 1))
+results$hit <- gsub(";;", ";", paste(results$hit, ";", sep="")) # temp fix to end with ;....get rid of sizeout!!!
+
+# Get percent identity of global alignment of each sequence with its best hit from database
+
+# align each otu sequence to its best needle hit, using same parameters as command line needle
[…]

### 2016-08-28 R/run_nw.R
change output filename to nw_tophits.tsv

@@ -60,6 +60,6 @@ results <- within(results, {
 })
 
 # Write results to .tsv file
-write.table(results, file=paste(dirname(args[1]), "/", gsub("\\..*$","", basename(args[1])),"_nw.tsv", sep=""), 
+write.table(results, file=paste(dirname(args[1]), "/", "nw_tophits.tsv", sep=""), 
             sep="\t", quote=FALSE)
 

### 2016-08-28 Shell/run_blast.sh
script to run BLAST only

@@ -1,0 +1,11 @@
+#!/bin/bash
+
+dir=$(dirname $1)
+
+### Get blast results (local alignment) for each sequence
+formatdb -p F -i $2
+blastall -p blastn -i $1 -d $2 -b 1 -m 8 > $dir/blast_results.txt
+# Sort by query sequence and then E-value in descending order, then get only the first hit for each query sequence
+sort -k1,1 -k11,11nr $dir/blast_results.txt | sort -u -k1,1 > $dir/blast_tophits.txt
+# cleanup 
+rm -f $2.*

### 2016-08-28 Shell/tax_local_global.sh
remove old files

@@ -1,38 +1,0 @@
-#!/bin/bash
-
-dir=$(dirname $1)
-
-### Get blast results (local alignment) for each sequence
-formatdb -p F -i data/ITS2db_trimmed.fasta
-blastall -p blastn -i $1 -d data/ITS2db_trimmed.fasta -b 1 -m 8 > $dir/blast_results.txt
-# Sort by query sequence and then E-value in descending order, then get only the first hist for each query sequence
-sort -k1,1 -k11,11nr $dir/blast_results.txt | sort -u -k1,1 > $dir/blast_results_filtered.txt
-# cleanup 
-rm -f data/ITS2db_trimmed.fasta.*
-
-### Get top needle hit (global alignment) for each sequence
-# split fasta file into separate file for each sequence to feed to get_needle
-rm -rf $dir/needle; mkdir $dir/needle
-split -a 10 -l 2 $1 $dir/needle/
-# rename files according to OTU name
-#for i in $dir/needle/*
-#do
-#d="$(head -1 "$i" | cut -d " " -f1 | cut -c 2-).fasta";
-#mv "$i" $dir/needle/$d
-#done
-# build function to get top needle hit from a sequence in above format
-get_needle() {
-	needle -gapopen 10.0 -gapextend 0.5 -endopen 0.0 -endextend 0.0 -asequence $1 -bsequence data/ITS2db_trimmed.fasta -outfile $1.out -aformat score 
-}
-export -f get_needle
-# feed each line of tab-separated sequences to the get_needle function
-parallel get_needle :::  $dir/needle/*
-# get top hit for each sequence and write to file
-parallel "sort -nt'(' -k2 {} | tail -n 1" ::: $dir/needle/*.out | sort -k1,1 > $dir/needle_results.txt
-
-
-## merge together blast results and needle results to compare
-join $dir/blast_results_filtered.txt $dir/needle_results.txt > $dir/tax_results.txt
-
-# Clean up
-rm -rf $dir/needle

### 2016-08-28 Shell/otus_97_byspecies.sh
reduce n cores to avert error

@@ -21,7 +21,7 @@ clust97() {
 	pick_otus.py -i $1 -s 0.97 --denovo_otu_id_prefix $s'_denovo' --optimal_uclust -o data/otus_97_byspecies
 }
 export -f clust97
-parallel -j5 clust97 ::: data/fasta/byspecies/*.fasta
+parallel -j3 clust97 ::: data/fasta/byspecies/*.fasta
 
 # Pick representative sequence set for each sample
 repset97() {
@@ -49,3 +49,8 @@ merge_otu_maps.py -i data/otus_97_byspecies/all_97_otus.txt,data/otus_97_byspeci
 
 # Remove singletons
 awk '$3 ~ /./ {print}' data/otus_97_byspecies/merged_otu_map.txt > data/otus_97_byspecies/nosingles_otus.txt
+
+# Make OTU table
+make_otu_table.py -i data/otus_97_byspecies/nosingles_otus.txt -o data/otus_97_byspecies/97_otus_byspecies.biom
+rm -f  data/otus_97_byspecies/97_otus_byspecies.tsv  # delete old OTU .tsv if present
+biom convert -i data/otus_97_byspecies/97_otus_byspecies.biom -o data/otus_97_byspecies/97_otus_byspecies.tsv --to-tsv

### 2016-08-28 R/run_nw.R
remove -sizeout from derep database

@@ -39,8 +39,8 @@ stopCluster(cl)  # Stop cluster
 # Collect results in a data frame
 results <- data.frame(otu=names(otus), hits=best, stringsAsFactors = FALSE)
 # If multiple hit, select first one
-results$hit <- unlist(lapply(strsplit(results$hits, split=";;"), "[", 1))
-results$hit <- gsub(";;", ";", paste(results$hit, ";", sep="")) # temp fix to end with ;....get rid of sizeout!!!
+results$hit <- unlist(lapply(strsplit(results$hits, split=";"), "[", 1))
+#results$hit <- gsub(";;", ";", paste(results$hit, ";", sep="")) # temp fix to end with ;....get rid of sizeout!!!
 
 # Get percent identity of global alignment of each sequence with its best hit from database
 

### 2016-08-28 Shell/prep_ITS2db.sh

@@ -94,7 +94,7 @@ pick_otus.py -i data/ITS2db_trimmed_notuniques.fasta -s 1.0 -o data
 # list of identical sequence groups is in output file: data/ITS2db_trimmed_notuniques_otus.txt
 
 # Dereplicate database
-usearch -derep_fulllength data/ITS2db_trimmed.fasta -output data/ITS2db_trimmed_derep.fasta -sizeout
+usearch -derep_fulllength data/ITS2db_trimmed.fasta -output data/ITS2db_trimmed_derep.fasta
 
 
 

### 2016-08-29 R/run_nw.R
reformat nw output

@@ -1,9 +1,8 @@
-# get command line arguments
+# Get command line arguments
 args = commandArgs(trailingOnly=TRUE)
-
-# test if there is at least one argument: if not, return an error
+# If two arguments not provided, return an error
 if (length(args) < 2) {
-  stop("must specify query sequences and reference sequences", call.=FALSE)
+  stop("must specify query sequences (.fasta) and reference sequences (.fasta)", call.=FALSE)
 }
 
 # Import query sequences
@@ -17,7 +16,13 @@ ref <- readDNAStringSet(args[2])
 # import substitution matrix
 EDNAFULL <- as.matrix(read.table("data/EDNAFULL", skip=8, header=T, row.names=1))
 
-# Build function to find which reference sequences produces best alignment score
+# Build function to find which reference sequences produces best alignment score using
+# Needleman-Wunsch algorithm with default EMBOSS parameters for DNA sequences.
+# -type="global-local" forces whole reference (pattern) sequence to be used, but subject/query may be 
+# a concescutive subsequence. This is because queries may include slightly longer region of gene 
+# than references because references were trimmed by o-smart-trim. This means that query sequence 
+# can have more bases than the reference at beginning or end of the sequence (=end gaps in 
+# alignment) without penalties.
 get.best <- function(x) {
   scores <- pairwiseAlignment(subject=x, pattern=ref, type="global-local", scoreOnly=TRUE,
                            substitutionMatrix=EDNAFULL,
@@ -37,26 +42,23 @@ best <- parSapply(cl, otus, FUN=get.best)
 stopCluster(cl)  # Stop cluster
 
 # Collect results in a data frame
-results <- data.frame(otu=names(otus), hits=best, stringsAsFactors = FALSE)
-# If multiple hit, select first one
-results$hit <- unlist(lapply(strsplit(results$hits, split=";"), "[", 1))
-#results$hit <- gsub(";;", ";", paste(results$hit, ";", sep="")) # temp fix to end with ;....get rid of sizeout!!!
+results <- data.frame(otu=names(otus), nw.hits=best, stringsAsFactors=FALSE)
 
-# Get percent identity of global alignment of each sequence with its best hit from database
-
-# align each otu sequence to its best needle hit, using same parameters as command line needle
[…]

### 2016-09-01 analysis/STJ2012_analysis.Rmd
update otu taxonomy data import and filtering

--- 2016-08-24 analysis/STJ2012_analysis.Rmd
@@
 # Load all R scripts in "R/" directory
-sapply(list.files(path="R", pattern="*.R", full.names=TRUE), source, .GlobalEnv)
+sapply(list.files(path="R", pattern="*[^w].R", full.names=TRUE), source, .GlobalEnv)
 # Load package libraries
@@
 ```{r data_import, cache=T}
+
+```
+
+```{r filter_taxa}
+# Import taxonomic assignment data from blast and nw
+blast.res.colnames <- c("otu", "BLAST.hit", "pct.id", "B.length", "B.mismatches", "B.gapopen", 
+                        "q.start", "q.end", "s.start", "s.end", "e.val", "bit.score")
+tax.import <- function(b, nw) {
+  m <- merge(read.table(b, col.names=blast.res.colnames, stringsAsFactors=FALSE),
+             read.table(nw, stringsAsFactors=FALSE), by="otu")
+  rownames(m) <- m$otu
+  m$Clade <- str_extract(m$nw.hits, pattern="^.")
+  return(m)
+}
+
+tax97 <- tax.import(b="data/otus_97/blast_tophits.txt", nw="data/otus_97/nw_tophits.tsv")
+tax100 <- tax.import(b="data/otus_100/blast_tophits.txt", nw="data/otus_100/nw_tophits.tsv")
+tax97bs <- tax.import(b="data/otus_97_bysample/blast_tophits.txt", nw="data/otus_97_bysample/nw_tophits.tsv")
+tax97bsp <- tax.import(b="data/otus_97_byspecies/blast_tophits.txt", nw="data/otus_97_byspecies/nw_tophits.tsv")
+
+# Deal with identical taxonomic assignments (because some reference sequences are not unique...)
+# Create names for groups of identical sequences based on members of group...
+ident <- readLines("data/ITS2db_trimmed_notuniques_otus.txt")
+ident <- gsub("denovo[0-9]*\t", "", ident)
+ident <- strsplit(ident, split="\t")
+ident2 <- lapply(ident, str_match_all, pattern="[A-I]{1}[0-9]{1,3}.*[_/]")
+ident2 <- lapply(ident2, function(g) gsub(pattern="\\..*_$", "", x=unlist(g)))
+ident2 <- lapply(ident2, function(g) gsub(pattern="_$", "", g))
+ident2 <- lapply(ident2, function(g) unlist(strsplit(g, split="/")))
+subtypes <- lapply(ident2, function(x) levels(factor(unlist(x))))
+subtypes <- lapply(subtypes, function(s) paste(paste(s, collapse="/"), "_multiple", sep=""))
+names(ident) <- subtypes
+ident <- melt(do.call(rbind, ident))
+ident <- unique(ident[order(ident[,1]), c(3,1)])
+
+# Replace any sequence in taxonomy assignment that is a member of a group of identical sequences with the name of the group
+collapse.idents <- function(df) {
+  within(df, {
+    for (i in 1:nrow(ident)) {
+      BLAST.hit <- gsub(ident[i,1], ident[i,2], BLAST.hit)
+      nw.hits <- gsub(ident[i,1], ident[i,2], nw.hits)
+      nw.hit1 <- gsub(ident[i,1], ident[i,2], nw.hit1)
+    }
+  })
+}
+
+tax97 <- collapse.idents(tax97)
+tax97bs <- collapse.idents(tax97bs)
+tax97bsp <- collapse.idents(tax97bsp)
+tax100 <- collapse.idents(tax100)
+
+# Get subtypes without accession numbers
+get.st <- function(df) {
+  within(df, {
+    bl.st <- gsub(BLAST.hit, pattern="_.*$", replacement="")
+    bl.st <- ifelse(nw.sim==100, bl.st, paste("'", bl.st, "'", sep=""))
+    nw.st <- gsub(nw.hits, pattern="_[A-Z]{2}[0-9]{6}", replacement="")
+    nw.st <- gsub(nw.st, pattern="_multiple", replacement="")
+    nw.st <- ifelse(nw.sim==100, nw.st, paste("'", nw.st, "'", sep=""))
+    Subtype <- nw.st
+  })
+}
+
+tax97 <- get.st(tax97)
+tax97bs <- get.st(tax97bs)
+tax97bsp <- get.st(tax97bsp)
+tax100 <- get.st(tax100)
+```
+
+```{r build_phyloseq}
 # Import OTU tables
-OTU97 <- otu_table(read.table("data/97_otus.tsv", header=T, check.names=F, row.names=1,
+otu97 <- otu_table(read.table("data/otus_97/97_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU97bs <- otu_table(read.table("data/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
+otu97bs <- otu_table(read.table("data/otus_97_bysample/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU97bsp <- otu_table(read.table("data/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
+otu97bsp <- otu_table(read.table("data/otus_97_byspecies/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-OTU100 <- otu_table(read.table("data/100_otus.tsv", header=T, check.names=F, row.names=1,
+otu100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
                                skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-# Import taxonomy
-TAX97 <- tax_table(as.matrix(read.table("data/97_otus_tax_assignments.txt", sep="\t", row.names=2,
-                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX97bs <- tax_table(as.matrix(read.table("data/97_otus_bysample_tax_assignments.txt", sep="\t", row.names=2,
-                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX97bsp <- tax_table(as.matrix(read.table("data/97_otus_byspecies_tax_assignments.txt", sep="\t", row.names=2,
-                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
-TAX100 <- tax_table(as.matrix(read.table("data/100_otus_tax_assignments.txt", sep="\t", row.names=2,
-                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
+
 # Import sample data
-SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
+sam <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
+
 # Build phyloseq objects
-phy97 <- phyloseq(OTU97, TAX97, SAM)
-phy97bs <- phyloseq(OTU97bs, TAX97bs, SAM)
-phy97bsp <- phyloseq(OTU97bsp, TAX97bsp, SAM)
-phy100 <- phyloseq(OTU100, TAX100, SAM)
+phy97 <- phyloseq(otu97, tax_table(as.matrix(tax97)), sam)
+phy97bs <- phyloseq(otu97bs, tax_table(as.matrix(tax97bs)), sam)
+phy97bsp <- phyloseq(otu97bsp, tax_table(as.matrix(tax97bsp)), sam)
+phy100 <- phyloseq(otu100, tax_table(as.matrix(tax100)), sam)
 ```
@@
 ```{r filter_dataset, cache=T}
+# FILTER OUT OTUS THAT ARE NOT SYMBIODINIUM
+# Build function to BLAST all sequences with less than 90% similarity to Symbiodinium to the NCBI nr database, and get the top hit.
+If the top hit does not contain the string "Symbiodinium", then this sequence is assumed to not be Symbiodinium.
+is.it.symbio <- function(df, repset) {
+  poor.otus <- df[which(df$nw.sim < 90), "otu"]  # Identify OTUs with <90% similarity to Symbiodinium
+  sapply(poor.otus, function(otu) {  # Get sequences for these OTUs to BLAST to NCBI
+    system(paste("awk '/", otu, " /{print;getline;print}' ", repset, " >> temp.fasta", sep=""))
+  })
+  blast.res <- system("blastn -db nr -remote -query temp.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u", intern=TRUE)  # Blast sequences to NCBI
+  blast.res <- data.frame(do.call(rbind, lapply(strsplit(blast.res, split="\t"), "[", c(1,3))),
+                          stringsAsFactors=FALSE)
+  blast.res$sym <- str_detect(blast.res[, 2], "Symbiodinium")  # See if Symbiodinium is in BLAST result
+  df2 <- setNames(merge(poor.otus, blast.res, by=1, all=T), nm=c("otu", "hit", "sym"))
+  df2[which(is.na(df2$hit)), c("hit", "sym")] <- c("No hit", FALSE)
+  system("rm temp.fasta")  # Cleanup temp file
+  return(df2)
+}
+
+tax97.blasts <- is.it.symbio(tax97, "data/otus_97/97_otus_rep_set.fasta")
+tax97.notsym <- as.character(tax97.blasts[which(tax97.blasts$sym==FALSE), "otu"])
+
+tax97bs.blasts <- is.it.symbio(tax97bs, "data/otus_97_bysample/all_rep_set_rep_set.fasta")
+tax97bs.notsym <- as.character(tax97bs.blasts[which(tax97bs.blasts$sym==FALSE), "otu"])
+
+tax97bsp.blasts <- is.it.symbio(tax97bsp, "data/otus_97_byspecies/all_rep_set_rep_set.fasta")
+tax97bsp.notsym <- as.character(tax97bsp.blasts[which(tax97bsp.blasts$sym==FALSE), "otu"])
+
+tax100.blasts <- is.it.symbio(tax100, "data/otus_100/100_otus_rep_set.fasta")
+tax100.notsym <- as.character(tax100.blasts[which(tax100.blasts$sym==FALSE), "otu"])
+
+# Remove OTUs that are not Symbiodinium
+phy97.f <- prune_taxa(!taxa_names(phy97) %in% tax97.notsym, phy97)
+phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% tax97bs.notsym, phy97bs)
+phy97bsp.f <- prune_taxa(!taxa_names(phy97bsp) %in% tax97bsp.notsym, phy97bsp)
+phy100.f <- prune_taxa(!taxa_names(phy100) %in% tax100.notsym, phy100)
+
 # Filter OTUs by minimum count
@@
 # Identify OTUs below threshold count
-taxa97 <- taxa_sums(phy97)[which(taxa_sums(phy97) >= n)]
-taxa97bs <- taxa_sums(phy97bs)[which(taxa_sums(phy97bs) >= n)]
-taxa97bsp <- taxa_sums(phy97bsp)[which(taxa_sums(phy97bsp) >= n)] 
-taxa100 <- taxa_sums(phy100)[which(taxa_sums(phy100) >= n)]
+taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
+taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
+taxa97bsp <- taxa_sums(phy97bsp.f)[which(taxa_sums(phy97bsp.f) >= n)] 
+taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
 # Remove taxa below threshold count
-phy97.f <- prune_taxa(names(taxa97), phy97)
-phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs)
-phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp)
-phy100.f <- prune_taxa(names(taxa100), phy100)
+phy97.f <- prune_taxa(names(taxa97), phy97.f)
+phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
+phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp.f)
+phy100.f <- prune_taxa(names(taxa100), phy100.f)
 
@@
 # Plot composition of 97% within-sample OTUs colored by clade
-composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
+composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$nw.st), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
 # Plot composition of 97% within-sample OTUs colored by OTU
@@
 ## *Diploria strigosa*  
-Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, denovo235.
+Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE samE!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, denovo235.
 In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- denovo73, denovo231, and denovo10.
@@
 V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
-                       unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
+                       unlist(lapply(strsplit(V(net)$nw.st, split="_"), "[", 1)))
 V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
@@
 # Find all nodes beginning with D, and give them the same node ID number
-nodemapnew[grep("^D", V(net)$Subtype)] <- min(grep("^D", V(net)$Subtype))
+nodemapnew[grep("^D", V(net)$nw.st)] <- min(grep("^D", V(net)$nw.st))
 # Replace the now missing values in the node map so node IDs remain sequential
@@
   text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
+       labels=levels(sam$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
   text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
@@
   text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
+       labels=levels(sam$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
   text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
@@
   text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
-       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
+       labels=levels(sam$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
   text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)

### 2016-09-06 R/run_nw.R
keep all best nw.hits, not just first

@@ -27,7 +27,7 @@ get.best <- function(x) {
   scores <- pairwiseAlignment(subject=x, pattern=ref, type="global-local", scoreOnly=TRUE,
                            substitutionMatrix=EDNAFULL,
                            gapOpening=10.0, gapExtension=0.5)
-  best <- paste(names(ref)[which(scores==max(scores))], collapse=";")
+  best <- cbind(names(x), names(ref)[which(scores==max(scores))])
   return(best)
 }
 
@@ -42,13 +42,12 @@ best <- parSapply(cl, otus, FUN=get.best)
 stopCluster(cl)  # Stop cluster
 
 # Collect results in a data frame
-results <- data.frame(otu=names(otus), nw.hits=best, stringsAsFactors=FALSE)
+results <- data.frame(otu=rep(names(best), sapply(best, length)),
+                      nw.hit=unlist(best))
 
-# Get percent similarity of global alignment of each sequence with its best hit from database
-# If multiple best hits, select first one for comparison
-results$nw.hit1 <- unlist(lapply(strsplit(results$nw.hits, split=";"), "[", 1))
-# Align each otu sequence to its best hit
-psa <- pairwiseAlignment(subject=otus[results$otu], pattern=ref[results$nw.hit1], type="global-local",
+# Get percent similarity of global alignment of each sequence with its bets hits from database
+# Align each otu sequence to its best hits
+psa <- pairwiseAlignment(subject=otus[results$otu], pattern=ref[results$nw.hit], type="global-local",
                          substitutionMatrix=EDNAFULL,
                          gapOpening=10.0, gapExtension=0.5)
 

### 2016-09-06 R/run_nw.R
modify output format

@@ -43,24 +43,28 @@ stopCluster(cl)  # Stop cluster
 
 # Collect results in a data frame
 results <- data.frame(otu=rep(names(best), sapply(best, length)),
-                      nw.hit=unlist(best))
+                      hit=unlist(best), stringsAsFactors = FALSE)
 
-# Get percent similarity of global alignment of each sequence with its bets hits from database
+# Align each sequence with its bets hits from database to get number of mismatches, indels, and % similarity
 # Align each otu sequence to its best hits
-psa <- pairwiseAlignment(subject=otus[results$otu], pattern=ref[results$nw.hit], type="global-local",
+psa <- pairwiseAlignment(subject=otus[results$otu], pattern=ref[results$hit], type="global-local",
                          substitutionMatrix=EDNAFULL,
                          gapOpening=10.0, gapExtension=0.5)
 
 # Calculate percent similarity to reference with indels counting as single difference regardless of length
 results <- within(results, {
-  nw.len <- width(pattern(psa))
-  nw.mis <- nmismatch(psa)
-  nw.ins <- unlist(lapply(as.list(insertion(psa)), length))
-  nw.del <- unlist(lapply(as.list(deletion(psa)), length))
-  nw.sim <- round((nw.len-nw.mis-nw.ins-nw.del)/nw.len, 4) * 100
+  score <- score(psa)
+  len <- width(pattern(psa))
+  mis <- nmismatch(psa)
+  ins <- unlist(lapply(as.list(insertion(psa)), length))
+  del <- unlist(lapply(as.list(deletion(psa)), length))
+  sim <- round((len-mis-ins-del)/len, 4) * 100
 })
 
+# Merge identical hits
+results2 <- with(results, aggregate(hit ~ otu + sim + del + ins + mis + len + score, FUN=paste0, collapse=";"))
+
 # Write results to .tsv file
-write.table(results, file=paste(dirname(args[1]), "/", "nw_tophits.tsv", sep=""), 
+write.table(results2, file=paste(dirname(args[1]), "/", "nw_tophits.tsv", sep=""), 
             sep="\t", quote=FALSE)
 

### 2016-09-06 Shell/run_blast.sh

@@ -2,10 +2,47 @@
 
 dir=$(dirname $1)
 
-### Get blast results (local alignment) for each sequence
+# BLAST each sequence to the reference ITS2 database (trimmed, dereplicated)
 formatdb -p F -i $2
-blastall -p blastn -i $1 -d $2 -b 1 -m 8 > $dir/blast_results.txt
+blastall -p blastn -i $1 -d $2 -e 0.00001 -b 10 -m 8 > $dir/blast_results.txt
 # Sort by query sequence and then E-value in descending order, then get only the first hit for each query sequence
-sort -k1,1 -k11,11nr $dir/blast_results.txt | sort -u -k1,1 > $dir/blast_tophits.txt
-# cleanup 
+# sort -k1,1 -k11,11g $dir/blast_results.txt | sort -u -k1,1 > $dir/blast_tophits.txt
+
+# Identify the lowest E-value obtained for each sequence
+sort -k1,1 -k11,11g $dir/blast_results.txt | sort -u -k1,1 | cut -f1,11 > $dir/lowevals.txt
+# Get all BLAST hits that obtained the lowest E-value for each sequence (keeps ties)
+rm -f $dir/blast_tophits_withties.txt
+while read n; do
+  otu=`echo $n | cut -d" " -f1`
+  eval=`echo $n | cut -d" " -f2`
+  awk '$1 ~ /'"$otu"'$/ && $11 ~ /'"$eval"'$/ {print}' $dir/blast_results.txt >> $dir/blast_tophits_withties.txt
+done < $dir/lowevals.txt
+
+# Merge hits with equivalent results...(if score/length/eval/mismatch/gaps are identical for multiple hits)
+# by appending multiple hits to the end of the line
+awk '
+BEGIN { FS="\t" }
+{
+    curr = $1 FS $3 FS $4 FS $5 FS $6 FS $7 FS $8 FS $9 FS $10 FS $11 FS $12
+    if (curr == prev) {
+        rec = rec "\t" $2
+    }
+    else {
+        if (rec) print rec
+        rec = $0
+    }
+    prev = curr
+}
+END { if (rec) print rec }
+' $dir/blast_tophits_withties.txt > $dir/blast_tophits_merged.txt
+
[…]

### 2016-09-06 analysis/STJ2012_analysis.Rmd
import both blast and nw

@@
 # Load all R scripts in "R/" directory
-sapply(list.files(path="R", pattern="*[^w].R", full.names=TRUE), source, .GlobalEnv)
+source("R/functions.R", .GlobalEnv)
 # Load package libraries
@@
 The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object (class 'phyloseq') to facilitate downstream analyses.
-```{r data_import, cache=T}
+```{r import_taxa_data, cache=T}
+# Import taxonomic assignment data from blast and nw
 
-```
+read.blast <- function(file) {
+  b <- read.table(file, stringsAsFactors=FALSE,
+                  col.names=c("otu", "pct.id", "length", "mismatches", "gapopen", "q.start", "q.end", 
+                              "s.start", "s.end", "e.val", "bit.score", "hit"))
+  b <- b[order(b$otu),]
+  return(b)
+}
+b <- read.blast("data/otus_97/blast_tophits.tsv")
 
-```{r filter_taxa}
-# Import taxonomic assignment data from blast and nw
-blast.res.colnames <- c("otu", "BLAST.hit", "pct.id", "B.length", "B.mismatches", "B.gapopen", 
-                        "q.start", "q.end", "s.start", "s.end", "e.val", "bit.score")
-tax.import <- function(b, nw) {
-  m <- merge(read.table(b, col.names=blast.res.colnames, stringsAsFactors=FALSE),
-             read.table(nw, stringsAsFactors=FALSE), by="otu")
+read.nw <- function(file) {
+  nw <- read.table(file, stringsAsFactors=FALSE)
+  nw <- nw[order(nw$otu),]
+  return(nw)
+}
+nw <- read.nw("data/otus_97/nw_tophits.tsv")
+
+tax.import <- function(bfp, nwfp) {
+  b <- read.table(bfp, col.names=blast.res.colnames, stringsAsFactors=FALSE)
+  b <- b[order(b$otu),]
+  ag <- aggregate(data.frame(BLAST.hit=b$BLAST.hit), by=list(otu=b$otu), FUN=function(x) paste0(x, collapse=";"))
+  b2 <- b[!duplicated(b$otu), ]
+  b2$BLAST.hit <- ag$BLAST.hit
+  nw <- read.table(nwfp, stringsAsFactors=FALSE)
+  m <- merge(b2, nw, by="otu")
   rownames(m) <- m$otu
@@
 
-tax97 <- tax.import(b="data/otus_97/blast_tophits.txt", nw="data/otus_97/nw_tophits.tsv")
-tax100 <- tax.import(b="data/otus_100/blast_tophits.txt", nw="data/otus_100/nw_tophits.tsv")
-tax97bs <- tax.import(b="data/otus_97_bysample/blast_tophits.txt", nw="data/otus_97_bysample/nw_tophits.tsv")
-tax97bsp <- tax.import(b="data/otus_97_byspecies/blast_tophits.txt", nw="data/otus_97_byspecies/nw_tophits.tsv")
+tax97 <- tax.import(bfp="data/otus_97/blast_tophits_withties.txt", nwfp="data/otus_97/nw_tophits.tsv")
+tax100 <- tax.import(bfp="data/otus_100/blast_tophits.txt", nwfp="data/otus_100/nw_tophits.tsv")
+tax97bs <- tax.import(bfp="data/otus_97_bysample/blast_tophits.txt", nwfp="data/otus_97_bysample/nw_tophits.tsv")
+tax97bsp <- tax.import(bfp="data/otus_97_byspecies/blast_tophits.txt", nwfp="data/otus_97_byspecies/nw_tophits.tsv")
 
@@
 
-# Replace any sequence in taxonomy assignment that is a member of a group of identical sequences with the name of the group
+# Replace any sequence name in taxonomy assignment that is a member of a group of identical sequences with the name of the group
 collapse.idents <- function(df) {

### 2016-09-06 Shell/run_blast_poormatches.sh
remove BLAST from build, but create script to BLAST poor matches to NCBI

@@ -1,0 +1,11 @@
+# $1 is path to nw_tophits, $2 is path to representative sequences
+
+dir=$(dirname $1)
+
+# Get names of otus whose representative sequence is <90% similar to best hit
+awk '$3 < 90 {print $2}' $1 > $dir/poor_matches.txt
+# Get corresponding sequences
+awk 'NR==FNR {a[">"$1]; next} $1 in a {print; getline; print}' $dir/poor_matches.txt $2 > $dir/poor_matches_seqs.fasta
+
+# Blast sequences to NCBI nr databse
+blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u > $dir/poormatch_IDs.txt

### 2016-09-07 Shell/run_blast_poormatches.sh
update blast poor matches

@@ -9,3 +9,7 @@ awk 'NR==FNR {a[">"$1]; next} $1 in a {print; getline; print}' $dir/poor_matches
 
 # Blast sequences to NCBI nr databse
 blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u > $dir/poormatch_IDs.txt
+
+# Clean up
+rm $dir/poor_matches.txt
+rm $dir/poor_matches_seqs.fasta

### 2016-09-07 Shell/run_blast_poormatches.sh
fix code to keep names of poor matches that do not get a successful BLAST hit

@@ -3,13 +3,16 @@
 dir=$(dirname $1)
 
 # Get names of otus whose representative sequence is <90% similar to best hit
-awk '$3 < 90 {print $2}' $1 > $dir/poor_matches.txt
+awk '$3 < 90 {print $2}' $1 | sort -u > $dir/poor_matches.txt
 # Get corresponding sequences
 awk 'NR==FNR {a[">"$1]; next} $1 in a {print; getline; print}' $dir/poor_matches.txt $2 > $dir/poor_matches_seqs.fasta
 
 # Blast sequences to NCBI nr databse
-blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u > $dir/poormatch_IDs.txt
+blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u > $dir/poormatch_IDs1.txt
+
+join -a 1 $dir/poor_matches.txt $dir/poormatch_IDs1.txt > $dir/poormatch_IDs.txt
 
 # Clean up
 rm $dir/poor_matches.txt
 rm $dir/poor_matches_seqs.fasta
+rm $dir/poormarch_IDs1.txt

### 2016-09-07 Shell/run_blast_poormatches.sh
fix poormatch blast to track those with no hits

@@ -8,11 +8,11 @@ awk '$3 < 90 {print $2}' $1 | sort -u > $dir/poor_matches.txt
 awk 'NR==FNR {a[">"$1]; next} $1 in a {print; getline; print}' $dir/poor_matches.txt $2 > $dir/poor_matches_seqs.fasta
 
 # Blast sequences to NCBI nr databse
-blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u > $dir/poormatch_IDs1.txt
+blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u -k1,1 > $dir/poormatch_IDs1.txt
 
 join -a 1 $dir/poor_matches.txt $dir/poormatch_IDs1.txt > $dir/poormatch_IDs.txt
 
 # Clean up
 rm $dir/poor_matches.txt
 rm $dir/poor_matches_seqs.fasta
-rm $dir/poormarch_IDs1.txt
+rm $dir/poormatch_IDs1.txt

### 2016-09-07 analysis/STJ2012_analysis.Rmd
remove blast results; assign subtype after filtering

@@
 # Import taxonomic assignment data from blast and nw
-
-read.blast <- function(file) {
-  b <- read.table(file, stringsAsFactors=FALSE,
-                  col.names=c("otu", "pct.id", "length", "mismatches", "gapopen", "q.start", "q.end", 
-                              "s.start", "s.end", "e.val", "bit.score", "hit"))
-  b <- b[order(b$otu),]
-  return(b)
-}
-b <- read.blast("data/otus_97/blast_tophits.tsv")
-
 read.nw <- function(file) {
@@
 }
-nw <- read.nw("data/otus_97/nw_tophits.tsv")
 
-tax.import <- function(bfp, nwfp) {
-  b <- read.table(bfp, col.names=blast.res.colnames, stringsAsFactors=FALSE)
-  b <- b[order(b$otu),]
-  ag <- aggregate(data.frame(BLAST.hit=b$BLAST.hit), by=list(otu=b$otu), FUN=function(x) paste0(x, collapse=";"))
-  b2 <- b[!duplicated(b$otu), ]
-  b2$BLAST.hit <- ag$BLAST.hit
-  nw <- read.table(nwfp, stringsAsFactors=FALSE)
-  m <- merge(b2, nw, by="otu")
-  rownames(m) <- m$otu
-  m$Clade <- str_extract(m$nw.hits, pattern="^.")
-  return(m)
-}
-
-tax97 <- tax.import(bfp="data/otus_97/blast_tophits_withties.txt", nwfp="data/otus_97/nw_tophits.tsv")
-tax100 <- tax.import(bfp="data/otus_100/blast_tophits.txt", nwfp="data/otus_100/nw_tophits.tsv")
-tax97bs <- tax.import(bfp="data/otus_97_bysample/blast_tophits.txt", nwfp="data/otus_97_bysample/nw_tophits.tsv")
-tax97bsp <- tax.import(bfp="data/otus_97_byspecies/blast_tophits.txt", nwfp="data/otus_97_byspecies/nw_tophits.tsv")
+tax97 <- read.nw("data/otus_97/nw_tophits.tsv")
+tax100 <- read.nw("data/otus_100/nw_tophits.tsv")
+tax97bs <- read.nw("data/otus_97_bysample/nw_tophits.tsv")
+tax97bsp <- read.nw("data/otus_97_byspecies/nw_tophits.tsv")
 
@@
     for (i in 1:nrow(ident)) {
-      BLAST.hit <- gsub(ident[i,1], ident[i,2], BLAST.hit)
-      nw.hits <- gsub(ident[i,1], ident[i,2], nw.hits)
-      nw.hit1 <- gsub(ident[i,1], ident[i,2], nw.hit1)
+      hit <- gsub(ident[i,1], ident[i,2], hit)
     }
@@
 
-# Get subtypes without accession numbers
-get.st <- function(df) {
-  within(df, {
-    bl.st <- gsub(BLAST.hit, pattern="_.*$", replacement="")
-    bl.st <- ifelse(nw.sim==100, bl.st, paste("'", bl.st, "'", sep=""))
-    nw.st <- gsub(nw.hits, pattern="_[A-Z]{2}[0-9]{6}", replacement="")
-    nw.st <- gsub(nw.st, pattern="_multiple", replacement="")
-    nw.st <- ifelse(nw.sim==100, nw.st, paste("'", nw.st, "'", sep=""))
-    Subtype <- nw.st
-  })
-}
+#tax97 <- get.st(tax97)
+tax97 <- with(tax97, tax97[order(otu, -sim), ])
+tax97 <- tax97[!duplicated(tax97$otu), ]
+rownames(tax97) <- tax97$otu
 
-tax97 <- get.st(tax97)
-tax97bs <- get.st(tax97bs)
-tax97bsp <- get.st(tax97bsp)
-tax100 <- get.st(tax100)
+#tax97bs <- get.st(tax97bs)
+tax97bs <- with(tax97bs, tax97bs[order(otu, -sim), ])
+tax97bs <- tax97bs[!duplicated(tax97bs$otu), ]
+rownames(tax97bs) <- tax97bs$otu
+
+#tax97bsp <- get.st(tax97bsp)
+tax97bsp <- with(tax97bsp, tax97bsp[order(otu, -sim), ])
+tax97bsp <- tax97bsp[!duplicated(tax97bsp$otu), ]
+rownames(tax97bsp) <- tax97bsp$otu
+
+#tax100 <- get.st(tax100)
+tax100 <- with(tax100, tax100[order(otu, -sim), ])
+tax100 <- tax100[!duplicated(tax100$otu), ]
+rownames(tax100) <- tax100$otu
 ```
@@
 # FILTER OUT OTUS THAT ARE NOT SYMBIODINIUM
-# Build function to BLAST all sequences with less than 90% similarity to Symbiodinium to the NCBI nr database, and get the top hit.
-If the top hit does not contain the string "Symbiodinium", then this sequence is assumed to not be Symbiodinium.
-is.it.symbio <- function(df, repset) {
-  poor.otus <- df[which(df$nw.sim < 90), "otu"]  # Identify OTUs with <90% similarity to Symbiodinium
-  sapply(poor.otus, function(otu) {  # Get sequences for these OTUs to BLAST to NCBI
-    system(paste("awk '/", otu, " /{print;getline;print}' ", repset, " >> temp.fasta", sep=""))
-  })
-  blast.res <- system("blastn -db nr -remote -query temp.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u", intern=TRUE)  # Blast sequences to NCBI
-  blast.res <- data.frame(do.call(rbind, lapply(strsplit(blast.res, split="\t"), "[", c(1,3))),
-                          stringsAsFactors=FALSE)
-  blast.res$sym <- str_detect(blast.res[, 2], "Symbiodinium")  # See if Symbiodinium is in BLAST result
-  df2 <- setNames(merge(poor.otus, blast.res, by=1, all=T), nm=c("otu", "hit", "sym"))
-  df2[which(is.na(df2$hit)), c("hit", "sym")] <- c("No hit", FALSE)
-  system("rm temp.fasta")  # Cleanup temp file
-  return(df2)
-}
+# If the top hit in NCBI does not contain the string "Symbiodinium", then this sequence is assumed to not be Symbiodinium.
+poormatch97 <- readLines("data/otus_97/poormatch_IDs.txt")
+poormatch97 <- data.frame(otu=str_extract(poormatch97, "denovo[^ ]*"),
+                          symbio=str_detect(poormatch97, "Symbiodinium"),   # TRUE if Symbiodinium
+                          stringsAsFactors = FALSE)
+notsymbio97 <- poormatch97[which(poormatch97$symbio==F), 1]
 
-tax97.blasts <- is.it.symbio(tax97, "data/otus_97/97_otus_rep_set.fasta")
-tax97.notsym <- as.character(tax97.blasts[which(tax97.blasts$sym==FALSE), "otu"])
+poormatch97bs <- readLines("data/otus_97_bysample/poormatch_IDs.txt")
+poormatch97bs <- data.frame(otu=str_extract(poormatch97bs, "denovo[^ ]*"),
+                            symbio=str_detect(poormatch97bs, "Symbiodinium"),   # TRUE if Symbiodinium
+                            stringsAsFactors = FALSE)
+notsymbio97bs <- poormatch97bs[which(poormatch97bs$symbio==F), 1]
 
-tax97bs.blasts <- is.it.symbio(tax97bs, "data/otus_97_bysample/all_rep_set_rep_set.fasta")
-tax97bs.notsym <- as.character(tax97bs.blasts[which(tax97bs.blasts$sym==FALSE), "otu"])
+poormatch97bsp <- readLines("data/otus_97_byspecies/poormatch_IDs.txt")
+poormatch97bsp <- data.frame(otu=str_extract(poormatch97bsp, "denovo[^ ]*"),
+                             symbio=str_detect(poormatch97bsp, "Symbiodinium"),   # TRUE if Symbiodinium
+                             stringsAsFactors = FALSE)
+notsymbio97bsp <- poormatch97bsp[which(poormatch97bsp$symbio==F), 1]
 
-tax97bsp.blasts <- is.it.symbio(tax97bsp, "data/otus_97_byspecies/all_rep_set_rep_set.fasta")
-tax97bsp.notsym <- as.character(tax97bsp.blasts[which(tax97bsp.blasts$sym==FALSE), "otu"])
-
-tax100.blasts <- is.it.symbio(tax100, "data/otus_100/100_otus_rep_set.fasta")
-tax100.notsym <- as.character(tax100.blasts[which(tax100.blasts$sym==FALSE), "otu"])
+poormatch100 <- readLines("data/otus_100/poormatch_IDs.txt")
+poormatch100 <- data.frame(otu=str_extract(poormatch100, "denovo[^ ]*"),
+                           symbio=str_detect(poormatch100, "Symbiodinium"),   # TRUE if Symbiodinium
+                           stringsAsFactors = FALSE)
+notsymbio100 <- poormatch100[which(poormatch100$symbio==F), 1]
 
 # Remove OTUs that are not Symbiodinium
-phy97.f <- prune_taxa(!taxa_names(phy97) %in% tax97.notsym, phy97)
-phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% tax97bs.notsym, phy97bs)
-phy97bsp.f <- prune_taxa(!taxa_names(phy97bsp) %in% tax97bsp.notsym, phy97bsp)
-phy100.f <- prune_taxa(!taxa_names(phy100) %in% tax100.notsym, phy100)
+phy97.f <- prune_taxa(!taxa_names(phy97) %in% notsymbio97, phy97)
+phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% notsymbio97bs, phy97bs)
+phy97bsp.f <- prune_taxa(!taxa_names(phy97bsp) %in% notsymbio97bsp, phy97bsp)
+phy100.f <- prune_taxa(!taxa_names(phy100) %in% notsymbio100, phy100)
 
@@
 phy100.f <- prune_taxa(names(taxa100), phy100.f)
+
+# Label clades and subtypes for filtered phyloseq object tax_tables
+get.st <- function(df) {
+  within(df, {
+    Clade <- substr(hit, 1, 1)
+    Subtype <- gsub(hit, pattern="_[A-Z]{2}[0-9]{6}", replacement="")
+    Subtype <- gsub(Subtype, pattern="_multiple", replacement="")
+    Subtype2 <- ifelse(as.numeric(sim)==100, Subtype,
+                       paste0("[", rep(rle(sort(Subtype))$values, times=rle(sort(Subtype))$lengths), "]^", 
+                              unlist(lapply(rle(sort(Subtype))$lengths, seq_len)))[order(order(Subtype))])
+    Subtype <- ifelse(as.numeric(sim)==100, Subtype, paste("*", Subtype, sep=""))
+  })
+}
+
+tax_table(phy97.f) <- as.matrix(get.st(data.frame(tax_table(phy97.f), stringsAsFactors=FALSE)))
+tax_table(phy97bs.f) <- as.matrix(get.st(data.frame(tax_table(phy97bs.f), stringsAsFactors=FALSE)))
+tax_table(phy97bsp.f) <- as.matrix(get.st(data.frame(tax_table(phy97bsp.f), stringsAsFactors=FALSE)))
+tax_table(phy100.f) <- as.matrix(get.st(data.frame(tax_table(phy100.f), stringsAsFactors=FALSE)))
+
 ```
@@
                         by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
-sum(cladeAbund$RelAbund)
 cladeAbund$Prop <- prop.table(cladeAbund$RelAbund)
-pie(cladeAbund$Prop, col=taxcolors)
-dotchart(cladeAbund$Prop)
+
 bars <- barplot(cladeAbund$Prop*100, col=taxcolors, space=0,
@@
 cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
-cladeAbund
 
@@
 # Plot composition of 97% within-sample OTUs colored by clade
-composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$nw.st), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
+composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
 # Plot composition of 97% within-sample OTUs colored by OTU
@@
 ## *Diploria strigosa*  
-Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE samE!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, denovo235.
+Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, denovo235.
 In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- denovo73, denovo231, and denovo10.
@@
 # Network analyses
-## "Resident" symbionts
+## Abundant symbionts
 #### (>1% relative abundance)
@@
 ```{r networks, fig.width=5, fig.height=5, fig.align="center"}
-# Get network with one node for each species.
-otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
-# Percent of samples of species that contain OTU at more than 1% relative abundance
-sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
-                   FUN=function(x) length(x[x>0.01])/length(x))
-rownames(sp.ag) <- sp.ag$Species
-sp.ag <- sp.ag[, -1]
-edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
-edges <- droplevels(edges[edges$weight>0, ])
-# Create "nodes" table from sample and tax data
-sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
-tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
-tdf$id <- rownames(tdf)
-nodes <- merge(sdf, tdf, all=T)
-nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
-# Create network
-net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+abunet <- sppnet(phy=phy97bs.f.p, plot=F,
+                 fun=function(x) length(x[x>0.01])/length(x),
+                 layout=layout.fruchterman.reingold)
 
-# Best one for now ---------
-V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
-V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))) #5*sqrt(degree(net))
-V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
-                       unlist(lapply(strsplit(V(net)$nw.st, split="_"), "[", 1)))
-V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
-                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
-                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
-                                                "furcata"="P.fur.", "radians"="S.rad.", 
-                                                "siderea"="S.sid.", "strigosa"="D.str."))
-V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
-                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
-E(net)$color <- "gray60"
-E(net)$width <- 15 * E(net)$weight
-set.seed(12347)
-set.seed(12364)
+par(mar=c(0,0,0,0))
+V(abunet)$size <- ifelse(is.na(V(abunet)$Clade), 15, 10*sqrt(degree(abunet)))
 set.seed(12374)
-l <- layout.fruchterman.reingold(net)
+l <- layout.fruchterman.reingold(abunet)
 l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
-par(mar=c(0,0,0,0))
-plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
+plot(abunet, rescale=F, layout=l, edge.curved=0.1, vertex.label.cex=V(abunet)$size/20,
      vertex.label.color="black")
-#legend(x=-1.05, y=1.1, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
-#     col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
 ```
@@
 ```{r dom_networks, fig.width=5, fig.height=5, fig.align="center"}
-# Get network with one node for each species.
-otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
-# Percent of samples of species that contain OTU at more than 50% relative abundance
-sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
-                   FUN=function(x) length(x[x>0.5])/length(x))
-rownames(sp.ag) <- sp.ag$Species
-sp.ag <- sp.ag[, -1]
-edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
-edges <- droplevels(edges[edges$weight>0, ])
-# Create "nodes" table from sample and tax data
-sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
-tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
-tdf$id <- rownames(tdf)
-nodes <- merge(sdf, tdf, all=T)
-nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
-# Create network
-net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
+# Create dominant symbionts network
+domnet <- sppnet(phy=phy97bs.f.p, plot=F,
+                 fun=function(x) length(x[x>0.5])/length(x), 
+                 layout=layout.fruchterman.reingold)
 
-# Best one for now ---------
-V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
-V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 10*(degree(net))^0.75) #5*sqrt(degree(net))
-V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
-                       unlist(lapply(strsplit(V(net)$nw.st, split="_"), "[", 1)))
-V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
-                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
-                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
-                                                "furcata"="P.fur.", "radians"="S.rad.", 
-                                                "siderea"="S.sid.", "strigosa"="D.str."))
-V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
-                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
-E(net)$color <- "gray60"
-E(net)$width <- 15 * E(net)$weight
-set.seed(12347)
-set.seed(12364)
+par(mar=c(0,0,0,0), lwd=1)
 set.seed(12374)
-l <- layout.fruchterman.reingold(net)
+l <- layout.fruchterman.reingold(domnet)
 l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
-par(mar=c(0,0,0,0))
-plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
+plot(domnet, rescale=F, layout=l, edge.curved=0.1, vertex.label.cex=V(domnet)$size/20,
      vertex.label.color="black")
-#legend(x=-1.05, y=1.1, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
-#     col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
+
 ```
@@
 ```{r bk.net, fig.width=5, fig.height=5, fig.align="center"}
+# Create background symbionts network
 # filter out OTUs that only occurred in one sample
 phy97bs.f.p2 <- filter_taxa(phy97bs.f.p, function(x) sum(x>0) > 1, prune=TRUE)
-# Get network with one node for each species.
-otudf <- data.frame(t(otu_table(phy97bs.f.p2)), check.names=F)
-# Whether any samples of species contain OTU at less than 1% relative abundance (but more than 0%)
-sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p2))$Species), 
-                   FUN=function(x) any(x>0 & x<0.01))
-rownames(sp.ag) <- sp.ag$Species
-sp.ag <- sp.ag[, -1]
-edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
-edges <- droplevels(edges[edges$weight>0, ])  # only keep if was observed in more than N colony of a spp
-# Create "nodes" table from sample and tax data
-sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p2))$Species))
-tdf <- data.frame(tax_table(phy97bs.f.p2))[rownames(data.frame(tax_table(phy97bs.f.p2))) %in% edges$target, ]
-tdf$id <- rownames(tdf)
-nodes <- merge(sdf, tdf, all=T)
-nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
-# Create network
-net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
-
-# Edit vertex attributes
-V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
-V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 5*degree(net)) #5*sqrt(degree(net))
-V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
-                       unlist(lapply(strsplit(V(net)$nw.st, split="_"), "[", 1)))
-V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
-                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
-                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
-                                                "furcata"="P.fur.", "radians"="S.rad.", 
-                                                "siderea"="S.sid.", "strigosa"="D.str."))
-V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
-                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
-E(net)$color <- "gray60"
-E(net)$width <- E(net)$weight * 2
-
+bkgnet <- sppnet(phy=phy97bs.f.p2, plot=F,
+                 fun=function(x) any(x>0 & x<0.01),
+                 layout=layout.fruchterman.reingold)
 # Merge D nodes (assuming they are all D1a/trenchii)
 # Get original node map (sequence starting at one to the total number of nodes)
-nodemap <- seq(1:length(V(net)))
+nodemap <- seq(1:length(V(bkgnet)))
 # Make a copy of the nodemap that will be modified
@@
 # Find all nodes beginning with D, and give them the same node ID number
-nodemapnew[grep("^D", V(net)$nw.st)] <- min(grep("^D", V(net)$nw.st))
+nodemapnew[grep("^D", V(bkgnet)$Subtype)] <- min(grep("^D", V(bkgnet)$Subtype))
 # Replace the now missing values in the node map so node IDs remain sequential
@@
 # Contract the network using the new nodemap (nodes with same node ID are merged)
-net2 <- contract(net, nodemapnew, vertex.attr.comb=list(size="sum", "last"))
+bkgnet2 <- contract(bkgnet, nodemapnew, vertex.attr.comb=list(size="sum", "last"))
 # Simplify network so that multiple edges are combined
-net2 <- simplify(net2, remove.multiple=T, edge.attr.comb=list(weight="mean", width="mean", "ignore"))
+bkgnet2 <- simplify(bkgnet2, remove.multiple=T, edge.attr.comb=list(weight="mean", width="mean", "ignore"))
+# remove symbionts that are only connected to one or two coral species
+bkgnet3 <- delete.vertices(bkgnet2, degree(bkgnet2)<=2) # & V(net2)$type==2)
+V(bkgnet3)$size <- ifelse(is.na(V(bkgnet3)$Clade), 15, 1.5*degree(bkgnet3)^1.5)
+E(bkgnet3)$width <- E(bkgnet3)$weight * 2
 
-# remove symbionts that are only connected to one or two coral species
-net3 <- delete.vertices(net2, degree(net2)<=2) # & V(net2)$type==2)
-V(net3)$size <- ifelse(is.na(V(net3)$Clade), 15, 1.5*degree(net3)^1.5)
+# Plot background symbionts network
 set.seed(42384)
-l3 <- layout.fruchterman.reingold(net3, weights=E(net3)$weight)
+l3 <- layout.fruchterman.reingold(bkgnet3)
 l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
 par(mar=c(0,0,0,0))
-plot(net3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(net3)$size/20,
+plot(bkgnet3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(bkgnet3)$size/20,
      vertex.label.color="black")
@@
 text(0.9,0.9, "O.ann.", cex=0.5)
-
-set.seed(42399)
-l3 <- layout_nicely(net3)
-l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
-par(mar=c(0,0,0,0))
-V(net3)$size <- ifelse(is.na(V(net3)$Clade), 15, 10*degree(net3)^0.2)
-plot(net3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(net3)$size/20,
-     vertex.label.color="black")
 

### 2016-09-07 analysis/STJ2012_analysis.Rmd
new labels ("Subtype2") on network vertices

@@
     Subtype2 <- ifelse(as.numeric(sim)==100, Subtype,
-                       paste0("[", rep(rle(sort(Subtype))$values, times=rle(sort(Subtype))$lengths), "]^", 
+                       paste0("'[", rep(rle(sort(Subtype))$values, times=rle(sort(Subtype))$lengths), "]'^", 
                               unlist(lapply(rle(sort(Subtype))$lengths, seq_len)))[order(order(Subtype))])
-    Subtype <- ifelse(as.numeric(sim)==100, Subtype, paste("*", Subtype, sep=""))
+    #Subtype <- ifelse(as.numeric(sim)==100, Subtype, paste("*", Subtype, sep=""))
   })
@@
 This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
-```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
+```{r cladecomp, cache=T, fig.height=3, fig.width=4}
 cladeAbund <- aggregate(data.frame(RelAbund=rowSums(otu_table(phy97bs.f.p))),
@@
 cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
+```
 
+```{r cladecomp_bysample, cache=T, fig.height=6, fig.width=4}
 par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)

### 2016-09-07 R/functions.R

@@ -1,0 +1,338 @@
+betad <- function(phy, group) {
+  # Calculate distances to species centroids for each sample in principal coordinate space
+  samdat <- data.frame(sample_data(phy))
+  braydist <- vegdist(t(otu_table(phy)), method="bray")
+  sambd <- betadisper(d=braydist, group=samdat[, group], 
+                      type="centroid", bias.adjust=TRUE)
+  # Calculate each species' mean distance to centroid and standard error
+  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
+  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
+                            FUN=function(x) sd(x)/sqrt(length(x)))$x
+  # Determine order of decreasing mean dispersion
+  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
+  # Update betadisper object with species in decreasing order of mean dispersion
+  sambd <- betadisper(d=braydist, 
+                      group=factor(samdat[, group], levels=as.character(sambdsumm.ord$group)), 
+                      type="centroid", bias.adjust=TRUE)
+  # Use permutations to perform pairwise comparisons of group mean dispersions
+  sampt <- permutest(sambd, pairwise=T)
+  saml <- multcompLetters(sampt$pairwise$permuted)
+  # Return results as a list
+  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
+}
+
+composition <- function(phy, col, legend=T) {
+  samdat <- data.frame(sample_data(phy))
+  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
+  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
+  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
+  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
+  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
+                                           rownames(samdat)])
[…]

### 2016-09-07 analysis/STJ2012_analysis.Rmd
fix labeling error

@@
     Subtype <- gsub(Subtype, pattern="_multiple", replacement="")
-    Subtype2 <- ifelse(as.numeric(sim)==100, Subtype,
+    Subtype2 <- ifelse(as.numeric(sim)==100, paste0("'", Subtype, "'"),
                        paste0("'[", rep(rle(sort(Subtype))$values, times=rle(sort(Subtype))$lengths), "]'^", 

### 2016-09-07 R/functions.R

@@ -88,7 +88,8 @@ otubarplot <- function(phy, main) {
   typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
                                                                 order(Shore, Site, InputFileName))])]
   # Get info for plotting OTU names and blast hits on top of barplot
-  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
+  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype2"])
+  blasthits <- parse(text=blasthits)
   names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
   samples <- colnames(typerelabund)
   shores <- data.frame(sample_data(phy))[samples, "Shore"]
@@ -107,8 +108,9 @@ otubarplot <- function(phy, main) {
   for (i in 1:length(bars)) {
     text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
          divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
-         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
+         labels=blasthits[which(!is.na(heights[,i]))+1],
          cex=0.75)
+    #paste(names[which(!is.na(heights[,i]))+1],
   }
   mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
 }

### 2016-09-07 R/betad.R
cleanup and reorganize

@@ -1,22 +1,0 @@
-betad <- function(phy, group) {
-  # Calculate distances to species centroids for each sample in principal coordinate space
-  samdat <- data.frame(sample_data(phy))
-  braydist <- vegdist(t(otu_table(phy)), method="bray")
-  sambd <- betadisper(d=braydist, group=samdat[, group], 
-                      type="centroid", bias.adjust=TRUE)
-  # Calculate each species' mean distance to centroid and standard error
-  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
-  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
-                            FUN=function(x) sd(x)/sqrt(length(x)))$x
-  # Determine order of decreasing mean dispersion
-  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
-  # Update betadisper object with species in decreasing order of mean dispersion
-  sambd <- betadisper(d=braydist, 
-                      group=factor(samdat[, group], levels=as.character(sambdsumm.ord$group)), 
-                      type="centroid", bias.adjust=TRUE)
-  # Use permutations to perform pairwise comparisons of group mean dispersions
-  sampt <- permutest(sambd, pairwise=T)
-  saml <- multcompLetters(sampt$pairwise$permuted)
-  # Return results as a list
-  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
-}

### 2016-09-07 R/composition.R

@@ -1,38 +1,0 @@
-composition <- function(phy, col, legend=T) {
-  samdat <- data.frame(sample_data(phy))
-  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
-  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
-  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
-  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
-  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
-                                           rownames(samdat)])
-  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
-  shorebreaks <- which(shorebreaks==F) - 1
-  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
-  # Make Barplot
-  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
-  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
-  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
-  mtext(side=1, "Relative abundance", cex=0.7, line=1)
-  # Add legend
-  if (legend==T) {
-    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
-           cex=0.7, legend=c("A", "B", "C", "D", "F", "G"), fill=taxcolors, x.intersp=0.5)
-    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
-           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
-           x.intersp=0.3)
-  }
-  # Add grouping bars for Shore
-  for (i in 1:length(shorebreaks)) {
-    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
-    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
[…]

### 2016-09-07 R/makenet.R

@@ -1,20 +1,0 @@
-makenet <- function(physeq, n) {
-  require(phyloseq)
-  require(reshape2)
-  require(igraph)
-  # Remove taxa that are not present in subset
-  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
-  # Convert OTU table to "edges" table (weight = relative abundance)
-  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
-  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
-  edges <- edges[edges$weight>n, ]
-  # Create "nodes" table from sample and tax data
-  sdf <- data.frame(sample_data(physeq))
-  sdf$id <- rownames(sdf)
-  tdf <- data.frame(tax_table(physeq))
-  tdf$id <- rownames(tdf)
-  nodes <- merge(sdf, tdf, all=T)
-  # Create network
-  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
-  return(net)
-}

### 2016-09-07 R/otubarplot.R

@@ -1,31 +1,0 @@
-otubarplot <- function(phy, main) {
-  samdat <- data.frame(sample_data(phy))
-  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
-                                           rownames(samdat)])
-  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
-                                                                order(Shore, Site, InputFileName))])]
-  # Get info for plotting OTU names and blast hits on top of barplot
-  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
-  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
-  samples <- colnames(typerelabund)
-  shores <- data.frame(sample_data(phy))[samples, "Shore"]
-  sites <- data.frame(sample_data(phy))[samples, "Site"]
-  divides <- apply(typerelabund, 2, cumsum)
-  heights <- diff(divides)
-  heights[heights < 0.04] <- NA
-  # Plot barplot and OTU names and blast hits
-  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
-                  xlab="", ylab="", xaxt="n")
-  text(bars, -0.06, labels=samples, xpd=T)
-  text(bars, -0.10, labels=shores, xpd=T)
-  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T)
-  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, pos=2)
-  mtext(side=2, text = "Relative Abundance", line=2)
-  for (i in 1:length(bars)) {
-    text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
-         divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
-         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
-         cex=0.75)
-  }
-  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
-}

### 2016-09-07 R/otubarplot2.R

@@ -1,109 +1,0 @@
-otubarplot2 <- function(samples) {
-  # Get phyloseq objects for selected samples from each clustering approach
-  samples <<- samples
-  phy100 <- subset_samples(phy100.f.p, rownames(sample_data(phy100.f.p)) %in% samples)
-  phy100 <- filter_taxa(phy100, function(x) sum(x)!=0, TRUE)
-  phy97 <- subset_samples(phy97.f.p, rownames(sample_data(phy97.f.p)) %in% samples)
-  phy97 <- filter_taxa(phy97, function(x) sum(x)!=0, TRUE)
-  phy97bs <- subset_samples(phy97bs.f.p, rownames(sample_data(phy97bs.f.p)) %in% samples)
-  phy97bs <- filter_taxa(phy97bs, function(x) sum(x)!=0, TRUE)
-  
-  # Get sample data
-  samdat <- data.frame(sample_data(phy100))
-  # Set sample order to go by dominant 97% OTU symbiont subtype then coral species
-  sord <- rownames(sample_data(phy97)[order(factor(tax_table(phy97)[apply(otu_table(phy97), MARGIN=2, FUN=which.max), "Subtype"]), sample_data(phy97)$Species), ])
-  samdat <- samdat[sord, ]
-  # Get tax table and otu table
-  taxdat <- data.frame(tax_table(phy100))
-  otutab <- as.matrix(otu_table(phy100)[, rownames(samdat)])
-
-  # Get top n most abundant sequence variants and unique colors for each one
-  ranks <- order(rowSums(otutab), decreasing=T)
-  n <- 45
-  topn <- rownames(otutab[ranks[1:n], ])
-  topn <- data.frame(tax_table(phy100)[topn, "Clade"])
-  # Get 'qualitative' color palettes from color brewer to assign to each sequence variant
-  qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',][-c(3:5), ]
-  col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
-  colors <- sample(col_vector, n)
-  topn$colors <- colors
-  # Plot bars
-  par(mgp=c(0,0.2,0), cex.axis=0.6, mar=c(6,2,3,1))
-  bars <- barplot(sort(otutab[, 1], decreasing=T), width=1/(ncol(otutab)*1), xlim=c(0, 1),
[…]

### 2016-09-07 R/perms.R

@@ -1,40 +1,0 @@
-# Function to test differences in community composition within and between groups by PERMANOVA
-
-perms <- function(phy, group, trt) {
-  groups <- data.frame(sample_data(phy))[, group]
-  # Create data frame to populate with results
-  resdf <- data.frame(
-    matrix(ncol=8, nrow=10,
-           dimnames=list(levels(data.frame(sample_data(phy))[, group]), 
-                         c(group, "n", "overall", "within", "between", "R2", "bd", "p")))
-    )
-  # Compute within- and between-treatment statistics for each group
-  for (i in 1:nlevels(groups)) {
-    group.name <<- levels(groups)[i]
-    resdf[i, group] <- group.name
-    # make subsetted phyloseq object for individual group
-    phy.group <- phy 
-    samdat <- data.frame(sample_data(phy))
-    sample_data(phy.group) <- sample_data(samdat[samdat[[group]]==group.name, ])
-    samdat <- as(sample_data(phy.group), "data.frame") # sample data for subsetted phyloseq object
-    # calculate distances
-    braydist <- phyloseq::distance(phy.group, "bray")
-    resdf$n[i] <- nsamples(phy.group)
-    md <- meandist(braydist, samdat[, trt])
-    resdf$within[i] <- summary(md)$W  # Weighted mean dissimilarity within treatments
-    resdf$between[i] <- summary(md)$B  # Mean dissimilarity between treatments
-    resdf$overall[i] <- summary(md)$D  # Overall dissimilarity
-    if (nlevels(as(sample_data(phy.group), "data.frame")[, trt]) > 1) {
-      permanova <- adonis(braydist ~ get(trt), data=samdat, permutations=99999)
-      resdf$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
-      resdf$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
-      bd <- betadisper(braydist, group=samdat[, trt])
-      resdf$bd[i] <- TukeyHSD(bd)$group[4]
-    } else {
-      resdf$R2[i] <- NA
-      resdf$p[i] <- NA
-      resdf$bd[i] <- NA
-    }
-  }
-  return(resdf)
-}

### 2016-09-07 R/phystats.R

@@ -1,18 +1,0 @@
-# Function to calculate summary statistics from phyloseq object
-
-phystats <- function(phy) {
-  return(
-    data.frame(
-      'Total count in OTU table'=sum(otu_table(phy)),
-      'Number of OTUs'=ntaxa(phy),
-      'Range of OTU counts'=paste0(range(taxa_sums(phy)), collapse=" - "),
-      'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
-      'Number of samples'=nsamples(phy),
-      'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
-      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
-                                                     as.integer(sd(sample_sums(phy))), sep=" ± "),
-      'Geometric mean (±SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
-                                                    as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "),
-      check.names=F)
-  )
-}

### 2016-09-07 R/plotnet.R

@@ -1,9 +1,0 @@
-plotnet <- function(net) {
-  plot(net,
-       edge.curved=0.1, 
-       edge.width=15*(E(net)$weight)^0.35,
-       vertex.label=unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)),
-       vertex.color=ifelse(is.na(V(net)$Clade), "white",
-                           taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))]),
-       vertex.size=ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))))
-}

### 2016-09-07 Shell/run_blast.sh

@@ -10,7 +10,7 @@ blastall -p blastn -i $1 -d $2 -e 0.00001 -b 10 -m 8 > $dir/blast_results.txt
 
 # Identify the lowest E-value obtained for each sequence
 sort -k1,1 -k11,11g $dir/blast_results.txt | sort -u -k1,1 | cut -f1,11 > $dir/lowevals.txt
-# Get all BLAST hits that obtained the lowest E-value for each sequence (keeps ties)
+# Get all BLAST hits that obtained the lowest E-value for each sequence (keeps ties/multiple hits with same E-val)
 rm -f $dir/blast_tophits_withties.txt
 while read n; do
   otu=`echo $n | cut -d" " -f1`

### 2016-09-07 analysis/STJ2012_analysis.Rmd
small formatting changes

@@
 OTUs comprising more than 4% of a sample are labeled with the unique OTU number and the *Symbiodinium* subtype and NCBI GenBank accession number of the closest BLAST hit for that OTU in the reference database.
-OTU numbers and barplot colors are NOT comparable across the three clustering methods.
+OTU numbers and barplot colors are NOT comparable across clustering methods.
 
 ## *Diploria strigosa*  
-Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, denovo235.
-In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- denovo73, denovo231, and denovo10.
+Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, B1.
+In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- B1, [B1]^6, and [B1]^3.
 Look at the 100% OTU dataset to see the composition of unique sequence variants in these samples.

### 2016-09-07 R/functions.R

@@ -109,7 +109,7 @@ otubarplot <- function(phy, main) {
     text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
          divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
          labels=blasthits[which(!is.na(heights[,i]))+1],
-         cex=0.75)
+         cex=0.9)
     #paste(names[which(!is.na(heights[,i]))+1],
   }
   mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
@@ -289,7 +289,8 @@ plotnet <- function(net) {
   plot(net,
        edge.curved=0.1, 
        edge.width=15*(E(net)$weight)^0.35,
-       vertex.label=parse(text=V(net)$Subtype2),
+       vertex.label=parse(text=ifelse(is.na(V(net)$Clade),
+                                      names(V(net)), V(net)$Subtype2)),
        vertex.color=ifelse(is.na(V(net)$Clade), "white",
                            taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))]),
        vertex.size=ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))))

### 2016-12-30 manuscript/manuscript.Rmd
Initial commit

---
title: Describing *Symbiodinium* metacommunities using high-throughput sequencing
  of ITS2
author:
- affiliation: University of Hawaii
  email: ross.cunning@gmail.com
  footnote: Corresponding Author
  name: Ross Cunning
- affiliation: University of Hawaii
  name: Ruth D. Gates
- affiliation: California State University Northridge
  name: Peter J. Edmunds
bibliography: library.bib
output:
  word_document: default
  pdf_document: default
  html_document:
    fig_caption: yes
subtitle: A case study from St. John, U.S. Virgin Islands
address:
- address: Hawaii Institute of Marine Biology, Kaneohe, HI 96744, USA
  code: University of Hawaii
- address: Department of Biology, Northridge, CA 91330, USA
  code: California State University Northridge
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = F, warning = F, message = F, cache = T)
options(scipen = 3, digits = 9)
```

```{r setup_2, cache=F, include=FALSE}
# Load all R scripts in "R/" directory
source("R/functions.R", .GlobalEnv)
# Load package libraries
library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr); library(RColorBrewer); library(pBrackets)
# Set colors for plotting clades
taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
```

```{r import_taxa_data, include=F}
# Import taxonomic assignment data from nw
read.nw <- function(file) {
  nw <- read.table(file, stringsAsFactors=FALSE)
  nw <- nw[order(nw$otu),]
  return(nw)
}

tax97 <- read.nw("data/otus_97/nw_tophits.tsv")
tax100 <- read.nw("data/otus_100/nw_tophits.tsv")
tax97bs <- read.nw("data/otus_97_bysample/nw_tophits.tsv")

# Deal with identical taxonomic assignments (because some reference sequences are not unique...)
# Create names for groups of identical sequences based on members of group...
ident <- readLines("data/ITS2db_trimmed_notuniques_otus.txt")
ident <- gsub("denovo[0-9]*\t", "", ident)
ident <- strsplit(ident, split="\t")
ident2 <- lapply(ident, str_match_all, pattern="[A-I]{1}[0-9]{1,3}.*[_/]")
ident2 <- lapply(ident2, function(g) gsub(pattern="\\..*_$", "", x=unlist(g)))
ident2 <- lapply(ident2, function(g) gsub(pattern="_$", "", g))
ident2 <- lapply(ident2, function(g) unlist(strsplit(g, split="/")))
subtypes <- lapply(ident2, function(x) levels(factor(unlist(x))))
subtypes <- lapply(subtypes, function(s) paste(paste(s, collapse="/"), "_multiple", sep=""))
names(ident) <- subtypes
ident <- melt(do.call(rbind, ident))
ident <- unique(ident[order(ident[,1]), c(3,1)])

# Replace any sequence name in taxonomy assignment that is a member of a group of identical sequences with the name of the group
collapse.idents <- function(df) {
  within(df, {
    for (i in 1:nrow(ident)) {
      hit <- gsub(ident[i,1], ident[i,2], hit)
    }
  })
}

tax97 <- collapse.idents(tax97)
tax97bs <- collapse.idents(tax97bs)
tax100 <- collapse.idents(tax100)

tax97 <- with(tax97, tax97[order(otu, -sim), ])
tax97 <- tax97[!duplicated(tax97$otu), ]
rownames(tax97) <- tax97$otu

tax97bs <- with(tax97bs, tax97bs[order(otu, -sim), ])
tax97bs <- tax97bs[!duplicated(tax97bs$otu), ]
rownames(tax97bs) <- tax97bs$otu

tax100 <- with(tax100, tax100[order(otu, -sim), ])
tax100 <- tax100[!duplicated(tax100$otu), ]
rownames(tax100) <- tax100$otu
```

```{r build_phyloseq, include=F}
# Import OTU tables
otu97 <- otu_table(read.table("data/otus_97/97_otus.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu97bs <- otu_table(read.table("data/otus_97_bysample/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)

# Import sample data
sam <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))

# Build phyloseq objects
phy97 <- phyloseq(otu97, tax_table(as.matrix(tax97)), sam)
phy97bs <- phyloseq(otu97bs, tax_table(as.matrix(tax97bs)), sam)
phy100 <- phyloseq(otu100, tax_table(as.matrix(tax100)), sam)
```

```{r filter_dataset, include=F}
# FILTER OUT OTUS THAT ARE NOT SYMBIODINIUM
# If the top hit in NCBI does not contain the string "Symbiodinium", then this sequence is assumed to not be Symbiodinium.
poormatch97 <- readLines("data/otus_97/poormatch_IDs.txt")
poormatch97 <- data.frame(otu=str_extract(poormatch97, "denovo[^ ]*"),
                          symbio=str_detect(poormatch97, "Symbiodinium"),   # TRUE if Symbiodinium
                          stringsAsFactors = FALSE)
notsymbio97 <- poormatch97[which(poormatch97$symbio==F), 1]

poormatch97bs <- readLines("data/otus_97_bysample/poormatch_IDs.txt")
poormatch97bs <- data.frame(otu=str_extract(poormatch97bs, "denovo[^ ]*"),
                            symbio=str_detect(poormatch97bs, "Symbiodinium"),   # TRUE if Symbiodinium
                            stringsAsFactors = FALSE)
notsymbio97bs <- poormatch97bs[which(poormatch97bs$symbio==F), 1]

poormatch100 <- readLines("data/otus_100/poormatch_IDs.txt")
poormatch100 <- data.frame(otu=str_extract(poormatch100, "denovo[^ ]*"),
                           symbio=str_detect(poormatch100, "Symbiodinium"),   # TRUE if Symbiodinium
                           stringsAsFactors = FALSE)
notsymbio100 <- poormatch100[which(poormatch100$symbio==F), 1]

# Remove OTUs that are not Symbiodinium
phy97.f <- prune_taxa(!taxa_names(phy97) %in% notsymbio97, phy97)
phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% notsymbio97bs, phy97bs)
phy100.f <- prune_taxa(!taxa_names(phy100) %in% notsymbio100, phy100)

# Filter OTUs by minimum count
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

# Filter samples by minimum count
# Set threshold number of reads
sn <- 200
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)

# Filter OTUs by minimum count again in case any dropped below threshold after filtering samples
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

# Label clades and subtypes for filtered phyloseq object tax_tables
get.st <- function(df) {
  within(df, {
    Clade <- substr(hit, 1, 1)
    Subtype <- gsub(hit, pattern="_[A-Z]{2}[0-9]{6}", replacement="")
    Subtype <- gsub(Subtype, pattern="_multiple", replacement="")
    Subtype2 <- ifelse(as.numeric(sim)==100, paste0("'", Subtype, "'"),
                       paste0("'", rep(rle(sort(Subtype))$values, times=rle(sort(Subtype))$lengths), "'^", 
                              unlist(lapply(rle(sort(Subtype))$lengths, seq_len)))[order(order(Subtype))])
    #Subtype <- ifelse(as.numeric(sim)==100, Subtype, paste("*", Subtype, sep=""))
  })
}

tax_table(phy97.f) <- as.matrix(get.st(data.frame(tax_table(phy97.f), stringsAsFactors=FALSE)))
tax_table(phy97bs.f) <- as.matrix(get.st(data.frame(tax_table(phy97bs.f), stringsAsFactors=FALSE)))
tax_table(phy100.f) <- as.matrix(get.st(data.frame(tax_table(phy100.f), stringsAsFactors=FALSE)))
```

```{r transform_dataset, include=F}
# TRANSFORM COUNT DATA
# Convert to proportion (relative abundance)
phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
phy100.f.t <- transform_sample_counts(phy100.f, transform)
```

### Abstract

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and susceptibility to stress of their coral hosts. Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic vs. interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity. Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St. John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses variability that is more likely to be intragenomic, while maintaining variability that is more likely interspecific. While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis. We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs. 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

### Introduction

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent outcomes such as photobiology, energetics, coral growth, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts within a coral may have important functional consequences [@Loram:2007p3688; @Cunning:2015ja]. Moreover, plasticity and flexibility of these communities within species and individuals over time has been identified as an important trait that may underlie a coral's sensitivity to stress [@Putnam:2012wv] and its ability to respond adaptively to environmental change by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, understanding and predicting ecological performance of corals depends on our ability to accurately and comprehensively describe their *Symbiodinium* communities.

  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we]. Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used. *Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation [@LaJeunesse:2014io]. However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes -- the one most commonly used is ITS2. High-throughput sequencing of ITS2 is now increasingly being utilized to characterize these communities at unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb]; however, these big datasets place even greater importance on issues surrounding the interpretation of ITS2 sequence diversity in relation to biological diversity.

  While ITS2 was initially investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not variable enough in all cases [@Finney:2010p8324; @Thornhill:2013dx]; for example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially others [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution. In fact, this process may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic vs. interspecific variation and preclude its use as a true species-level marker. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.

  Here, we describe a *Symbiodinium* metacommunity in St. John, U.S. Virgin Islands, with the objectives of 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical approaches uniquely enabled by such datasets to advance understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' to extract meaningful information from ITS2 metabarcoding data given its inherent limitations. Specifically, this approach seeks to cope with the fact that sequences within a genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt]. Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates diversity by collapsing species known to be different. True biological diversity may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.

  These assumptions are that 1) most individual colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions together suggest that closely related ITS2 sequences (>97% similar; @Arif:2014gk) occurring within the same sample are more likely to arise from a single taxon rather than multiple, but closely related ITS2 sequences occurring as numerically dominant variants in different samples are more likely to represent distinct *Symbiodinium* taxa. These assumptions can be leveraged to bioinformatically parse intragenomic and interspecific variation by clustering sequences at 97% similarity within each sample independently, in order to collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St. John. We then explore ecological patterns only made possible by large metabarcoding datasets, including 1) testing for whole-community differences corresponding to environmental variability between the north and south shores, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying and testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species. The volume of data and multitude of analysis approaches facilitated by ITS2 metabarcoding has the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology. More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.

### Materials and Methods

#### Sample collection and processing

  A total of 87 tissue samples from scleractinian and milleporine corals were collected from various sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Small tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay. Samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

```{r Fig1, fig.height=3, fig.width=4, fig.cap="Figure 1. Sampling locations in St. John, U.S. Virgin Islands."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

sitecoords <- data.frame(
  name=c("Booby Rock", "Tektite", "Kiddel", "Groot Pan", "Yawzi Point", "Anna's Point", "Haulover Bay"),
  lat=c(18.30327, 18.31057, 18.30770, 18.31067, 18.31515, 18.36710, 18.35008),
  lon=c(-64.71057, -64.72203, -64.71398, -64.71813, -64.72513, -64.73337, -64.67933)
)

viers <- c(-64.722697, 18.322277)

#pdf(file="figures/map.pdf", height=3, width=4)
par(mar=c(0,0,0,0))
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
box()
text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
})
points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
#dev.off()
```

#### Bioinformatics

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was less than 90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".

  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI. These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev]. Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file. The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce this analysis.

#### *Symbiodinium* data analysis

  OTU tables, sample metadata, and taxonomic data were imported into R using the phyloseq package [@McMurdie:2013ky] to facilitate downstream analyses. OTU tables were filtered to remove any OTU that was not observed at least `r n` times, or any sample with less than `r sn` sequences, and counts were transformed to relative abundance. Permutational analysis of variance (PERMANOVA) was used to test for differences in *Symbiodinium* community composition between the north and south shores of St. John within each coral species.

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. For the dominant and abundant symbionts networks, weighted edges were defined as the proportion of individuals of a species in which a symbiont OTU occurred at greater than 50% or greater than 1% relative abundance, respectively. For the background symbionts network, unweighted edges were defined based on the presence of a symbiont at less than 1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm.

  Beta diversity was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

#### Environmental data

  Sea Surface Temperature (SST) and chlorophyll concentration (chla): nighttime sea surface temperature and ocean colour data at 1 km spatial resolution were extracted from the sensor MODIS aboard the Aqua satellite. We used as input the monthly climatologies available at the IMaRS website. From the 12 monthly climatological values, we calculated the median, mean, standard deviation and minimum and maximum monthly mean. Climatologies for MODIS cover the period 2003 to 2010.

  Wind-driven wave exposure for each station is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves. To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these datasets wave exposure was calculated using the methods based on wave theory described in Chollett et al. (2012) for 32 fetch directions and the coastline data at full resolution. Average wave exposure at each station was calculated in R with the aid of the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

### Results

#### Comparison of clustering approaches

  The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving n=`r nsamples(phy97.f)` samples for downstream analysis. The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n`.

```{r Table1}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f)[c(2, 3, 6, 8), ], align="c",
             caption = "Table 1. Summary statistics for each clustering approach.")
```

  A subset of samples were selected for comparative analysis of the three clustering approaches (Fig. 1). 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples. Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), we present the remainder of the results using the 97% within-sample clustering approach. Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the closest BLAST hit to the representative sequence of the OTU; those with the same annotation but different fill colors represent unique sequence variants with the same closest BLAST hit."}
set.seed(42432)
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

#### Symbiodinium community composition

```{r cladeAbund, include=F}
# Calculate relative abundance of each clade
cladeAbund <- aggregate(data.frame(RelAbund=rowSums(otu_table(phy97bs.f.p))),
                        by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
cladeAbund$Prop <- prop.table(cladeAbund$RelAbund) * 100
cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
clademax <- aggregate(data.frame(max=apply(otu_table(phy97bs.f.p), 1, FUN=max)),
                      by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=max)
```

  Within the set of coral samples obtained in this study, clade B had the highest relative abundance (`r with(cladeAbund, Prop[Clade=="B"])`%), followed by clade C (`r with(cladeAbund, Prop[Clade=="C"])`%), A (`r with(cladeAbund, Prop[Clade=="A"])`%), D (`r with(cladeAbund, Prop[Clade=="D"])`%), and G (`r with(cladeAbund, Prop[Clade=="G"])`%. The number of OTUs within each clade followed a similar pattern, with the highest number in clade B (`r with(cladeAbund, Notus[Clade=="B"])`), followed by clade C (`r with(cladeAbund, Notus[Clade=="C"])`), A (`r with(cladeAbund, Notus[Clade=="A"])`), D (`r with(cladeAbund, Notus[Clade=="D"])`), and G (`r with(cladeAbund, Notus[Clade=="G"])`). The distribution of these clades within each sample is shown in Fig. 3.

```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore. Individual segments of each bar represent distinct 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
```

#### Differences between shores

  The north shore of St. John was characterized by lower wave exposure, slightly higher and more variable temperatures, and much higher and more variable chlorophyll A, relative to the south shore (Table 2).

```{r Table2}
envdata <- read.csv("data/STJ_envdata.csv", check.names=F, row.names=1)
sites <- envdata[-1,4:5]
colnames(sites) <- c("North", "South")
sites$North <- as.numeric(as.character(sites$North))
sites$South <- as.numeric(as.character(sites$South))
knitr::kable(sites, digits=3, caption="Table 2. Environmental differences between the north and south shores of St. John, U.S. Virgin Islands. (North shore=64.711967°W, 18.368141°N; South shore=64.726840°W,	18.315465°N).")
```

  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any coral species (Table 3). However, some species showed trends that may have been significant with larger sample sizes. For example, *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056). *Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.071).

```{r Table3, cache=T} 
set.seed(43789) 
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 3. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

#df <- data.frame(otu_table(phy97bs.f.p), check.names=F)

#clade.otutab <- aggregate(df, by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
#rownames(clade.otutab) <- clade.otutab[,1]
#clade.otutab <- clade.otutab[,-1]

#clade.phy <- phyloseq(otu_table(clade.otutab, taxa_are_rows=T), sample_data(sam))
#perms(clade.phy, group="Species", trt="Shore")

```

#### Network analyses of *Symbiodinium* meta-community
```{r networks,  cahce=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

# Create abundant symbionts network
set.seed(12374)
abunet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.01])/length(x),
                 layout=layout.fruchterman.reingold)
# Calculate number of abundant symbionts connected to each species
abu.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(abunet)[from(x)]))
abu.Nedges <- sort(abu.Nedges, decreasing=T)

# Create background symbionts network
# filter out OTUs that only occurred in one sample
phy97bs.f.p2 <- filter_taxa(phy97bs.f.p, function(x) sum(x>0) > 1, prune=TRUE)
bkgnet <- sppnet(phy=phy97bs.f.p2, plot=F,
                 fun=function(x) any(x>0 & x<0.01),
                 layout=layout.fruchterman.reingold)
# Merge D nodes (assuming they are all D1a/trenchii)
# Get original node map (sequence starting at one to the total number of nodes)
nodemap <- seq(1:length(V(bkgnet)))
# Make a copy of the nodemap that will be modified
nodemapnew <- nodemap
# Find all nodes beginning with D, and give them the same node ID number
nodemapnew[grep("^D", V(bkgnet)$Subtype)] <- min(grep("^D", V(bkgnet)$Subtype))
# Replace the now missing values in the node map so node IDs remain sequential
while (max(setdiff(nodemap, nodemapnew)) < max(nodemap)) {
  nodemapnew[which(nodemapnew==max(setdiff(nodemap, nodemapnew))+1)] <- min(setdiff(nodemap, nodemapnew))
}
# Contract the network using the new nodemap (nodes with same node ID are merged)
bkgnet2 <- contract(bkgnet, nodemapnew, vertex.attr.comb=list(size="sum", "last"))
# Simplify network so that multiple edges are combined
bkgnet2 <- simplify(bkgnet2, remove.multiple=T, edge.attr.comb=list(weight="mean", width="mean", "ignore"))
# remove symbionts that are only connected to one or two coral species
bkgnet3 <- delete.vertices(bkgnet2, degree(bkgnet2)<=2) # & V(net2)$type==2)
V(bkgnet3)$size <- ifelse(is.na(V(bkgnet3)$Clade), 15, 1.5*degree(bkgnet3)^1.5)
E(bkgnet3)$width <- E(bkgnet3)$weight * 2

# Calculate number of edges for background symbionts
nspp <- sapply(V(bkgnet3)[V(bkgnet3)$type==2], function(x) length(E(bkgnet3)[from(x)]))
names(nspp) <- V(bkgnet3)$label[match(names(nspp), V(bkgnet3)$name)]
nspp <- sort(nspp, decreasing=T)
```

  We analyzed patterns of association between coral species and *Symbiodinium* by constructing three networks focusing on "abundant" (>1% relative abundance; Fig. 4), "dominant" (>50% relative abundance; Fig. 5A), and "background" (<1% relative abundance; Fig. 5B) symbionts in each coral species, based on 97% within-sample clustering. (Networks for individuals within each species are presented in Supplementary Materials). In the abundant symbiont network (Fig. 4), `r sum(V(abunet)$type==2)` distinct OTUs were observed. *D. strigosa* associated with the greatest number of distinct OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F. fragum* (n=`r abu.Nedges["fragum"]`), *M. alcicornis* (n=`r abu.Nedges["alcicornis"]`), and *D. cylindrus* (n=`r abu.Nedges["cylindrus"]`); these primarily comprised OTUs closely related to B1 and B19. In contrast, only a single OTU occurred at >1% relative abundance in both *P. astreoides* (A4) and *M. cavernosa* (C3). All coral species hosted at least one "abundant" *Symbiodinium* OTU that was also abundant in at least one other coral species, except for *S. radians*. This species was heavily dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.

  A total of `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs occurred as "dominant" symbionts, meaning they were present at >50% relative abundance within a sample (Fig. 5A). A single *Symbiodinium* B1 OTU was the most prevalent dominant symbiont, dominating all *D. cylindrus* samples and many *D. strigosa*, *O. annularis*, *M. alcicornis*, and *F. fragum* samples. Other closely related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *D. strigosa*, *M. alcicornis*, and *F. fragum*, while *O. annularis* was dominated just as often by *Symbiodinium* C7. After B1, the next most prevalent dominant symbionts were *Symbiodinium* C3, which dominated all *M. cavernosa* and most *S. siderea* samples, and *Symbiodinium* A4, which dominated all *P. astreoides* and most *P. furcata* samples. *P. furcata* and *S. siderea* were each occasionally dominated by other symbiont OTUs, including C45 or C45a (*P. furcata*), and C1 or D1 (*S. siderea*). All coral species were dominated by at least one *Symbiodinium* OTU that also dominated at least one other coral species, except for *S. radians*, which was exclusively dominated by C46.

  There were `r sum(V(bkgnet3)$type==2)` symbionts that occurred at background levels (<1% relative abundance) in at least three coral species (Fig. 5B). *Symbiodinium* D1a/*S. trenchii* was detected in the highest number of species (n=`r nspp["D1a"]`), followed by A4 (n=`r nspp["A4"]`) and C3 (n=`r nspp["C3"]`). C31, C46, and B1 were each detected at background levels in 4 different species.

```{r Fig4, fig.width=5, fig.height=5, fig.align="center", fig.retina=1, out.width=5, fig.cap="<br>Figure 4. Network analysis of abundant *Symbiodinium* taxa in each coral species. *Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at greater than 1% relative abundance within an individual; thickness of edges is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%. Symbiont nodes are colored by clade identity, sized proportionally to the number of coral species they dominated."}
# Plot abundant symbionts network
par(mfrow=c(1,1), mar=c(0,0,0,0), lwd=1)
V(abunet)$size <- ifelse(is.na(V(abunet)$Clade), 10, 13*(degree(abunet))^0.4)
set.seed(12374)
l <- layout.fruchterman.reingold(abunet)
l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
plot(abunet, rescale=F, layout=l, edge.curved=0.1, 
     vertex.label.cex=ifelse(is.na(V(abunet)$Clade), V(abunet)$size/15, V(abunet)$size/18),
     vertex.label.color="black") 
```

```{r Fig5, fig.width=7.20472, fig.height=3.60236, fig.align="center", fig.cap="Figure 5. Network analysis of dominant (A) and background (B) *Symbiodinium* taxa in each coral species (see Materials and Methods for details on network construction). Symbiont nodes are colored by clade identity, sized proportionally to the number of coral species to which they are connected."}

par(mfrow=c(1,2), mar=c(0.1,1,1.9,1), lwd=1, xpd=NA) 
# Plot dominant symbionts network
set.seed(12374)
l <- layout.fruchterman.reingold(domnet)
l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
V(domnet)$size <- ifelse(is.na(V(domnet)$Clade), 16, 18*(degree(domnet))^0.5)
plot(domnet, rescale=F, layout=l, edge.curved=0.1, vertex.label.cex=V(domnet)$size/25,
     vertex.label.color="black")
title("A.) Dominant symbionts", adj=0)

# Plot background symbionts network
set.seed(42384)
l3 <- layout.fruchterman.reingold(bkgnet3)
l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
V(bkgnet3)$size <- ifelse(is.na(V(bkgnet3)$Clade), 16, 8*(degree(bkgnet3))^0.8)
plot(bkgnet3, rescale=F, layout=l3, edge.curved=0.1, vertex.label.cex=V(bkgnet3)$size/25,
     vertex.label.color="black")
points(0.9,0.9, pch=0, cex=3.5)
text(0.9,0.9, "Oa", cex=1)
title("B.) Background symbionts", adj=0)
```

#### Symbiotic flexibility of coral species

  Significant differences in symbiotic flexibility (i.e. multivariate dispersion) were detected among coral species. The highest flexibility was found in *D. strigosa*, *P. furcata*, *M. alcicornis*, *S. siderea*, and *M. annularis*, and the lowest was found in *M. cavernosa*. Flexibility was also low in *D. cylindrus*, *S. radians*, and *P. astreoides*, and intermediate in *F. fragum*.

```{r Fig6, cache=F, fig.width=3.5, fig.height=3.5, fig.cap="Figure 6. Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition."}
betad97bs <- betad(phy97bs.f.t, group="Species")

par(mar=c(3,3,1,1), mgp=c(1.8,0.1,0), tcl=0.25)
with(betad97bs$sambdsumm.ord, {
  plot(mean, type="n", ylim=c(0, 0.65), ylab="Distance to centroid", xaxt="n", xlab="Coral species")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  axis(side=1, at=1:10, cex.axis=0.75,
       labels=str_replace_all(levels(sam$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)],
                              c("alcicornis"="Ma", "annularis"="Oa",
                                "astreoides"="Pa", "cavernosa"="Mc",
                                "cylindrus"="Dc", "fragum"="Ff", 
                                "furcata"="Pf", "radians"="Sr", 
                                "siderea"="Ss", "strigosa"="Ds")))
  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
})
```

### Discussion

#### Bioinformatic analysis of ITS2 diversity

  The perceived structure of corals' *Symbiodinium* communities was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data; specifically, the method of clustering sequence diversity into operational taxonomic units (OTUs). Clustering sequences from all samples together at 97% similarity reduced the total number of OTUs by 98% relative to no clustering (Table 1). Intermediate diversity was resolved by within-sample clustering, with 13% more OTUs than across-sample clustering. By comparing the qualitative outcomes of these approaches (Fig. 2), it is apparent that across-sample clustering may assign the same dominant OTU to samples with dissimilar sequence assemblages. This results from the fact that an OTU's representative sequence (and basis for taxonomic assignment) is selected as the numerically dominant member; when these member sequences occur across many samples, the most abundant may simply reflect the number and composition of samples in the dataset. Consequently, closely related sequences arising from different *Symbiodinium* may be subsumed under the dominant variant that happens to be most well-represented (i.e., most heavily sampled). Therefore, while 97% similarity may successfully collapse intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when clustering is applied across a dataset potentially comprised of many closely related types, it may also collapse interspecific variation and underestimate true diversity.

  Meanwhile, clustering within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the size or structure of the sample set. The outcomes of within-sample clustering more accurately reflect knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that considerable species diversity exists within clades B and C in the Caribbean that associates strongly with host specificity. Indeed, only within-sample clustering assigned different dominant clade C types to samples of *O. annularis* and *S. siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx]. While multi-locus approaches clearly provide a superior description of *Symbiodinium* diversity, within-sample clustering of ITS2 metabarcoding data may be a good approximation of diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding). Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa. Therefore, we recommend within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.

  Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* clone that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met. In fact, multiple types co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming variation within a sample is intragenomic. However, co-occurring types in many cases are members of different clades, with ITS2 sequences divergent enough to still be resolved separately by 97% clustering. Thus, only when very closely related types co-occur [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail, requiring population genetic markers to parse interspecific variation. Other problematic cases may occur when multiple ITS2 sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that a slight numerical dominance of one or the other may lead to different OTU assigments in samples with very similar sequence assemblages (e.g., Fig. SX). These challenges may be overcome by 1) analyzing read abundance correlations across samples to inform OTU assignments, or 2) using alternative criteria for taxonomic assignment, such as the most evolutionarily derived sequence variant.

  While within-sample clustering of ITS2 data cannot provide species-level descriptions of *Symbiodinium* communities (since the ITS2 marker itself does not provide this resolution), this nuanced bioinformatic approach nevertheless accommodates known complexities in the evolution of the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.

#### *Symbiodinium* metacommunity ecology in St. John

  When treated with a bioinformatic approach that reflects the complexity of the marker, ITS2 metabarcoding remains a very powerful way to describe *Symbiodinium* communities - it provides a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample, has a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species and scaled up to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.

  In the ten host species sampled at St. John USVI, the most prevalent *Symbiodinium* were members of clades B and C, consistent with previous work in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324]. Less frequent associations with members of clades A and D were observed, though these types did dominate colonies of certain species (clade A dominated *P. astreoides* and *P. furcata*, while clade D dominated *S. siderea*). Finally, clade G was also observed at low relative abundances (<`r round(clademax[5,2], 1)` within samples), which likely would not have been detected without high-throughput sequencing.

  When the associations between *Symbiodinium* types and coral species are visualized by network analysis, features of both symbiont and host ecology emerge. First, ITS2 type B1 was "abundant" (i.e. >1%) in individuals of seven out of ten coral species, and "dominant" (i.e. >50%) in five out of ten, suggesting it is a more generalist symbiont lineage. However, analysis of more variable chloroplast and microsatellite loci has revealed that ITS2 type B1 in the Caribbean is comprised of numerous specialized lineages that demonstrate specific associations with different coral host species [@Santos:2004p4044; @Finney:2010p8324; @Parkinson:2015gt]. Likewise, the apparent generalist *Symbiodinium* C3 (which dominated colonies of both *S. siderea* and *M. cavernosa*) and *Symbiodinium* A4 (which dominated both *P. astreoides* and *P. furcata*) may also be comprised of multiple specialized lineages [@Thornhill:2013dx], which will require evaluation with population genetic markers.

  In addition to hosting apparent generalist ITS2 lineages, most coral species (except for *M. cavernosa* and *P. astreoides*) hosted abundant OTUs that were not abundant in any other corals, and are thus apparently more host-specific. Unique OTUs in the B1-radiation were abundant in *M. alcicornis* and *D. cylindrus*, while others from both the B1- and B19-radiations [see @LaJeunesse:2004p122] were abundant in *F. fragum*, *S. radians*, and *D. strigosa*. Several clade C OTUs were similarly only abundant in one species: C1144 in *D. strigosa*, C7 in *O. annularis*, C1 in *S. siderea*, C1c/C45 and C45a in *P. furcata*, and C46 in *S. radians*. Interestingly, *S. radians* was the only coral species in which none of the abundant symbionts occurred in any other corals, which may reflect the unique ecology of *S. radians* as the only study species that typically forms small, encrusting colonies.

  Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at less than 1% relative abundance in at least three different host species. Such a distribution suggests that these symbionts can occupy a range of hosts, but functionally do not outperform other resident symbionts to become dominant under prevailing environmental conditions. Their presence, however, indicates that they may be able to achieve dominance within a colony under certain environmental conditions. Indeed, the most prevalent 'background' symbiont in St. John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], in seven out of ten species), a thermotolerant symbiont that has been shown to proliferate within coral hosts during and after thermal stress [@LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790]. Association with these symbionts may therefore be indicative of a broader fundamental niche space and resilience to environmental change. *Symbiodinium* A4 and C3 also commonly occurred as background symbionts, suggesting they may perform analogous functional and ecological roles within the metacommunity.

  Focusing on coral hosts, there was a clear difference among species in the number of OTUs that were abundant (Fig. 4) and dominant (Fig. 5A), indicating different levels of variability in the symbiosis ecology of different coral species. *D. strigosa* individuals were dominated by four different OTUs, while *D. cylindrus*, *S. radians*, *M. cavernosa*, and *P. astreoides* were dominated by a single OTU. In fact, metabarcoding data enables quantifying symbiosis flexibility as multivariate dispersion (a metric of beta diversity), which can be statistically compared among species [@Anderson:2006fs]. The flexibility of symbiosis is important as a potential indicator of the environmental sensitivity of a species [@Putnam:2012wv], and its ability to switch or shuffle its symbionts. This metric revealed that *M. cavernosa*, *P. astreoides*, *S. radians*, and *D. cylindrus* all had the lowest symbiotic flexibility. On the other hand, *D. strigosa*, *P. furcata*, *M. alcicornis*, and *S. siderea* had the most variable symbiont communities, indicating that these species may be more 'flexible' in their associations, and that their symbiont community structure may be more responsive to variability in the natural environment.

  To analyze potential influence of the environment at a larger spatial scale, we tested whether symbiont community composition of each species differed between the north and south shores. While environmental data revealed a significant biophysical differences between these two sides of the island (e.g., wave height, temperature, and chlorophyll A), there were no significant differences in symbiont community composition. This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S. siderea* and *P. furcata* tended to be dominated more frequently by *Symbiodinium* in clades D and A, respectively, on the south shore, compared to the north shore (Fig. 2). Since these symbionts are typically associated with higher temperature environments, it is counter to expectation that they were more prevalent on the south shore, which was characterized by slightly lower and more stable temperatures. This pattern may indicate that other factors that we did not measure (or did not measure with high enough spatial resolution) may be more important in structuring these communities.

#### Conclusions

  While not providing species-level resolution, ITS2 metabarcoding represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs. With proper bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, changes in dominant or background symbiont abundance over environmental gradients of over time, with important implications for understanding coral responses to environmental change.

### Acknowledgements
Funding for this work was provided by _________. RC was supported by a NSF Postdoctoral Research Fellowship in Biology (NSF-PRFB 1400787). Remote sensing observations of sea surface temperature and phytoplankton pigment concentration were processed at the Institute for Marine Remote Sensing at the College of Marine Science, University of South Florida, in St. Petersburg, Florida, and provided courtesy of Frank Müller-Karger.

### References

### 2017-01-02 manuscript/manuscript.Rmd
incorporate Pete comments

@@
 
-  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent outcomes such as photobiology, energetics, coral growth, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
+**Be very explicit up front about what is meant by type vs species**
+
+  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
 Even slight differences in the relative abundance of different symbionts within a coral may have important functional consequences [@Loram:2007p3688; @Cunning:2015ja].
-Moreover, plasticity and flexibility of these communities within species and individuals over time has been identified as an important trait that may underlie a coral's sensitivity to stress [@Putnam:2012wv] and its ability to respond adaptively to environmental change by 'shuffling' or 'switching' symbionts [@Baker:2003p200].
-Therefore, understanding and predicting ecological performance of corals depends on our ability to accurately and comprehensively describe their *Symbiodinium* communities.
+Moreover, plasticity and flexibility of these communities within species and individuals over time is an important trait modulating the sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond adaptively by 'shuffling' or 'switching' symbionts [@Baker:2003p200].
+Therefore, understanding the ecological performance of corals depends on accurately and comprehensively describing their *Symbiodinium* communities.
 
   *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we].
-Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used.
-*Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation [@LaJeunesse:2014io].
-However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes -- the one most commonly used is ITS2.
-High-throughput sequencing of ITS2 is now increasingly being utilized to characterize these communities at unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb]; however, these big datasets place even greater importance on issues surrounding the interpretation of ITS2 sequence diversity in relation to biological diversity.
+Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them.
+*Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io].
+However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2.
+High-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb].
+However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
 
-  While ITS2 was initially investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not variable enough in all cases [@Finney:2010p8324; @Thornhill:2013dx]; for example, the 'B1' ITS2 sequence variant is shared by *S.
+  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within this genus [@Finney:2010p8324; @Thornhill:2013dx].
+For example, the 'B1' ITS2 sequence variant is shared by *S.
 minutum*, *S.
 pseudominutum*, and *S.
-antillogorgium*, and potentially others [@Parkinson:2015gt].
-Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution.
-In fact, this process may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx].
-Together, these features of ITS2 complicate the interpretation of intragenomic vs.
-interspecific variation and preclude its use as a true species-level marker.
-Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker.
+antillogorgium*, and potentially other species [@Parkinson:2015gt].
+Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (sensu XXX).
+In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx].
+Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker.
+Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis (e.g., GeoSymbio, etc.), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms (XXX), it remains an essential and powerful marker for the *Symbiodinium* system.
 Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.
 
-  Here, we describe a *Symbiodinium* metacommunity in St.
+  Here, we describe a *Symbiodinium* metacommunity with scleractinians and a *Millepora* hydrocoral in St.
 John, U.S.
-Virgin Islands, with the objectives of 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical approaches uniquely enabled by such datasets to advance understanding of *Symbiodinium* metacommunity ecology.
-In addressing the first objective, we test the ability of 'within-sample clustering' to extract meaningful information from ITS2 metabarcoding data given its inherent limitations.
-Specifically, this approach seeks to cope with the fact that sequences within a genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt].
-Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates diversity by collapsing species known to be different.
-True biological diversity may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.
-
-  These assumptions are that 1) most individual colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422].
-These assumptions together suggest that closely related ITS2 sequences (>97% similar; @Arif:2014gk) occurring within the same sample are more likely to arise from a single taxon rather than multiple, but closely related ITS2 sequences occurring as numerically dominant variants in different samples are more likely to represent distinct *Symbiodinium* taxa.
-These assumptions can be leveraged to bioinformatically parse intragenomic and interspecific variation by clustering sequences at 97% similarity within each sample independently, in order to collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).
+Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications by which such datasets may advance our understanding of *Symbiodinium* metacommunity ecology.
+In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations.
+Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt].
+Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates species diversity by collapsing species known to be different.
+True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.
+These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422].
+These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to represent a single taxon rather than multiple, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa.
+Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).
 
@@
 John.
-We then explore ecological patterns only made possible by large metabarcoding datasets, including 1) testing for whole-community differences corresponding to environmental variability between the north and south shores, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying and testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species.
-The volume of data and multitude of analysis approaches facilitated by ITS2 metabarcoding has the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology.
+We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species.
+The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts.
 More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.
@@
 
-  A total of 87 tissue samples from scleractinian and milleporine corals were collected from various sites around St.
+  A total of 87 tissue samples from scleractinian and milleporine corals were collected from seven sites around St.
 John, US Virgin Islands, USA (Fig.
 1), between August 7th and 9th, 2012.
-Small tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.
+Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth.
+Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.
 Samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB).
@@
 
-```{r Fig1, fig.height=3, fig.width=4, fig.cap="Figure 1. Sampling locations in St. John, U.S. Virgin Islands."}
+#### Environmental data
+
+  Broad-scale differences in the hydrographic regimes of the north and south shores of St.
+John were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave regime.
+These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [xxx] and wave regime [xxx]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition (xxx).
+SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite [xxx].
+From the 12 monthly climatological values obtained for each sampling year (2003-2010, n=96 months), the median, mean, standard deviation and minimum and maximum monthly mean were calculated.
+
+  Wind-driven wave exposure for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves.
+To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz].
+Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk].
+From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al.
+2012) for 32 fetch directions and coastline data at full resolution.
+Mean wave exposure (J m^-3^) at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].
+
+#### Bioinformatics
+
+  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence.
+Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf].
+Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained.
+Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently).
+For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical.
+After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf].
+Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript.
+If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".
+
+  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI.
+These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev].
+Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file.
+The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce the present analysis.
+
+#### *Symbiodinium* data analysis
+
+  OTU tables, sample metadata, and taxonomic data were imported into R using the phyloseq package [@McMurdie:2013ky] to facilitate downstream analyses.
+OTU tables were filtered to remove any OTU that was not observed at least `r n` times, or any sample with < `r sn` sequences, and counts were transformed to relative abundance.
+Permutational analysis of variance (PERMANOVA) was used to test for differences in *Symbiodinium* community composition between the north and south shores of St.
+John within each coral species.
+
+  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh].
+Networks were created featuring 'dominant' (>50% relative abundance) and 'abundant' (>1% relative abundance) OTUs, with weighted edges proportional to the number of individuals within a species in which a symbiont OTU occurred.
+A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species.
+To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network.
+All network layouts were constructed based on the Fruchterman-Reingold algorithm (XXX).
+
+  Beta diversity (sensu XXX) was calculated as the multivariate dispersion of samples within a coral species.
+Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test.
+This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].
+
+### Results
+
+#### Environmental and sequence data overview
+
+  The environmental conditions broadly characterizing the north and south shores are presented in Figure 1.
+From 2003 to 2010, the north shore of St.
+John was characterized by lower wave exposure, slightly higher and more variable sea surface temperatures, and higher and more variable chlorophyll a concentrations, relative to the south shore.
+(Statistical tests?)
+
+```{r Fig1, fig.height=3, fig.width=5, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
 library(rgdal)
@@
 
-#pdf(file="figures/map.pdf", height=3, width=4)
-par(mar=c(0,0,0,0))
+#pdf(file="figures/map.pdf", height=3, width=6)
+layout(mat=matrix(c(1,1,1,2,1,1,1,3,1,1,1,4), nrow=3, byrow=T))
+par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1)
 plot(stjmap, lwd=1, col="gray40",
-     xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
-box()
-text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
-axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
-axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
+     xlim=c(-64.74, -64.68), ylim=c(18.303, 18.370))
+#box()
+text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
+axis(side=3, at=seq(-64.77, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
+axis(side=3, at=seq(-64.77, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
 axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
@@
 text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
+NS <- c(-64.711967, 18.368141)
+SS <- c(-64.726840, 18.315465)
+points(NS[1], NS[2], pch="*", cex=2, col="red")
+points(SS[1], SS[2], pch="*", cex=2, col="red")
+#dev.off()
+
+# Plot environmental data
+envdata <- read.csv("data/STJ_envdata.csv", check.names=F, row.names=1)
+sites <- envdata[-1,4:5]
+colnames(sites) <- c("North", "South")
+sites$North <- as.numeric(as.character(sites$North))
+sites$South <- as.numeric(as.character(sites$South))
+sites <- data.frame(t(sites))
+
+#SST
+sst <- sites[, 2:6]
+sst <- within(sst, {
+  `Mean+S.D.` <- Mean.SST + S.D..SST
+  `Mean-S.D.` <- Mean.SST - S.D..SST
+})
+
+par(cex=0.7, mar=c(2,0,1,2), mgp=c(1.5,0.5,0), xpd=NA)
+plot(NA, xlim=c(26,29), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
+axis(side=1, at=seq(26,29,1))
+bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
+bxp(list(stats=matrix(as.numeric(sst[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)
+
+#chlA
+chla <- sites[, 7:11]
+chla <- within(chla, {
+  `Mean+S.D.` <- Mean.Chla + S.D..Chla
+  `Mean-S.D.` <- Mean.Chla - S.D..Chla
+})
+
+#par(new=T, mar=c(7,14,10,8), cex=0.7)
+plot(NA, xlim=c(0.03,0.09), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
+axis(side=1, at=seq(0.03,0.09,0.01))
+bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
+bxp(list(stats=matrix(as.numeric(chla[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)
+
+t(nsamp)[2,]
 #dev.off()
@@
 
-#### Bioinformatics
-
-  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence.
-Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf].
-Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained.
-Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently).
-For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical.
-After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf].
-Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript.
-If the match was less than 90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".
-
-  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI.
-These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev].
-Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file.
-The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce this analysis.
-
-#### *Symbiodinium* data analysis
-
-  OTU tables, sample metadata, and taxonomic data were imported into R using the phyloseq package [@McMurdie:2013ky] to facilitate downstream analyses.
-OTU tables were filtered to remove any OTU that was not observed at least `r n` times, or any sample with less than `r sn` sequences, and counts were transformed to relative abundance.
-Permutational analysis of variance (PERMANOVA) was used to test for differences in *Symbiodinium* community composition between the north and south shores of St.
-John within each coral species.
-
-  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh].
-For the dominant and abundant symbionts networks, weighted edges were defined as the proportion of individuals of a species in which a symbiont OTU occurred at greater than 50% or greater than 1% relative abundance, respectively.
-For the background symbionts network, unweighted edges were defined based on the presence of a symbiont at less than 1% relative abundance in at least one individual within the species.
-To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network.
-All network layouts were constructed based on the Fruchterman-Reingold algorithm.
-
-  Beta diversity was calculated as the multivariate dispersion of samples within a coral species.
-Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test.
-This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].
-
-#### Environmental data
-
-  Sea Surface Temperature (SST) and chlorophyll concentration (chla): nighttime sea surface temperature and ocean colour data at 1 km spatial resolution were extracted from the sensor MODIS aboard the Aqua satellite.
-We used as input the monthly climatologies available at the IMaRS website.
-From the 12 monthly climatological values, we calculated the median, mean, standard deviation and minimum and maximum monthly mean.
-Climatologies for MODIS cover the period 2003 to 2010.
-
-  Wind-driven wave exposure for each station is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves.
-To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz].
-Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database which provides global coastline at 1:250,000 scale [@Wessel:1996wk].
-From these datasets wave exposure was calculated using the methods based on wave theory described in Chollett et al.
-(2012) for 32 fetch directions and the coastline data at full resolution.
-Average wave exposure at each station was calculated in R with the aid of the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].
-
-### Results
+  Sequence data were successfully generated from 84 coral samples collected across two sites on the north shore and five sites on the south shore of St.
+John (Fig.
+1).
+The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
+After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.
 
@@
 
-  The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
-After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving n=`r nsamples(phy97.f)` samples for downstream analysis.
    The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig.
@@
 
-```{r Table1}
+```{r Table2}
 # Compute summary statistics 
@@
 
-  A subset of samples were selected for comparative analysis of the three clustering approaches (Fig.
+  A subset of samples was selected for comparative analysis of the three clustering approaches (Fig.
 1).
 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples.
-Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), we present the remainder of the results using the 97% within-sample clustering approach.
+Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), the remainder of the results is presented using the 97% within-sample clustering approach.
 Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.
@@
 set.seed(42432)
+par(mfrow=c(1,1))
 otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
@@
                         by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
-cladeAbund$Prop <- prop.table(cladeAbund$RelAbund) * 100
+cladeAbund$Prop <- round(prop.table(cladeAbund$RelAbund) * 100, 1)
 cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
@@
 
-#### Differences between shores
+#### Differences in *Symbiodinium* between shores
 
-  The north shore of St.
-John was characterized by lower wave exposure, slightly higher and more variable temperatures, and much higher and more variable chlorophyll A, relative to the south shore (Table 2).
-
-```{r Table2}
-envdata <- read.csv("data/STJ_envdata.csv", check.names=F, row.names=1)
-sites <- envdata[-1,4:5]
-colnames(sites) <- c("North", "South")
-sites$North <- as.numeric(as.character(sites$North))
-sites$South <- as.numeric(as.character(sites$South))
-knitr::kable(sites, digits=3, caption="Table 2.
-Environmental differences between the north and south shores of St.
-John, U.S.
-Virgin Islands.
-(North shore=64.711967°W, 18.368141°N; South shore=64.726840°W,	18.315465°N).")
-```
-
-  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any coral species (Table 3).
-However, some species showed trends that may have been significant with larger sample sizes.
-For example, *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056).
+  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any cnidarian species (Table 3).
+However, some qualitative differences were apparent: *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056).
 *Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.071).
@@
 
-  We analyzed patterns of association between coral species and *Symbiodinium* by constructing three networks focusing on "abundant" (>1% relative abundance; Fig.
+  Patterns of association between hosts and *Symbiodinium* were analyzed by constructing three networks focusing on "abundant" (>1% relative abundance; Fig.
 4), "dominant" (>50% relative abundance; Fig.
 5A), and "background" (<1% relative abundance; Fig.
-5B) symbionts in each coral species, based on 97% within-sample clustering.
+5B) symbionts in each host species, based on 97% within-sample clustering.
 (Networks for individuals within each species are presented in Supplementary Materials).
 In the abundant symbiont network (Fig.
-4), `r sum(V(abunet)$type==2)` distinct OTUs were observed.
+4), `r sum(V(abunet)$type==2)` OTUs were observed.
 *D.
-strigosa* associated with the greatest number of distinct OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F.
+strigosa* associated with the greatest number of OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F.
 fragum* (n=`r abu.Nedges["fragum"]`), *M.
@@
 radians*.
-This species was heavily dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.
+This species was dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.
 
-  A total of `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs occurred as "dominant" symbionts, meaning they were present at >50% relative abundance within a sample (Fig.
+  `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs occurred as "dominant" symbionts in the corals sampled, defined by >50% relative abundance within a host (Fig.
 5A).
-A single *Symbiodinium* B1 OTU was the most prevalent dominant symbiont, dominating all *D.
+A single OTU identical to *Symbiodinium* B1 was the most prevalent dominant symbiont, dominating all *D.
 cylindrus* samples and many *D.
-strigosa*, *O.
-annularis*, *M.
-alcicornis*, and *F.
-fragum* samples.
-Other closely related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *D.
+strigosa* (4/9), *O.
+annularis* (2/4), *M.
+alcicornis* (6/10), and *F.
+fragum* (8/9) samples.
+Other closely-related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *D.
 strigosa*, *M.
@@
 cavernosa* and most *S.
-siderea* samples, and *Symbiodinium* A4, which dominated all *P.
-astreoides* and most *P.
+siderea* samples (7/10), and *Symbiodinium* A4, which dominated all *P.
+astreoides* and 3/10 *P.
 furcata* samples.
@@
 5B).
-*Symbiodinium* D1a/*S.
-trenchii* was detected in the highest number of species (n=`r nspp["D1a"]`), followed by A4 (n=`r nspp["A4"]`) and C3 (n=`r nspp["C3"]`).
-C31, C46, and B1 were each detected at background levels in 4 different species.
+*Symbiodinium* D1a (which has been formally described as *S.
+trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1a"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa).
+C31, C46, and B1 were each detected at background levels in 4 host species.
 
@@
 
-#### Symbiotic flexibility of coral species
+#### Symbiotic flexibility of host species
 
-  Significant differences in symbiotic flexibility (i.e.
-multivariate dispersion) were detected among coral species.
-The highest flexibility was found in *D.
+  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals (i.e., average distance to centroid) of each host species was calculated as a metric of symbiosis 'flexibility'; more flexible host species will show greater variability in symbiont community composition among individuals, while less flexible host species will show greater uniformity.
+Significant differences in symbiotic flexibility were detected among host species (Fig.
+6).
+The highest distance to centroid was found in *D.
 strigosa*, *P.
@@
 
-  The perceived structure of corals' *Symbiodinium* communities was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data; specifically, the method of clustering sequence diversity into operational taxonomic units (OTUs).
+The observed structure of *Symbiodinium* communities found in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data.
+Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs.
 Clustering sequences from all samples together at 97% similarity reduced the total number of OTUs by 98% relative to no clustering (Table 1).
 Intermediate diversity was resolved by within-sample clustering, with 13% more OTUs than across-sample clustering.
-By comparing the qualitative outcomes of these approaches (Fig.
-2), it is apparent that across-sample clustering may assign the same dominant OTU to samples with dissimilar sequence assemblages.
-This results from the fact that an OTU's representative sequence (and basis for taxonomic assignment) is selected as the numerically dominant member; when these member sequences occur across many samples, the most abundant may simply reflect the number and composition of samples in the dataset.
-Consequently, closely related sequences arising from different *Symbiodinium* may be subsumed under the dominant variant that happens to be most well-represented (i.e., most heavily sampled).
-Therefore, while 97% similarity may successfully collapse intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when clustering is applied across a dataset potentially comprised of many closely related types, it may also collapse interspecific variation and underestimate true diversity.
+Comparison of the outcomes of these approaches (Fig.
+2) demonstrates that across-sample clustering assigns the same dominant OTU to samples with dissimilar sequence assemblages.
+This results from the selection of an OTU's most abundant member sequence as its basis for taxonomic assignment: when member sequences occur across many samples (i.e., during across-sample clustering), the most abundant depends strongly on the number of samples of different hosts in the dataset.
+Consequently, closely-related sequences arising from different *Symbiodinium* types may be clustered into the same OTU which is then named according to the sequence variant that is most well-represented (i.e., most heavily sampled) in the dataset.
+Therefore, while clustering based on 97% similarity may successfully collapse intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when clustering is applied across a dataset potentially comprised of many closely related types, it may also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present.
 
-  Meanwhile, clustering within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the size or structure of the sample set.
-The outcomes of within-sample clustering more accurately reflect knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that considerable species diversity exists within clades B and C in the Caribbean that associates strongly with host specificity.
+  Meanwhile, clustering within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the data set.
+The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa.
 Indeed, only within-sample clustering assigned different dominant clade C types to samples of *O.
@@
 siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx].
-While multi-locus approaches clearly provide a superior description of *Symbiodinium* diversity, within-sample clustering of ITS2 metabarcoding data may be a good approximation of diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding).
+Therefore, while multi-locus approaches may provide a superior description of *Symbiodinium* diversity, within-sample clustering of ITS2 metabarcoding data may be a good approximation of diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods exploit the numerically abundant sequence variants as taxonomically informative.
 Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa.
@@
   Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* clone that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met.
-In fact, multiple types co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming variation within a sample is intragenomic.
+Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming variation within a sample is intragenomic.
 However, co-occurring types in many cases are members of different clades, with ITS2 sequences divergent enough to still be resolved separately by 97% clustering.
@@
 
-  When treated with a bioinformatic approach that reflects the complexity of the marker, ITS2 metabarcoding remains a very powerful way to describe *Symbiodinium* communities - it provides a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample, has a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species and scaled up to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
+  When treated with a bioinformatic approach that reflects the complexity of the marker, ITS2 metabarcoding remains a powerful way to describe *Symbiodinium* communities.
+It provides a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample, has a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
 
   In the ten host species sampled at St.
-John USVI, the most prevalent *Symbiodinium* were members of clades B and C, consistent with previous work in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
-Less frequent associations with members of clades A and D were observed, though these types did dominate colonies of certain species (clade A dominated *P.
+John, USVI, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
+Less frequent associations between hosts and members of clades A and D were observed, though these types dominated colonies of some species (clade A dominated *P.
 astreoides* and *P.
@@
 siderea*).
-Finally, clade G was also observed at low relative abundances (<`r round(clademax[5,2], 1)` within samples), which likely would not have been detected without high-throughput sequencing.
+Finally, clade G was also observed at low relative abundances (<`r round(clademax[5,2], 2)` within samples), which likely would not have been detected without high-throughput sequencing.
 
-  When the associations between *Symbiodinium* types and coral species are visualized by network analysis, features of both symbiont and host ecology emerge.
+  When the associations between *Symbiodinium* types and coral species were visualized by network analysis, features of both symbiont and host ecology emerge.
 First, ITS2 type B1 was "abundant" (i.e.
@@
 
-  Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at less than 1% relative abundance in at least three different host species.
-Such a distribution suggests that these symbionts can occupy a range of hosts, but functionally do not outperform other resident symbionts to become dominant under prevailing environmental conditions.
+  Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at <1% relative abundance in at least three different host species.
+Such a distribution suggests that these symbionts can occupy a range of hosts, but do not often become dominant under prevailing environmental conditions.
 Their presence, however, indicates that they may be able to achieve dominance within a colony under certain environmental conditions.
@@
 
-  Focusing on coral hosts, there was a clear difference among species in the number of OTUs that were abundant (Fig.
-4) and dominant (Fig.
+  Focusing on hosts, there was a clear difference among species in the number of OTUs that were abundant (>1% relative abundance, Fig.
+4) and dominant (>50% relative abundance, Fig.
 5A), indicating different levels of variability in the symbiosis ecology of different coral species.
@@
   To analyze potential influence of the environment at a larger spatial scale, we tested whether symbiont community composition of each species differed between the north and south shores.
-While environmental data revealed a significant biophysical differences between these two sides of the island (e.g., wave height, temperature, and chlorophyll A), there were no significant differences in symbiont community composition.
+While environmental data revealed significant biophysical differences between these two sides of the island (e.g., wave height, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled.
 This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S.
@@
   While not providing species-level resolution, ITS2 metabarcoding represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs.
-With proper bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, changes in dominant or background symbiont abundance over environmental gradients of over time, with important implications for understanding coral responses to environmental change.
+With proper bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, changes in dominant or background symbiont abundance over environmental gradients or over time, with important implications for understanding coral responses to environmental change.
 

### 2017-01-04 manuscript/manuscript.Rmd
update map figure

@@
 
-  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and susceptibility to stress of their coral hosts.
+  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-susceptibility of their coral hosts.
 Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research.
-High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic vs.
-interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity.
+High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic versus interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity.
 Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St.
@@
 
-**Be very explicit up front about what is meant by type vs species**
-
   The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
@@
 Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them.
-*Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io].
+*Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io].
 However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2.
-High-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb].
+Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs].
 However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
@@
 antillogorgium*, and potentially other species [@Parkinson:2015gt].
-Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (sensu XXX).
+Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g.
+[@Dover:1986ju]).
 In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx].
 Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker.
-Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis (e.g., GeoSymbio, etc.), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms (XXX), it remains an essential and powerful marker for the *Symbiodinium* system.
+Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level.
+Moreover, due to the high volume of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker for the *Symbiodinium* system.
 Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.
@@
 John, U.S.
-Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications by which such datasets may advance our understanding of *Symbiodinium* metacommunity ecology.
+Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance our understanding of *Symbiodinium* metacommunity ecology.
 In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations.
-Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt].
-Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates species diversity by collapsing species known to be different.
+Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt].
+Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity may underestimate diversity by collapsing species known to be different.
 True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.
@@
 John.
-We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species.
+We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species.
 The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts.
@@
 
-#### Sample collection and processing
+#### Sample collection and environmental conditions
 
@@
 Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.
-Samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB).
-Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.
-
-#### Environmental data
 
@@
 
-#### Bioinformatics
+#### ITS2 sequencing and bioinformatics
+
+  Coral tissue samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB).
+Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.
 
@@
 
-#### Environmental and sequence data overview
+#### Sample collection and environmental data
+
+  Sequence data were successfully generated from 84 coral samples collected across two sites on the north shore and five sites on the south shore of St.
+John (Fig.
+1).
+The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
+After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.
 
@@
 
-```{r Fig1, fig.height=3, fig.width=5, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
+```{r Fig1, fig.height=3, fig.width=6, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
 library(rgdal)
@@
 
-#pdf(file="figures/map.pdf", height=3, width=6)
-layout(mat=matrix(c(1,1,1,2,1,1,1,3,1,1,1,4), nrow=3, byrow=T))
-par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1)
+pdf(file="figures/map.pdf", height=3, width=6)
+layout(mat=matrix(c(1,1,6,2,1,1,5,3,1,1,5,4), nrow=3, byrow=T))
+
+par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
 plot(stjmap, lwd=1, col="gray40",
-     xlim=c(-64.74, -64.68), ylim=c(18.303, 18.370))
+     xlim=c(-64.75, -64.69), ylim=c(18.303, 18.370))
 #box()
@@
   points(lon, lat, pch=21, bg="black")
-  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
+  #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
 })
-points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
-text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
+#points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
+#text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
 NS <- c(-64.711967, 18.368141)
 SS <- c(-64.726840, 18.315465)
-points(NS[1], NS[2], pch="*", cex=2, col="red")
-points(SS[1], SS[2], pch="*", cex=2, col="red")
+points(NS[1], NS[2], pch="N", cex=0.9, bg="gray80", col="black", lwd=1)
+points(SS[1], SS[2], pch="S", cex=0.9, bg="gray40", col="black", lwd=1)
+points(NS[1], NS[2], pch=1, cex=2.2, bg="gray80", col="black", lwd=1)
+points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)
 #dev.off()
@@
 axis(side=1, at=seq(26,29,1))
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
 bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
@@
 axis(side=1, at=seq(0.03,0.09,0.01))
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
 bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
@@
 
-t(nsamp)[2,]
-#dev.off()
+plot(NA, xlim=c(0,1), ylim=c(0.5,2.5), xlab="", ylab="", bty="n", yaxt="n")
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
+#text(par("usr")[1], c(1,2), c("O","O"), cex=1.8, pos=2, offset=0.7, lwd=0.1)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
+
+#### histogram of samples
+nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
+nsam <- dcast(nsam, Species~Shore)
+rownames(nsam) <- nsam$Species
+
+rownames(nsam) <- str_replace_all(rownames(nsam), 
+                c("alcicornis"="M.alc.", "annularis"="O.ann.", "astreoides"="P.ast.","cavernosa"="M.cav.",
+                  "cylindrus"="D.cyl.", "fragum"="F.fra.", "furcata"="P.fur.", "radians"="S.rad.",
+                  "siderea"="S.sid.", "strigosa"="P.str."))
+nsam <- nsam[,2:3]
+par(mar=c(4,0,6,4))
+barplot(t(as.matrix(nsam)), las=2)
+legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1.5, legend=c("N","S"), bty="n", inset=c(0.1,-0.2), xpd=T, y.intersp=0.8, x.intersp=0.5)
+dev.off()
 ```
-
-  Sequence data were successfully generated from 84 coral samples collected across two sites on the north shore and five sites on the south shore of St.
-John (Fig.
-1).
-The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
-After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.
 
@@
 Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU.
-Rectangles are annotated with the closest BLAST hit to the representative sequence of the OTU; those with the same annotation but different fill colors represent unique sequence variants with the same closest BLAST hit."}
+Rectangles are annotated with the taxonomic assignment of the OTU (see Methods)."}
 set.seed(42432)
@@
 
-```{r Table3, cache=T} 
+```{r Table3, cache=T } 
 set.seed(43789) 
@@
 #### Network analyses of *Symbiodinium* meta-community
-```{r networks,  cahce=F, include=FALSE}
+```{r networks,  cache=F, include=FALSE}
 # Create dominant symbionts network
@@
 
-The observed structure of *Symbiodinium* communities found in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data.
+  The observed structure of *Symbiodinium* communities in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data.
 Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs.
-Clustering sequences from all samples together at 97% similarity reduced the total number of OTUs by 98% relative to no clustering (Table 1).
-Intermediate diversity was resolved by within-sample clustering, with 13% more OTUs than across-sample clustering.
-Comparison of the outcomes of these approaches (Fig.
-2) demonstrates that across-sample clustering assigns the same dominant OTU to samples with dissimilar sequence assemblages.
-This results from the selection of an OTU's most abundant member sequence as its basis for taxonomic assignment: when member sequences occur across many samples (i.e., during across-sample clustering), the most abundant depends strongly on the number of samples of different hosts in the dataset.
-Consequently, closely-related sequences arising from different *Symbiodinium* types may be clustered into the same OTU which is then named according to the sequence variant that is most well-represented (i.e., most heavily sampled) in the dataset.
-Therefore, while clustering based on 97% similarity may successfully collapse intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when clustering is applied across a dataset potentially comprised of many closely related types, it may also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present.
+Clustering sequences from all samples together reduced the total number of OTUs by 98% relative to no clustering (Table 1), while within-sample clustering resolved intermediate diversity, with 13% more OTUs than across-sample clustering.
+Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering in many cases assigns the same dominant OTU to samples with different dominant sequence variants (Fig.
+2), while within-sample clustering assigns different OTUs that reflect the dominant variant within that particular sample.
+This occurs because an OTU's assigned identity is based on its most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented.
+The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, despite the fact that these other samples may not contain the named sequences.
+For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species.
+Therefore, even though clustering at 97% similarity successfully collapses intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related types, it is likely to also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present.
+Indeed, this limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples: despite being comprised mostly of C41 variants, ~3k sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7k sequences from *Pocillopora verrucosa* with which they were clustered.
+These cases clearly demonstrate how clustering sequences across samples causes the identity of OTUs in any given sample to be driven largely by the other samples in the dataset.
 
-  Meanwhile, clustering within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the data set.
+  Meanwhile, clustering sequences independently within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset.
 The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa.
-Indeed, only within-sample clustering assigned different dominant clade C types to samples of *O.
+Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O.
 annularis* and *S.
 siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx].
-Therefore, while multi-locus approaches may provide a superior description of *Symbiodinium* diversity, within-sample clustering of ITS2 metabarcoding data may be a good approximation of diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods exploit the numerically abundant sequence variants as taxonomically informative.
+Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically abundant sequence variants to assign taxonomy.
 Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa.
@@
 Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming variation within a sample is intragenomic.
-However, co-occurring types in many cases are members of different clades, with ITS2 sequences divergent enough to still be resolved separately by 97% clustering.
-Thus, only when very closely related types co-occur [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail, requiring population genetic markers to parse interspecific variation.
-Other problematic cases may occur when multiple ITS2 sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that a slight numerical dominance of one or the other may lead to different OTU assigments in samples with very similar sequence assemblages (e.g., Fig.
+However, co-occurring types in many cases are members of different clades, whose ITS2 sequences have diverged enough to be resolved separately by 97% clustering.
+Thus, only when very closely related types co-occur in a sample [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail, requiring population genetic markers to parse interspecific variation.
+Other problematic cases may occur when multiple intragenomic sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that slight differences in relative abundance may lead to different OTUs being assigned to samples with otherwise highly similar sequence assemblages (e.g., Fig.
 SX).
@@
 
-  While within-sample clustering of ITS2 data cannot provide species-level descriptions of *Symbiodinium* communities (since the ITS2 marker itself does not provide this resolution), this nuanced bioinformatic approach nevertheless accommodates known complexities in the evolution of the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.
+  While within-sample clustering of ITS2 data cannot provide species-level descriptions of *Symbiodinium* communities (since the ITS2 marker itself does not provide this resolution), this nuanced bioinformatic approach nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.
+When treated with an appropriate bioinformatic approach, ITS2 metabarcoding is a powerful tool to describe *Symbiodinium* communities.
+It provides a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample, has a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
 
@@
 
-  When treated with a bioinformatic approach that reflects the complexity of the marker, ITS2 metabarcoding remains a powerful way to describe *Symbiodinium* communities.
-It provides a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample, has a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
+  In the ten host species sampled in St.
+John, USVI, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
+Less frequent associations between hosts and members of clades A and D were observed, though clade A dominated *P.
+astreoides* and some *P.
+furcata* colonies, while clade D dominated one *S.
+siderea* colony.
+Finally, clade G was observed at low relative abundances (<`r round(clademax[5,2], 2)`) in some samples, which likely would not have been detected without high-throughput sequencing.
 
-  In the ten host species sampled at St.
-John, USVI, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
-Less frequent associations between hosts and members of clades A and D were observed, though these types dominated colonies of some species (clade A dominated *P.
-astreoides* and *P.
-furcata*, while clade D dominated *S.
-siderea*).
-Finally, clade G was also observed at low relative abundances (<`r round(clademax[5,2], 2)` within samples), which likely would not have been detected without high-throughput sequencing.
+  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave height, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled.
+This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S.
+siderea* and *P.
+furcata* were dominated more frequently by *Symbiodinium* in clades D and A, respectively, on the south shore, compared to the north shore (Fig.
+2).
+Since these symbionts are typically associated with warmer and more variable temperature regimes, it is counter to expectation that they were more prevalent on the south shore, where temperatures were slightly lower and more stable.
+This pattern may indicate that other factors that we did not measure (or did not measure with high enough spatial resolution) may be more important in structuring these communities.
 
@@
 Interestingly, *S.
-radians* was the only coral species in which none of the abundant symbionts occurred in any other corals, which may reflect the unique ecology of *S.
+radians* was the only coral species in which none of the abundant symbionts occurred in any other host, which may reflect the unique ecology of *S.
 radians* as the only study species that typically forms small, encrusting colonies.
@@
   Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at <1% relative abundance in at least three different host species.
-Such a distribution suggests that these symbionts can occupy a range of hosts, but do not often become dominant under prevailing environmental conditions.
-Their presence, however, indicates that they may be able to achieve dominance within a colony under certain environmental conditions.
+Such a distribution suggests that these symbionts can occupy a range of hosts, but are unlikely to dominate them under prevailing environmental conditions.
+Their presence, however, indicates that they may be able to achieve dominance within a colony under certain conditions.
 Indeed, the most prevalent 'background' symbiont in St.
-John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], in seven out of ten species), a thermotolerant symbiont that has been shown to proliferate within coral hosts during and after thermal stress [@LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790].
-Association with these symbionts may therefore be indicative of a broader fundamental niche space and resilience to environmental change.
+John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], present in seven out of ten host species), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790].
+Association with these symbionts may therefore broaden a host's fundamental niche space and increase its resilience to environmental change.
 *Symbiodinium* A4 and C3 also commonly occurred as background symbionts, suggesting they may perform analogous functional and ecological roles within the metacommunity.
@@
 
-  To analyze potential influence of the environment at a larger spatial scale, we tested whether symbiont community composition of each species differed between the north and south shores.
-While environmental data revealed significant biophysical differences between these two sides of the island (e.g., wave height, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled.
-This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S.
-siderea* and *P.
-furcata* tended to be dominated more frequently by *Symbiodinium* in clades D and A, respectively, on the south shore, compared to the north shore (Fig.
-2).
-Since these symbionts are typically associated with higher temperature environments, it is counter to expectation that they were more prevalent on the south shore, which was characterized by slightly lower and more stable temperatures.
-This pattern may indicate that other factors that we did not measure (or did not measure with high enough spatial resolution) may be more important in structuring these communities.
-
 #### Conclusions

### 2017-01-05 manuscript/manuscript.Rmd
revision 1

@@
 Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St.
-John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses variability that is more likely to be intragenomic, while maintaining variability that is more likely interspecific.
+John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while maintaining diversity that is more likely interspecific.
 While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis.
@@
 
-  A total of 87 tissue samples from scleractinian and milleporine corals were collected from seven sites around St.
+  A total of 84 tissue samples from scleractinian and milleporine corals were collected from seven sites around St.
 John, US Virgin Islands, USA (Fig.
@@
 
-  Broad-scale differences in the hydrographic regimes of the north and south shores of St.
-John were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave regime.
-These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [xxx] and wave regime [xxx]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition (xxx).
-SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite [xxx].
-From the 12 monthly climatological values obtained for each sampling year (2003-2010, n=96 months), the median, mean, standard deviation and minimum and maximum monthly mean were calculated.
+  Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St.
+John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave exposure (J m^-3^).
+These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition [@Houlbreque:2009gh].
+SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite for each month of the year, averaged over the period 2003-2010.
+We used as input the monthly climatologies available at the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates, values from the nearest available pixel were used.
+Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig.
+1).
 
@@
 2012) for 32 fetch directions and coastline data at full resolution.
-Mean wave exposure (J m^-3^) at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].
+Mean wave exposure at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].
 
@@
 To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network.
-All network layouts were constructed based on the Fruchterman-Reingold algorithm (XXX).
+All network layouts were constructed based on the Fruchterman-Reingold algorithm [@Fruchterman:1991ko].
 
-  Beta diversity (sensu XXX) was calculated as the multivariate dispersion of samples within a coral species.
+  Beta diversity (sensu [@Anderson:2006fs]) was calculated as the multivariate dispersion of samples within a coral species.
 Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test.
@@
 
-  Sequence data were successfully generated from 84 coral samples collected across two sites on the north shore and five sites on the south shore of St.
+  A total of 84 coral samples representing ten different host species were collected across two sites on the north shore and five sites on the south shore of St.
 John (Fig.
 1).
-The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
+The sequence data generated from these samples comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
 After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.
@@
 From 2003 to 2010, the north shore of St.
-John was characterized by lower wave exposure, slightly higher and more variable sea surface temperatures, and higher and more variable chlorophyll a concentrations, relative to the south shore.
-(Statistical tests?)
+John was characterized by slightly lower sea surface temperatures, higher chlorophyll a concentrations, and lower wave exposure, relative to the south shore.
 
-```{r Fig1, fig.height=3, fig.width=6, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
+```{r Fig1, fig.height=3, fig.width=6.85, fig.cap="Figure 1.
+Sampling of *Symbiodinium* communities from ten different host species (embedded histogram) on the north and south shores of St.
+John, US Virgin Islands, and the broad-scale physical conditions associated with these two environments.
+Mapped points along each shore indicate locations from which biological samples were collected, while the 'N' and 'S' points indicate locations from which environmental data representative of the north and south shores were obtained.
+Boxplots show the sea-surface temperatures (top), chlorophyll A concentrations (middle), and wave exposure (bottom) recorded at these two locations."}
 library(rgdal)
@@
 
-pdf(file="figures/map.pdf", height=3, width=6)
+#pdf(file="figures/map.pdf", height=3, width=6.85)
 layout(mat=matrix(c(1,1,6,2,1,1,5,3,1,1,5,4), nrow=3, byrow=T))
 
+# PLOT MAP
 par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
-plot(stjmap, lwd=1, col="gray40",
-     xlim=c(-64.75, -64.69), ylim=c(18.303, 18.370))
-#box()
-text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
-axis(side=3, at=seq(-64.77, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
-axis(side=3, at=seq(-64.77, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
+plot(stjmap, lwd=1.5, col="black",
+     xlim=c(-64.77, -64.70), ylim=c(18.303, 18.370))
+#text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
+axis(side=3, at=seq(-64.78, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
+axis(side=3, at=seq(-64.78, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
 axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
@@
 with(sitecoords, {
-  points(lon, lat, pch=21, bg="black")
+  points(lon, lat, pch=21, bg="black", cex=0.8)
   #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
 })
-#points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
-#text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
 NS <- c(-64.711967, 18.368141)
@@
 points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)
-#dev.off()
 
-# Plot environmental data
-envdata <- read.csv("data/STJ_envdata.csv", check.names=F, row.names=1)
-sites <- envdata[-1,4:5]
-colnames(sites) <- c("North", "South")
-sites$North <- as.numeric(as.character(sites$North))
-sites$South <- as.numeric(as.character(sites$South))
-sites <- data.frame(t(sites))
-
-#SST
-sst <- sites[, 2:6]
-sst <- within(sst, {
-  `Mean+S.D.` <- Mean.SST + S.D..SST
-  `Mean-S.D.` <- Mean.SST - S.D..SST
-})
-
-par(cex=0.7, mar=c(2,0,1,2), mgp=c(1.5,0.5,0), xpd=NA)
-plot(NA, xlim=c(26,29), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
-axis(side=1, at=seq(26,29,1))
-text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
-points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
-bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
-bxp(list(stats=matrix(as.numeric(sst[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)
-
-#chlA
-chla <- sites[, 7:11]
-chla <- within(chla, {
-  `Mean+S.D.` <- Mean.Chla + S.D..Chla
-  `Mean-S.D.` <- Mean.Chla - S.D..Chla
-})
-
-#par(new=T, mar=c(7,14,10,8), cex=0.7)
-plot(NA, xlim=c(0.03,0.09), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
-axis(side=1, at=seq(0.03,0.09,0.01))
-text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
-points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
-bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
-bxp(list(stats=matrix(as.numeric(chla[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)
-
-plot(NA, xlim=c(0,1), ylim=c(0.5,2.5), xlab="", ylab="", bty="n", yaxt="n")
-text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
-#text(par("usr")[1], c(1,2), c("O","O"), cex=1.8, pos=2, offset=0.7, lwd=0.1)
-points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
-
-#### histogram of samples
+# Plot histogram of sample collections
+par(new=T)
+samdat <- data.frame(sample_data(phy97))
 nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
@@
 rownames(nsam) <- nsam$Species
-
 rownames(nsam) <- str_replace_all(rownames(nsam), 
@@
 nsam <- nsam[,2:3]
-par(mar=c(4,0,6,4))
-barplot(t(as.matrix(nsam)), las=2)
-legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1.5, legend=c("N","S"), bty="n", inset=c(0.1,-0.2), xpd=T, y.intersp=0.8, x.intersp=0.5)
-dev.off()
+par(mar=c(7,5,5,6), mgp=c(0.5,0.2,0))
+barplot(t(as.matrix(nsam)), las=2, ylab="# samples", xpd=NA, cex.axis=0.5, cex.names=0.5, cex.lab=0.5, tcl=0.1)
+legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1, cex=0.5, legend=c("N","S"), bty="n", inset=c(0.12,-0.1), xpd=T, y.intersp=0.8, x.intersp=0.75)
+
+# Plot environmental data
+envdata <- read.csv("data/STJ_envdata2.csv")
+envdata$Site <- factor(envdata$Site, levels=c("S", "N"))
+
+#SST
+par(cex=0.7, mar=c(3,0,1,2), mgp=c(1,0.25,0), xpd=NA, cex.axis=0.7, cex.lab=0.7, tcl=-0.25)
+plot(NA, xlim=c(25,30), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
+axis(side=1, at=seq(25,30,1))
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
+boxplot(envdata$SST ~ envdata$Site, range=0, add=T, axes=F, horizontal=T)
+
+#chlA
+plot(NA, xlim=c(0,0.1), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
+axis(side=1, at=seq(0,0.1,0.01))
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
+boxplot(envdata$chlA ~ envdata$Site, range=0, add=T, axes=F, horizontal=T)
+
+#Wind exposure
+we <- list(N=3.234787362,	S=4.025022708)
+plot(NA, xlim=c(3,4.2), ylim=c(0.5,2.5), xlab="Wave exposure (J m-3)", ylab="", bty="n", yaxt="n")
+points(c(we$N, we$S), c(2,1), pch=19)
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
+
+#dev.off()
 ```
@@
 More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1).
-The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n`.
+The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n` reads per OTU.
 
-```{r Table2}
+```{r Table1}
 # Compute summary statistics 
@@
 
-  A subset of samples was selected for comparative analysis of the three clustering approaches (Fig.
+  A subset of samples was selected for comparative analysis of the community structure resolved by the three clustering approaches (Fig.
 1).
 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples.
-Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), the remainder of the results is presented using the 97% within-sample clustering approach.
+Because the outcomes of the latter approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), the remainder of the results is presented using the 97% within-sample clustering approach.
 Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.
@@
 *Symbiodinium* community composition of each sample.
-Samples are plotted as horizontal bars, sorted by species and shore.
+Samples are plotted as horizontal bars, sorted by species and shore of collection (north vs.
+south).
 Individual segments of each bar represent distinct 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
@@
 
-  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any cnidarian species (Table 3).
+  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any host species (Table 2).
 However, some qualitative differences were apparent: *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056).
@@
 
-```{r Table3, cache=T } 
+```{r Table2, cache=T } 
 set.seed(43789)  
@@
 knitr::kable(shorestats97bs, digits=3, row.names=F, 
-             caption="Table 3. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")
+             caption="Table 2. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")
 
@@
 
-#### Network analyses of *Symbiodinium* meta-community
+#### Network analyses of *Symbiodinium* metacommunity
 ```{r networks,  cache=F, include=FALSE}
@@
 
-  `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs occurred as "dominant" symbionts in the corals sampled, defined by >50% relative abundance within a host (Fig.
+  The dominant symbionts network contained `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs, defined as those occurring at >50% relative abundance within a host (Fig.
 5A).
@@
 
-  There were `r sum(V(bkgnet3)$type==2)` symbionts that occurred at background levels (<1% relative abundance) in at least three coral species (Fig.
+  There were `r sum(V(bkgnet3)$type==2)` symbionts that occupied a 'background' niche, defined as <1% relative abundance in at least three coral species (Fig.
 5B).
@@
 
-  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals (i.e., average distance to centroid) of each host species was calculated as a metric of symbiosis 'flexibility'; more flexible host species will show greater variability in symbiont community composition among individuals, while less flexible host species will show greater uniformity.
-Significant differences in symbiotic flexibility were detected among host species (Fig.
+  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals within each host species (i.e., average distance to centroid) was calculated as a metric of symbiosis 'flexibility'; more flexible host species will show greater variability in symbiont community composition among individuals, while less flexible host species will show greater uniformity.
+This metric revealed significant differences in symbiotic flexibility among host species (Fig.
 6).
@@
 This occurs because an OTU's assigned identity is based on its most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented.
-The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, despite the fact that these other samples may not contain the named sequences.
+The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, even if these other samples do not actually contain the named sequences.
 For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species.
@@
 
-  Meanwhile, clustering sequences independently within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset.
+  Meanwhile, clustering sequences within each sample independently assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset.
 The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa.
@@
 Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa.
-Therefore, we recommend within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.
+Therefore, we recommend a within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.
 
   Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* clone that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met.
-Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming variation within a sample is intragenomic.
-However, co-occurring types in many cases are members of different clades, whose ITS2 sequences have diverged enough to be resolved separately by 97% clustering.
-Thus, only when very closely related types co-occur in a sample [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail, requiring population genetic markers to parse interspecific variation.
-Other problematic cases may occur when multiple intragenomic sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that slight differences in relative abundance may lead to different OTUs being assigned to samples with otherwise highly similar sequence assemblages (e.g., Fig.
+Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming that variation within a sample is intragenomic.
+However, co-occurring types in many cases are members of different clades, whose ITS2 sequences are divergent enough to be resolved separately by 97% clustering.
+Thus, only when very closely related types co-occur in a sample [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail.
+Other problematic cases may occur when multiple intragenomic sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that slight differences in their relative abundance may lead to different OTU assignments in samples with otherwise highly similar sequence assemblages (e.g., Fig.
 SX).
-These challenges may be overcome by 1) analyzing read abundance correlations across samples to inform OTU assignments, or 2) using alternative criteria for taxonomic assignment, such as the most evolutionarily derived sequence variant.
+These challenges may be overcome by developing criteria for taxonomic assignment that are more complex than choosing the single most abundant variant, for example, by taking into account the presence and abundance of multiple variants and/or the phylogenetic relationships among sequences (e.g.
+to use more evolutionarily derived variants for taxonomic assignment).
 
   While within-sample clustering of ITS2 data cannot provide species-level descriptions of *Symbiodinium* communities (since the ITS2 marker itself does not provide this resolution), this nuanced bioinformatic approach nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.
-When treated with an appropriate bioinformatic approach, ITS2 metabarcoding is a powerful tool to describe *Symbiodinium* communities.
-It provides a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample, has a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
+When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample with a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
 
@@
 
-  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave height, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled.
+  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave exposure, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled.
 This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S.
@@
 2).
-Since these symbionts are typically associated with warmer and more variable temperature regimes, it is counter to expectation that they were more prevalent on the south shore, where temperatures were slightly lower and more stable.
-This pattern may indicate that other factors that we did not measure (or did not measure with high enough spatial resolution) may be more important in structuring these communities.
+Since these symbionts are typically associated with warmer and more variable temperature regimes [e.g., @Oliver:2011cr] and high-light habitats [@Byler], it is consistent with expectation that they were more prevalent on the south shore, where temperatures were slightly higher and chlorophyll levels were lower, suggesting light penetration is likely greater.
+However, the lack of differentiation in the symbiont communities of most host species suggests their structure is not sensitive to the level of environmental variation between shores.
 
@@
 However, analysis of more variable chloroplast and microsatellite loci has revealed that ITS2 type B1 in the Caribbean is comprised of numerous specialized lineages that demonstrate specific associations with different coral host species [@Santos:2004p4044; @Finney:2010p8324; @Parkinson:2015gt].
-Likewise, the apparent generalist *Symbiodinium* C3 (which dominated colonies of both *S.
+Likewise, the apparent generalists *Symbiodinium* C3 (which dominated colonies of both *S.
 siderea* and *M.

### 2017-01-05 manuscript/manuscript.Rmd
prepare draft to share

@@
 Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering in many cases assigns the same dominant OTU to samples with different dominant sequence variants (Fig.
-2), while within-sample clustering assigns different OTUs that reflect the dominant variant within that particular sample.
-This occurs because an OTU's assigned identity is based on its most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented.
+2), while within-sample clustering assigns them distinct OTUs that reflect the dominant variant in the sample.
+This occurs because OTU identity is based on the most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented.
 The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, even if these other samples do not actually contain the named sequences.
 For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species.
+
   Therefore, even though clustering at 97% similarity successfully collapses intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related types, it is likely to also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present.
@@
   Meanwhile, clustering sequences within each sample independently assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset.
-The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa.
+The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects patterns of *Symbiodinium* diversity and ecology that have been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa.
 Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O.
@@
 siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx].
-Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically abundant sequence variants to assign taxonomy.
+Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically dominant sequence variants to assign taxonomy.
 Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa.
@@
 SX).
-These challenges may be overcome by developing criteria for taxonomic assignment that are more complex than choosing the single most abundant variant, for example, by taking into account the presence and abundance of multiple variants and/or the phylogenetic relationships among sequences (e.g.
-to use more evolutionarily derived variants for taxonomic assignment).
+These challenges may be overcome by basing taxonomic assignment on more complex criteria than just the most abundant variant, such as multiple co-dominant variants, or more evolutionarily derived variants.
 
-  While within-sample clustering of ITS2 data cannot provide species-level descriptions of *Symbiodinium* communities (since the ITS2 marker itself does not provide this resolution), this nuanced bioinformatic approach nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.
-When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample with a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
+  While no treatment of ITS2 data can provide species-level descriptions of *Symbiodinium* communities (since the marker itself does not provide this resolution), within-sample clustering nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.
+When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative picture of *Symbiodinium* diversity, and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
 
@@
 
-  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave exposure, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled.
+  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave exposure, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the host species sampled on either shore.
 This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S.
@@
 2).
-Since these symbionts are typically associated with warmer and more variable temperature regimes [e.g., @Oliver:2011cr] and high-light habitats [@Byler], it is consistent with expectation that they were more prevalent on the south shore, where temperatures were slightly higher and chlorophyll levels were lower, suggesting light penetration is likely greater.
-However, the lack of differentiation in the symbiont communities of most host species suggests their structure is not sensitive to the level of environmental variation between shores.
+Since these symbionts are typically associated with warmer and more variable temperature regimes [e.g., @Baker:2004p21; @Oliver:2011cr] and high-light habitats [@LaJeunesse:2002p3917], it is consistent with expectations that they were more prevalent on the south shore, where temperatures were slightly higher, and chlorophyll levels were lower, suggesting greater light penetration.
+However, the lack of differentiation in the symbiont communities of most host species suggests their structure is not sensitive to environmental differences between shores.
 
@@
 First, ITS2 type B1 was "abundant" (i.e.
->1%) in individuals of seven out of ten coral species, and "dominant" (i.e.
+>1%) in seven out of ten coral species, and "dominant" (i.e.
 >50%) in five out of ten, suggesting it is a more generalist symbiont lineage.
@@
 astreoides* and *P.
-furcata*) may also be comprised of multiple specialized lineages [@Thornhill:2013dx], which will require evaluation with population genetic markers.
+furcata*) may also be comprised of multiple specialized lineages [@Thornhill:2013dx], which will require evaluation with higher resolution genetic markers.
 
@@
 Such a distribution suggests that these symbionts can occupy a range of hosts, but are unlikely to dominate them under prevailing environmental conditions.
-Their presence, however, indicates that they may be able to achieve dominance within a colony under certain conditions.
+Their presence, however, indicates that they may be able to achieve dominance under certain conditions.
 Indeed, the most prevalent 'background' symbiont in St.
 John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], present in seven out of ten host species), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790].
-Association with these symbionts may therefore broaden a host's fundamental niche space and increase its resilience to environmental change.
+Association with these symbionts may therefore broaden a host's fundamental niche space and increase its resilience to environmental change [@Correa:2010p7787].
 *Symbiodinium* A4 and C3 also commonly occurred as background symbionts, suggesting they may perform analogous functional and ecological roles within the metacommunity.
@@
 *D.
-strigosa* individuals were dominated by four different OTUs, while *D.
+strigosa* individuals could be dominated by four different OTUs, while all *D.
 cylindrus*, *S.
@@
 cavernosa*, and *P.
-astreoides* were dominated by a single OTU.
-In fact, metabarcoding data enables quantifying symbiosis flexibility as multivariate dispersion (a metric of beta diversity), which can be statistically compared among species [@Anderson:2006fs].
+astreoides* individuals were dominated by a single OTU.
+In fact, this variability, or 'flexibility' in the symbiosis ecology of a host species, can be quantified as multivariate dispersion (a metric of beta diversity) using metabarcoding data, which can be statistically compared among species [@Anderson:2006fs].
 The flexibility of symbiosis is important as a potential indicator of the environmental sensitivity of a species [@Putnam:2012wv], and its ability to switch or shuffle its symbionts.
@@
   While not providing species-level resolution, ITS2 metabarcoding represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs.
-With proper bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, changes in dominant or background symbiont abundance over environmental gradients or over time, with important implications for understanding coral responses to environmental change.
+With appropriate bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, and changes in dominant or background symbiont abundance over environmental gradients or over time, with important implications for understanding coral responses to environmental change.
 

### 2017-01-05 R/functions.R
remove bd column from perms

@@ -231,9 +231,9 @@ perms <- function(phy, group, trt) {
   groups <- data.frame(sample_data(phy))[, group]
   # Create data frame to populate with results
   resdf <- data.frame(
-    matrix(ncol=8, nrow=10,
+    matrix(ncol=7, nrow=10,
            dimnames=list(levels(data.frame(sample_data(phy))[, group]), 
-                         c(group, "n", "overall", "within", "between", "R2", "bd", "p")))
+                         c(group, "n", "overall", "within", "between", "R2", "p")))
   )
   # Compute within- and between-treatment statistics for each group
   for (i in 1:nlevels(groups)) {
@@ -255,12 +255,12 @@ perms <- function(phy, group, trt) {
       permanova <- adonis(braydist ~ get(trt), data=samdat, permutations=99999)
       resdf$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
       resdf$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
-      bd <- betadisper(braydist, group=samdat[, trt])
-      resdf$bd[i] <- TukeyHSD(bd)$group[4]
+      #bd <- betadisper(braydist, group=samdat[, trt])
+      #resdf$bd[i] <- TukeyHSD(bd)$group[4]
     } else {
       resdf$R2[i] <- NA
       resdf$p[i] <- NA
-      resdf$bd[i] <- NA
+      #resdf$bd[i] <- NA
     }
   }
   return(resdf)

### 2017-02-03 R/functions.R
change Dstr to Pstr

@@ -163,7 +163,7 @@ otubarplot2 <- function(samples) {
        xpd=NA, adj=c(0,0))
   # Add species names
   spnames <- as.character(sample_data(phy97)$Species[match(sord, rownames(sample_data(phy97)))])
-  fspnames <- c("D. strigosa", "M. alcicornis", "P. furcata", "O. annularis", "S. siderea", "F. fragum", "S. radians", "P. astreoides", "D. cylindrus", "M. cavernosa")
+  fspnames <- c("P. strigosa", "M. alcicornis", "P. furcata", "O. annularis", "S. siderea", "F. fragum", "S. radians", "P. astreoides", "D. cylindrus", "M. cavernosa")
   spnames <- sapply(spnames, function(x) str_subset(fspnames, x))
   newsp <- c(1, which(diff(as.numeric(factor(spnames)))!=0)+1, nsamples(phy100)+1)
   i <- 1

### 2017-02-03 R/functions.R
formatting tweaks

@@ -54,7 +54,7 @@ composition <- function(phy, col, legend=T) {
   # Add lines to separate species and species names
   for (i in 1:length(spbreaks)) {
     lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
-    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
+    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6, font = 3,
          labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                       samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
   }
@@ -277,9 +277,9 @@ phystats <- function(phy) {
       'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
       'Number of samples'=nsamples(phy),
       'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
-      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
+      'Arithmetic mean (± SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
                                                      as.integer(sd(sample_sums(phy))), sep=" ± "),
-      'Geometric mean (*/SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
+      'Geometric mean (*/ GSD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                                                     as.integer(exp(sd(log(sample_sums(phy))))), sep=" */ "),
       check.names=F)
   )

### 2017-02-03 manuscript/manuscript.Rmd
Pete round 2 revisions

@@
 
-  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-susceptibility of their coral hosts.
-Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research.
-High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic versus interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity.
-Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St.
-John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while maintaining diversity that is more likely interspecific.
-While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis.
-We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St.
-John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs.
-'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments.
+  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-tolerance of their coral hosts, making the description of *Symbiodinium* communities in corals (and metacommunities on reefs) a routine and essential goal in coral reef research.
+High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented scale, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity.
+Using samples of *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St.
+John, US Virgin Islands, we demonstrate that generating operational taxonomic units by clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific.
+We utilized this ‘within-sample clustering’ approach to create metacommunity networks of host-symbiont associations in St.
+John, which revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations with *Symbiodinium*, although no differences in symbiont communities were observed between north and south shore environments.
 These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.
@@
 
-  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
-Even slight differences in the relative abundance of different symbionts within a coral may have important functional consequences [@Loram:2007p3688; @Cunning:2015ja].
-Moreover, plasticity and flexibility of these communities within species and individuals over time is an important trait modulating the sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond adaptively by 'shuffling' or 'switching' symbionts [@Baker:2003p200].
-Therefore, understanding the ecological performance of corals depends on accurately and comprehensively describing their *Symbiodinium* communities.
+  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building scleractinian and milleporine corals plays a major role in their biology and ecology, as the identity and functional performance of the symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
+Even slight differences in the relative abundance of different symbionts may have important functional consequences for the host [@Loram:2007p3688; @Cunning:2015ja].
+Moreover, variation in these communities among individuals within a host species, and within individuals over time, is an important trait modulating sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond in beneficial ways to changing environmental conditions by 'shuffling' or 'switching' symbionts [@Baker:2003p200].
+Therefore, understanding the ecological performance of corals will greatly benefit from accurate and comprehensive descriptions of their *Symbiodinium* communities.
 
-  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we].
-Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them.
+  *Symbiodinium* identity is primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we].
+Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them.
 *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io].
-However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2.
-Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs].
-However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
+However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2.
+Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs].
+However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
 
-  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within this genus [@Finney:2010p8324; @Thornhill:2013dx].
+  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within the genus [@Finney:2010p8324; @Thornhill:2013dx].
 For example, the 'B1' ITS2 sequence variant is shared by *S.
@@
 antillogorgium*, and potentially other species [@Parkinson:2015gt].
-Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g.
+Moreover, the position of ITS2 within the tandemly-repeating ribosomal DNA array creates multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g.
 [@Dover:1986ju]).
-In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx].
-Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker.
-Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level.
-Moreover, due to the high volume of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker for the *Symbiodinium* system.
-Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.
+In fact, concerted evolution may mask species divergence within *Symbiodinium* by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx].
+Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker for *Symbiodinium*.
+Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level.
+Moreover, due to the large quantity of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2, it remains an essential and powerful marker for the *Symbiodinium* system.
+Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys of *Symbiodinium*.
 
-  Here, we describe a *Symbiodinium* metacommunity with scleractinians and a *Millepora* hydrocoral in St.
+  Here, we describe a *Symbiodinium* metacommunity associated with scleractinians and a *Millepora* hydrocoral in St.
 John, U.S.
-Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance our understanding of *Symbiodinium* metacommunity ecology.
-In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations.
-Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt].
-Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity may underestimate diversity by collapsing species known to be different.
-True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.
+Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance understanding of *Symbiodinium* metacommunity ecology.
+In addressing the first objective, we test the ability of 'within-sample clustering' (i.e.
+clustering sequences at 97% similarity within each sample independently) to generate biologically relevant operational taxonomic units (OTUs) from ITS2 metabarcoding data.
+Specifically, this approach addresses the fact that dominant ITS2 sequences from different *Symbiodinium* species may be more similar to one another than intragenomic variants within one *Symbiodinium* [@Thornhill:2013dx; @Arif:2014gk; @Parkinson:2015gt].
+Therefore, clustering all sequences together underestimates diversity by collapsing different species into the same OTU, while treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to intragenomic variation.
+Within-sample clustering may better approximate true diversity by exploiting key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.
 These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422].
-These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to represent a single taxon rather than multiple, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa.
+These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to be intragenomic variants, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa.
 Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).
 
-  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St.
+  Using within-sample clustering, we analyzed *Symbiodinium* communities in ten host species across the north and south shores of St.
 John.
 We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species.
-The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts.
-More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.
+The large quantity of data, and the kinds of analytical approaches facilitated by ITS2 metabarcoding, have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts.
+More generally, the present study demonstrates how ecological knowledge can inform bioinformatic analyses using marker genes for which sequence diversity does not map directly to species.
 
@@
 
-  A total of 84 tissue samples from scleractinian and milleporine corals were collected from seven sites around St.
+  Eighty-four tissue samples were collected from scleractinian and milleporine corals sampled at seven sites around St.
 John, US Virgin Islands, USA (Fig.
 1), between August 7th and 9th, 2012.
-Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth.
-Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.
+Samples were collected on snorkel from 3-7m depth, and all colonies to be sampled were selected haphazardly in upward-facing locations exposed to ambient irradiance.
+Tissue biopsies (~5-mm diameter and ~5-mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.
 
   Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St.
-John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave exposure (J m^-3^).
-These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition [@Houlbreque:2009gh].
-SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite for each month of the year, averaged over the period 2003-2010.
-We used as input the monthly climatologies available at the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates, values from the nearest available pixel were used.
+John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and integrated wave exposure on a power scale (J m^-3^).
+These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and conditions known to affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and mediate the supply of food resources that play important roles in coral nutrition [@Houlbreque:2009gh].
+SST and chlorophyll were determined using data from the MODIS sensor on the Aqua satellite.
+SST was evaluated from nighttime records, and chlorophyll from ocean colour, both of which were obtained at 1 km spatial resolution for each month, averaged over the years 2003-2010, from the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates (above), values from the nearest available pixel were used.
 Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig.
@@
 
-  Wind-driven wave exposure for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves.
-To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz].
+  Wind-driven wave exposure on a power scale for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch.
+To calculate wave exposure, wind speed and direction at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz].
 Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk].
 From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al.
-2012) for 32 fetch directions and coastline data at full resolution.
-Mean wave exposure at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].
+2012) for 32 fetch directions (equally distributed through 360°) and coastline data at full resolution (xx km).
+Total wave exposure (summed over all directions) was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].
 
@@
 Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript.
-If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".
+If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "*Symbiodinium*".
 
@@
 Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test.
-This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].
+This analysis was implemented using the betadisper function in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].
 
@@
 1).
-The sequence data generated from these samples comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers.
+The sequence data generated from these samples comprised 1,490,813 sequences after merging paired reads, removing chimeric sequences, and trimming primers.
 After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.
 
-  The environmental conditions broadly characterizing the north and south shores are presented in Figure 1.
+  The environmental conditions broadly characterizing the north and south shores are presented in Fig.
+1.
 From 2003 to 2010, the north shore of St.
@@
 par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
-plot(stjmap, lwd=1.5, col="black",
+plot(stjmap, lwd=1.5, col="gray40",
      xlim=c(-64.77, -64.70), ylim=c(18.303, 18.370))
@@
   points(lon, lat, pch=21, bg="black", cex=0.8)
-  #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
+  text(lon, lat, name, pos=c(4,2,4,4,4,1,3), cex=0.8)
 })
@@
 
-#Wind exposure
+#Wave exposure
 we <- list(N=3.234787362,	S=4.025022708)
-plot(NA, xlim=c(3,4.2), ylim=c(0.5,2.5), xlab="Wave exposure (J m-3)", ylab="", bty="n", yaxt="n")
-points(c(we$N, we$S), c(2,1), pch=19)
-text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
-points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
+barplot(c(we$N, we$S), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
+text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1, xpd=NA)
+points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5, xpd=NA)
 
@@
 
-```{r Table1}
+```{r Table1, cache=F}
 # Compute summary statistics 
@@
 
-#### Symbiodinium community composition
+#### *Symbiodinium* community composition
 
@@
 south).
-Individual segments of each bar represent distinct 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
-par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
+Individual segments of each bar represent 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
+#png(file="crap.png", width = 4, height=6, units="in", res=300)
+par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=1)
 # Plot composition of 97% within-sample OTUs colored by clade
 composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
+#dev.off()
 ```
@@
 shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
+shorestats97bs$Species <- str_replace_all(
+  shorestats97bs$Species,
+  c("alcicornis"="*Millepora alcicornis*",
+    "annularis"="*Orbicella annularis*",
+    "astreoides"="*Porites astreoides*",
+    "cavernosa"="*Montastraea cavernosa*",
+    "cylindrus"="*Dendrogyra cylindrus*",
+    "fragum"="*Favia fragum*",
+    "furcata"="*Porites furcata*",
+    "radians"="*Siderastrea radians*",
+    "siderea"="*Siderastrea siderea*",
+    "strigosa"="*Pseudodiplora strigosa*")
+)
 knitr::kable(shorestats97bs, digits=3, row.names=F, 
@@
 (Networks for individuals within each species are presented in Supplementary Materials).
-In the abundant symbiont network (Fig.
+In the network for abundant symbionts (Fig.
 4), `r sum(V(abunet)$type==2)` OTUs were observed.
-*D.
+*P.
 strigosa* associated with the greatest number of OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F.
@@
 
-  The dominant symbionts network contained `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs, defined as those occurring at >50% relative abundance within a host (Fig.
+  In the network for dominant symbionts, `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs were included, defined as those occurring at >50% relative abundance within a host (Fig.
 5A).
 A single OTU identical to *Symbiodinium* B1 was the most prevalent dominant symbiont, dominating all *D.
-cylindrus* samples and many *D.
-strigosa* (4/9), *O.
-annularis* (2/4), *M.
-alcicornis* (6/10), and *F.
-fragum* (8/9) samples.
-Other closely-related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *D.
+cylindrus* samples and many *P.
+strigosa* (4 of 9), *O.
+annularis* (2 of 4), *M.
+alcicornis* (6 of 10), and *F.
+fragum* (8 of 9) samples.
+Other closely-related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *P.
 strigosa*, *M.
@@
 cavernosa* and most *S.
-siderea* samples (7/10), and *Symbiodinium* A4, which dominated all *P.
-astreoides* and 3/10 *P.
-furcata* samples.
+siderea* (7 of 10), and *Symbiodinium* A4, which dominated all *P.
+astreoides* and 3 of 10 *P.
+furcata*.
 *P.
@@
 
-  There were `r sum(V(bkgnet3)$type==2)` symbionts that occupied a 'background' niche, defined as <1% relative abundance in at least three coral species (Fig.
+  There were `r sum(V(bkgnet3)$type==2)` symbionts that occupied a 'background' niche, defined as <1% relative abundance in three or more coral species (Fig.
 5B).
-*Symbiodinium* D1a (which has been formally described as *S.
-trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1a"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa).
+A member of *Symbiodinium* clade D (which has been formally described as *S.
+trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa).
 C31, C46, and B1 were each detected at background levels in 4 host species.
@@
 Network analysis of abundant *Symbiodinium* taxa in each coral species.
-*Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at greater than 1% relative abundance within an individual; thickness of edges is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%.
-Symbiont nodes are colored by clade identity, sized proportionally to the number of coral species they dominated."}
+*Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at >1% relative abundance within an individual; thickness of edges (i.e., links between coral and symbiont nodes) is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%.
+Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species they dominated."}
+
 # Plot abundant symbionts network
@@
 
-  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals within each host species (i.e., average distance to centroid) was calculated as a metric of symbiosis 'flexibility'; more flexible host species will show greater variability in symbiont community composition among individuals, while less flexible host species will show greater uniformity.
+  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals within each host species (i.e., average distance to centroid) was calculated as a metric of symbiosis 'flexibility', with higher values indicating hosts with greater variability in symbiont community composition among individuals, and lower values indicating hosts with greater uniformity.
 This metric revealed significant differences in symbiotic flexibility among host species (Fig.
 6).
-The highest distance to centroid was found in *D.
+The highest distance to centroid was found in *P.
 strigosa*, *P.
@@
 
-```{r Fig6, cache=F, fig.width=3.5, fig.height=3.5, fig.cap="Figure 6. Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition."}
+```{r Fig6, cache=F, fig.width=3.5, fig.height=3.5, fig.cap="Figure 6.
+Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition (mean distance to centroid ± SE.
+Host taxa that do not share a letter are significantly different (p < 0.05)."}
 betad97bs <- betad(phy97bs.f.t, group="Species")
 
-par(mar=c(3,3,1,1), mgp=c(1.8,0.1,0), tcl=0.25)
+par(mar=c(7,3,1,1), mgp=c(1.8,0.1,0), tcl=0.25)
 with(betad97bs$sambdsumm.ord, {
-  plot(mean, type="n", ylim=c(0, 0.65), ylab="Distance to centroid", xaxt="n", xlab="Coral species")
+  plot(mean, type="n", ylim=c(0, 0.65), ylab="Distance to centroid", xaxt="n", xlab="")
   arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
   points(1:10, mean, cex=1, pch=21, bg="white")
-  axis(side=1, at=1:10, cex.axis=0.75,
+  axis(side=1, at=1:10, cex.axis=0.75, labels=NA)
+  text(1.4:10.4, -0.04, srt=45, xpd=T, pos=2, font=3,
        labels=str_replace_all(levels(sam$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)],
-                              c("alcicornis"="Ma", "annularis"="Oa",
-                                "astreoides"="Pa", "cavernosa"="Mc",
-                                "cylindrus"="Dc", "fragum"="Ff", 
-                                "furcata"="Pf", "radians"="Sr", 
-                                "siderea"="Ss", "strigosa"="Ds")))
+                              c("alcicornis"="M. alcicornis",
+                                "annularis"="O. annularis",
+                                "astreoides"="P. astreoides",
+                                "cavernosa"="M. cavernosa",
+                                "cylindrus"="D. cylindrus",
+                                "fragum"="F. fragum",
+                                "furcata"="P. furcata",
+                                "radians"="S. radians",
+                                "siderea"="S. siderea",
+                                "strigosa"="P. strigosa")))
   text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
@@
 
-  The observed structure of *Symbiodinium* communities in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data.
-Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs.
-Clustering sequences from all samples together reduced the total number of OTUs by 98% relative to no clustering (Table 1), while within-sample clustering resolved intermediate diversity, with 13% more OTUs than across-sample clustering.
-Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering in many cases assigns the same dominant OTU to samples with different dominant sequence variants (Fig.
+  The observed structure of *Symbiodinium* communities in a variety of host species was impacted by the bioinformatic approach used to analyze ITS2 sequence data.
+Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset, or independently within each sample, affected the number and taxonomic assignment of OTUs.
+Clustering sequences across samples reduced the total number of OTUs by 98% relative to the number of OTUs detected with no clustering (Table 1).
+Within-sample clustering resolved intermediate diversity, with 13% more OTUs than across-sample clustering.
+Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering often assigns the same dominant OTU to samples with different dominant sequence variants (Fig.
 2), while within-sample clustering assigns them distinct OTUs that reflect the dominant variant in the sample.
-This occurs because OTU identity is based on the most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented.
-The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, even if these other samples do not actually contain the named sequences.
-For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species.
+This occurs because OTU identity is determined by the most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it.
+Therefore, a particular sequence variant from a more extensively-sampled and/or deeply-sequenced host may determine the identity of an OTU that includes closely-related but different sequences in other samples, even though they do not actually contain the given sequence.
+Such merging of OTUs across samples is undesirable, since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species [@Sampayo:2009p3422].
 
-  Therefore, even though clustering at 97% similarity successfully collapses intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related types, it is likely to also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present.
-Indeed, this limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples: despite being comprised mostly of C41 variants, ~3k sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7k sequences from *Pocillopora verrucosa* with which they were clustered.
-These cases clearly demonstrate how clustering sequences across samples causes the identity of OTUs in any given sample to be driven largely by the other samples in the dataset.
+  Therefore, even though clustering at 97% similarity collapses intragenomic variation within a single *Symbiodinium* [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related *Symbiodinium*, it is likely to also collapse interspecific variation and underestimate diversity.
+This limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples from multiple hosts: despite being comprised mostly of C41 variants, ~3,000 ITS2 sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7,000 sequences from *Pocillopora verrucosa* with which they were clustered.
+In this case (and as shown in Fig.
+2), clustering sequences across samples leads to the conclusion that samples with different dominant sequence variants are dominated by the same *Symbiodinium* OTU.
+This outcome occurs because the identity of OTUs in any given sample may be determined by other samples in the dataset; indeed, using this approach, the same sample and sequence assemblage may receive different OTU assignments if it were part of a different set of samples.
 
-  Meanwhile, clustering sequences within each sample independently assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset.
-The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects patterns of *Symbiodinium* diversity and ecology that have been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa.
+  Meanwhile, clustering sequences within each sample independently assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples included in the analysis.
+The outcome of within-sample clustering of ITS2 sequences as applied herein better reflects patterns of *Symbiodinium* diversity and ecology that have been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different coral host taxa.
 Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O.
 annularis* and *S.
-siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx].
-Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically dominant sequence variants to assign taxonomy.
-Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa.
+siderea* (Fig.
+2), which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx].
+Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases.
+In theory, this approach should also reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically dominant sequence variants to assign taxonomy.
+Furthermore, metabarcoding overcomes the primary limitations of the DGGE method by providing more quantitative data and highly sensitive detection of low abundance taxa [@Quigley:2014dv].
 Therefore, we recommend a within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.
 
-  Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* clone that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met.
-Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming that variation within a sample is intragenomic.
-However, co-occurring types in many cases are members of different clades, whose ITS2 sequences are divergent enough to be resolved separately by 97% clustering.
+  Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* species that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met.
+Indeed, multiple *Symbiodinium* frequently co-occur in single coral colonies [@Silverstein:2012ub], which undermines support for assuming that variation within a sample is intragenomic.
+However, co-occurring *Symbiodinium* in many cases are members of different clades, whose ITS2 sequences are divergent enough to be resolved separately by 97% clustering.
 Thus, only when very closely related types co-occur in a sample [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail.
@@
 SX).
-These challenges may be overcome by basing taxonomic assignment on more complex criteria than just the most abundant variant, such as multiple co-dominant variants, or more evolutionarily derived variants.
+These challenges may be overcome by basing taxonomic assignment on more complex criteria than just the most abundant sequence variant, such as multiple co-dominant variants, or more evolutionarily derived variants.
 
-  While no treatment of ITS2 data can provide species-level descriptions of *Symbiodinium* communities (since the marker itself does not provide this resolution), within-sample clustering nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all.
-When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative picture of *Symbiodinium* diversity, and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
+  While no treatment of ITS2 data can provide species-level descriptions of *Symbiodinium* communities (since the marker itself does not provide this resolution), within-sample clustering nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better reflect species-level diversity than either across-sample clustering or no clustering at all.
+When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative analysis of *Symbiodinium* diversity, and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.
 
@@
   In the ten host species sampled in St.
-John, USVI, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
-Less frequent associations between hosts and members of clades A and D were observed, though clade A dominated *P.
+John, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
+Less frequent associations between hosts and members of clades A and D were observed, although clade A dominated *P.
 astreoides* and some *P.
-furcata* colonies, while clade D dominated one *S.
+furcata* colonies, and clade D dominated one *S.
 siderea* colony.
-Finally, clade G was observed at low relative abundances (<`r round(clademax[5,2], 2)`) in some samples, which likely would not have been detected without high-throughput sequencing.
+Finally, clade G was observed at low relative abundances (<`r 100*round(clademax[5,2], 2)`%) in some samples, which likely would not have been detected without high-throughput sequencing.
 
-  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave exposure, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the host species sampled on either shore.
-This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S.
+  While the environmental conditions differed between the north and south shores of St.
+John (e.g., wave exposure, temperature, and chlorophyll a), there were no statistically significant differences in symbiont community composition in the corals sampled on either shore.
+This may partly be due to small sample sizes, since two species showed a trend for differences between shores: *S.
 siderea* and *P.
-furcata* were dominated more frequently by *Symbiodinium* in clades D and A, respectively, on the south shore, compared to the north shore (Fig.
+furcata* were more frequently dominated by *Symbiodinium* in clades D and A (respectively) on the south shore, compared to the north shore (Fig.
 2).
-Since these symbionts are typically associated with warmer and more variable temperature regimes [e.g., @Baker:2004p21; @Oliver:2011cr] and high-light habitats [@LaJeunesse:2002p3917], it is consistent with expectations that they were more prevalent on the south shore, where temperatures were slightly higher, and chlorophyll levels were lower, suggesting greater light penetration.
-However, the lack of differentiation in the symbiont communities of most host species suggests their structure is not sensitive to environmental differences between shores.
+*Symbiodinium* in clades D and A are typically associated with warm and variable temperature regimes [e.g., @Baker:2004p21; @Oliver:2011cr] and shallow habitats with high light intensities [@LaJeunesse:2002p3917].
+Therefore, it is consistent with expectations that they were more prevalent on the south shore, where temperatures were slightly higher, and chlorophyll levels were lower, suggesting greater light penetration.
+However, the lack of differentiation between shores in the symbiont communities of most host species suggests that environmental differences at this scale are not important drivers of community structure.
 
-  When the associations between *Symbiodinium* types and coral species were visualized by network analysis, features of both symbiont and host ecology emerge.
+  Network analysis of associations between *Symbiodinium* types and coral species revealed metacommunity-level patterns in *Symbiodinium* ecology.
 First, ITS2 type B1 was "abundant" (i.e.
->1%) in seven out of ten coral species, and "dominant" (i.e.
->50%) in five out of ten, suggesting it is a more generalist symbiont lineage.
-However, analysis of more variable chloroplast and microsatellite loci has revealed that ITS2 type B1 in the Caribbean is comprised of numerous specialized lineages that demonstrate specific associations with different coral host species [@Santos:2004p4044; @Finney:2010p8324; @Parkinson:2015gt].
-Likewise, the apparent generalists *Symbiodinium* C3 (which dominated colonies of both *S.
+>1% relative abundance in a sample) in seven of ten host species (Fig.
+4), and "dominant" (i.e.
+>50%) in five of ten (Fig.
+5), suggesting it is a generalist symbiont.
+However, analysis of other markers such as chloroplast and microsatellite loci has revealed that *Symbiodinium* ITS2 type B1 in the Caribbean is comprised of multiple lineages that show high host species fidelity [@Santos:2004p4044; @Finney:2010p8324; @Parkinson:2015gt].
+Likewise, the apparent generalists *Symbiodinium* C3 (which dominated colonies of *S.
 siderea* and *M.
-cavernosa*) and *Symbiodinium* A4 (which dominated both *P.
+cavernosa*) and *Symbiodinium* A4 (which dominated *P.
 astreoides* and *P.
-furcata*) may also be comprised of multiple specialized lineages [@Thornhill:2013dx], which will require evaluation with higher resolution genetic markers.
+furcata*, Fig.
+5) may also be comprised of multiple host-specialized lineages [@Thornhill:2013dx], which could be revealed using higher resolution genetic markers.
 
-  In addition to hosting apparent generalist ITS2 lineages, most coral species (except for *M.
+  In addition to hosting apparent generalist ITS2 lineages, most coral species sampled in St.
+John (except for *M.
 cavernosa* and *P.
-astreoides*) hosted abundant OTUs that were not abundant in any other corals, and are thus apparently more host-specific.
+astreoides*) hosted abundant *Symbiodinium* OTUs that were not abundant in any other corals, and are thus apparently more host-specific.
 Unique OTUs in the B1-radiation were abundant in *M.
@@
 fragum*, *S.
-radians*, and *D.
+radians*, and *P.
 strigosa*.
-Several clade C OTUs were similarly only abundant in one species: C1144 in *D.
+Several clade C OTUs similarly were only abundant in one species: C1144 in *P.
 strigosa*, C7 in *O.
@@
 Interestingly, *S.
-radians* was the only coral species in which none of the abundant symbionts occurred in any other host, which may reflect the unique ecology of *S.
-radians* as the only study species that typically forms small, encrusting colonies.
+radians* was the only species in which none of the abundant symbionts occurred in any other host, which may reflect the unique ecology of *S.
+radians* as the only study species that typically forms small, encrusting colonies as adults (XXX).
 
-  Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at <1% relative abundance in at least three different host species.
-Such a distribution suggests that these symbionts can occupy a range of hosts, but are unlikely to dominate them under prevailing environmental conditions.
-Their presence, however, indicates that they may be able to achieve dominance under certain conditions.
-Indeed, the most prevalent 'background' symbiont in St.
-John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], present in seven out of ten host species), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790].
-Association with these symbionts may therefore broaden a host's fundamental niche space and increase its resilience to environmental change [@Correa:2010p7787].
-*Symbiodinium* A4 and C3 also commonly occurred as background symbionts, suggesting they may perform analogous functional and ecological roles within the metacommunity.
+  To reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at <1% relative abundance in samples of three or more host species.
+This distribution suggests that these symbionts can occupy a range of hosts, but are unlikely to be dominant symbionts, at least under prevailing environmental conditions.
+The most prevalent 'background' symbiont in St.
+John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io]), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in marginal environments [@LaJeunesse:2010p7790].
+Other symbionts occupying a background niche in St.
+John, such as *Symbiodinium* A4 and C3, may similarly be able to achieve dominance under different sets of environmental conditions.
+These background symbionts, therefore, may perform an important functional role within the metacommunity by broadening the fundamental niche space that a host may occupy (*sensu* @Bruno:2003p148), thereby increasing its resilience to environmental change [@Correa:2010p7787].
 
-  Focusing on hosts, there was a clear difference among species in the number of OTUs that were abundant (>1% relative abundance, Fig.
-4) and dominant (>50% relative abundance, Fig.
-5A), indicating different levels of variability in the symbiosis ecology of different coral species.
-*D.
-strigosa* individuals could be dominated by four different OTUs, while all *D.
-cylindrus*, *S.
-radians*, *M.
-cavernosa*, and *P.
-astreoides* individuals were dominated by a single OTU.
-In fact, this variability, or 'flexibility' in the symbiosis ecology of a host species, can be quantified as multivariate dispersion (a metric of beta diversity) using metabarcoding data, which can be statistically compared among species [@Anderson:2006fs].
-The flexibility of symbiosis is important as a potential indicator of the environmental sensitivity of a species [@Putnam:2012wv], and its ability to switch or shuffle its symbionts.
+  Indeed, the ability of corals to associate with different symbionts may be an important trait that mediates their sensitivity to stress [@Putnam:2012wv], and their ability to change symbionts over time in response to environmental change [@Baker:2003p200].
+While flexibility in symbiosis ecology has been quantified previously for individual corals (i.e., based on the diversity of symbionts co-occurring within a single host colony [@Putnam:2012wv]), here we quantify flexibility at the host species level based on the variability of symbiont community structure among multiple colonies (Fig.
+6), a metric of beta diversity that can be statistically compared among host species [@Anderson:2006fs].
 This metric revealed that *M.
@@
 radians*, and *D.
-cylindrus* all had the lowest symbiotic flexibility.
-On the other hand, *D.
+cylindrus* had the lowest symbiotic flexibility, meaning that all sampled individuals had similar symbiont community structure.
+In these host species, symbiont communities may be more constrained by host biology and/or less responsive to environmental variation.
+On the other hand, *P.
 strigosa*, *P.
@@
 alcicornis*, and *S.
-siderea* had the most variable symbiont communities, indicating that these species may be more 'flexible' in their associations, and that their symbiont community structure may be more responsive to variability in the natural environment.
+siderea* had high flexibility, meaning that sampled individuals displayed more divergent symbiont communities.
+Community structure in these hosts may be more responsive to variability in the environment [e.g., @Kennedy:2016do], or subject to greater stochasticity.
+While future work should investigate whether the scope for symbiont community change over time within individuals is linked to variability among individuals, we suggest that the latter represents a useful metric of symbiosis flexibility that can be easily quantified using metabarcoding data.
 
@@
 
-  While not providing species-level resolution, ITS2 metabarcoding represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs.
-With appropriate bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, and changes in dominant or background symbiont abundance over environmental gradients or over time, with important implications for understanding coral responses to environmental change.
+  ITS2 metabarcoding and within-sample OTU clustering represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs.
+The scale and resolution of datasets generated in this way facilitate new analytical applications that can address critical topics in coral symbiosis ecology, including changes in dominant and background symbionts across environmental gradients and over time, and the role of metacommunity processes in shaping *Symbiodinium* communities.
+Describing these trends has the potential to greatly advance understanding of coral responses to environmental change.
 
 ### Acknowledgements
-Funding for this work was provided by _________.
+
+  Funding for this work was provided in part by the Long Term Research in Environmental Biology program of the US National Science Foundation (NSF) (DEB 13-50146).
 RC was supported by a NSF Postdoctoral Research Fellowship in Biology (NSF-PRFB 1400787).

### 2017-02-08 manuscript/manuscript.Rmd
Pete revisions round 3

@@
 
-  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-tolerance of their coral hosts, making the description of *Symbiodinium* communities in corals (and metacommunities on reefs) a routine and essential goal in coral reef research.
-High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented scale, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity.
-Using samples of *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St.
-John, US Virgin Islands, we demonstrate that generating operational taxonomic units by clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific.
-We utilized this ‘within-sample clustering’ approach to create metacommunity networks of host-symbiont associations in St.
-John, which revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations with *Symbiodinium*, although no differences in symbiont communities were observed between north and south shore environments.
+  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance and stress-tolerance of their coral hosts, making the analysis of *Symbiodinium* communities in corals (and metacommunities on reefs) advantageous for many aspects of coral reef research.
+High-throughput sequencing of ITS2 nrDNA offers unprecedented scale in describing these communities, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity.
+Here, we demonstrate that generating operational taxonomic units by clustering ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific.
+We utilize this 'within-sample clustering' to analyze *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St.
+John, US Virgin Islands.
+While *Symbiodinium* communities did not differ between shores, metacommunity network analysis of host-symbiont associations revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are more 'flexible' or 'specific' in their associations with *Symbiodinium*.
 These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.
@@
 However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2.
-Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs].
+Consequently, high-throughput sequencing of ITS2 is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs].
 However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
@@
 Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level.
-Moreover, due to the large quantity of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2, it remains an essential and powerful marker for the *Symbiodinium* system.
+Moreover, due to the large quantity of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2, it remains an essential and powerful marker for *Symbiodinium*.
 Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys of *Symbiodinium*.
@@
 John, U.S.
-Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance understanding of *Symbiodinium* metacommunity ecology.
+Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that can advance understanding of *Symbiodinium* metacommunity ecology.
 In addressing the first objective, we test the ability of 'within-sample clustering' (i.e.
@@
 Specifically, this approach addresses the fact that dominant ITS2 sequences from different *Symbiodinium* species may be more similar to one another than intragenomic variants within one *Symbiodinium* [@Thornhill:2013dx; @Arif:2014gk; @Parkinson:2015gt].
-Therefore, clustering all sequences together underestimates diversity by collapsing different species into the same OTU, while treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to intragenomic variation.
+Therefore, clustering all sequences together underestimates diversity by collapsing different species into the same OTU.
+Conversely, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to intragenomic variation.
 Within-sample clustering may better approximate true diversity by exploiting key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.
@@
 We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species.
-The large quantity of data, and the kinds of analytical approaches facilitated by ITS2 metabarcoding, have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts.
+The large quantity of data, and the kinds of analytical approaches facilitated by ITS2 metabarcoding, have the potential to revolutionize understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts.
 More generally, the present study demonstrates how ecological knowledge can inform bioinformatic analyses using marker genes for which sequence diversity does not map directly to species.
@@
 
-  Eighty-four tissue samples were collected from scleractinian and milleporine corals sampled at seven sites around St.
+  Eighty-four tissue samples were collected from scleractinian and milleporine corals at seven sites around St.
 John, US Virgin Islands, USA (Fig.
 1), between August 7th and 9th, 2012.
-Samples were collected on snorkel from 3-7m depth, and all colonies to be sampled were selected haphazardly in upward-facing locations exposed to ambient irradiance.
-Tissue biopsies (~5-mm diameter and ~5-mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.
+Samples were collected on snorkel from 3-7m depth, and colonies to be sampled were selected haphazardly in upward-facing locations exposed to ambient irradiance.
+Tissue biopsies (~5-mm diameter and ~5-mm depth) were removed from coral colonies with metal borers, and the holes were plugged with non-toxic modeling clay.
 
@@
 John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and integrated wave exposure on a power scale (J m^-3^).
-These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and conditions known to affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and mediate the supply of food resources that play important roles in coral nutrition [@Houlbreque:2009gh].
+These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and conditions known to affect coral performance, *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and the supply of food resources that play important roles in coral nutrition [@Houlbreque:2009gh].
 SST and chlorophyll were determined using data from the MODIS sensor on the Aqua satellite.
@@
 
-  A total of 84 coral samples representing ten different host species were collected across two sites on the north shore and five sites on the south shore of St.
+  A total of 84 coral samples representing ten host species were collected at two sites on the north shore and five sites on the south shore of St.
 John (Fig.
 1).
-The sequence data generated from these samples comprised 1,490,813 sequences after merging paired reads, removing chimeric sequences, and trimming primers.
+These samples generated 1,490,813 sequences after merging paired reads, removing chimeric sequences, and trimming primers.
 After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.
@@
 ```{r Fig1, fig.height=3, fig.width=6.85, fig.cap="Figure 1.
-Sampling of *Symbiodinium* communities from ten different host species (embedded histogram) on the north and south shores of St.
+Sampling of *Symbiodinium* communities from ten host species (embedded histogram) on the north and south shores of St.
 John, US Virgin Islands, and the broad-scale physical conditions associated with these two environments.
-Mapped points along each shore indicate locations from which biological samples were collected, while the 'N' and 'S' points indicate locations from which environmental data representative of the north and south shores were obtained.
-Boxplots show the sea-surface temperatures (top), chlorophyll A concentrations (middle), and wave exposure (bottom) recorded at these two locations."}
+Mapped points along each shore indicate locations from which biological samples were collected, while 'N' and 'S' indicate locations from which environmental data representative of the north and south shores were obtained.
+Boxplots show the sea-surface temperatures (top), chlorophyll *a* concentrations (middle), and wave exposure (bottom) recorded at these two locations."}
 library(rgdal)
@@
 rownames(nsam) <- str_replace_all(rownames(nsam), 
-                c("alcicornis"="M.alc.", "annularis"="O.ann.", "astreoides"="P.ast.","cavernosa"="M.cav.",
-                  "cylindrus"="D.cyl.", "fragum"="F.fra.", "furcata"="P.fur.", "radians"="S.rad.",
-                  "siderea"="S.sid.", "strigosa"="P.str."))
+                c("alcicornis"="M.alcicornis", "annularis"="O.annularis", "astreoides"="P.astreoides","cavernosa"="M.cavernosa",
+                  "cylindrus"="D.cylindrus", "fragum"="F.fragum", "furcata"="P.furcata", "radians"="S.radians",
+                  "siderea"="S.siderea", "strigosa"="P.strigosa"))
 nsam <- nsam[,2:3]
 par(mar=c(7,5,5,6), mgp=c(0.5,0.2,0))
-barplot(t(as.matrix(nsam)), las=2, ylab="# samples", xpd=NA, cex.axis=0.5, cex.names=0.5, cex.lab=0.5, tcl=0.1)
+b <- barplot(t(as.matrix(nsam)), las=2, ylab="# samples", xpd=NA, names.arg=rep("",10), cex.axis=0.5, cex.names=0.5, cex.lab=0.5, tcl=0.1)
+text(b+1.2, -0.5, labels=rownames(nsam), srt=45, pos=2, cex=0.5, adj=1)
 legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1, cex=0.5, legend=c("N","S"), bty="n", inset=c(0.12,-0.1), xpd=T, y.intersp=0.8, x.intersp=0.75)
@@
 
-   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig.
+   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach (Table 1 and Fig.
 S1).
@@
 Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches.
-A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants.
-The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU.
+A) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants.
+The dominant OTUs within each sample assigned by B) 97% across-sample clustering and C) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A) to indicate the representative sequence of the assigned OTU.
 Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU.
@@
 
-  Within the set of coral samples obtained in this study, clade B had the highest relative abundance (`r with(cladeAbund, Prop[Clade=="B"])`%), followed by clade C (`r with(cladeAbund, Prop[Clade=="C"])`%), A (`r with(cladeAbund, Prop[Clade=="A"])`%), D (`r with(cladeAbund, Prop[Clade=="D"])`%), and G (`r with(cladeAbund, Prop[Clade=="G"])`%.
+  Within the set of coral samples, clade B had the highest relative abundance (`r with(cladeAbund, Prop[Clade=="B"])`%), followed by clade C (`r with(cladeAbund, Prop[Clade=="C"])`%), A (`r with(cladeAbund, Prop[Clade=="A"])`%), D (`r with(cladeAbund, Prop[Clade=="D"])`%), and G (`r with(cladeAbund, Prop[Clade=="G"])`%).
 The number of OTUs within each clade followed a similar pattern, with the highest number in clade B (`r with(cladeAbund, Notus[Clade=="B"])`), followed by clade C (`r with(cladeAbund, Notus[Clade=="C"])`), A (`r with(cladeAbund, Notus[Clade=="A"])`), D (`r with(cladeAbund, Notus[Clade=="D"])`), and G (`r with(cladeAbund, Notus[Clade=="G"])`).
@@
   No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any host species (Table 2).
-However, some qualitative differences were apparent: *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056).
-*Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.071).
+However, qualitative differences were apparent: *Porites furcata* was dominated by either clade A or clade C *Symbiodinium* on the south shore, but only by clade C on the north shore (difference between shores PERMANOVA: p=0.056).
+*Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but only by clade C on the north shore (difference between shores PERMANOVA: p=0.071).
 
@@
 knitr::kable(shorestats97bs, digits=3, row.names=F, 
-             caption="Table 2. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")
+             caption="Table 2.
+Mean overall, within-shore ('within'), and between-shore ('between') Bray-Curtis dissimilarities of the *Symbiodinium* communities in each host species, and PERMANOVA tests (partial R^2^ and p-values) for a difference between shores.
+Between-shore tests were not possible for *O.
+annularis* and *P.
+astreoides* since they were only sampled from one shore.")
 
@@
 
-#### Network analyses of *Symbiodinium* metacommunity
+#### Network analysis of *Symbiodinium* metacommunity
 ```{r networks,  cache=F, include=FALSE}
@@
 
-  Patterns of association between hosts and *Symbiodinium* were analyzed by constructing three networks focusing on "abundant" (>1% relative abundance; Fig.
+  Patterns of association between hosts and *Symbiodinium* were analyzed using networks focusing on "abundant" (>1% relative abundance; Fig.
 4), "dominant" (>50% relative abundance; Fig.
@@
 5B).
-A member of *Symbiodinium* clade D (which has been formally described as *S.
-trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa).
+A member of *Symbiodinium* clade D (which has been described as *S.
+trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species (`r nspp["D1"]`), followed by A4 and C3 (both in `r nspp["C3"]` taxa).
 C31, C46, and B1 were each detected at background levels in 4 host species.
@@
 *Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at >1% relative abundance within an individual; thickness of edges (i.e., links between coral and symbiont nodes) is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%.
-Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species they dominated."}
+Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species in which they were found."}
 
@@
 
-  The observed structure of *Symbiodinium* communities in a variety of host species was impacted by the bioinformatic approach used to analyze ITS2 sequence data.
-Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset, or independently within each sample, affected the number and taxonomic assignment of OTUs.
+  The structure of *Symbiodinium* communities that was detected in host species was impacted by the bioinformatic approach used to analyze ITS2 sequence data.
+Whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset, or independently within each sample, affected the number and taxonomic assignment of OTUs.
 Clustering sequences across samples reduced the total number of OTUs by 98% relative to the number of OTUs detected with no clustering (Table 1).
@@
 
-  Therefore, even though clustering at 97% similarity collapses intragenomic variation within a single *Symbiodinium* [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related *Symbiodinium*, it is likely to also collapse interspecific variation and underestimate diversity.
-This limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples from multiple hosts: despite being comprised mostly of C41 variants, ~3,000 ITS2 sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7,000 sequences from *Pocillopora verrucosa* with which they were clustered.
+  Therefore, even though clustering at 97% similarity may be needed to collapse intragenomic variation within a single *Symbiodinium* [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related *Symbiodinium*, it is likely to also collapse interspecific variation and underestimate diversity.
+This limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples from multiple hosts.
+Despite being comprised mostly of C41 variants, ~3,000 ITS2 sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7,000 sequences from *Pocillopora verrucosa* with which they were clustered.
 In this case (and as shown in Fig.
@@
 The outcome of within-sample clustering of ITS2 sequences as applied herein better reflects patterns of *Symbiodinium* diversity and ecology that have been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different coral host taxa.
-Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O.
+Indeed, only within-sample clustering assigned different dominant clade C OTUs to samples of *O.
 annularis* and *S.
 siderea* (Fig.
-2), which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx].
-Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases.
-In theory, this approach should also reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically dominant sequence variants to assign taxonomy.
-Furthermore, metabarcoding overcomes the primary limitations of the DGGE method by providing more quantitative data and highly sensitive detection of low abundance taxa [@Quigley:2014dv].
+2), which have been differentiated based on other markers [@Thornhill:2013dx].
+The within-sample clustering approach should also reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically dominant sequence variants to assign taxonomy.
+Furthermore, metabarcoding overcomes the primary limitations of the DGGE method by providing more quantitative data and sensitive detection of low abundance taxa [@Quigley:2014dv].
 Therefore, we recommend a within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.
@@
 
-  In the ten host species sampled in St.
-John, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
+  In the ten host species sampled, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324].
 Less frequent associations between hosts and members of clades A and D were observed, although clade A dominated *P.
 astreoides* and some *P.
-furcata* colonies, and clade D dominated one *S.
-siderea* colony.
-Finally, clade G was observed at low relative abundances (<`r 100*round(clademax[5,2], 2)`%) in some samples, which likely would not have been detected without high-throughput sequencing.
+furcata*, and clade D dominated one *S.
+siderea*.
+Finally, clade G was observed at low relative abundances (<`r 100*round(clademax[5,2], 2)`%) in some coral colonies, which likely would not have been detected without high-throughput sequencing.
 
@@
 2).
-*Symbiodinium* in clades D and A are typically associated with warm and variable temperature regimes [e.g., @Baker:2004p21; @Oliver:2011cr] and shallow habitats with high light intensities [@LaJeunesse:2002p3917].
-Therefore, it is consistent with expectations that they were more prevalent on the south shore, where temperatures were slightly higher, and chlorophyll levels were lower, suggesting greater light penetration.
-However, the lack of differentiation between shores in the symbiont communities of most host species suggests that environmental differences at this scale are not important drivers of community structure.
+*Symbiodinium* in clades D and A are typically associated with warm and variable temperatures [e.g., @Baker:2004p21; @Oliver:2011cr] and shallow habitats with high light intensities [@LaJeunesse:2002p3917].
+Therefore, it is consistent with expectations that they were more prevalent on the south shore, where temperatures were slightly higher, and chlorophyll levels were lower, suggesting greater light penetration into the water column.
+However, the lack of differentiation between shores in the symbiont communities of most host species suggests that environmental differences at this scale are not major drivers of *Symbiodinium* community structure.
 
@@
   In addition to hosting apparent generalist ITS2 lineages, most coral species sampled in St.
-John (except for *M.
+John (except *M.
 cavernosa* and *P.
@@
 The most prevalent 'background' symbiont in St.
-John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io]), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in marginal environments [@LaJeunesse:2010p7790].
+John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io]), which proliferates within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in marginal environments [@LaJeunesse:2010p7790].
 Other symbionts occupying a background niche in St.
-John, such as *Symbiodinium* A4 and C3, may similarly be able to achieve dominance under different sets of environmental conditions.
-These background symbionts, therefore, may perform an important functional role within the metacommunity by broadening the fundamental niche space that a host may occupy (*sensu* @Bruno:2003p148), thereby increasing its resilience to environmental change [@Correa:2010p7787].
+John, such as *Symbiodinium* A4 and C3, may similarly become dominant under different sets of environmental conditions.
+These background symbionts, therefore, may perform an important functional role within the metacommunity by broadening the fundamental niche that a host may occupy (*sensu* @Bruno:2003p148), thereby increasing its resilience to environmental change [@Correa:2010p7787].
 

### 2017-03-26 manuscript/manuscript.Rmd
submission

@@
 ---
-title: Describing *Symbiodinium* metacommunities using high-throughput sequencing
-  of ITS2
+title: 'Describing *Symbiodinium* metacommunities using high-throughput sequencingof ITS2: A case study from St. John, U.S. Virgin Islands'
 author:
-- affiliation: University of Hawaii
+- affiliation: HIMB
   email: ross.cunning@gmail.com
@@
   name: Ross Cunning
-- affiliation: University of Hawaii
+- affiliation: HIMB
   name: Ruth D. Gates
-- affiliation: California State University Northridge
+- affiliation: CSUN
   name: Peter J. Edmunds
+address:
+- address: Hawaii Institute of Marine Biology, School of Ocean and Earth Science and
+    Technology, University of Hawaii at Manoa, USA
+  code: HIMB
+- address: Department of Biology, California State University, Northridge, CA, USA
+  code: CSUN
+output: word_document
+#output:
+#  rticles::elsevier_article:
+#    fig_caption: yes
+csl: the-isme-journal.csl
 bibliography: library.bib
-output:
-  word_document: default
-  pdf_document: default
-  html_document:
-    fig_caption: yes
-subtitle: A case study from St. John, U.S. Virgin Islands
-address:
-- address: Hawaii Institute of Marine Biology, Kaneohe, HI 96744, USA
-  code: University of Hawaii
-- address: Department of Biology, Northridge, CA 91330, USA
-  code: California State University Northridge
+
+abstract: |
+  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance and stress-tolerance of their coral hosts, making the analysis of *Symbiodinium* communities in corals (and metacommunities on reefs) advantageous for many aspects of coral reef research.
+High-throughput sequencing of ITS2 nrDNA offers unprecedented scale in describing these communities, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity.
+Here, we demonstrate that generating operational taxonomic units by clustering ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific.
+We utilize this 'within-sample clustering' to analyze *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St.
+John, US Virgin Islands.
+While *Symbiodinium* communities did not differ between shores, metacommunity network analysis of host-symbiont associations revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are more 'flexible' or 'specific' in their associations with *Symbiodinium*.
+These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.
 ---
+
+**Conflict of interest:** The authors declare no conflict of interest.  
+**Subject category:** Microbial population and community ecology
+
 ```{r setup, include=FALSE}
@@
 
-### Abstract
+# Introduction
 
-  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance and stress-tolerance of their coral hosts, making the analysis of *Symbiodinium* communities in corals (and metacommunities on reefs) advantageous for many aspects of coral reef research.
-High-throughput sequencing of ITS2 nrDNA offers unprecedented scale in describing these communities, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity.
-Here, we demonstrate that generating operational taxonomic units by clustering ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific.
-We utilize this 'within-sample clustering' to analyze *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St.
-John, US Virgin Islands.
-While *Symbiodinium* communities did not differ between shores, metacommunity network analysis of host-symbiont associations revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are more 'flexible' or 'specific' in their associations with *Symbiodinium*.
-These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.
-
-### Introduction
-
-  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building scleractinian and milleporine corals plays a major role in their biology and ecology, as the identity and functional performance of the symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
+  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building scleractinian and milleporine corals plays a major role in their biology and ecology, as identity and the functional performance of symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt].
 Even slight differences in the relative abundance of different symbionts may have important functional consequences for the host [@Loram:2007p3688; @Cunning:2015ja].
 Moreover, variation in these communities among individuals within a host species, and within individuals over time, is an important trait modulating sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond in beneficial ways to changing environmental conditions by 'shuffling' or 'switching' symbionts [@Baker:2003p200].
-Therefore, understanding the ecological performance of corals will greatly benefit from accurate and comprehensive descriptions of their *Symbiodinium* communities.
+Therefore, an accurate and comprehensive description of *Symbiodinium* communities informs understanding of the ecological performance of corals.
 
-  *Symbiodinium* identity is primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we].
+  *Symbiodinium* identity is primarily described using genetic sequences of chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we].
 Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them.
 *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io].
-However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2.
+However, ecological surveys of *Symbiodinium* diversity still generally rely on commonly-used marker genes, such as ITS2.
 Consequently, high-throughput sequencing of ITS2 is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs].
-However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
+However, such analyses often create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.
 
@@
 
-### Materials and Methods
+# Materials and Methods
 
-#### Sample collection and environmental conditions
+## Sample collection and environmental conditions
 
@@
 
-#### ITS2 sequencing and bioinformatics
+## ITS2 sequencing and bioinformatics
 
-  Coral tissue samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB).
+  Coral tissue samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 $\mu$M EDTA; 4.2% w/v sarkosyl; 2.1% v/v $\beta$-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB).
 Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.
 
-  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence.
+  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap $\geq$ 150 bp with $\leq$ 3 mismatches required to retain a sequence.
 Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf].
-Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained.
+Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length $\geq$ 250 bp after trimming were retained.
 Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently).
@@
 Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file.
-The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce the present analysis.
+The bioinformatic pipeline used here was implemented using a series of bash scripts which can be found in the data archive accompanying this paper (http://github.com/jrcunning/STJ2012).
+These scripts are provided along with all raw data and a Makefile which can be executed to fully reproduce the present analysis.
 
-#### *Symbiodinium* data analysis
+## *Symbiodinium* data analysis
 
@@
 A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species.
-To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network.
+To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to $\leq$ 2 coral species were removed from the network.
 All network layouts were constructed based on the Fruchterman-Reingold algorithm [@Fruchterman:1991ko].
@@
 
-### Results
+# Results
 
-#### Sample collection and environmental data
+## Sample collection and environmental data
 
@@
 we <- list(N=3.234787362,	S=4.025022708)
-barplot(c(we$N, we$S), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
+barplot(c(we$S, we$N), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
 text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1, xpd=NA)
@@
 
-#### Comparison of clustering approaches
+## Comparison of clustering approaches
 
@@
 
-knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f)[c(2, 3, 6, 8), ], align="c",
-             caption = "Table 1. Summary statistics for each clustering approach.")
+knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f)[c(2, 3, 6, 8), ],
+             caption = "Summary statistics for each clustering approach.")
 ```
@@
 
-```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2.
+```{r Fig2, fig.width=6, fig.height=4, fig.cap="Figure 2.
 Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches.
@@
 
-#### *Symbiodinium* community composition
+## *Symbiodinium* community composition
 
@@
 
-```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3.
+```{r Fig3, fig.height=6, fig.width=4, fig.cap="Figure 3.
 *Symbiodinium* community composition of each sample.
@@
 
-#### Differences in *Symbiodinium* between shores
+## Differences in *Symbiodinium* between shores
 
@@
 knitr::kable(shorestats97bs, digits=3, row.names=F, 
-             caption="Table 2.
-Mean overall, within-shore ('within'), and between-shore ('between') Bray-Curtis dissimilarities of the *Symbiodinium* communities in each host species, and PERMANOVA tests (partial R^2^ and p-values) for a difference between shores.
+             caption="Mean overall, within-shore ('within'), and between-shore ('between') Bray-Curtis dissimilarities of the *Symbiodinium* communities in each host species, and PERMANOVA tests (partial R^2^ and p-values) for a difference between shores.
 Between-shore tests were not possible for *O.
@@
 
-#### Network analysis of *Symbiodinium* metacommunity
+## Network analysis of *Symbiodinium* metacommunity
 ```{r networks,  cache=F, include=FALSE}
@@
 
-```{r Fig4, fig.width=5, fig.height=5, fig.align="center", fig.retina=1, out.width=5, fig.cap="<br>Figure 4.
+```{r Fig4, fig.width=5, fig.height=5, fig.cap="Figure 4.
 Network analysis of abundant *Symbiodinium* taxa in each coral species.
@@
 
-```{r Fig5, fig.width=7.20472, fig.height=3.60236, fig.align="center", fig.cap="Figure 5.
+```{r Fig5, fig.width=7.20472, fig.height=3.60236, fig.cap="Figure 5.
 Network analysis of dominant (A) and background (B) *Symbiodinium* taxa in each coral species (see Materials and Methods for details on network construction).
@@
 
-#### Symbiotic flexibility of host species
+## Symbiotic flexibility of host species
 
@@
 ```{r Fig6, cache=F, fig.width=3.5, fig.height=3.5, fig.cap="Figure 6.
-Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition (mean distance to centroid ± SE.
+Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition (mean distance to centroid) $\\pm$ SE.
 Host taxa that do not share a letter are significantly different (p < 0.05)."}
@@
 
-### Discussion
+# Discussion
 
-#### Bioinformatic analysis of ITS2 diversity
+## Bioinformatic analysis of ITS2 diversity
 
@@
 
-#### *Symbiodinium* metacommunity ecology in St. John
+## *Symbiodinium* metacommunity ecology in St. John
 
@@
 radians* was the only species in which none of the abundant symbionts occurred in any other host, which may reflect the unique ecology of *S.
-radians* as the only study species that typically forms small, encrusting colonies as adults (XXX).
+radians* as the only study species that typically forms small, encrusting colonies as adults.
 
@@
 
-#### Conclusions
+## Conclusions
 
@@
 
-### Acknowledgements
+# Acknowledgements
 
@@
 
-### References
+# References

### 2017-03-26 supp/STJ2012_analysis.Rmd
submission

--- 2016-09-07 analysis/STJ2012_analysis.Rmd
@@
 ```{r setup_2, include=FALSE}
-# Load all R scripts in "R/" directory
+# Load custom R functions
 source("R/functions.R", .GlobalEnv)
@@
 ```{r import_taxa_data, cache=T}
-# Import taxonomic assignment data from blast and nw
+# Import taxonomic assignment data from nw
 read.nw <- function(file) {
@@
 cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
+
 ```
@@
 ```
-
-##### Results:
-- __*P. furcata* is the only species whose symbiont communities are significantly different between the north and south shores__, at both the 97% within-sample OTU and 100% OTU scales.
-
-# Main findings
-1.
-Symbiont communities are generally consistent with previous work on these species in the Caribbean, but high-throughput sequencing provides a much **more complete picture** of these communities’ composition.
-2. South shore *P. furcata* are more likely to have clade A instead of C, and *S. siderea* more likely to have clade D instead of C.
-3. **Within-sample OTU clustering** is more appropriate than clustering across samples.
-4. We can apply **new types of analysis** with this much data, including network analysis and statistical tests on beta diversity.

### 2017-03-26 Shell/otus_97_bysample.sh

@@ -27,7 +27,7 @@ cat data/otus_97_bysample/[0-9]*_rep_set.fasta > data/otus_97_bysample/all_rep_s
 # Cluster all 97% rep sets at 100% identity
 pick_otus.py -i data/otus_97_bysample/all_rep_set.fasta -s 1.0 --optimal_uclust -o data/otus_97_bysample
 
-# Get rep set of 100% OTUs and assign taxonomy
+# Get rep set of 100% OTUs
 pick_rep_set.py -i data/otus_97_bysample/all_rep_set_otus.txt -f data/otus_97_bysample/all_rep_set.fasta \
 -o data/otus_97_bysample/all_rep_set_rep_set.fasta
 

### 2017-06-05 R/functions.R
Figures for revision

@@ -184,8 +184,10 @@ otubarplot2 <- function(samples) {
   
   
   # Add title
-  text(-0.03, 1.18, label=expression(paste(bold("A. "), "Relative abundance of unique sequence variants in samples:", sep=" ")), 
+  text(-0.03, 1.18, label=expression(paste(bold("A. "), "", sep=" ")), 
        xpd=NA, adj=c(0,0))
+  #text(-0.03, 1.18, label=expression(paste(bold("A. "), "Relative abundance of unique sequence variants in samples:", sep=" ")), 
+  #     xpd=NA, adj=c(0,0))
   # Add species names
   spnames <- as.character(sample_data(phy97)$Species[match(sord, rownames(sample_data(phy97)))])
   fspnames <- c("P. strigosa", "M. alcicornis", "P. furcata", "O. annularis", "S. siderea", "F. fragum", "S. radians", "P. astreoides", "D. cylindrus", "M. cavernosa")
@@ -198,7 +200,9 @@ otubarplot2 <- function(samples) {
     i <- i + 1
   }
   # Add dominant OTU assigned by 97% clustering across samples
-  text(-0.03, -0.125, label=expression(paste(bold("B. "), "Dominant OTU assigned by 97% clustering across samples:")), xpd=NA, adj=c(0,0))
+  #text(-0.03, -0.125, label=expression(paste(bold("B. "), "Dominant OTU assigned by 97% clustering across samples:")), xpd=NA, adj=c(0,0))
+  text(-0.03, -0.125, label=expression(paste(bold("B. "), "")), xpd=NA, adj=c(0,0))
+  
   names97 <- unlist(lapply(strsplit(as.character(data.frame(tax_table(phy97)[apply(data.frame(otu_table(phy97), check.names=F)[,sord], MARGIN=2, FUN=which.max), "Subtype3"])$Subtype3), split="_"), "[[", 1))
   #names97 <- taxdat[taxdat$Subtype3 %in% names97, "Subtype3"]
   otus97 <- rownames(tax_table(phy97)[apply(data.frame(otu_table(phy97), check.names=F)[,sord], MARGIN=2, FUN=which.max), ])
@@ -223,7 +227,9 @@ otubarplot2 <- function(samples) {
     i <- i + 1
   }
   # Add dominant OTU assigned by 97% clustering within samples
[…]

### 2017-06-05 R/analysis.R
cleanup

@@ -1,0 +1,441 @@
+# Set options
+knitr::opts_knit$set(root.dir = normalizePath(".."))
+knitr::opts_chunk$set(echo = F, warning = F, message = F, cache = T)
+options(scipen = 3, digits = 9)
+
+# Load functions
+source("R/functions.R", .GlobalEnv)
+
+# Load package libraries
+library(phyloseq); library(rgdal)
+library(vegan); library(multcompView)
+library(reshape2); library(igraph)
+library(stringr); library(RColorBrewer)
+library(pBrackets)
+
+# Set colors for plotting clades
+taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
+                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
+
+# ==========================
+# Import data
+# ==========================
+# Import taxonomic assignment data
+read.nw <- function(file) {
+  nw <- read.table(file, stringsAsFactors=FALSE)
+  nw <- nw[order(nw$otu),]
+  return(nw)
+}
+
+tax97 <- read.nw("data/otus_97/nw_tophits.tsv")
+tax100 <- read.nw("data/otus_100/nw_tophits.tsv")
+tax97bs <- read.nw("data/otus_97_bysample/nw_tophits.tsv")
+
+# Deal with identical taxonomic assignments (because some reference sequences are not unique...)
+# Create names for groups of identical sequences based on members of group...
+ident <- readLines("data/ITS2db_trimmed_notuniques_otus.txt")
+ident <- gsub("denovo[0-9]*\t", "", ident)
+ident <- strsplit(ident, split="\t")
+ident2 <- lapply(ident, str_match_all, pattern="[A-I]{1}[0-9]{1,3}.*[_/]")
+ident2 <- lapply(ident2, function(g) gsub(pattern="\\..*_$", "", x=unlist(g)))
+ident2 <- lapply(ident2, function(g) gsub(pattern="_$", "", g))
+ident2 <- lapply(ident2, function(g) unlist(strsplit(g, split="/")))
+subtypes <- lapply(ident2, function(x) levels(factor(unlist(x))))
[…]

### 2017-06-05 supp/STJ2012_supp.Rmd
cleanup supp

--- 2017-03-26 supp/STJ2012_analysis.Rmd
@@
 ---
-title: "*Symbiodinium* communities in reef corals of St. John, U.S.V.I. (2012)"
-subtitle: "Data analysis report"
+title: "Using high-throughput sequencing of ITS2 to describe *Symbiodinium* metacommunities in St. John, U.S. Virgin Islands"
+subtitle: "Supplementary information and data analysis walk-through"
 author: "Ross Cunning"
@@
 knitr::opts_knit$set(root.dir = normalizePath(".."))
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
 options(scipen = 3, digits = 9)
@@
 ```{r STJmap, echo=F, message=F, cache=T, fig.height=3, fig.width=4, fig.align="center", fig.cap="Sampling locations in St. John"}
-source("R/STJ_map.R")
+library(rgdal)
+stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
+stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long
+
+sitecoords <- data.frame(
+  name=c("Booby Rock", "Tektite", "Kiddel", "Groot Pan", "Yawzi Point", "Anna's Point", "Haulover Bay"),
+  lat=c(18.30327, 18.31057, 18.30770, 18.31067, 18.31515, 18.36710, 18.35008),
+  lon=c(-64.71057, -64.72203, -64.71398, -64.71813, -64.72513, -64.73337, -64.67933)
+)
+
+viers <- c(-64.722697, 18.322277)
+
+par(mar=c(0,0,0,0))
+plot(stjmap, lwd=1, col="gray40",
+     xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
+box()
+text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
+axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
+axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
+axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
+axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
+with(sitecoords, {
+  points(lon, lat, pch=21, bg="black")
+  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
+})
+points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
+text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
 ```
@@
 Therefore, selecting the most abundant sequence variant _within_ a sample prevents a potentially distinct _Symbiodinium_ type from being collapsed into an OTU with a different representative sequence that happens to be more abundant in other samples within the dataset.
-For more discussion on this approach, see [this entry in my lab notebook](http://jrcunning.github.io/labnotebook//2016/02/12/ITS2_clustering.html).
 
@@
 tax97bs <- read.nw("data/otus_97_bysample/nw_tophits.tsv")
-tax97bsp <- read.nw("data/otus_97_byspecies/nw_tophits.tsv")
 
@@
 tax97bs <- collapse.idents(tax97bs)
-tax97bsp <- collapse.idents(tax97bsp)
 tax100 <- collapse.idents(tax100)
@@
 rownames(tax97bs) <- tax97bs$otu
-
-#tax97bsp <- get.st(tax97bsp)
-tax97bsp <- with(tax97bsp, tax97bsp[order(otu, -sim), ])
-tax97bsp <- tax97bsp[!duplicated(tax97bsp$otu), ]
-rownames(tax97bsp) <- tax97bsp$otu
 
@@
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
-otu97bsp <- otu_table(read.table("data/otus_97_byspecies/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
-                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
 otu100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
@@
 phy97bs <- phyloseq(otu97bs, tax_table(as.matrix(tax97bs)), sam)
-phy97bsp <- phyloseq(otu97bsp, tax_table(as.matrix(tax97bsp)), sam)
 phy100 <- phyloseq(otu100, tax_table(as.matrix(tax100)), sam)
@@
 
-poormatch97bsp <- readLines("data/otus_97_byspecies/poormatch_IDs.txt")
-poormatch97bsp <- data.frame(otu=str_extract(poormatch97bsp, "denovo[^ ]*"),
-                             symbio=str_detect(poormatch97bsp, "Symbiodinium"),   # TRUE if Symbiodinium
-                             stringsAsFactors = FALSE)
-notsymbio97bsp <- poormatch97bsp[which(poormatch97bsp$symbio==F), 1]
-
 poormatch100 <- readLines("data/otus_100/poormatch_IDs.txt")
@@
 phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% notsymbio97bs, phy97bs)
-phy97bsp.f <- prune_taxa(!taxa_names(phy97bsp) %in% notsymbio97bsp, phy97bsp)
 phy100.f <- prune_taxa(!taxa_names(phy100) %in% notsymbio100, phy100)
@@
 taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
-taxa97bsp <- taxa_sums(phy97bsp.f)[which(taxa_sums(phy97bsp.f) >= n)] 
 taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
@@
 phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
-phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp.f)
 phy100.f <- prune_taxa(names(taxa100), phy100.f)
@@
 phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
-phy97bsp.f <- prune_samples(sample_sums(phy97bsp.f)>=sn, phy97bsp.f)
 phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
@@
 tax_table(phy97bs.f) <- as.matrix(get.st(data.frame(tax_table(phy97bs.f), stringsAsFactors=FALSE)))
-tax_table(phy97bsp.f) <- as.matrix(get.st(data.frame(tax_table(phy97bsp.f), stringsAsFactors=FALSE)))
 tax_table(phy100.f) <- as.matrix(get.st(data.frame(tax_table(phy100.f), stringsAsFactors=FALSE)))
@@
 stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
-stats97bsp.f <- data.frame(`97% within-species OTUs`=t(phystats(phy97bsp.f)), check.names=F)
 stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)
@@
 taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
-taxhist97bsp <- hist(log10(taxa_sums(phy97bsp.f)), plot=F)
 taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
@@
 samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
-samhist97bsp <- hist(log10(sample_sums(phy97bsp.f)), plot=F)
 samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)
 
-par(mfrow=c(2, 4), mar=c(3,3,1,1))
+par(mfrow=c(2, 3), mar=c(3,3,1,1))
 plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
 plot(taxhist97bs, col="black", main="97% sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(taxhist97bsp, col="black", main="97% species OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
@@
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-plot(samhist97bsp, col="black", main="97% species OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
 plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
 # Create stats table
-knitr::kable(cbind(stats97.f, stats97bs.f, stats97bsp.f, stats100.f))
+knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
 ```
@@
 stats97bs <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs)), check.names=F)
-stats97bsp <- data.frame(`97% within-species OTUs`=t(phystats(phy97bsp)), check.names=F)
 stats100 <- data.frame(`100% OTUs`=t(phystats(phy100)), check.names=F)
@@
 
-taxhist97bsp <- hist(log10(taxa_sums(phy97bsp)), plot=F)
-samhist97bsp <- hist(log10(sample_sums(phy97bsp)), plot=F)
-
 taxhist100 <- hist(log10(taxa_sums(phy100)), plot=F)
@@
 
-par(mfrow=c(2, 4), mar=c(3,3,1,1))
+par(mfrow=c(2, 3), mar=c(3,3,1,1))
 plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
      xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
-plot(taxhist97bsp, col="black", main="97% within-species OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
+
 plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
      xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
-plot(samhist97bs, col="black", main="97% within-species OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
-     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
+
 plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
@@
 # Create stats table
-knitr::kable(cbind(stats97, stats97bs, stats97bsp, stats100))
+knitr::kable(cbind(stats97, stats97bs, stats100))
 ```
@@
 phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
-phy97bsp.f.p <- transform_sample_counts(phy97bsp.f, function(x) x/sum(x))
 phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
@@
 phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
-phy97bsp.f.t <- transform_sample_counts(phy97bsp.f, transform) 
 phy100.f.t <- transform_sample_counts(phy100.f, transform)
@@
 This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
-```{r cladecomp, cache=T, fig.height=3, fig.width=4}
-cladeAbund <- aggregate(data.frame(RelAbund=rowSums(otu_table(phy97bs.f.p))),
-                        by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
-cladeAbund$Prop <- prop.table(cladeAbund$RelAbund)
-
-bars <- barplot(cladeAbund$Prop*100, col=taxcolors, space=0,
-                names.arg=cladeAbund$Clade, xlab=expression(paste(italic('Symbiodinium'), " Clade")),
-                ylab="Relative abundance (%)")
-text(bars, cladeAbund$Prop*100+2, labels=round(cladeAbund$Prop*100, 1), xpd=T)
-
-cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
-
-```
 
@@
 composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
-# Plot composition of 97% within-sample OTUs colored by OTU
-#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)
-
 ```
@@
 ## *Diploria strigosa*
-Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, B1.
-In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- B1, [B1]^6, and [B1]^3.
-Look at the 100% OTU dataset to see the composition of unique sequence variants in these samples.
-Indeed, these samples are dominated by **different sequence variants**.
-However, in the 97% OTU approach, they have all been collapsed into the same OTU, while in the 97% within-sample clustering approach, samples with different dominant sequence variants are maintained as separate OTUs, even though these different sequence variants are more than 97% similar to each other.
-I suggest that the  within-sample clustering approach is a better way of treating *Symbiodinium* ITS2 sequence data.
-```{r dstr, cache=T, fig.height=16, fig.width=8}
+
+```{r dstr, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Diploria strigosa
@@
 dstr97bs.f.p <- subset_samples(phy97bs.f.p, Species=="strigosa")
-dstr97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="strigosa")
 dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
 # Plot custom barplots for Diploria strigosa
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(dstr97.f.p, main="97% OTUs")
-otubarplot(dstr97bsp.f.p, main="97% within-species OTUs")
 otubarplot(dstr97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Porites furcata*
-```{r pfur, cache=T, fig.height=16, fig.width=8}
+```{r pfur, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Porites furcata
 pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
-pfur97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="furcata")
 pfur97bs.f.p <- subset_samples(phy97bs.f.p, Species=="furcata")
@@
 # Plot custom barplots for Porites furcata
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(pfur97.f.p, main="97% OTUs")
-otubarplot(pfur97bsp.f.p, main="97% within-species OTUs")
 otubarplot(pfur97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Orbicella annularis*
-```{r oann, cache=F, fig.height=16, fig.width=8}
+```{r oann, cache=F, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Orbicella annularis
 oann97.f.p <- subset_samples(phy97.f.p, Species=="annularis")
-oann97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="annularis")
 oann97bs.f.p <- subset_samples(phy97bs.f.p, Species=="annularis")
@@
 # Plot custom barplots for Orbicella annularis
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(oann97.f.p, main="97% OTUs")
-otubarplot(oann97bsp.f.p, main="97% within-species OTUs")
 otubarplot(oann97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Millepora alcicornis*
-```{r malc, cache=T, fig.height=16, fig.width=8}
+```{r malc, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Millepora alcicornis
 malc97.f.p <- subset_samples(phy97.f.p, Species=="alcicornis")
-malc97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="alcicornis")
 malc97bs.f.p <- subset_samples(phy97bs.f.p, Species=="alcicornis")
@@
 # Plot custom barplots for Millepora alcicornis
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(malc97.f.p, main="97% OTUs")
-otubarplot(malc97bsp.f.p, main="97% within-sample OTUs")
 otubarplot(malc97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Siderastrea siderea*
-```{r ssid, cache=T, fig.height=16, fig.width=8}
+```{r ssid, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Siderastrea siderea
 ssid97.f.p <- subset_samples(phy97.f.p, Species=="siderea")
-ssid97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="siderea")
 ssid97bs.f.p <- subset_samples(phy97bs.f.p, Species=="siderea")
@@
 # Plot custom barplots for Siderastrea siderea
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(ssid97.f.p, main="97% OTUs")
-otubarplot(ssid97bsp.f.p, main="97% within-species OTUs")
 otubarplot(ssid97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Favia fragum*
-```{r ffra, cache=T, fig.height=16, fig.width=8}
+```{r ffra, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Favia fragum
 ffra97.f.p <- subset_samples(phy97.f.p, Species=="fragum")
-ffra97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="fragum")
 ffra97bs.f.p <- subset_samples(phy97bs.f.p, Species=="fragum")
@@
 # Plot custom barplots for Favia fragum
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(ffra97.f.p, main="97% OTUs")
-otubarplot(ffra97bsp.f.p, main="97% within-species OTUs")
 otubarplot(ffra97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Siderastrea radians*
-```{r srad, cache=T, fig.height=16, fig.width=8}
+```{r srad, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Siderastrea radians
 srad97.f.p <- subset_samples(phy97.f.p, Species=="radians")
-srad97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="radians")
 srad97bs.f.p <- subset_samples(phy97bs.f.p, Species=="radians")
@@
 # Plot custom barplots for Siderastrea radians
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(srad97.f.p, main="97% OTUs")
-otubarplot(srad97bsp.f.p, main="97% within-species OTUs")
 otubarplot(srad97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Porites astreoides*
-```{r past, cache=T, fig.height=16, fig.width=8}
+```{r past, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Porites astreoides
 past97.f.p <- subset_samples(phy97.f.p, Species=="astreoides")
-past97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="astreoides")
 past97bs.f.p <- subset_samples(phy97bs.f.p, Species=="astreoides")
@@
 # Plot custom barplots for Porites astreoides
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(past97.f.p, main="97% OTUs")
-otubarplot(past97bsp.f.p, main="97% within-species OTUs")
 otubarplot(past97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Dendrogyra cylindrus*
-```{r dcyl, cache=T, fig.height=16, fig.width=8}
+```{r dcyl, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Dendrogyra cylindrus
 dcyl97.f.p <- subset_samples(phy97.f.p, Species=="cylindrus")
-dcyl97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="cylindrus")
 dcyl97bs.f.p <- subset_samples(phy97bs.f.p, Species=="cylindrus")
@@
 # Plot custom barplots for Dendrogyra cylindrus
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(dcyl97.f.p, main="97% OTUs")
-otubarplot(dcyl97bsp.f.p, main="97% within-species OTUs")
 otubarplot(dcyl97bs.f.p, main="97% within-sample OTUs")
@@
 ## *Montastraea cavernosa*
-```{r mcav, cache=T, fig.height=16, fig.width=8}
+```{r mcav, cache=T, fig.height=12, fig.width=8}
 # Create subsetted phyloseq objects for Montastraea cavernosa
 mcav97.f.p <- subset_samples(phy97.f.p, Species=="cavernosa")
-mcav97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="cavernosa")
 mcav97bs.f.p <- subset_samples(phy97bs.f.p, Species=="cavernosa")
@@
 # Plot custom barplots for Montastraea cavernosa
-par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
+par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
 otubarplot(mcav97.f.p, main="97% OTUs")
-otubarplot(mcav97bsp.f.p, main="97% within-species OTUs")
 otubarplot(mcav97bs.f.p, main="97% within-sample OTUs")
@@
 
-# Differences between species
-Whether *Symbiodinium* communities differ among species is evaluated using permutational analysis of variance (PERMANOVA).
-
-```{r permanovas, cache=T}
-# PERMANOVA for difference among species
-sppdiffs <- function(phy) {
-  permanova.spp.t <- adonis(phyloseq::distance(phy, "bray") ~ Species, 
-                          data=as(sample_data(phy), "data.frame"), permutations=9999)
-  permanova.spp.t
-  # Pairwise PERMANOVA tests for differences between species
-  dmat <- as(phyloseq::distance(phy, "bray"), "matrix")
-  df <- as(sample_data(phy), "data.frame")
-  pairs <- data.frame(t(combn(levels(df$Species), 2)), stringsAsFactors=F)
-  p.pairs <- setNames(rep(NA, nrow(pairs)), with(pairs, paste(X1, X2, sep='-')))
-  for (i in 1:nrow(pairs)) {
-    dfs <- subset(df, Species %in% c(pairs[i,1], pairs[i,2]))
-    dmats <- dmat[rownames(dfs), rownames(dfs)]
-    set.seed(152470)
-    permanova <- adonis(dmats ~ Species, data=dfs, permutations=999)
-    p.pairs[i] <- permanova$aov.tab$`Pr(>F)`[1]
-  }
-  # Return letters signifying differences 
-  return(multcompLetters(p.pairs))
-}
-
-sppdiffs97 <- sppdiffs(phy97.f.t)
-sppdiffs97bs <- sppdiffs(phy97bs.f.t)
-sppdiffs100 <- sppdiffs(phy100.f.t)
-
-sppdiffs <- data.frame("97% within-sample OTUs"=sppdiffs97bs$Letters,
-                       "100% OTUs"=sppdiffs100$Letters, 
-                       "97% OTUs"=sppdiffs97$Letters, check.names=F)
-knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
-```
-
-##### Results:
-- At the 97% within-sample OTU scale, *Symbiodinium* communities are **not significantly different** in the corals *M.
-alcicornis*, *O.
-annularis*, and *F.
-fragum*, as these corals are all dominantly associated with the same *Symbiodinium* OTU (BLAST identity=B1).
-These corals are **also not different** from *D.
-cylindrus* or *D.
-strigosa*, which also dominantly associated with the same B1 OTU, although the latter two species are different from each other.
-Meanwhile, *M.
-cavernosa* and *S.
-siderea* also share similar symbiont communities dominated by *Symbiodinium* C3.
-Finally, *S.
-radians*, *P.
-furcata*, and *P.
-astreoides* each have distinct, **significantly different** *Symbiodinium* communities.
-- At the 100% OTU scale, *Symbiodinium* communities were **significantly differentiated** in all species except for *O.
-annularis* and *D.
-strigosa* (sharing B1-dominated communities), and *S.
-siderea* and *M.
-cavernosa* (sharing C3-dominated communities).
-
 # Differences between shores {.tabset}

### 2017-06-06 Shell/fetch_seqs.sh
update for NCBI https

@@ -5,7 +5,7 @@ getseq() {
 	n=$1
 	acn=${n##*_}
 	echo $1 > data/dbseqs/${acn}.fasta
-	curl "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta" >> data/dbseqs/${acn}.fasta
+	curl "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&id=${acn}&rettype=fasta" >> data/dbseqs/${acn}.fasta
 }
 export -f getseq
 rm -rf data/dbseqs; mkdir data/dbseqs
@@ -14,7 +14,7 @@ parallel -a data/accn_nos.txt getseq
 # Put all sequences together and reformat to keep only subtype_accn as sequence name, and put sequence all on one line
 cat data/dbseqs/*.fasta | \
 awk '!/^>/ { printf "%s", $0; n = "\n" } /^>/ { print n $0; n = "" } END { printf "%s", n }' | \
-awk '!/^>gi/ {print}' > data/ITS2db_raw.fasta
+awk '!/\ / {print}' > data/ITS2db_raw.fasta
 
 # Clean up
 rm -rf data/dbseqs

### 2017-06-06 Shell/run_blast_poormatches.sh
use local nt blastdb

@@ -7,8 +7,10 @@ awk '$3 < 90 {print $2}' $1 | sort -u > $dir/poor_matches.txt
 # Get corresponding sequences
 awk 'NR==FNR {a[">"$1]; next} $1 in a {print; getline; print}' $dir/poor_matches.txt $2 > $dir/poor_matches_seqs.fasta
 
-# Blast sequences to NCBI nr databse
-blastn -db nr -remote -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u -k1,1 > $dir/poormatch_IDs1.txt
+# Blast sequences to NCBI nt database
+### Must point -db to local copy of the nt database
+### or enable remote blast using '-remote' flag
+blastn -db ~/ncbi_nt/nt -query $dir/poor_matches_seqs.fasta -outfmt '6 qseqid qcovs stitle sseqid' -max_target_seqs 1 | sort -u -k1,1 > $dir/poormatch_IDs1.txt
 
 join -a 1 $dir/poor_matches.txt $dir/poormatch_IDs1.txt > $dir/poormatch_IDs.txt
 

### 2017-06-07 Shell/prep_ITS2db.sh
do not write id_to_tax file

@@ -29,8 +29,8 @@ cutadapt -a AAGCATATAAGTAAGCGGAGG -e 0.15 data/ITS2db_trimF2_trimR.fasta -o data
 
 
 # Generate id_to_taxonomy file
-TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
-sed -e 's/>\([A-Z]\)\(.*\)\(_.*$\)/\1\2\3'"${TAB}"'Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
+#TAB=$'\t'  # ( because \t is not recognized by sed on mac os x )
+#sed -e 's/>\([A-Z]\)\(.*\)\(_.*$\)/\1\2\3'"${TAB}"'Symbiodiniaceae;Symbiodinium;Clade\1;\1\2;_;_/' -e 'tx' -e 'd' -e ':x' data/ITS2db.fasta > data/id_to_taxonomy.txt
 
 
 

### 2017-06-07 R/STJ_map.R
cleanup

@@ -1,34 +1,0 @@
-# St. John Map
-
-library(rgdal)
-stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
-stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long
-
-sitecoords <- data.frame(
-  name=c("Booby Rock", "Tektite", "Kiddel", "Groot Pan", "Yawzi Point", "Anna's Point", "Haulover Bay"),
-  lat=c(18.30327, 18.31057, 18.30770, 18.31067, 18.31515, 18.36710, 18.35008),
-  lon=c(-64.71057, -64.72203, -64.71398, -64.71813, -64.72513, -64.73337, -64.67933)
-)
-
-viers <- c(-64.722697, 18.322277)
-
-#pdf(file="figures/map.pdf", height=3, width=4)
-par(mar=c(0,0,0,0))
-plot(stjmap, lwd=1, col="gray40",
-     xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
-box()
-text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
-axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
-axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
-axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
-axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
-with(sitecoords, {
-  points(lon, lat, pch=21, bg="black")
-  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
-})
-points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
-text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
-#dev.off()
-
-
-

### 2017-06-07 Shell/otus_97_byspecies.sh
cleanup

@@ -1,56 +1,0 @@
-#!/bin/bash
-
-rm -rf data/otus_97_byspecies
-# Split into individual sample files
-rm -rf data/fasta/bysample
-split_sequence_file_on_sample_ids.py -i data/fasta/combined_seqs_trimmed.fasta -o data/fasta/bysample
-
-rm -rf data/fasta/byspecies; mkdir data/fasta/byspecies
-# Get SampleIDs for each species
-cd data/fasta/bysample
-awk -F"\t" '{print $1".fasta" > $6"_IDs.txt"}' ../../mapping_file.txt
-rm -f Species_IDs.txt
-for spp in `ls *_IDs.txt`; do
-	cat $(cat $spp) > ../byspecies/${spp%%_*}.fasta
-done
-cd ../../..
-
-# OTU clustering at 97% similarity for each species (in parallel)
-clust97() {
-	s=`basename $1`
-	pick_otus.py -i $1 -s 0.97 --denovo_otu_id_prefix $s'_denovo' --optimal_uclust -o data/otus_97_byspecies
-}
-export -f clust97
-parallel -j3 clust97 ::: data/fasta/byspecies/*.fasta
-
-# Pick representative sequence set for each sample
-repset97() {
-	filename=`basename $1`
-	s=${filename%_*}
-	pick_rep_set.py -i $1 -m most_abundant -f 'data/fasta/byspecies/'$s'.fasta' \
-	-o 'data/otus_97_byspecies/'$s'_rep_set.fasta'
-}
-export -f repset97
-parallel repset97 ::: data/otus_97_byspecies/*_otus.txt
-
-cat data/otus_97_byspecies/*otus.txt > data/otus_97_byspecies/all_97_otus.txt
-# Merge all rep sets together
-cat data/otus_97_byspecies/*_rep_set.fasta > data/otus_97_byspecies/all_rep_set.fasta
-
-# Cluster all 97% rep sets at 100% identity
-pick_otus.py -i data/otus_97_byspecies/all_rep_set.fasta -s 1.0 --optimal_uclust -o data/otus_97_byspecies
-
-# Get rep set of 100% OTUs and assign taxonomy
-pick_rep_set.py -i data/otus_97_byspecies/all_rep_set_otus.txt -f data/otus_97_byspecies/all_rep_set.fasta \
--o data/otus_97_byspecies/all_rep_set_rep_set.fasta
-
[…]

### 2017-06-07 Shell/run_blast.sh

@@ -1,48 +1,0 @@
-#!/bin/bash
-
-dir=$(dirname $1)
-
-# BLAST each sequence to the reference ITS2 database (trimmed, dereplicated)
-formatdb -p F -i $2
-blastall -p blastn -i $1 -d $2 -e 0.00001 -b 10 -m 8 > $dir/blast_results.txt
-# Sort by query sequence and then E-value in descending order, then get only the first hit for each query sequence
-# sort -k1,1 -k11,11g $dir/blast_results.txt | sort -u -k1,1 > $dir/blast_tophits.txt
-
-# Identify the lowest E-value obtained for each sequence
-sort -k1,1 -k11,11g $dir/blast_results.txt | sort -u -k1,1 | cut -f1,11 > $dir/lowevals.txt
-# Get all BLAST hits that obtained the lowest E-value for each sequence (keeps ties/multiple hits with same E-val)
-rm -f $dir/blast_tophits_withties.txt
-while read n; do
-  otu=`echo $n | cut -d" " -f1`
-  eval=`echo $n | cut -d" " -f2`
-  awk '$1 ~ /'"$otu"'$/ && $11 ~ /'"$eval"'$/ {print}' $dir/blast_results.txt >> $dir/blast_tophits_withties.txt
-done < $dir/lowevals.txt
-
-# Merge hits with equivalent results...(if score/length/eval/mismatch/gaps are identical for multiple hits)
-# by appending multiple hits to the end of the line
-awk '
-BEGIN { FS="\t" }
-{
-    curr = $1 FS $3 FS $4 FS $5 FS $6 FS $7 FS $8 FS $9 FS $10 FS $11 FS $12
-    if (curr == prev) {
-        rec = rec "\t" $2
-    }
-    else {
-        if (rec) print rec
-        rec = $0
-    }
-    prev = curr
-}
-END { if (rec) print rec }
-' $dir/blast_tophits_withties.txt > $dir/blast_tophits_merged.txt
-
-# Move column two (first blast hit) to last column position so all hits are together, and collapse multiple hits into a single field separated by ";"
-awk 'BEGIN{FS=OFS="\t"} {a=$2; for (i=2;i<NF; i++) $i=$(i+1); $NF=a}1' $dir/blast_tophits_merged.txt | \
[…]

### 2017-06-07 Shell/tax_results.sh

@@ -1,21 +1,0 @@
-#!/bin/bash
-
-DIR=$(dirname $1)
-OUT=$2
-
-
-# Assign taxonomy
-assign_taxonomy.py -i $1 -r data/ITS2db_trimmed.fasta -t data/id_to_taxonomy.txt -m blast -e 1e-30 -o $DIR/blast_taxonomy
-
-# Make list of "no blast hits"
-awk '/No blast hit/' $DIR/blast_taxonomy/*_tax_assignments.txt > $DIR/blast_taxonomy/no_blast_hits.txt
-
-# Make OTU table excluding no blast hits
-make_otu_table.py -i $DIR/nosingles_otus.txt \
--t $DIR/blast_taxonomy/*_tax_assignments.txt \
--e $DIR/blast_taxonomy/no_blast_hits.txt -o $DIR/$OUT.biom
-rm -f  $DIR/*.tsv  # delete old OTU .tsv if present
-biom convert -i $DIR/$OUT.biom -o data/$OUT.tsv --to-tsv
-
-# Add clade to tax table and copy to data directory
-cut -d$'\t' -f4- $DIR/blast_taxonomy/*_tax_assignments.txt | cut -c-1 | paste - $DIR/blast_taxonomy/*_tax_assignments.txt > 'data/'$OUT'_tax_assignments.txt'

### 2017-06-07 R/analysis.R
add wave data source

@@ -247,7 +247,7 @@ boxplot(envdata$chlA ~ envdata$Site, range=0, add=T, axes=F, horizontal=T)
 text(par("usr")[1], par("usr")[4], "c)", adj=3)
 
 #Wave exposure
-we <- list(N=3.234787362,	S=4.025022708)
+we <- list(N=3.234787362,	S=4.025022708)  # Wave exposure data provided by Iliana Chollett
 barplot(c(we$S, we$N), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
 text(par("usr")[1], c(0.7,1.9), c("S","N"), pos=2, offset=0.5, xpd=NA)
 points(rep(par("usr")[1]-strwidth("cc"),2), c(0.7,1.9), pch=1, cex=2.5, xpd=NA)
